---
title: "Manuscript synthetic simulation scenarios"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Manuscript synthetic simulation scenarios}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup-ms-scenarios
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
# Executable DeCovarT chunks stay eval: false in this CRAN vignette:
# quarto.exe on Windows does not inherit R CMD build's temporary library
# (https://github.com/quarto-dev/quarto-r/issues/217).
mu_small <- matrix(
  c(20, 22, 22, 20),
  nrow = 2,
  dimnames = list(c("gene_1", "gene_2"), c("ct1", "ct2"))
)
p_bal <- c(0.5, 0.5)
```

This vignette documents the **synthetic scenarios that enter the methods
paper**: a bivariate Gaussian-convolution toy ($G=2$, $J=2$) that isolates
gene–gene correlation, and a high-dimensional hybrid reference
($G=50$, $J=3$) in which two cell types are mean-collinear and must be
separated by network topology.

# Simulation benchmark wrapper {#sec-simulation-benchmark}

[`run_simulation_benchmark()`](https://bastienchassagnol.github.io/DeCovarT/reference/run_simulation_benchmark.html) wraps `simulate_bulk_mixture()`, `deconvolute_ratios()`, and `compute_benchmark_metrics()`. Generative grids (proportions, correlation sweeps, mean signatures, covariance tensors) should be assembled in scripts—see `scripts/configure_bivariate_toy_scenarios.R` for the bivariate toy study—and passed as a tibble with a `true_theta` list column. Deconvolution solvers are supplied as a named list of `FUN` plus optional `additional_parameters` for `do.call()`. Optional scenario-level parallelism uses `furrr` and `future` (Suggests), capped by default at half of detected cores, with per-sample parallelism disabled inside `deconvolute_ratios()` to avoid nested workers.

# Bivariate toy model {#sec-bivariate-toy}

The manuscript study is implemented by
`run_simulation_benchmark()`, with scenario grids built in
`scripts/configure_bivariate_toy_scenarios.R`. Even when mean profiles
are similar, gene–gene correlation can degrade mean-only solvers, and
is partly recovered once
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$
enters the likelihood.

With $p_1+p_2=1$, only one free ALR coordinate is estimated (see
[derivatives under simplex transforms](generative-model-derivatives.html)).
@fig-toy-design combines the factorial design with a gene-space density
at fixed means, variances, and $\boldsymbol{p}$.

::: {#fig-toy-design layout-ncol=2}

![Hyperparameter grid used in the bivariate manuscript study: two
centroid geometries, balanced versus unbalanced $\boldsymbol{p}$, a
gene–gene correlation sweep, and the first-generation plus DeCovarT
solvers.](figures/fig_toy_model_parameters.svg){#fig-toy-hyperparams}

![AI-generated 2D density of the two-gene mixture while only the
pairwise correlation changes, at fixed means, marginal variances, and
composition. The ggplot panel below recomputes the same contrast and
annotates MixSim overlap.](figures/fig_toy_2d_density.svg){#fig-toy-density-ai}

Factorial design (left) and correlation-only 2D densities (right) for
the bivariate toy. {#fig-toy-design}
:::

```{r}
#| label: fig-toy-density-ggplot
#| fig-cap: >-
#|   Two-gene mixture densities at fixed means
#|   $\boldsymbol{\mu}_{\cdot 1}=(20,22)$, $\boldsymbol{\mu}_{\cdot 2}=(22,20)$,
#|   $\boldsymbol{p}=(1/2,1/2)$ and unit diagonals. Only the common
#|   gene–gene correlation $\rho$ changes. Labels report MixSim
#|   $\overline{\omega}$ (`compute_average_overlap()`).
#| fig-width: 8.4
#| fig-height: 4.4
#| echo: false
#| eval: false
set.seed(20260828)
mu_small <- matrix(
  c(20, 22, 22, 20),
  nrow = 2,
  dimnames = list(c("gene_1", "gene_2"), c("ct1", "ct2"))
)
p_bal <- c(0.5, 0.5)
rho_grid <- c(-0.6, 0, 0.6)
n_draw <- 400L
dens_df <- list()
ovl_df <- list()
for (rho in rho_grid) {
  R <- matrix(c(1, rho, rho, 1), nrow = 2)
  Sigma <- array(c(R, R), dim = c(2, 2, 2))
  dimnames(Sigma) <- list(
    rownames(mu_small),
    rownames(mu_small),
    colnames(mu_small)
  )
  sim <- DeCovarT::simulate_bulk_mixture(
    mu_small,
    Sigma,
    p = p_bal,
    n = n_draw
  )
  dens_df[[length(dens_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    gene_1 = sim$Y[1, ],
    gene_2 = sim$Y[2, ]
  )
  theta <- list(p = p_bal, mu = mu_small, sigma = Sigma)
  ovl_df[[length(ovl_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    ovl = DeCovarT::compute_average_overlap(theta)
  )
}
dens_df <- do.call(rbind, dens_df)
ovl_df <- do.call(rbind, ovl_df)
ovl_df$lab <- sprintf("overlap = %.2f", ovl_df$ovl)
ggplot2::ggplot(dens_df, ggplot2::aes(x = gene_1, y = gene_2)) +
  ggplot2::stat_density_2d_filled(contour_var = "ndensity", bins = 12) +
  ggplot2::geom_label(
    data = ovl_df,
    ggplot2::aes(x = Inf, y = Inf, label = lab),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.1,
    fill = "#fffbe9",
    label.size = 0.15
  ) +
  ggplot2::facet_wrap(~rho, nrow = 1) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Expression of gene 1",
    y = "Expression of gene 2"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

## Log-likelihood along the simplex

For $J=2$ the free coordinate is $p_1\in(0,1)$. The listing below
evaluates
$\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}(p_1,1-p_1)$ on one bulk draw
(and would feed `plotly` when that Suggests package is installed).
It is not executed during `R CMD build`: Quarto's Windows CLI starts a
new R process that cannot see the temporary library where the package
under construction was just installed
([quarto-r#217](https://github.com/quarto-dev/quarto-r/issues/217)).
The same likelihood path is checked in `tests/testthat/`.

```{r}
#| label: fig-toy-loglik
#| echo: true
#| eval: false
set.seed(20260828)
rho <- 0.6
R <- matrix(c(1, rho, rho, 1), nrow = 2)
Sigma <- array(c(R, R), dim = c(2, 2, 2))
dimnames(Sigma) <- list(
  rownames(mu_small),
  rownames(mu_small),
  colnames(mu_small)
)
y_one <- DeCovarT::simulate_bulk_mixture(
  mu_small,
  Sigma,
  p = p_bal,
  n = 1
)$Y[, 1]
p1_grid <- seq(0.02, 0.98, by = 0.01)
ll_grid <- vapply(
  p1_grid,
  function(p1) {
    DeCovarT::loglik_multivariate(c(p1, 1 - p1), y_one, mu_small, Sigma)
  },
  numeric(1)
)
ll_df <- data.frame(p1 = p1_grid, loglik = ll_grid)
gg_ll <- ggplot2::ggplot(ll_df, ggplot2::aes(x = p1, y = loglik)) +
  ggplot2::geom_line(colour = "#1B4F72", linewidth = 0.8) +
  ggplot2::geom_vline(
    xintercept = 0.5,
    linetype = "dashed",
    colour = "grey30"
  ) +
  ggplot2::labs(
    x = "p1 (p2 = 1 - p1)",
    y = "Log-likelihood"
  ) +
  ggplot2::theme_bw(base_size = 11)
gg_ll
```

```{r}
#| label: fig-toy-loglik-plotly
#| echo: false
#| eval: false
if (requireNamespace("plotly", quietly = TRUE)) {
  plotly::ggplotly(gg_ll, tooltip = c("x", "y"))
}
```

## How to run the factorial grid

```{r}
#| label: lst-bivariate-call
#| eval: false
#| echo: true
library(DeCovarT)
source("scripts/configure_bivariate_toy_scenarios.R")

deconvolution_functions <- bivariate_toy_deconvolution_functions(
  itmax = 200L,
  epsilon = 1e-4
)

scenario_config <- build_bivariate_scenario_config(
  proportions = list(
    "balanced" = c(0.5, 0.5),
    "highly unbalanced" = c(0.95, 0.05)
  ),
  signature_matrices = list(
    "small CLD" = matrix(c(20, 22, 22, 20), nrow = 2),
    "large CLD" = matrix(c(20, 40, 40, 20), nrow = 2)
  ),
  corr_sequence = seq(-0.8, 0.8, by = 0.2),
  diagonal_terms = list("homoscedastic" = c(1, 1))
)

bivariate <- run_simulation_benchmark(
  scenario_config = scenario_config,
  deconvolution_functions = deconvolution_functions,
  n = 500,
  cores = 1,
  parallel_scenarios = FALSE
)
```

`config` stores Shannon entropy of $\boldsymbol{p}$ and MixSim overlap;
`simulations` stores per-sample $\hat{\boldsymbol{p}}$ and
`compute_benchmark_metrics()`. The article compares DeCovarT with
DeconRNASeq / NNLS [@gongDeconRNASeqStatisticalFramework2013].

# High-dimensional hybrid scenario {#sec-scenario-grid}

`scripts/generate_random_markov_network.R` (seed `20260807`) builds one
$G=50$, $J=3$ reference that mean-only methods cannot fully resolve:
types 1 and 2 share a high local cosine on 30 genes and differ only
through $\boldsymbol{\Omega}$; type 3 has a compact marker block; 10 genes
are a null (`equal_all`) block.

$$
\underbrace{\mathcal{G}_{12}}_{30\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{3}}_{10\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{\mathrm{eq}}}_{10\text{ genes}},
\qquad
\boldsymbol{\Omega}_j
=
\mathrm{blockdiag}\bigl(
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{12})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{3})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{\mathrm{eq}})}
\bigr).
$$

$\boldsymbol{\mu}$ uses two calls to `generate_mean_signature_matrix()`
([mean signature construction](https://bastienchassagnol.github.io/DeCovarT/articles/synthetic-scenarios.html#sec-lr-block-mu)):
$\rho_{12}=0.95$ on $\mathcal{G}_{12}$, then $\rho_{3}=0.05$ on a
marker pool for type 3. Local topologies follow the BLGGM two-scale
pattern (@tbl-hybrid-topology-design;
[GGM networks](https://bastienchassagnol.github.io/DeCovarT/articles/synthetic-scenarios.html#sec-ggm-networks)).

| Gene block | # genes | Cell type 1 | Cell type 2 | Cell type 3 |
|---|---|---|---|---|
| `shared_12_vs_3` | 30 | SBM (hub-like modules) | star (one driver) | Erdős–Rényi |
| `marker_3` | 10 | Erdős–Rényi | Erdős–Rényi | scale-free |
| `equal_all` | 10 | Erdős–Rényi | Erdős–Rényi | Erdős–Rényi |

: Local topology per (gene block, cell type); block-diagonal precision support. {#tbl-hybrid-topology-design}

| pair | cosine | Euclidean |
|:-----|-------:|----------:|
| celltype_1 vs celltype_2 | 0.973 | 2.765 |
| celltype_1 vs celltype_3 | 0.562 | 11.219 |
| celltype_2 vs celltype_3 | 0.562 | 11.219 |

: Realised pairwise geometry of the $50\times 3$ mean signature (seed 20260807). {#tbl-mean-geometry}

| cell type | topology | $\lambda_{\min}$ | $\lambda_{\max}$ | $\kappa(\Omega)$ | prop. inhib. |
|:----------|:---------|-----------------:|-----------------:|-----------------:|-----------:|
| celltype_1 | SBM | 0.100 | 1.685 | 16.846 | 0.500 |
| celltype_2 | star | 0.100 | 3.331 | 33.311 | 0.489 |
| celltype_3 | scale-free | 0.100 | 2.096 | 20.961 | 0.500 |

: Precision spectra of each cell type's $\boldsymbol{\Omega}_j$ (seed 20260807). {#tbl-topology-diagnostics}

Types 1 and 2 remain near-collinear on the full $G=50$ vector (cosine
$\approx 0.97$). Type 3 is only *moderately* separated (cosine
$\approx 0.56$) because the shared $\mathcal{G}_{12}$ baseline and the
flat `equal_all` block inflate every pairwise inner product—the
finite-$J$ cosine bias of
[asymptotic cosine control](https://bastienchassagnol.github.io/DeCovarT/articles/synthetic-scenarios.html#sec-asymptotic-cosine).
@tbl-topology-diagnostics shows that types 1 and 2 still differ in
$\boldsymbol{\Omega}$.

![Cell-type-specific block topologies: SBM/hub-like modules (type 1) and
a star (type 2) on the 30 `shared_12_vs_3` genes; scale-free wiring
(type 3) on `marker_3`. The `equal_all` block is Erdős–Rényi in every
type. Edge colour: precision sign (red inhibitory, teal activatory).](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

# Implemented solvers {#sec-hybrid-deconvolution}

@tbl-solver-diagnostics lists the **shipped** algorithms and their
default controls (first-generation linear methods in
`R/03_01_first_generation_deconvolution.R`; covariance-aware maps in
`R/03_03_DeCovarT_estimate_ratios_frequentist.R`). All start from the
open simplex via `.starting_simplex()` unless `initial_p` is supplied.
Reproduce the hybrid comparison with
`source("scripts/deconvolute_simulated_scenario.R")` ($N=20$ bulks,
true $\boldsymbol{p}=(0.4,0.4,0.2)$).

| Algorithm | Function | Defaults | Constraint |
|:----------|:---------|:---------|:-----------|
| $\nu$-SVR (`CIBERSORT`-style) | `deconvolute_ratios_cibersort` | linear kernel; $\nu\in\{0.2,0.5,0.8\}$ if $G\ge 4$, else $\nu=0.5$ | clip negatives, `repair_simplex()` |
| OLS | `deconvolute_ratios_lsfit` | `stats::lsfit(..., intercept = FALSE)` | `repair_simplex()` |
| Robust linear (`ABIS`-style) | `deconvolute_ratios_rlm` | `MASS::rlm(..., method = "M")` | `repair_simplex()` |
| NNLS | `deconvolute_ratios_nnls` | `nnls::nnls()` | `repair_simplex()` |
| QP (`DeconRNASeq`-style) | `deconvolute_ratios_deconrnaseq` | `limSolve::lsei()` with $\mathbf{1}^{\mathsf{T}}\boldsymbol{p}=1$, $\boldsymbol{p}\ge\mathbf{0}$ | simplex equality / inequality |
| Marquardt–Levenberg | `deconvolute_ratios_Marquardt_Levenberg` | `epsilon = 1e-4`, `itmax = 200`, ALR | ALR $\to$ open simplex |
| Newton–Raphson | `deconvolute_ratios_Newton_Raphson` | `nlminb`; `rel.tol = x.tol = 1e-4`, `iter.max = 200` | ALR |
| L-BFGS-B | `deconvolute_ratios_L_BFGS_B` | box $[0,1]^J$; `maxit = 200`; then $p/\sum p$ | box + closure |
| BFGS (ALR) | `deconvolute_ratios_gradient_descent` | `method = "BFGS"`, `maxit = 200` | ALR |
| Simulated annealing | `deconvolute_ratios_simulated_annealing` | `method = "SANN"`, `maxit = 200` | ALR |

: Shipped deconvolution algorithms and default hyperparameters. {#tbl-solver-diagnostics}

# Benchmark metrics {#sec-benchmark-metrics}

When true $\boldsymbol{p}^{\star}$ is known,
`compute_benchmark_metrics()` currently returns MSE, RMSE, MAE, a
pseudo-$R^{2}$, and Pearson correlation between
$\hat{\boldsymbol{p}}$ and $\boldsymbol{p}^{\star}$. Reconstruction-only
scores (no $\boldsymbol{p}^{\star}$) compare
$\hat{\boldsymbol{y}}=\boldsymbol{\mu}\hat{\boldsymbol{p}}$ with $\boldsymbol{y}$.
@fig-metrics-families summarises the broader families used in recent
bulk and spatial benchmarks.

::: {#fig-metrics-families layout-ncol=2}

![Simplex geometry of $L_1$, $L_2$, $L_{\infty}$, Aitchison, and
angular discrepancies between $\boldsymbol{p}^{\star}$ and
$\hat{\boldsymbol{p}}$.](figures/fig_compositional_metrics.svg){#fig-comp-metrics}

![How those distances behave near the centre versus near a vertex of
the simplex.](figures/fig_compositional_metrics_interpretation.svg){#fig-comp-interp}

Compositional error metrics for $\boldsymbol{p}$ on the simplex.
:::

![Hierarchical relative RMSE (hrRMSE) used when some reference types
are missing: residual error is scaled by the biological variance of
$\boldsymbol{p}^{\star}$ rather than by raw proportion units
[@baWhenLessNot2026].](figures/fig_hierarchical_rmse_metrics.png){#fig-hrrmse width=70%}

Notation matches the article: $\boldsymbol{p}$ on $\Delta^{J-1}$,
$\hat{\boldsymbol{p}}$ the estimator, $\boldsymbol{y}$ the bulk, and
$\boldsymbol{\mu}$ the mean signature.

| Family | Metrics | Role for bulk $\boldsymbol{p}$ |
|:-------|:--------|:------------------------|
| Implemented now | MSE, RMSE, MAE, Pearson $r$, pseudo-$R^{2}$ | `compute_benchmark_metrics()` |
| Simplex / compositional | $L_1$, $L_2$, $L_{\infty}$, KL, Aitchison, Jensen–Shannon | HADACA3 error family [@barbotPromisesLimitsMultimodal2026]; $L_p$ and KL also appear in `imply` / RNA-Sieve evaluations [@mengImplyImprovingCelltype2024; @erdmann-phamLikelihoodbasedDeconvolutionBulk2021] |
| Concordance | Lin CCC, sample- and type-wise Pearson / Spearman | Scale-free agreement; hrRMSE is the relative counterpart [@baWhenLessNot2026] |
| Spatial reconstruction | RMSE, PCC, JSD, SSIM, cosine similarity, average ranking score | DANST-style spot reconstruction ([Commun. Biol.](https://www.nature.com/articles/s42003-026-09659-y#Sec9)); SSIM is image-valued and is not a simplex metric |
| Uncertainty | Wald / profile / bootstrap coverage of $\boldsymbol{p}$ | Only when $\boldsymbol{p}^{\star}$ is known (simulations) |

: Metric families for comparing $\hat{\boldsymbol{p}}$ with
$\boldsymbol{p}^{\star}$ (or $\hat{\boldsymbol{y}}$ with $\boldsymbol{y}$). {#tbl-metric-families}

::: {#nte-metric-choice .callout-note}
## Which metric to report

RMSE and MAE remain the default scalar scores in the package. Add an
$L_1$ / Aitchison / JSD panel when rare types matter (vertices of the
simplex), Lin CCC or type-wise correlation when a scale-free ranking is
needed, and interval coverage whenever the simulation knows
$\boldsymbol{p}^{\star}$. Do not treat SSIM as a drop-in replacement for
simplex $L_p$ scores.
:::

::: {#nte-related-dd .callout-info}
## Related literature (`muscat` / `scDD`)

Pseudobulk tools such as `muscat`
[@crowellMuscatDetectsSubpopulationspecific2020] aggregate single-cell
counts across samples. `scDD`
[@korthauerStatisticalApproachIdentifying2016] partitions genes into
EE, DE, DP, DM, and DB patterns. The hybrid scenario is a
network-level analogue of three of these (@fig-dd-taxonomy):
`equal_all` is EE; `marker_3` is DE for type 3; `shared_12_vs_3` has
**no mean shift between types 1 and 2**, only a precision-topology
contrast (DM-like for deconvolution, not scDD's original single-gene
modality test).
:::

![Schematic EE / DE / DM-like taxonomy for the three gene
blocks.](figures/fig_dd_taxonomy_ee_de_dm.jpg){#fig-dd-taxonomy width="100%"}

# References
