---
title: "Deconvolution use cases with DeCovarT"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Deconvolution use cases with DeCovarT}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup-use-cases
#| include: false
options(tinytable_html_mathjax = TRUE)
```

This vignette walks through two complementary simulation use cases for `DeCovarT`:

1. a **bivariate toy model** ($G=2$ genes, $J=2$ cell types) that isolates the effect of gene–gene correlation on mean-only versus covariance-aware deconvolution;
2. a **high-dimensional hybrid scenario** ($G=50$, $J=3$) with two mean-collinear cell types that must be separated by network topology, plus an orthogonal third type—then a head-to-head comparison of the package's frequentist solvers on imbalanced mixtures.

# Bivariate toy model: when covariance hurts mean-only deconvolution {#sec-bivariate-toy}

The manuscript's bivariate study (two genes, two cell populations) is implemented by [`benchmark_bivariate_gaussian_convolutions()`](../reference/benchmark_bivariate_gaussian_convolutions.html) in `R/02_01_toy_simulation.R`. The purpose is deliberately narrow: show that **even when mean expression profiles are similar**, accounting for the covariance between genes can (a) **degrade** standard mean-only deconvolution algorithms, and (b) be **partly alleviated** by DeCovarT's native inclusion of $\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$.

## Generative design

With the simplex constraint $p_1+p_2=1$, only one free unconstrained coordinate $\rho_1$ needs to be estimated (via the additive logistic map; see the [softmax / ALR vignette](softmax-alr-derivatives.html)). Bulk mixtures $\boldsymbol{Y}\in\mathcal{M}_{G\times N}$ are simulated without observation noise beyond the Gaussian convolution itself:

* **Proportions.** Equi-balanced $\boldsymbol{p}=(1/2,1/2)$ and highly unbalanced $\boldsymbol{p}=(0.95,0.05)$ (the exported helper also accepts intermediate unbalanced levels such as $(0.6,0.4)$).
* **Means.** Two centroid geometries: a **small** mean separation $\boldsymbol{\mu}_{.1}=(20,22)$, $\boldsymbol{\mu}_{.2}=(22,20)$ (high overlap) versus a **large** separation $\boldsymbol{\mu}_{.1}=(20,40)$, $\boldsymbol{\mu}_{.2}=(40,20)$.
* **Covariances.** Homoscedastic diagonals $\mathrm{diag}(\boldsymbol{\Sigma}_j)=\boldsymbol{I}_2$, with the pairwise gene–gene correlation $\mathrm{Cor}(x_1,x_2)$ swept over $\{-0.8,-0.6,\ldots,0.8\}$ **independently** for each cell type.

`benchmark_bivariate_gaussian_convolutions()` enumerates this factorial design with `tidyr::expand_grid()`, tags each scenario with a unique `ID`, simulates $N$ bulk replicates via `simulate_bulk_mixture()`, and calls `deconvolute_ratios()` for every algorithm in `deconvolution_functions`.

## How to run it

```{r}
#| label: lst-bivariate-call
#| eval: false
#| echo: true
library(DeCovarT)

deconvolution_functions <- list(
  "nnls" = list(FUN = deconvolute_ratios_nnls),
  "lsei" = list(FUN = deconvolute_ratios_deconrnaseq),
  "Marquardt-Levenberg" = list(
    FUN = deconvolute_ratios_Marquardt_Levenberg,
    additional_parameters = list(epsilon = 1e-3, itmax = 200)
  )
)

bivariate <- benchmark_bivariate_gaussian_convolutions(
  proportions = list(
    "balanced" = c(0.5, 0.5),
    "highly unbalanced" = c(0.95, 0.05)
  ),
  signature_matrices = list(
    "small CLD" = matrix(c(20, 22, 22, 20), nrow = 2),
    "large CLD" = matrix(c(20, 40, 40, 20), nrow = 2)
  ),
  corr_sequence = seq(-0.8, 0.8, by = 0.2),
  diagonal_terms = list("homoscedastic" = c(1, 1)),
  deconvolution_functions = deconvolution_functions,
  n = 500,
  cores = 1
)
```

The returned list has two tibbles: `config` (design, Shannon entropy of $\boldsymbol{p}$, MixSim overlap) and `simulations` (per-sample estimates and error metrics). The article compares DeCovarT against DeconRNASeq / NNLS [@gongDeconRNASeqStatisticalFramework2013]: overlap of the two cell-type Gaussians is a strong proxy for estimation quality—the lower the overlap, the lower the MSE—while **strong gene–gene correlation at fixed means** inflates mean-only error and is partially recovered once $\boldsymbol{\Sigma}_j$ enters the likelihood (ComplexHeatmap panels in the manuscript).


# High-dimensional hybrid scenario and deconvolution {#sec-scenario-grid}

The earlier factorial design swept one topology and one cosine level at a time, keeping a single **shared** covariance across all $J$ cell types (`simulate_hierarchical_grn_moments()`; see [GGM networks](synthetic-scenarios.html#sec-ggm-networks)). `scripts/generate_random_markov_network.R` instead builds **one** scenario that stresses several axes of the feature-selection pipeline simultaneously: two cell types that are unrecoverable from $\boldsymbol{\mu}$ alone and must be told apart from **topology**, a third cell type set apart by a compact marker block, and a null block that should survive every selection stage as *discarded*. Notation matches the manuscript: $G=50$ genes, $J=3$ cell types, $N$ bulk samples for $\boldsymbol{Y}\in\mathcal{M}_{G\times N}$.

## Gene and cell-type design

The $G=50$ genes are partitioned into three blocks, and each cell type gets its **own** $G\times G$ precision $\boldsymbol{\Omega}_j$ (not a shared one):

$$
\underbrace{\mathcal{G}_{12}}_{30\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{3}}_{10\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{\mathrm{eq}}}_{10\text{ genes}},
\qquad
\boldsymbol{\Omega}_j
=
\mathrm{blockdiag}\bigl(
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{12})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{3})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{\mathrm{eq}})}
\bigr),
\qquad j=1,2,3.
$$

$\boldsymbol{\mu}$ is built by **two** iterative calls to `generate_mean_signature_matrix()` ([Eq. mean cosine](synthetic-scenarios.html#eq-mean-cosine)): first on $\mathcal{G}_{12}$ for the pair (cell type 1, cell type 2) with a **high** target cosine $\rho_{12}=0.95$ (near-collinear, per [mean signature construction](synthetic-scenarios.html#sec-lr-block-mu)); then on a pool of $2\times10$ genes for the pair (merged $\{1,2\}$, cell type 3) with a **low** target cosine $\rho_{3}=0.05$ (near-orthogonal), keeping only the 10 genes whose private block belongs to cell type 3. $\mathcal{G}_{\mathrm{eq}}$ gets a flat baseline directly (identical mean in all three types by construction). Only $\mathcal{G}_{12}$ and $\mathcal{G}_3$ carry a mean signal; cell types 1 and 2 are, by design, mean-indistinguishable on $\mathcal{G}_{12}$.

Following the BLGGM hybrid design ([GGM networks](synthetic-scenarios.html#sec-ggm-networks)), each block's **local** topology is set independently per cell type, with no edges between blocks:

| Gene block | # genes | Cell type 1 | Cell type 2 | Cell type 3 |
|---|---|---|---|---|
| `shared_12_vs_3` | 30 | stochastic block model (hub-like modules) | star (single key-driver gene) | Erdős–Rényi (background) |
| `marker_3` | 10 | Erdős–Rényi (background) | Erdős–Rényi (background) | scale-free |
| `equal_all` | 10 | Erdős–Rényi | Erdős–Rényi | Erdős–Rényi |

: Local topology assigned to each (gene block, cell type) pair; no edges connect different blocks (block-diagonal precision support). {#tbl-hybrid-topology-design}

Cell types 1 and 2 therefore differ **only** on `shared_12_vs_3`, and only through $\boldsymbol{\Omega}$: hub/stochastic-block modules (cascading-pathway-like) for cell type 1 versus a single star (one key-driver gene) for cell type 2. `equal_all` is wired identically (Erdős–Rényi) in every cell type, matching its complete lack of mean signal.

```{r}
#| label: setup-hybrid-scenario
#| warning: false
#| message: false
#| eval: true
#| echo: false
#| output: false

if (requireNamespace("DeCovarT", quietly = TRUE)) {
  library(DeCovarT)
} else {
  stop("Package 'DeCovarT' is not installed.")
}

set.seed(20260807)

n_g12 <- 30L
n_g3 <- 10L
n_geq <- 10L
n_genes <- n_g12 + n_g3 + n_geq
celltype_names <- c("celltype_1", "celltype_2", "celltype_3")
idx_g12 <- seq_len(n_g12)
idx_g3 <- n_g12 + seq_len(n_g3)
idx_geq <- n_g12 + n_g3 + seq_len(n_geq)

mean_scale <- 10
mu_12 <- generate_mean_signature_matrix(
  n_genes = n_g12,
  n_celltypes = 2L,
  mean_scale = mean_scale,
  target_cosine = 0.95,
  celltype_names = c("celltype_1", "celltype_2")
)
mu_3_pool <- generate_mean_signature_matrix(
  n_genes = 2L * n_g3,
  n_celltypes = 2L,
  mean_scale = mean_scale,
  target_cosine = 0.05,
  celltype_names = c("celltype_12_merged", "celltype_3")
)
mu_3_block <- mu_3_pool[(n_g3 + 1L):(2L * n_g3), , drop = FALSE]

mu <- matrix(NA_real_, n_genes, 3L, dimnames = list(NULL, celltype_names))
mu[idx_g12, "celltype_1"] <- mu_12[, "celltype_1"]
mu[idx_g12, "celltype_2"] <- mu_12[, "celltype_2"]
mu[idx_g12, "celltype_3"] <- 0.5
mu[idx_g3, "celltype_1"] <- mu_3_block[, "celltype_12_merged"]
mu[idx_g3, "celltype_2"] <- mu_3_block[, "celltype_12_merged"]
mu[idx_g3, "celltype_3"] <- mu_3_block[, "celltype_3"]
mu[idx_geq, ] <- 2

precision_shift <- 0.1
precision_scale <- 0.3
prop_inhibitory <- 0.5

block_defs <- list(
  celltype_1 = list(
    list(idx_g12, "stochastic_block_model", list(block_prob = c(0.4, 0.3, 0.3))),
    list(idx_g3, "erdos_renyi", list()),
    list(idx_geq, "erdos_renyi", list())
  ),
  celltype_2 = list(
    list(idx_g12, "hub", list(n_hubs = 1L)),
    list(idx_g3, "erdos_renyi", list()),
    list(idx_geq, "erdos_renyi", list())
  ),
  celltype_3 = list(
    list(idx_g12, "erdos_renyi", list()),
    list(idx_g3, "scale_free", list(power = 1, edges_per_node = 1L)),
    list(idx_geq, "erdos_renyi", list())
  )
)
topology_label <- c(
  celltype_1 = "SBM (hub-like modules)",
  celltype_2 = "star (single key driver)",
  celltype_3 = "scale-free (marker genes)"
)

spectral_rows <- list()
for (ct in celltype_names) {
  adjacency <- matrix(0L, n_genes, n_genes)
  for (blk in block_defs[[ct]]) {
    idx <- blk[[1]]
    adjacency[idx, idx] <- DeCovarT:::generate_random_network_skeleton(
      n_genes = length(idx),
      graph_model = blk[[2]],
      graph_params = blk[[3]]
    )
  }
  W <- DeCovarT:::assign_iid_signed_weights(
    adjacency,
    prop_inhibitory = prop_inhibitory,
    weight_magnitude = precision_scale
  )
  Omega <- DeCovarT:::build_normalised_precision(W, precision_shift)
  evals <- eigen(Omega, symmetric = TRUE, only.values = TRUE)$values
  upper_w <- W[upper.tri(W) & adjacency != 0L]
  spectral_rows[[ct]] <- data.frame(
    `cell type` = ct,
    topology = topology_label[[ct]],
    lambda_min = min(evals),
    lambda_max = max(evals),
    kappa_Omega = max(evals) / min(evals),
    prop_inhib = if (length(upper_w)) mean(upper_w > 0) else NA_real_,
    check.names = FALSE
  )
}
spectral_df <- do.call(rbind, spectral_rows)
rownames(spectral_df) <- NULL

pairs <- utils::combn(celltype_names, 2L)
pairwise_rows <- list()
for (k in seq_len(ncol(pairs))) {
  ct_j <- pairs[1L, k]
  ct_l <- pairs[2L, k]
  mu_j <- mu[, ct_j]
  mu_l <- mu[, ct_l]
  pairwise_rows[[k]] <- data.frame(
    pair = paste(ct_j, "vs", ct_l),
    cosine = sum(mu_j * mu_l) / sqrt(sum(mu_j^2) * sum(mu_l^2)),
    euclidean = sqrt(sum((mu_j - mu_l)^2))
  )
}
pairwise_df <- do.call(rbind, pairwise_rows)
```

```{r}
#| label: tbl-mean-geometry
#| tbl-cap: "Realised pairwise cosine and Euclidean distance of the 50 x 3 mean signature, over the full G = 50 genes."
#| warning: false
#| message: false
#| echo: false
tinytable::tt(pairwise_df, digits = 3) |>
  tinytable::style_tt(j = 2:3, align = "c")
```

```{r}
#| label: tbl-topology-diagnostics
#| tbl-cap: "Per-cell-type precision spectrum for the 50 x 50 hybrid-topology Omega_j."
#| warning: false
#| message: false
#| echo: false
# Math-friendly column names for HTML (markdown + KaTeX via format_tt).
spectral_tt <- spectral_df
names(spectral_tt) <- c(
  "cell type",
  "topology",
  "$\\lambda_{\\min}$",
  "$\\lambda_{\\max}$",
  "$\\kappa(\\Omega)$",
  "prop inhib"
)
tinytable::tt(spectral_tt, digits = 3) |>
  tinytable::style_tt(j = 3:6, align = "c") |>
  tinytable::format_tt(j = 3:5, markdown = TRUE)
```

Column guide: $\lambda_{\min}$, $\lambda_{\max}$, and $\kappa(\boldsymbol{\Omega})=\lambda_{\max}/\lambda_{\min}$ summarise the precision spectrum of each cell type's *own* $\boldsymbol{\Omega}_j$; `prop_inhib` is the realised fraction of inhibitory precision edges. @tbl-mean-geometry shows cell types 1 and 2 near-collinear (cosine $\approx 0.97$), tracking the local target $\rho_{12}=0.95$ used on `shared_12_vs_3`. Cell type 3, however, sits at only a *moderate* cosine ($\approx0.56$) from the other two—well above the local target $\rho_{3}=0.05$ used on the `marker_3` pool. The gap is exactly the finite-sample effect of [asymptotic cosine control](synthetic-scenarios.html#sec-asymptotic-cosine): the flat `celltype_12_merged` background on `shared_12_vs_3` and the identical `equal_all` baseline both contribute a strictly positive term to *every* pairwise inner product over the full $G=50$ vector, diluting the block-local orthogonality of `marker_3` alone. @tbl-topology-diagnostics shows that, despite the shared mean on `shared_12_vs_3`, cell types 1 and 2 carry distinct, well-conditioned precision spectra—exactly the topology-only separation the scenario is designed to require.

@fig-network-topologies renders the three realised networks with `igraph`, coloured by gene block and by precision-edge sign.

![Cell-type-specific block topologies for the 50-gene hybrid scenario: a stochastic-block/hub-like module structure for cell type 1, restricted to the 30 `shared_12_vs_3` genes; a single star (one key-driver gene) for cell type 2 on the same 30 genes; and a scale-free network for cell type 3 restricted to its own 10 `marker_3` genes. The `equal_all` block (grey) is wired as Erdős–Rényi in every cell type. Edge colour encodes the sign of the precision entry (red = inhibitory, teal = activatory); node colour encodes the gene block.](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

The full pipeline built on this scenario—mean signature, per-cell-type topologies, $N=200$ bulk simulation, and the pre-selection / NSGA-II feature-selection stages it stresses—is reproduced end-to-end in `scripts/generate_random_markov_network.R`.

## Deconvolution on imbalanced mixtures {#sec-hybrid-deconvolution}

`scripts/deconvolute_simulated_scenario.R` loads the NSGA-II-curated moments from `data/synthetic_networks/true_grn_moments.rds`, restricts to the final gene panel, simulates $N=20$ bulk mixtures with **imbalanced** true composition $\boldsymbol{p}=(0.4,0.4,0.2)$ (equal mass on the two mean-collinear types; $0.2$ on the orthogonal type), and runs every frequentist solver from an equi-balanced start. @tbl-solver-diagnostics summarises a diagnostic run **before** the numerical fixes documented in the [softmax / ALR vignette](softmax-alr-derivatives.html#sec-numerical-speedups): stalled Newton–Raphson (`eval.max = 1`), slow Marquardt–Levenberg (wrong Hessian sign under `marqLevAlg`), and the benefit of polishing simulated annealing with a local Newton step.

```{r}
#| label: tbl-solver-diagnostics
#| tbl-cap: "Per-solver summary on 20 imbalanced hybrid bulk mixtures (true p = (0.4, 0.4, 0.2)), diagnostic run before numerical fixes in R/03_03_DeCovarT_estimate_ratios_frequentist.R."
#| echo: false
solver_diag <- data.frame(
  algorithm = c(
    "gradient_descent (BFGS)",
    "Newton_Raphson (as shipped)",
    "Newton_Raphson_fixed",
    "sann_then_polish",
    "simulated_annealing",
    "L_BFGS_B",
    "Marquardt_Levenberg"
  ),
  n_ok = c(20, 0, 20, 20, 18, 20, 18),
  n_flagged = c(0, 20, 0, 0, 2, 0, 2),
  `mean max|error|` = c(0.145, 0.133, 0.102, 0.102, 0.110, 0.100, 0.102),
  `mean time/sample` = c(
    "2.6 ms", "8.6 ms", "39 ms", "55.6 ms", "58.7 ms", "86.9 ms", "1662 ms"
  ),
  check.names = FALSE
)
tinytable::tt(solver_diag, digits = 3) |>
  tinytable::style_tt(j = 2:5, align = "c")
```

Key reading of @tbl-solver-diagnostics:

* **`Newton_Raphson` (as shipped)** never left the equi-balanced start (`n_ok = 0`): `stats::nlminb()` was capped at a single objective evaluation (`eval.max = 1`).
* **`Newton_Raphson_fixed`** (same code without that control) and **`sann_then_polish`** both recover mean max$|$error$|\approx 0.102$, matching a correctly configured second-order step.
* **`Marquardt_Levenberg`** already reached similar accuracy but at $\sim$1.7 s/sample because the relative-distance-to-minimum (RDM) criterion never fired (Hessian sign convention under `marqLevAlg`; see [numerical notes](softmax-alr-derivatives.html#sec-numerical-speedups)).
* **`simulated_annealing`** stalled on 2/20 samples; polishing with Newton after a SANN warm start removed those stalls without a proximal simplex projection (the ALR map already lands on the open simplex).

After the package fixes (removed `eval.max = 1`, analytic gradient for L-BFGS-B, cached Cholesky factorisation of $\boldsymbol{\Sigma}(\boldsymbol{p})$, negated Hessian for `marqLevAlg`), Newton–Raphson, Marquardt–Levenberg and `sann_then_polish` agree to about four decimals, and Marquardt–Levenberg drops to tens of milliseconds per sample. Reproduce the current comparison with:

```{r}
#| label: lst-hybrid-deconv
#| eval: false
#| echo: true
source("scripts/deconvolute_simulated_scenario.R")
```

## Related literature

Pseudobulk simulation tools such as **`muscat`** [@crowellMuscatDetectsSubpopulationspecific2020] aggregate single-cell counts across samples and conditions for differential-state testing; **`scDD`** [@korthauerStatisticalApproachIdentifying2016] partitions genes, relative to a reference condition, into five differential-distribution (DD) patterns: equivalent expression (**EE**), differential expression (**DE**, a mean shift with one mode preserved per condition), differential proportion (**DP**, a shift in the mixing weights of a shared bimodal pattern), differential modality (**DM**, a change in the number of modes without necessarily shifting the overall mean), and differential expression **and** modality combined (**DB**).

The hybrid scenario above instantiates a loose, network-level analogue of three of these patterns, tailored to deconvolution rather than to single-cell distribution testing directly (@fig-dd-taxonomy):

- the 10 `equal_all` genes are **EE**: identical mean and identically Erdős–Rényi-wired second-order structure in every cell type;
- the 10 `marker_3` genes are **DE** against the merged $\{1,2\}$ background: a mean shift confined to cell type 3, layered on a scale-free local topology;
- the 30 `shared_12_vs_3` genes are **DE** against cell type 3 but, between cell types 1 and 2 specifically, carry **no mean shift at all**—only the precision-matrix topology (@tbl-hybrid-topology-design) differs. This is not scDD's original single-gene DM test (unimodal versus bimodal *marginal* densities within one gene), but it plays a structurally similar role here: two populations a mean-only model cannot separate, which a covariance-aware model such as DeCovarT is designed to exploit.

![Schematic differential-distribution taxonomy for the three gene blocks: equivalent expression (EE, `equal_all`), differential expression marking cell type 3 (DE, `marker_3`), and a differential-modality-like contrast between cell types 1 and 2 that is invisible at the mean level and only resolved by network topology (DM-like, `shared_12_vs_3`).](figures/fig_dd_taxonomy_ee_de_dm.png){#fig-dd-taxonomy width="100%"}

DeCovarT's generator is narrower—it fixes $\boldsymbol{\mu}$ through [Eq. mean cosine](synthetic-scenarios.html#eq-mean-cosine) and [Eq. low-rank block](synthetic-scenarios.html#eq-lr-block) rather than estimating signatures from real scRNA-seq—but the same low-rank-plus-block intuition underpins many reference-matrix constructions cited above.

# References
