---
title: "MLE properties and asymptotic inference for the DeCovarT log-likelihood"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{MLE properties and asymptotic inference for the DeCovarT log-likelihood}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup
#| include: false
library(DeCovarT)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

DeCovarT estimates a composition $\boldsymbol{p}\in\Delta^{J-1}$ by
maximising the Gaussian-convolution log-likelihood

$$
\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}(\boldsymbol{p})
=
-\tfrac{1}{2}\log\det\boldsymbol{\Sigma}(\boldsymbol{p})
-\tfrac{1}{2}
(\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}
(\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})
+C,
\qquad
\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_{j=1}^{J}p_j^{2}\boldsymbol{\Sigma}_j .
$$ {#eq-loglik}

The derivatives of @eq-loglik are collected in the
[derivatives vignette](generative-model-derivatives.html). This companion
article asks what the resulting estimator is actually entitled to claim.
Two answers are uncomfortable and worth stating plainly: the realised
log-likelihood is **not** concave and its maximiser **need not be
unique**, and the Wald intervals of
[`confint.decovart_fit()`](../reference/fit_decovart.html) are
**undefined** on a simplex face. Both have concrete remedies, implemented
in [`lrt_decovart()`](../reference/lrt_decovart.html),
[`confint_profile_decovart()`](../reference/confint_profile_decovart.html),
[`bootstrap_decovart()`](../reference/bootstrap_decovart.html) and
[`reference_bootstrap_decovart()`](../reference/reference_bootstrap_decovart.html).

# Theoretical properties of the DeCovarT MLE {#sec-mle-properties}

## Identifiability {#sec-identifiability}

::: {#prp-identifiability}
## Identifiability of the composition

Two compositions induce the same law
$\mathcal{N}_G(\boldsymbol{\mu}\boldsymbol{p},
\boldsymbol{\Sigma}(\boldsymbol{p}))$ only if they coincide, provided the
mean map is injective on the simplex. Because
$\mathbf{1}^{\top}\boldsymbol{p}=1$ removes one degree of freedom,
injectivity of $\boldsymbol{p}\mapsto\boldsymbol{\mu}\boldsymbol{p}$
restricted to $\Delta^{J-1}$ requires the $J-1$ **contrast** columns
$\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot J}$,
$j=1,\ldots,J-1$, to be linearly independent, i.e.
$\operatorname{rank}(\boldsymbol{\mu})=J$ with $G\ge J$ in the generic
case.
:::

Identifiability is a property of the *reference*, not of the data: no
amount of bulk sequencing rescues a signature whose contrast columns are
collinear. Two cell types with proportional mean profiles are separated
only through $\boldsymbol{\Sigma}_j$, which is exactly the regime
DeCovarT targets — and exactly the regime where the mean-based condition
number $\kappa(\boldsymbol{\mu})$ is a misleading diagnostic (see the
[feature-selection vignette](feature-selection.html#sec-weighted-kappa)).
Near-collinearity is reported as a warning by
[`fit_decovart()`](../reference/fit_decovart.html) and studied
numerically in the
[appendix simulations](DeCovarT_appendix_simualtion_frameworks.html).

## Label permutation is not a bootstrap {#sec-label-switching}

@prp-identifiability identifies $\boldsymbol{p}$ *given a labelled
reference*. It does not identify the labels themselves. Write
$\pi$ for a permutation of $\{1,\ldots,J\}$, and write
$\boldsymbol{\mu}_{\pi}$, $\boldsymbol{\Sigma}_{\pi}$ and
$\boldsymbol{p}_{\pi}$ for the corresponding reordering of columns,
slices and coordinates. The Gaussian convolution is unchanged:

$$
\boldsymbol{\mu}_{\pi}\boldsymbol{p}_{\pi}
=
\boldsymbol{\mu}\boldsymbol{p},
\qquad
\boldsymbol{\Sigma}(\boldsymbol{p}_{\pi};\boldsymbol{\Sigma}_{\pi})
=
\boldsymbol{\Sigma}(\boldsymbol{p}).
$$ {#eq-label-equivariance}

::: {#thm-label-equivariance}
## Equivariance under cell-type relabelling

For every permutation $\pi$ of the cell types and every bulk profile
$\boldsymbol{y}$,

$$
\ell(\boldsymbol{p};\,\boldsymbol{y},\,\boldsymbol{\mu},\,\boldsymbol{\Sigma})
=
\ell(\boldsymbol{p}_{\pi};\,\boldsymbol{y},\,\boldsymbol{\mu}_{\pi},
\boldsymbol{\Sigma}_{\pi}).
$$

In particular the maximised log-likelihood does not depend on how the
purified columns are named, so a permutation of those names is not a
useful exchangeability null for $\boldsymbol{p}$.
:::

This is the same permutation symmetry that makes finite mixture models
identifiable **only up to labelling**. Ordinary $K$-component mixtures
are typically identified up to permutation of the component labels, which
is distinct from degeneracies such as coincident components or zero
mixing weights
[@xiaoBayesianIdentificationEstimation2025]. Under an exchangeable prior
the posterior inherits that symmetry: there are up to $K!$
permutation-equivalent representations, and component-specific labelled
summaries are meaningless until a relabelling convention is imposed
[@grunBayesianFiniteMixture2022; @newmanFastSamplingModel2025]. The
canonical frequentist and MCMC accounts are
@stephensDealingLabelSwitching2000, who showed that artificial
identifiability constraints do not in general solve the inferential
problem, @jasraMarkovChainMonte2005 for the MCMC review, and
@papastamoulisLabelSwitchingPackage2016 for systematic post-processing.

DeCovarT is *not* an unlabelled mixture of the bulk. Reviews of
transcriptomic deconvolution distinguish **reference-based** methods,
where a named signature supplies the cell-type identities, from
**reference-free** factorisations in which both the signature and the
proportions must be inferred
[@avilacobosComputationalDeconvolutionTranscriptomics2018;
@nguyenFourteenYearsCellular2024;
@sturmComprehensiveEvaluationTranscriptomebased2019;
@gaspard-boulincCelltypeDeconvolutionMethods2025]. Writing $Y\approx SP$
with a known, named $S$ already anchors “B cell”, “monocyte”, and so on.
Permuting the columns of $S$ together with the corresponding rows of $P$
is algebraically harmless; permuting $P$ alone, or shuffling the names of
an already averaged reference $X$, changes the biological meaning of the
fit and is not a valid bootstrap.

Gene permutation is not a bootstrap either. Genes are the coordinates of
the expression space. Reordering them jointly in the bulk *and* the
reference changes nothing; reordering them in the bulk alone against a
frozen signature destroys the gene-wise pairing and is not a source of
sampling uncertainty.

```{r}
#| label: label-vs-gene-permutation
mu_lab <- matrix(
  c(20, 40, 15, 40, 20, 25, 25, 30, 35),
  nrow = 3,
  dimnames = list(paste0("g", 1:3), paste0("ct", 1:3))
)
sig_lab <- array(
  0,
  dim = c(3, 3, 3),
  dimnames = list(rownames(mu_lab), rownames(mu_lab), colnames(mu_lab))
)
for (j in 1:3) {
  sig_lab[, , j] <- diag(3) * c(1, 1.5, 1.2)[j]
}
p_lab <- c(0.5, 0.3, 0.2)
y_lab <- drop(mu_lab %*% p_lab)
perm <- c(3L, 1L, 2L)
mu_perm <- mu_lab[, perm, drop = FALSE]
colnames(mu_perm) <- colnames(mu_lab)
sig_perm <- sig_lab[, , perm, drop = FALSE]
dimnames(sig_perm) <- dimnames(sig_lab)

c(
  original = loglik_multivariate(p_lab, y_lab, mu_lab, sig_lab),
  cell_type_relabel = loglik_multivariate(
    p_lab[perm],
    y_lab,
    mu_perm,
    sig_perm
  ),
  gene_shuffle = loglik_multivariate(
    p_lab,
    y_lab[c(3L, 1L, 2L)],
    mu_lab,
    sig_lab
  )
)
```

The first two values agree to machine precision; the gene shuffle does
not. That is why
[`equivariance_check_decovart()`](../reference/equivariance_check_decovart.html)
only verifies the labelled relabelling, and why
[`reference_bootstrap_decovart()`](../reference/reference_bootstrap_decovart.html)
resamples donors or cells rather than names (@sec-reference-bootstrap).

True label switching returns if *both* $S$ and $P$ are unknown: then
$SP=(SQ)(Q^{-1}P)$ for every permutation matrix $Q$, and component 1 has
no intrinsic biological meaning across bootstrap or MCMC runs. The remedy
is to **align inferred signatures to an external anchor** after each
run, not to permute genes. Near-collinear cell types are a different
failure: if $S_{\cdot A}\approx S_{\cdot B}$, many splits with
$p_A+p_B$ fixed produce almost the same bulk, so individual fractions
wander while their sum is stable. Relabelling does not cure that
weak identification (@sec-identifiability; the
[appendix simulations](DeCovarT_appendix_simualtion_frameworks.html)
study collinear signatures numerically).

| Operation | Appropriate? | What it tests |
|:--|:--|:--|
| Bootstrap **cells within cell type** | Yes | Reference-signature uncertainty |
| Bootstrap **donors within cell type** | Favoured default | Biological / reference-population uncertainty |
| Resample **compositions** (Dirichlet) | Yes, as a composition sweep | Performance over mixture weights |
| Permute **cell-type labels** | Only as an equivariance check | Whether $\hat P$ relabels with $S$ |
| Permute **genes** | No, for ordinary inference | Destroys or trivialises expression structure |

: Reference-based resampling versus label or gene permutation. Only the first three rows estimate uncertainty; label permutation is a software check. {#tbl-reference-ops}

## Existence {#sec-existence}

::: {#prp-existence}
## Existence of a maximiser

If every $\boldsymbol{\Sigma}_j$ is symmetric positive definite, then
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^{2}\boldsymbol{\Sigma}_j$
is positive definite for every $\boldsymbol{p}\in\Delta^{J-1}$, since at
least one $p_j>0$ on the simplex. Hence @eq-loglik is finite and
continuous on the **compact** set $\Delta^{J-1}$, and attains its
supremum.
:::

Two practical consequences. First, positive definiteness is preserved
even when some proportions are exactly zero, which is what allows the
restricted maximisation of
[`restricted_mle_decovart()`](../reference/restricted_mle_decovart.html)
to represent a null such as $p_j=0$ exactly instead of pushing it through
a logarithm. Second, the maximiser may lie on the **boundary** of the
compact set, so first-order conditions $\nabla\ell=\boldsymbol{0}$ are
necessary only at interior optima.

## Uniqueness fails for one observed sample {#sec-uniqueness}

Existence is not uniqueness. The DeCovarT likelihood is a
*heteroscedastic* Gaussian model: the noise covariance depends on the
same parameter as the mean, so @eq-loglik is not the concave objective of
an ordinary linear regression. Even in the isotropic case
$\boldsymbol{\Sigma}_1=\boldsymbol{\Sigma}_2=\mathbf{I}_G$ with $J=2$ and
without the simplex constraint,

$$
\ell(\boldsymbol{p})
=
-\frac{G}{2}\log(p_1^{2}+p_2^{2})
-
\frac{\lVert\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p}\rVert^{2}}
{2(p_1^{2}+p_2^{2})}
+C ,
$$ {#eq-isotropic}

whose variance term moves with the coefficients. A single evaluation of
the analytic Hessian settles the question.

```{r}
#| label: nonconcavity
mean_signature <- matrix(c(2, 1, 1, 1, 2, 6 / 5), nrow = 3)
colnames(mean_signature) <- c("ct1", "ct2")
rownames(mean_signature) <- paste0("gene_", 1:3)
Sigma_iso <- array(c(diag(3), diag(3)), dim = c(3, 3, 2))
bulk <- drop(mean_signature %*% c(0.5, 0.5))

hessian_interior <- hessian_loglik_unconstrained(
  p = c(1 / 5, 3 / 10),
  y = bulk,
  mean_signature_matrix = mean_signature,
  Sigma = Sigma_iso
)
round(hessian_interior, 4)
c(
  determinant = det(hessian_interior),
  eigen(hessian_interior, symmetric = TRUE)$values
)
```

Every entry of $\boldsymbol{y}$ and $\boldsymbol{\mu}$ is strictly
positive, $\operatorname{rank}(\boldsymbol{\mu})=2$, $G=3>J=2$, and both
covariances are positive definite. Yet the Hessian at the strictly
interior point $\boldsymbol{p}=(1/5,3/10)$ has one positive and one
negative eigenvalue.

::: {#nte-nonconcave .callout-important}
## One indefinite Hessian disproves global concavity

A negative determinant at a single admissible point is a counterexample,
so the unconstrained DeCovarT log-likelihood is **not globally concave**
under the stated positivity and rank assumptions. The same conclusion
transfers to the additive log-ratio chart, because a $C^{2}$
diffeomorphism preserves the existence of directions of positive
curvature. Consequently:

- a converged optimiser certifies a **local** maximum at best;
- several starts may reach different modes (see @sec-diagnostics);
- convex-optimisation guarantees — a unique global optimum reachable by
  descent, and duality certificates — do not apply
  [@boydConvexOptimization2004]. The Marquardt–Levenberg damping of
  `marqLevAlg` [@R-marqLevAlg; @philippsRobustEfficientOptimization2021]
  and the relative-distance-to-minimum criterion of
  @commengesNewtonLikeAlgorithmLikelihood2006 buy robustness against
  saddle points, not global optimality.
:::

## Population consistency and efficiency {#sec-population}

The failure of concavity is a statement about *one realised* data set.
The population criterion behaves much better. Let $\boldsymbol{p}_0$ be
the truth and define
$Q(\boldsymbol{p};\boldsymbol{p}_0)
=\mathbb{E}_{\boldsymbol{p}_0}\{\ell_{\boldsymbol{Y}}(\boldsymbol{p})\}$.

::: {#thm-population}
## Population uniqueness through the Kullback–Leibler divergence

$$
Q(\boldsymbol{p}_0;\boldsymbol{p}_0)-Q(\boldsymbol{p};\boldsymbol{p}_0)
=
D_{\mathrm{KL}}\bigl(
  P_{\boldsymbol{p}_0}\,\Vert\,P_{\boldsymbol{p}}
\bigr)
\ \ge\ 0 ,
$$ {#eq-kl}

with equality if and only if
$P_{\boldsymbol{p}}=P_{\boldsymbol{p}_0}$. Under
@prp-identifiability that happens only at
$\boldsymbol{p}=\boldsymbol{p}_0$, so $\boldsymbol{p}_0$ is the
**unique** maximiser of the expected log-likelihood.
:::

@eq-kl is an identifiability statement, not a curvature statement: it
survives the non-concavity of @nte-nonconcave. It is also the engine of
the usual asymptotics — consistency, asymptotic normality and efficiency
of the MLE [@vaartAsymptoticStatistics2000] — because the local geometry
of the divergence *is* the Fisher information. By the
[Fisher information metric theorem](https://en.wikipedia.org/wiki/Kullback%E2%80%93Leibler_divergence#Fisher_information_metric_theorem),

$$
D_{\mathrm{KL}}\bigl(
  P_{\boldsymbol{p}_0}\,\Vert\,P_{\boldsymbol{p}_0+\boldsymbol{d}}
\bigr)
=
\tfrac{1}{2}\,
\boldsymbol{d}^{\top}I(\boldsymbol{p}_0)\,\boldsymbol{d}
+
o(\lVert\boldsymbol{d}\rVert^{2}) ,
$$ {#eq-fisher-metric}

with $I$ the expected information of
[`expected_fisher_unconstrained()`](../reference/expected_fisher_unconstrained.html);
the Gaussian case is treated in information-geometric terms by
@malagoInformationGeometryGaussian2015. The identity is worth checking
numerically, since it ties the plug-in `vcov` of a DeCovarT fit to the
divergence that makes the MLE consistent.

```{r}
#| label: kl-fisher
kl_gaussian_convolution <- function(p_ref, p_alt, mu, Sigma) {
  mean_ref <- drop(mu %*% p_ref)
  mean_alt <- drop(mu %*% p_alt)
  cov_ref <- .compute_global_variance(p_ref, Sigma)
  cov_alt <- .compute_global_variance(p_alt, Sigma)
  precision_alt <- solve(cov_alt)
  shift <- mean_alt - mean_ref
  0.5 *
    (sum(diag(precision_alt %*% cov_ref)) -
      nrow(cov_ref) +
      drop(t(shift) %*% precision_alt %*% shift) +
      as.numeric(determinant(cov_alt, logarithm = TRUE)$modulus) -
      as.numeric(determinant(cov_ref, logarithm = TRUE)$modulus))
}

p_reference <- c(0.4, 0.6)
information <- expected_fisher_unconstrained(
  p_reference,
  mean_signature,
  Sigma_iso
)
steps <- c(1e-2, 1e-3, 1e-4)
data.frame(
  step = steps,
  kullback_leibler = vapply(
    steps,
    function(h) {
      kl_gaussian_convolution(
        p_reference,
        p_reference + c(h, -h / 2),
        mean_signature,
        Sigma_iso
      )
    },
    numeric(1)
  ),
  quadratic_form = vapply(
    steps,
    function(h) {
      d <- c(h, -h / 2)
      0.5 * drop(t(d) %*% information %*% d)
    },
    numeric(1)
  )
)
```

The two columns agree to the order of $\lVert\boldsymbol{d}\rVert^{3}$,
as @eq-fisher-metric predicts.

## Equivariance: affine yes, logarithmic no {#sec-equivariance}

::: {#prp-equivariance}
## Gene-wise affine equivariance

Let $D=\operatorname{diag}(s_1,\ldots,s_G)$ with $s_g\neq 0$ and
$\boldsymbol{m}\in\mathbb{R}^{G}$. Applying the **same** gene-wise map to
all three inputs,

$$
\boldsymbol{y}^{\star}=D^{-1}(\boldsymbol{y}-\boldsymbol{m}),
\quad
\boldsymbol{\mu}^{\star}
=D^{-1}(\boldsymbol{\mu}-\boldsymbol{m}\mathbf{1}_J^{\top}),
\quad
\boldsymbol{\Sigma}_j^{\star}=D^{-1}\boldsymbol{\Sigma}_j D^{-1},
$$

leaves the MLE unchanged, because $\mathbf{1}_J^{\top}\boldsymbol{p}=1$
makes the recentring cancel and the Mahalanobis residual is invariant.
:::

A z-score computed **once per gene from $\boldsymbol{\mu}$** is such a
map, which is why `fit_decovart(standardise = TRUE)` is safe. Per-cell-type
or per-sample scaling is not, and neither is a logarithm: $\log$ is
concave, so Jensen's inequality shifts both moments and
$\log(\boldsymbol{\mu}\boldsymbol{p})
\neq(\log\boldsymbol{\mu})\boldsymbol{p}$. Reference-based deconvolution
therefore expects a non-log linear scale
[@newmanRobustEnumerationCell2015]: **keep raw counts** (or a linear
normalisation such as CPM/TPM) rather than $\log_2$ values. Monotonicity
of $\log$ preserves gene-wise rankings, but not the convolution MLE. The
numerical demonstration lives in the
[appendix simulations](DeCovarT_appendix_simualtion_frameworks.html#sec-affine).

# Asymptotic inference {#sec-inference}

## Where the estimate sits decides the null distribution {#sec-strata}

The admissible set is a polytope, and a composition can sit in its
interior or on a stratum of its boundary. With $J=4$ cell types
$\Delta^{3}$ is a tetrahedron whose strata are the interior, four
triangular faces, six edges and four vertices, corresponding to $q=0,1,2$
or $3$ proportions equal to zero (@fig-simplex-strata).

![Strata of the tetrahedron $\Delta^{3}$ for $J=4$ cell types. Vertices are pure populations, edges and faces are compositions with two or one absent cell types, and the interior holds compositions with every type present. Dashed lines are the edges hidden behind the far vertex `ct3`. Below each panel, $q$ counts the constraints active at the estimate and the red expression is the resulting null law of the likelihood-ratio statistic: the ordinary $\chi^2_1$ of @wilksLargeSampleDistributionLikelihood1938 in the interior, and binomially weighted chi-square mixtures on the boundary strata [@hermanDistributionLikelihoodRatio1954; @selfAsymptoticPropertiesMaximum1987].](figures/simplex_boundary_tangent_cones.png){#fig-simplex-strata width=100%}

The reason the calibration changes is geometric.
@selfAsymptoticPropertiesMaximum1987 show that the boundary problem is
asymptotically equivalent to estimating the **restricted mean** of a
single multivariate Gaussian draw: the limiting law of the normalised MLE
is the distribution of the *projection* of a Gaussian score onto the
**tangent cone** that approximates the admissible set at
$\boldsymbol{p}_0$. In the interior that cone is the whole of
$\mathbb{R}^{J-1}$ and the projection is the identity, recovering
classical theory; at a face it is a half-space, at an edge a
two-dimensional wedge, and at a vertex a simplicial cone. A visual
introduction to tangent and normal cones is available in these
[ESIEE optimisation notes](https://perso.esiee.fr/~chierchg/optimization/content/05/normal_cone.html).

## Interior nulls: Wald and profile likelihood {#sec-interior}

For an interior null $H_0:p_j=c$ with $0<c<1$, two statistics are
available.

The **Wald** statistic is the cheap one:
$(\hat{p}_j-c)/\mathrm{SE}_j$ with $\mathrm{SE}_j$ the delta-method
standard error of [`vcov_alr_delta()`](../reference/vcov_alr_delta.html),
giving the symmetric interval of
[`confint.decovart_fit()`](../reference/fit_decovart.html). It costs one
matrix inversion, but it linearises the additive log-ratio chart at
$\hat{\boldsymbol{p}}$ [@oehlertNoteDeltaMethod1992] and is therefore
*not* invariant to reparametrisation and can place limits outside
$[0,1]$.

The **profile likelihood-ratio** statistic

$$
D_j(c)
=
2\bigl\{
  \ell(\hat{\boldsymbol{p}})
  -
  \max_{\boldsymbol{p}\in\Delta^{J-1},\,p_j=c}\ell(\boldsymbol{p})
\bigr\}
\ \rightsquigarrow\ \chi^{2}_{1}
$$ {#eq-lrt}

concentrates out the other $J-2$ proportions instead. Because a smooth
bijection leaves likelihood ratios unchanged, @eq-lrt needs **no** delta
method and respects the simplex automatically. Inverting it gives the
asymmetric intervals of
[`confint_profile_decovart()`](../reference/confint_profile_decovart.html).

```{r}
#| label: interior-inference
signature <- matrix(
  c(20, 40, 15, 40, 20, 25, 25, 30, 35),
  nrow = 3,
  dimnames = list(paste0("gene_", 1:3), paste0("ct", 1:3))
)
Sigma <- array(
  0,
  dim = c(3, 3, 3),
  dimnames = list(
    rownames(signature),
    rownames(signature),
    colnames(signature)
  )
)
for (j in 1:3) {
  Sigma[, , j] <- diag(3) * c(1, 1.5, 1.2)[j]
}

# Eight technical replicates of ONE composition, so the asymptotics of
# Sec. 2.4 are taken in a meaningful direction.
set.seed(11)
replicates <- simulate_bulk_mixture(
  signature_matrix = signature,
  Sigma = Sigma,
  p = c(0.45, 0.35, 0.20),
  n = 8
)$Y

round(
  confint_profile_decovart(
    bulk_expression = replicates,
    mean_signature_matrix = signature,
    Sigma = Sigma
  ),
  4
)
```

```{r}
#| label: interior-lrt
lrt_decovart(
  bulk_expression = replicates,
  mean_signature_matrix = signature,
  Sigma = Sigma,
  null_value = c(ct1 = 0.30)
)
```

## Boundary nulls: chi-bar-square mixtures {#sec-boundary}

The scientifically interesting null is usually **absence**:
$H_0:p_j=0$ against $H_1:p_j\ge 0$. Here the parameter may move away from
the null in one direction only, so the local parameter space is a
half-line rather than a vector space, and Wilks' theorem does not apply.

::: {#thm-chibar}
## Boundary calibration

With $q$ constraints active on the boundary and $s$ interior constraints
tested simultaneously, and the corresponding information block
orthogonal to the nuisance block,

$$
D
\ \rightsquigarrow\
\bar{\chi}^{2}
=
\sum_{i=0}^{q}
\binom{q}{i}2^{-q}\,
\chi^{2}_{s+i},
$$ {#eq-chibar}

where $\chi^{2}_{0}$ denotes a point mass at zero
[@selfAsymptoticPropertiesMaximum1987, Case 9]. For a single active
constraint this is the $\tfrac{1}{2}\chi^{2}_{0}+\tfrac{1}{2}\chi^{2}_{1}$
law of @hermanDistributionLikelihoodRatio1954, so the correct one-sided
p-value is $\tfrac{1}{2}\Pr(\chi^{2}_{1}>D)$ — **half** the naive value.
:::

[`chi_bar_square_pvalue()`](../reference/chi_bar_square_pvalue.html)
evaluates @eq-chibar, and
[`lrt_decovart()`](../reference/lrt_decovart.html) selects it
automatically when a tested value reaches 0 or 1.

```{r}
#| label: boundary-lrt
set.seed(11)
replicates_absent <- simulate_bulk_mixture(
  signature_matrix = signature,
  Sigma = Sigma,
  p = c(0.5, 0.5, 0),
  n = 8
)$Y

lrt_decovart(
  bulk_expression = replicates_absent,
  mean_signature_matrix = signature,
  Sigma = Sigma,
  null_value = c(ct3 = 0)
)
```

::: {#nte-chibar-caveats .callout-warning}
## Three caveats on @eq-chibar

1. **Binomial weights are conditional.** They require the active block of
   the information matrix to be orthogonal to the nuisance block. When
   several absent cell types share signal, the mixing weights depend on
   the geometry of the tangent cone; @selfAsymptoticPropertiesMaximum1987
   exhibit a configuration (their Case 8, a *nuisance* parameter on the
   boundary) whose limit is not a chi-square mixture at all. Pass explicit
   `weights`, or calibrate by simulation (@sec-bootstrap).
2. **Do not encode the null in ALR coordinates.** $p_j=0$ corresponds to
   $\rho_j=-\infty$, so the chart cannot represent the hypothesis. The
   restricted fits substitute the constrained value directly.
3. **Wald is unusable here.** The delta-method covariance is undefined on
   the face; `vcov_alr_delta()` returns `NA` with a warning rather than a
   plausible-looking number.
:::

## Replication is what makes the asymptotics meaningful {#sec-replication}

@eq-lrt and @eq-chibar are limits as the amount of information about
**one** composition grows. That is a strong requirement in deconvolution.

For each bulk sample $i$, DeCovarT observes a single $G$-vector
$\boldsymbol{Y}_{\cdot i}\sim\mathcal{N}_G(\boldsymbol{\mu}
\boldsymbol{p}_{\cdot i},\boldsymbol{\Sigma}(\boldsymbol{p}_{\cdot i}))$
and estimates its own $\boldsymbol{p}_{\cdot i}$. Increasing the number
of samples $N$ therefore adds parameters exactly as fast as
observations, and provides **no** extra replication for any single
$\boldsymbol{p}_{\cdot i}$. The $G$ genes are not $G$ independent
observations either: their dependence through
$\boldsymbol{\Sigma}(\boldsymbol{p})$ is the whole point of the model.

::: {#nte-replication .callout-important}
## Per-sample intervals are plug-in approximations

The default per-sample `vcov` and `confint` of a `decovart_fit` are
one-observation plug-in quantities. Treat them as descriptive curvature
summaries, not as calibrated frequentist intervals.

Two things restore the asymptotics, and both are implemented:

- **Pool genuine replicates.** Supply several bulk columns that really
  do share one composition (technical replicates, or aliquots of one
  tissue) through the `bulk_expression` argument of
  `lrt_decovart()` / `confint_profile_decovart()`. Those functions
  maximise the *summed* log-likelihood over the replicates, so the
  limit is taken in the number of replicates.
- **Simulate instead of appealing to a limit** (@sec-bootstrap).

Reference uncertainty is a third, separate gap: $\boldsymbol{\mu}$ and
$\boldsymbol{\Sigma}_j$ are plug-in estimates from purified populations,
and none of the intervals above propagate their sampling variability.
[`reference_bootstrap_decovart()`](../reference/reference_bootstrap_decovart.html)
resamples those profiles (@sec-reference-bootstrap); see
@vellameUncertaintyQuantificationReferencebased2023 for the analogous
accuracy metric in methylation deconvolution.
:::

## Parametric bootstrap calibration {#sec-bootstrap}

The conditional law of the bulk profile is fully specified, so the
asymptotic mixture can be replaced by simulation.
[`bootstrap_decovart()`](../reference/bootstrap_decovart.html) draws
$\boldsymbol{Y}^{(b)}\sim\mathcal{N}_G(\boldsymbol{\mu}\boldsymbol{p},
\boldsymbol{\Sigma}(\boldsymbol{p}))$, refits, and — when `null_value` is
supplied — regenerates under the **restricted** MLE so the
likelihood-ratio statistic is recalibrated without invoking the
$50{:}50$ weights [@rochModernDiscreteProbability2024 reviews the
concentration tools behind such Monte Carlo guarantees].

The atom at zero is directly observable: under a true boundary null,
about half of the replicates should return $D=0$, because the
unrestricted MLE itself falls on the face.

```{r}
#| label: bootstrap-calibration
set.seed(5)
calibration <- bootstrap_decovart(
  bulk_expression = replicates_absent,
  mean_signature_matrix = signature,
  Sigma = Sigma,
  null_value = c(ct3 = 0),
  n_boot = 150
)
c(
  observed_statistic = calibration$statistic,
  monte_carlo_p = calibration$p_value,
  share_of_zero_statistics = mean(calibration$null_statistics < 1e-8),
  asymptotic_p = chi_bar_square_pvalue(
    calibration$statistic,
    n_boundary = 1L
  )
)
```

The observed share of zero statistics is the empirical counterpart of the
$\tfrac{1}{2}\chi^{2}_{0}$ atom in @eq-chibar. A restricted bootstrap is
still only *approximate* for a composite null, because the nuisance
proportions are held at their restricted estimates rather than profiled
over; it is exact only when $\boldsymbol{p}_0$ is fully specified.

## Reference-sample bootstrap {#sec-reference-bootstrap}

[`bootstrap_decovart()`](../reference/bootstrap_decovart.html) treats
$\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}_j$ as known. They are not:
both are estimated from purified profiles.
[`reference_bootstrap_decovart()`](../reference/reference_bootstrap_decovart.html)
resamples the **experimental units of that labelled reference**, rebuilds
the plug-in moments, and refits, so reference uncertainty appears in the
percentile intervals [@efronBootstrapMethodsAnother1979]. Cell-type names
stay attached throughout. The three `method` options match
@tbl-reference-ops:

- **`donors`** (default). Cluster bootstrap of donors *within each cell
  type*. A donor drawn twice contributes its purified columns twice.
  This is the favoured option when donor labels exist: it captures
  biological variation in the reference population rather than treating
  cells from one individual as independent.
- **`cells`**. Column resampling within each cell type, for when donor
  labels are unavailable.
- **`dirichlet`**. Draw $\boldsymbol{p}^{(b)}\sim\mathrm{Dirichlet}(\boldsymbol{\alpha})$,
  simulate $\boldsymbol{Y}^{(b)}$ from the Gaussian convolution at the
  original plug-in moments, and refit. This is a composition sweep, not
  a confidence interval for one observed bulk.

By default the observed bulk is held fixed for `donors` and `cells`
(reference uncertainty for a given mixture). Setting
`regenerate_bulk = TRUE` follows the fuller pipeline
$\text{reference units}\to S^{(b)}\to Y^{(b)}\to\hat{\boldsymbol{p}}^{(b)}$.

A related design point for **simulation studies**, not implemented as an
argument here, is to split reference-building cells from
pseudobulk-generating cells, ideally at the donor level. Using the same
cells, or cells from the same individual, in both the signature and the
artificial bulk makes a benchmark unrealistically easy
[@sturmComprehensiveEvaluationTranscriptomebased2019;
@avilacobosBenchmarkingCellType2020;
@zhaoEvaluatingReferencemixtureMatching2026].
@nguyenFourteenYearsCellular2024 review 53 bulk methods and
stress incomplete references and missing cell types; @avilacobosBenchmarkingCellType2020
show that omitted types and the expression scale dominate error on
pseudobulk mixtures. The simulation generator used to evaluate `imply`
is a useful reference-based template: cell-type parameters are drawn
from a multivariate Gaussian, purified-like expression is realised from
a Gamma (negative-binomial-like) law fitted to RNA-seq counts, and bulk
mixtures are then formed from those profiles with Dirichlet
compositions [@mengImplyImprovingCelltype2024]. That is distinct from
DeCovarT's Gaussian convolution of the *bulk*; it is a design for how to
build a labelled $S$ and a synthetic $Y$ with known $p$.

```{r}
#| label: reference-bootstrap
set.seed(9)
refs <- lapply(seq_len(ncol(signature)), function(j) {
  draws <- MASS::mvrnorm(
    n = 12L,
    mu = signature[, j],
    Sigma = Sigma[, , j]
  )
  out <- t(draws)
  rownames(out) <- rownames(signature)
  out
})
names(refs) <- colnames(signature)
donor_ids <- lapply(refs, function(x) {
  rep(paste0("d", 1:3), length.out = ncol(x))
})
ref_boot <- reference_bootstrap_decovart(
  bulk_expression = replicates,
  reference_profiles = refs,
  donor_ids = donor_ids,
  method = "donors",
  n_boot = 40,
  itmax = 80
)
ref_boot$interval
```

[`equivariance_check_decovart()`](../reference/equivariance_check_decovart.html)
is the software counterpart of @thm-label-equivariance: permute the
signature columns, refit, and confirm that $\hat{\boldsymbol{p}}$ merely
relabels. It is not the distribution used for confidence intervals.

```{r}
#| label: equivariance-check
eq_check <- equivariance_check_decovart(
  y = replicates[, 1],
  mean_signature_matrix = signature,
  Sigma = Sigma,
  perm = c(3L, 1L, 2L),
  itmax = 80
)
c(
  max_abs_diff = eq_check$max_abs_diff,
  loglik_diff = eq_check$loglik_diff
)
```

## RNA-Sieve: CLT likelihood, Fisher and Godambe {#sec-rna-sieve}

Align notation with @eq-loglik. `RNA-Sieve` writes $M$, $\alpha$ and $b$
for DeCovarT's $\boldsymbol{\mu}$, $\boldsymbol{p}$ and
$\boldsymbol{y}$ [@erdmann-phamLikelihoodbasedDeconvolutionBulk2021].
The bulk is the sum of $n$ independent cells. A gene-wise central-limit
theorem then supplies a Gaussian likelihood whose variance
$s_{g}^{2}(M,\alpha,S)$ depends on $\alpha$. The estimated mean matrix
is itself noisy,
$\varepsilon_{M,gk}\sim\mathcal{N}(0,s_{gk}/c_{k})$, so the design is
an errors-in-variables (total-least-squares) problem, not a convolution
of $J$ latent type profiles. Genes are treated as independent;
coexpression is left for later work.

DeCovarT draws one latent profile per cell type, $\boldsymbol{x}_{\cdot j}\sim\mathcal{N}_{G}(\boldsymbol{\mu}_{\cdot j},
\boldsymbol{\Sigma}_{j})$, and observes the weighted sum
$\boldsymbol{y}=\sum_{j}p_{j}\boldsymbol{x}_{\cdot j}$, hence
$\boldsymbol{y}\sim\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_{j}p_{j}^{2}\boldsymbol{\Sigma}_{j})$ with a *full* gene--gene
covariance. The plug-in $(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$ is
treated as known in the likelihood. RNA-Sieve's $n$-cell average is a
univariate CLT on each gene; DeCovarT's $p_{j}^{2}\boldsymbol{\Sigma}_{j}$
term is a finite-$J$ multivariate convolution.

The interval constructions differ in the same way. When the CLT model
holds, RNA-Sieve uses the inverse Fisher information of
$(\hat\alpha,\hat n)$ after a simplex reparametrisation, and reports
$\chi^{2}_{K-1}$ ellipsoids. When protocol shift takes the truth
outside the model family, they replace Fisher by the **Godambe**
sandwich, which describes Gaussian fluctuations around the
Kullback--Leibler projection onto the fitted family. In the correctly
specified case the two matrices coincide. Near-collinear cell types
make the information nearly singular; RNA-Sieve then subsamples genes.

DeCovarT already has the Fisher analogue: the expected information of
the convolution mapped through the ALR (@sec-interior; the derivatives
vignette). That Wald interval is the cheap interior tool, and it is
**undefined** on a simplex face (@sec-boundary). The Godambe sandwich
would be relevant *if* the multivariate Gaussian convolution is
misspecified (counts, protocol shift, omitted types). It does not
replace the chi-bar-square LRT or the restricted parametric bootstrap
on the boundary, and it does not replace
`reference_bootstrap_decovart()`, which is the resampling counterpart of
RNA-Sieve's noisy-$M$ term. Implementing a Godambe correction for
DeCovarT is an outlook item, not a present interval.

# Convergence and boundary diagnostics {#sec-diagnostics}

Because @nte-nonconcave rules out global guarantees, a single optimiser
return code carries less information than it appears to.
[`fit_decovart()`](../reference/fit_decovart.html) therefore records
three logically distinct claims, and refuses to conflate them:

| Field | Established by | Does **not** establish |
|:--|:--|:--|
| `converged` (`istop` / `code`, `rdm`, `iterations`) | numerical stopping rules met | that the point is a maximum |
| `local_maximum` (small ALR score norm, negative $\lambda_{\max}(\mathbf{H}_{\boldsymbol{\rho}})$) | stationarity with the right local curvature | global optimality |
| `near_boundary` ($\min_j\hat{p}_j$ small) | proximity to a simplex face | optimiser failure |
| `multimodal` (spread of log-likelihoods over random starts) | evidence *for* several modes | absence of further modes |

: What each DeCovarT convergence field does and does not certify. {#tbl-diagnostics}

[`boundary_diagnostics()`](../reference/boundary_diagnostics.html)
computes the first three per sample; they are stored in
`fit$diagnostics`.

```{r}
#| label: diagnostics
fit <- fit_decovart(
  signature_matrix = signature,
  bulk_expression = replicates[, 1:3],
  Sigma = Sigma,
  itmax = 100
)
do.call(rbind, fit$diagnostics)
```

Every sample above converged and every curvature is negative, yet
`local_maximum` is not uniformly `TRUE`: one iterate stopped with an ALR
score norm just above the default `score_tol = 1e-4`. That is the
distinction of @tbl-diagnostics doing its job rather than a failure —
`score_tol` bounds the *score*, whereas the solver's `epsilon` bounds the
*objective*, so the two should be tightened together.

::: {#nte-boundary-wording .callout-note}
## Near-boundary is not failure

Suppose the truth is $\boldsymbol{p}_0=(0,0.3,0.7)$. A fit returning
$\hat{\boldsymbol{p}}=(10^{-10},0.3001,0.6999)$ is numerically
*excellent*: the optimiser is correctly approaching a genuine boundary
optimum, and an interior ALR solver cannot settle at a finite
$\boldsymbol{\rho}$ there. So $\min_j\hat{p}_j\ll 1$ should not be read
as "optimisation failed" but as

> the estimate is close to a simplex face, so ALR coordinates and
> interior Wald inference may be unreliable.

`boundary_tol` is therefore a **statistical** warning threshold for Wald
linearisation, deliberately far larger than the machine-precision guard
that decides whether a logarithm is representable. The two serve
different purposes and should not share a value.
:::

Multimodality is the one property that restarts address directly.
[`multistart_decovart()`](../reference/multistart_decovart.html) refits
from independent Dirichlet$(1,\ldots,1)$ starts and reports the spread of
attained optima; `fit_decovart(n_starts = )` wires it into the S3
workflow and keeps the best mode.

```{r}
#| label: multistart
set.seed(3)
restarts <- multistart_decovart(
  y = replicates[, 1],
  mean_signature_matrix = signature,
  Sigma = Sigma,
  n_starts = 6L,
  itmax = 100
)
c(
  loglik_range = restarts$loglik_range,
  multimodal = restarts$multimodal
)
```

A small range is reassuring but not a proof: a converged code, a tiny
relative-distance-to-minimum and a negative-definite local Hessian
together support a **local** maximum, and none of them rules out a
second, equally good or better mode elsewhere on the simplex.

# Recap: which interval or test to report {#sec-recap}

The procedures below all target the same composition
$\boldsymbol{p}\in\Delta^{J-1}$, but they rest on different
approximations. @tbl-inference-recap is the practical counterpart of
the Wald warning in the derivatives vignette: use the cheapest method
that the geometry of the estimate actually supports.

| Procedure | What it returns | Calibration | Simplex faces | Replication | Function |
|:--|:--|:--|:--|:--|:--|
| Wald (ALR delta method) | per-sample CI | $\mathrm{N}(\boldsymbol{0},I(\hat{\boldsymbol{\rho}})^{-1})$ mapped by $\mathbf{J}_{\boldsymbol{\psi}}$ | unusable (`NA`) | one bulk column is a plug-in approximation (@nte-replication) | `confint()` / `vcov_alr_delta()` |
| Profile likelihood | CI in $[0,1]$ | Wilks $\chi^2_1$ on the deviance (@wilksLargeSampleDistributionLikelihood1938) | stays inside the simplex; endpoints on a face need the next row | pooled replicates, or treat as approximate | `confint_profile_decovart()` |
| Interior LRT | test of $p_j=c\in(0,1)$ | $\chi^2_1$ | not for $c=0$ or $c=1$ | same as profile | `lrt_decovart()` |
| Boundary LRT | test of $p_j=0$ | chi-bar-square $\tfrac12\chi^2_0+\tfrac12\chi^2_1$ [@hermanDistributionLikelihoodRatio1954; @selfAsymptoticPropertiesMaximum1987] | designed for this | same as profile | `lrt_decovart()` (automatic) |
| Parametric bootstrap | percentile CI and Monte Carlo $p$-value | $\boldsymbol{Y}^{(b)}\sim\mathcal{N}_G(\boldsymbol{\mu}\boldsymbol{p},\boldsymbol{\Sigma}(\boldsymbol{p}))$ at the MLE or restricted MLE | automatic (the $\tfrac12\chi^2_0$ atom is visible) | known plug-in $(\boldsymbol{\mu},\boldsymbol{\Sigma}_j,\boldsymbol{p})$ | `bootstrap_decovart()` |
| Reference-sample bootstrap | percentile CI | resample **donors** (default) or **cells** within each labelled type, rebuild $(\boldsymbol{\mu},\boldsymbol{\Sigma}_j)$, refit [@efronBootstrapMethodsAnother1979] | automatic | uncertainty in the reference, not in $\boldsymbol{Y}$ unless `regenerate_bulk` | `reference_bootstrap_decovart()` |
| Dirichlet composition sweep | percentile of $\hat{\boldsymbol{p}}$ over random mixtures | $\boldsymbol{p}^{(b)}\sim\mathrm{Dirichlet}(\boldsymbol{\alpha})$, then simulate $\boldsymbol{Y}$ | automatic | performance over compositions, not a CI for one bulk | `reference_bootstrap_decovart(method = "dirichlet")` |

: Inferential procedures implemented for DeCovarT proportions. Wald is the default of `confint.decovart_fit`; everything else is an explicit alternative. {#tbl-inference-recap}

A short decision rule, in the order the table suggests abandoning the
previous row:

1. If $\hat{\boldsymbol{p}}$ is interior and several bulk columns share
   one composition, a profile interval or an interior LRT is the
   reparametrisation-invariant upgrade of Wald.
2. If the scientific null is absence ($p_j=0$), do not report Wald;
   use the chi-bar-square LRT or a restricted parametric bootstrap.
3. If the dominant uncertainty is the purified reference, use
   `reference_bootstrap_decovart()` with `method = "donors"` when donor
   labels exist, otherwise `method = "cells"`. Do not permute genes or
   cell-type names: @thm-label-equivariance makes a label shuffle
   vacuous, and a gene shuffle is not a sampling model
   (@tbl-reference-ops). Use `equivariance_check_decovart()` only as a
   software check.
4. If the question is performance across mixture weights rather than a
   CI for one bulk, use
   `reference_bootstrap_decovart(method = "dirichlet")`.

# Perspectives {#sec-perspectives}

::: {#nte-perspectives .callout-tip}
## Beyond first-order asymptotics

Directions that would strengthen DeCovarT uncertainty quantification,
in rough order of implementation cost.

- **Godambe sandwich for a misspecified convolution.** `RNA-Sieve`
  replaces Fisher by the Godambe information when the CLT model is
  wrong [@erdmann-phamLikelihoodbasedDeconvolutionBulk2021]
  (@sec-rna-sieve). The same sandwich on DeCovarT's score would
  widen Wald intervals under protocol shift without discarding the
  multivariate covariance. It remains a first-order interior device:
  faces still need the chi-bar-square LRT or a restricted bootstrap.
- **Score-test inversion.** Inverting a one-sided score test avoids
  refitting under every candidate value, so it is cheaper than the
  profile scan of `confint_profile_decovart()` while keeping the
  one-sided boundary geometry.
- **Higher-order likelihood corrections.** Bartlett-type adjustments of
  the likelihood-ratio statistic, or modified profile likelihoods,
  improve the $\chi^{2}$ approximation in the small-replication regime
  that @nte-replication describes — precisely DeCovarT's regime.
- **Bayesian compositional models.** A logistic-normal or Dirichlet prior
  on $\boldsymbol{p}$ regularises the boundary: the posterior stays
  proper where the ALR Wald interval degenerates, and credible intervals
  are usually more stable near a face. A credible interval is of course
  not an exact frequentist interval, and the prior must be defensible
  scientifically. This sits naturally alongside the MAP treatment of the
  latent profiles sketched in the
  [perspectives vignette](decovart-perspectives.html).
- **Information geometry of the composition.** Treating $\Delta^{J-1}$
  with the Fisher–Rao metric of @eq-fisher-metric rather than as a subset
  of $\mathbb{R}^{J}$ would give natural-gradient updates and
  reparametrisation-invariant confidence regions
  [@malagoInformationGeometryGaussian2015; @aitchisonStatisticalAnalysisCompositional1982].
:::

# References
