---
title: "DeCovarT appendix: affine equivariance, collinearity and perturbations"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{DeCovarT appendix: affine equivariance, collinearity and perturbations}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup
#| include: false
options(tinytable_html_mathjax = TRUE)
```

This appendix records four numerical facts that sit behind the
`decovart_fit` API: gene-wise affine equivariance of the MLE, failure
under collinear signatures, small bulk perturbations when the covariance
is full, cell-type diagonal, or global, and the effect of random ALR
starts on second-order solvers. CPM / log2 normalisation, optimiser
tolerance, and scaling with $G$, $J$ and overlap are left to a later
benchmark issue.

# Gene-wise affine equivariance {#sec-affine}

Let $D=\mathrm{diag}(s_1,\ldots,s_G)$ with $s_g\neq 0$ and
$\boldsymbol{m}\in\mathbb{R}^{G}$. The same gene-wise map applied to bulk,
means and covariances,
$$
\boldsymbol{y}^\star_{\cdot i}
= D^{-1}(\boldsymbol{y}_{\cdot i}-\boldsymbol{m}),
\quad
\boldsymbol{\mu}^\star
= D^{-1}(\boldsymbol{\mu}-\boldsymbol{m}\boldsymbol{1}_J^\top),
\quad
\boldsymbol{\Sigma}_j^\star
= D^{-1}\boldsymbol{\Sigma}_j D^{-1},
$$
preserves the linear convolution because
$\boldsymbol{1}_J^\top\boldsymbol{p}_{\cdot i}=1$:
$$
\boldsymbol{y}^\star_{\cdot i}
=\boldsymbol{\mu}^\star\boldsymbol{p}_{\cdot i}
+\boldsymbol{\varepsilon}^\star_{\cdot i}.
$$
The Mahalanobis residual is invariant, so
$\hat{\boldsymbol{p}}(\boldsymbol{Y},\boldsymbol{\mu},\boldsymbol{\Sigma})
=\hat{\boldsymbol{p}}(\boldsymbol{Y}^\star,\boldsymbol{\mu}^\star,
\boldsymbol{\Sigma}^\star)$ in exact arithmetic. A z-score computed
**once per gene from $\boldsymbol{\mu}$** (not per cell type, not per bulk
sample) is that map. Sample-level covariates that shift gene-wise means
are included provided the same $D$ and $\boldsymbol{m}$ are applied to the
adjusted $\boldsymbol{\mu}_j(\boldsymbol{z}_i)$.

A logarithm is not affine. It is concave, so Jensen's inequality
changes first and second moments, and
$\log(\boldsymbol{\mu}\boldsymbol{p})\neq(\log\boldsymbol{\mu})\boldsymbol{p}$.
`CIBERSORT` therefore requires non-negative expression, no missing
values, and a non-log linear scale
[@newmanRobustEnumerationCell2015]. The logarithm is monotone on
$(0,\infty)$, so gene-wise orderings are preserved, but the
Gaussian-convolution MLE of $\boldsymbol{p}$ is not.

```{r}
#| label: affine-invariance
#| warning: false
library(DeCovarT)

set.seed(1)
genes <- paste0("g", 1:3)
cts <- paste0("ct", 1:2)
mu <- matrix(
  c(20, 40, 25, 35, 30, 22),
  nrow = 3,
  dimnames = list(genes, cts)
)
Sigma <- array(0, dim = c(3, 3, 2), dimnames = list(genes, genes, cts))
Sigma[, , 1] <- diag(c(1.0, 1.2, 0.8))
Sigma[, , 2] <- diag(c(1.5, 0.9, 1.1))
p_true <- c(0.65, 0.35)
y <- matrix(
  drop(mu %*% p_true + rnorm(3, sd = 0.05)),
  ncol = 1,
  dimnames = list(genes, "s1")
)

fit_raw <- fit_decovart(
  mu,
  y,
  Sigma = Sigma,
  method = "Marquardt-Levenberg",
  itmax = 80
)

centre <- c(2, -1, 0.5)
scale <- c(2, 0.5, 4)
D_inv <- 1 / scale
y_star <- (y - centre) / scale
mu_star <- (mu - centre) / scale
Sigma_star <- Sigma
for (j in seq_len(2)) {
  Sigma_star[, , j] <- Sigma[, , j] * tcrossprod(D_inv)
}
# Solvers accept the starred arrays directly (z-scores need not be
# non-negative). The public gate still requires non-negative raw counts.
p_star <- deconvolute_ratios_Marquardt_Levenberg(
  drop(y_star),
  mu_star,
  Sigma_star,
  itmax = 80
)
fit_z <- fit_decovart(
  mu,
  y,
  Sigma = Sigma,
  method = "Marquardt-Levenberg",
  itmax = 80,
  standardise = TRUE
)

rbind(
  raw = drop(coef(fit_raw)),
  general_affine = p_star,
  gene_wise_z = drop(coef(fit_z))
)
```

The two rows should agree up to optimiser noise: `standardise = TRUE`
is a gene-wise affine map. Cell-type-wise `scale()` of $\boldsymbol{\mu}$
would estimate a different $D$ per column and is not supported.

# Collinearity {#sec-collinearity}

If two signature columns are identical, or a parent profile is an
affine combination of children, $\boldsymbol{\mu}$ is rank-deficient.
That is **not** by itself a proof that $\boldsymbol{p}$ is unidentifiable
(RE2.4b): the convolution law also depends on
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_{j}p_{j}^{2}\boldsymbol{\Sigma}_{j}$.
Identical means with *distinct* covariances remain identifiable; the
negative control is identical means *and* proportional covariances. The
chunk below still warns, because `fit_decovart()` flags collinear mean
columns as a numerical hazard even when the covariance tensor separates
the types.

```{r}
#| label: collinear-two
#| warning: true
mu2 <- mu
mu2[, 2] <- mu2[, 1]
tryCatch(
  fit_decovart(
    mu2,
    y,
    Sigma = Sigma,
    method = "Marquardt-Levenberg",
    itmax = 20
  ),
  warning = function(w) conditionMessage(w),
  error = function(e) conditionMessage(e)
)
```

```{r}
#| label: collinear-parent
#| warning: true
genes3 <- paste0("g", 1:3)
cts3 <- c("child_1", "child_2", "parent")
mu3 <- cbind(
  child_1 = c(20, 40, 15),
  child_2 = c(40, 20, 25),
  parent = 0.5 * c(20, 40, 15) + 0.5 * c(40, 20, 25)
)
rownames(mu3) <- genes3
Sigma3 <- array(0, dim = c(3, 3, 3), dimnames = list(genes3, genes3, cts3))
for (j in 1:3) {
  Sigma3[, , j] <- diag(3)
}
y3 <- matrix(mu3 %*% c(0.2, 0.3, 0.5), ncol = 1)
dimnames(y3) <- list(genes3, "s1")
qr(mu3)$rank
tryCatch(
  fit_decovart(
    mu3,
    y3,
    Sigma = Sigma3,
    method = "Marquardt-Levenberg",
    itmax = 20
  ),
  warning = function(w) conditionMessage(w),
  error = function(e) conditionMessage(e)
)
```

Missing values are not a separate predictor / response policy (RE2.2):
quantified bulk RNA-seq matrices are complete, and the API errors on
`NA` / `NaN` / `Inf`.

# Small perturbations of the bulk {#sec-perturb}

With known $\boldsymbol{p}$, a small Gaussian perturbation of
$\boldsymbol{y}_{\cdot i}$ should move Marquardt--Levenberg estimates only
slightly. Three covariance specifications are compared:

1. Full cell-type tensor $\boldsymbol{\Sigma}_j$.
2. Cell-type diagonal: off-diagonals of each $\boldsymbol{\Sigma}_j$ set to
   zero (gene-wise weighted convolution).
3. Global competitor: every slice replaced by the convolution covariance
   $\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^{2}\boldsymbol{\Sigma}_j$
   evaluated at the known mixing weights.

```{r}
#| label: perturbation
#| warning: false
set.seed(2)
p_true <- c(0.6, 0.4)
y0 <- matrix(drop(mu %*% p_true), ncol = 1, dimnames = list(genes, "s1"))
y_pert <- y0 + rnorm(3, sd = 0.2)

Sigma_diag <- Sigma
for (j in 1:2) {
  Sigma_diag[, , j] <- diag(diag(Sigma[, , j]))
}
sigma_bar <- .compute_global_variance(p_true, Sigma)
Sigma_global <- Sigma
Sigma_global[, , 1] <- sigma_bar
Sigma_global[, , 2] <- sigma_bar

fit_spec <- function(y, S) {
  drop(coef(fit_decovart(
    mu,
    y,
    Sigma = S,
    method = "Marquardt-Levenberg",
    itmax = 80
  )))
}

specs <- list(
  full = Sigma,
  celltype_diagonal = Sigma_diag,
  global_weighted = Sigma_global
)
est <- vapply(specs, function(S) fit_spec(y_pert, S), numeric(2))
rmse <- apply(est, 2, function(p_hat) sqrt(mean((p_hat - p_true)^2)))
list(estimates = est, rmse_to_truth = rmse)
```

## Random ILR starts {#sec-init-seeds}

Marquardt--Levenberg and Newton--Raphson maximise the convolution
log-likelihood with analytic Hessians. That is cheaper per step than a
first-order quasi-Newton map, but a start near a simplex face can stall
on a plateau. Three Dirichlet$(1,1)$ draws of $\boldsymbol{p}^{(0)}$
(seeds 11, 22 and 33) are passed as `initial_p` on the same perturbed
bulk. On this well-conditioned toy the second-order estimates should
agree; a spread of several percentage points is a warning that the
start has trapped the solver.

```{r}
#| label: initial-seeds
#| warning: false
dirichlet_start <- function(seed) {
  set.seed(seed)
  g <- stats::rgamma(2, shape = 1, rate = 1)
  g / sum(g)
}
seeds <- c(11L, 22L, 33L)
starts <- lapply(seeds, dirichlet_start)
names(starts) <- paste0("seed_", seeds)

fit_from_start <- function(fun, p0) {
  fun(drop(y_pert), mu, Sigma, itmax = 80, initial_p = p0)
}

ml <- vapply(
  starts,
  function(p0) fit_from_start(deconvolute_ratios_Marquardt_Levenberg, p0),
  numeric(2)
)
nr <- vapply(
  starts,
  function(p0) fit_from_start(deconvolute_ratios_Newton_Raphson, p0),
  numeric(2)
)
list(
  starts = starts,
  marquardt = ml,
  newton_raphson = nr
)
```

A fuller comparison of normalisation (CPM, log2), tolerance, and
dimension $(G,J)$ belongs in a later issue.
