---
title: "Feature selection for reference-based deconvolution"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Feature selection for reference-based deconvolution}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  eq-title: "Equation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)
```

# Scope {#sec-scope}

This vignette surveys **feature selection for reference-based
(signature-based) deconvolution**—methods that recover cellular
proportions $\boldsymbol{p}$ from a bulk mixture using a purified
expression design matrix
[@gaspard-boulincCelltypeDeconvolutionMethods2025;
@avilacobosComputationalDeconvolutionTranscriptomics2018;
@sturmComprehensiveEvaluationTranscriptomebased2019]. Feature selection
itself is a classical pre-modelling filter
[@guyonIntroductionVariableFeature2003]. We deliberately exclude
**marker-based** abundance scoring (enrichment / GSEA-style modules),
which returns proxies of presence rather than a consensual gene panel
suited to linear or probabilistic mixture models
[@aranXCellDigitallyPortraying2017; @zhang_etal17].

The focus is **second-generation** signatures built from single-cell
pseudobulks (CIBERSORTx, MuSiC, DWLS, Kassandra and related pipelines)
[@newmanDeterminingCellType2019; @wangBulkTissueCell2019;
@tsoucasAccurateEstimationCelltype2019;
@zaitsevPreciseReconstructionTME2022], where the selected genes must
discriminate **all** cell types in the signature jointly, not only
one-versus-rest markers. Correlation-aware compact panels such as
DUBStepR illustrate the same goal from a single-cell feature-selection
angle [@ranjanDUBStepRScalableCorrelationbased2021].

Interactive signature refinement utilities (zero filtering,
specificity binning, Gini / entropy compaction) are implemented in
[DeconvExplorer](https://github.com/omnideconv/deconvExplorer)
(omnideconv ecosystem; cite as software once added to
`packages.bib`).

```{mermaid}
%%| label: fig-feature-selection-layout
%%| fig-cap: "Layout of this vignette: univariate and multivariate selection, each split into one-versus-others then all-versus-all, followed by perspectives for continuous cell states."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  A["Reference μ, Σ from sc / bulk"] --> B["Univariate"]
  A --> C["Multivariate"]
  B --> B1["One vs others"]
  B --> B2["All vs all"]
  C --> C1["One vs others\n(GGM / DiffNet)"]
  C --> C2["All vs all\n(κ, D/E-opt, overlap)"]
  B2 --> D["Consensual panel G"]
  C2 --> D
  D --> E["Perspectives:\nmechanistic / potency / archetypes"]
```

Sections are cross-linked below: @sec-preprocess builds moments;
@sec-univariate and @sec-multivariate score genes; @sec-wrapper and
@sec-recommended assemble a pipeline; @sec-perspectives covers
continua, Markov blankets, and mechanistic priors
(@sec-mechanistic).

# Notation {#sec-notation}

We align notation with the DeCovarT generative model. Let $G$ genes and
$J$ cell types index the purified means
$\boldsymbol{\mu}=(\boldsymbol{\mu}_{.j})_{j=1}^{J}\in\mathbb{R}^{G\times J}$
and (optional) covariances $\{\boldsymbol{\Sigma}_{j}\}_{j=1}^{J}$. A bulk
observation obeys

$$
\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})
\sim
\mathcal{N}_{G}\!\left(
\boldsymbol{\mu}\boldsymbol{p},\,
\sum_{j=1}^{J}p_{j}^{2}\boldsymbol{\Sigma}_{j}
\right),
\quad
\boldsymbol{p}\in\Delta^{J-1},
$$ {#eq-mixture}

with $\boldsymbol{\zeta}=(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$
and simplex
$\Delta^{J-1}=\{\boldsymbol{p}:p_{j}\ge 0,\sum_{j}p_{j}=1\}$. For a
gene subset $\mathcal{G}\subseteq\{1,\ldots,G\}$, write
$\boldsymbol{\mu}_{\mathcal{G}}$ for the corresponding rows. The
condition number of a (possibly rectangular) design matrix
$\boldsymbol{X}$ is defined via singular values

$$
\kappa_2(\boldsymbol{X})
=
\frac{\sigma_{\max}(\boldsymbol{X})}{\sigma_{\min}(\boldsymbol{X})},
$$ {#eq-condition-number}

with large $\kappa_2$ indicating multicollinearity and numerical
instability [@gongDeconRNASeqStatisticalFramework2013;
@newmanRobustEnumerationCell2015]. For a Gram matrix one has
$\kappa_2(\boldsymbol{X}^{\top}\boldsymbol{X})=\kappa_2(\boldsymbol{X})^{2}$,
so do not conflate the condition number of a signature with that of
its Gram matrix. Eigenvalue ratios apply directly only to an
appropriate symmetric positive-definite matrix (see also the
[synthetic scenarios](synthetic-scenarios.html#sec-geometric-score)
vignette).

# Preprocessing shared by all strategies {#sec-preprocess}

Before scoring genes:

- ① Aggregate **donor-level** cell-type pseudobulks (not individual
  cells as independent replicates). Pseudobulking reduces sparsity and
  avoids treating cells as independent biological replicates in
  differential testing [@squairConfrontingFalseDiscoveries2021].
- ② Estimate gene–cell-type means $\mu_{gj}$ and variance components on
  a **linear** scale (counts, CPM, or TPM). Log-space breaks the
  mixture linearity in @eq-mixture
  [@sturmComprehensiveEvaluationTranscriptomebased2019;
  @jinBenchmarkRNAseqDeconvolution2021;
  @hoffmannRobustComputationalReconstitution2006]. Variance
  stabilisation for *single-cell* QC is handled separately below
  (see @sec-sctransform).
- ③ Filter genes with poor bulk detectability, extreme zero inflation
  after pseudobulking, strong donor instability, or systematic
  sc–bulk discordance.
- ④ When the biological target is **cell number** rather than RNA mass,
  correct transcriptome-size differences across cell types
  [@luTranscriptomeSizeMatters2025].
- ⑤ Apply the **same** normalisation to bulk and signature
  [@jinBenchmarkRNAseqDeconvolution2021;
  @racleSimultaneousEnumerationCancer2017].

## Variance stabilisation and highly variable genes {#sec-sctransform}

Single-cell count matrices mix **technical** depth noise with
**biological** cell-type variability. Regularised negative-binomial
regression as in `Seurat::SCTransform` / `sctransform` separates the
two via **Pearson residuals**
[@hafemeisterNormalizationVarianceStabilization2019]:

1. Infer a depth-aware mean $\hat{\mu}_{ig}$ for cell $i$ and gene $g$
   (library size enters as an offset; optional covariates can be
   included in the design).
2. Infer a gene-wise dispersion $\hat{\theta}_{g}$ with empirical-Bayes
   shrinkage and a local mean–dispersion trend across genes.
3. Form Pearson residuals
   $$
   r_{ig}
   =
   \frac{y_{ig}-\hat{\mu}_{ig}}
   {\sqrt{\hat{\mu}_{ig}+\hat{\mu}_{ig}^{2}/\hat{\theta}_{g}}},
   $$ {#eq-pearson-residual}
   which are approximately homoscedastic and closer to Gaussian than
   raw counts—useful for PCA / HVG ranking, **not** as the linear
   mixture scale in @eq-mixture.

@fig-sc-transform summarises the negative-binomial observation model,
depth regression, residualisation, and empirical-Bayes shrinkage.

![SCTransform workflow: negative-binomial mean–variance model, depth
regression, Pearson residuals, and empirical-Bayes regularisation
across genes.](figures/sc_transform.png){#fig-sc-transform}

After residualisation, retain a compact **highly variable gene (HVG)**
universe before signature scoring. SCTransform / Pearson residuals
correct for sequencing depth while **preserving** biological
heterogeneity
[@hafemeisterNormalizationVarianceStabilization2019]. Do **not**
regress out cell type at this HVG stage: that would remove exactly the
inter-type signal the panel is meant to capture. A practical design
that emphasises cell-type structure (rather than a global intercept)
is

$$
\sim\ \texttt{CellType}
$$

**without intercept**, so the estimated coefficients contrast each
labelled type against the **global mean expression across cell types
and samples**. Rank genes by residual variance (or the corresponding
model $F$ / deviance explained) and keep the top $2{,}000$ HVGs as the
working universe for the selectors in @sec-univariate and
@sec-multivariate. Downstream deconvolution still uses
**linear-scale** $\mu_{gj}$ on that gene set
[@sturmComprehensiveEvaluationTranscriptomebased2019].

::: {#nte-sctransform-check .callout-tip}
## :information_source: Sanity check after SCTransform

Per gene, residual histograms and mean–variance plots of $r_{ig}$
should look roughly Gaussian and flat. Strong remaining depth trends
or heavy tails indicate that the offset / covariate design is
misspecified before any marker ranking.
:::

# Univariate approaches {#sec-univariate}

Univariate selectors score each gene marginally (or with a simple
contrast) and ignore partial correlations among genes.

## One versus others {#sec-univ-one-vs-others}

Classical signature pipelines score gene $g$ for cell type $j$ against
the remaining types, then concatenate per-type shortlists
[@abbasDeconvolutionBloodMicroarray2009;
@newmanRobustEnumerationCell2015; @finotello_etal19].

### Differential expression and fold-change ranks

Limma–voom, edgeR and DESeq2 remain the workhorses for
one-versus-rest contrasts
[@ritchieLimmaPowersDifferential2015a;
@robinsonEdgeRBioconductorPackage2010;
@loveModeratedEstimationFold2014a]. CIBERSORT retains genes with
adjusted $p\le 0.3$ ordered by decreasing fold-change before a global
condition-number prune [@newmanRobustEnumerationCell2015]. Abbas et
al. keep genes that exceed the second- and third-ranked cell types,
then choose the panel with smallest $\kappa$
[@abbasDeconvolutionBloodMicroarray2009]. quanTIseq / xCell-style
filters further drop non-haematopoietic contaminants
[@finotello_etal19; @aranXCellDigitallyPortraying2017].

### ANOVA $F$-test {#sec-anova-f}

Wang et al. rank genes by the nested-model $F$ statistic comparing a
cell-type factor to an intercept-only model
[@wangSilicoEstimatesTissue2010]. For gene $g$ with expression
$x_{gi}$ in sample $i$ and cell-type labels $z_{i}\in\{1,\ldots,J\}$,

$$
F_{g}
=
\frac{
\bigl(\mathrm{SSR}_{0}-\mathrm{SSR}_{1}\bigr)/(J-1)
}{
\mathrm{SSR}_{1}/(n-J)
},
$$ {#eq-anova-f}

where $\mathrm{SSR}_{0}=\sum_{i}(x_{gi}-\bar{x}_{g})^{2}$ and
$\mathrm{SSR}_{1}=\sum_{j}\sum_{i:z_{i}=j}(x_{gi}-\bar{x}_{gj})^{2}$.
Large $F_{g}$ indicates that cell-type means explain a substantial
share of gene variance.

Classical ANOVA inference for @eq-anova-f assumes, among other things:

- independent observations (hence donor-level pseudobulks rather than
  cells as replicates; @sec-preprocess);
- roughly **Gaussian conditional residuals**
  $x_{gi}\mid z_{i}=j\sim\mathcal{N}(\mu_{gj},\sigma_{g}^{2})$ with
  common within-type variance (homoscedasticity);
- no strong outliers or unmodelled batch structure.

When residuals are heavy-tailed or variance scales with the mean,
prefer a GLM / limma–voom pipeline
[@ritchieLimmaPowersDifferential2015a] and treat $F_{g}$ as a ranking
score rather than a calibrated $p$-value.

### Gini and entropy specificity {#sec-gini-entropy}

BioQC and DeconvExplorer compact signatures with Gini or entropy
scores over the row $\boldsymbol{\mu}_{g\cdot}$ [@zhang_etal17]. Write
non-negative mean expression as relative **shares**

$$
q_{gj}
=
\frac{\mu_{gj}}{\sum_{\ell=1}^{J}\mu_{g\ell}},
\qquad
\sum_{j}q_{gj}=1,
$$ {#eq-shares}

so $\boldsymbol{q}_{g}$ is a discrete distribution over cell types.
The **Gini coefficient** of inequality among ordered shares
$q_{g(1)}\le\cdots\le q_{g(J)}$ and the **Shannon entropy** are

$$
\mathrm{Gini}(g)
=
\frac{J}{J-1}
\left(
1-2\sum_{j=1}^{J}
\frac{J-j+\tfrac{1}{2}}{J}\,q_{g(j)}
\right),
\quad
H(g)
=
-\sum_{j=1}^{J}q_{gj}\log q_{gj}.
$$ {#eq-gini-entropy}

High Gini (or low entropy) flags genes whose mass concentrates on few
types—useful for one-versus-rest compaction. @fig-gini-entropy
contrasts the two scores.

![Gini versus Shannon entropy for cell-type specificity: formulas,
simplified share plots, and panel-wise pros/cons for one-versus-rest
compaction.](figures/gini_vs_entropy_specificity.png){#fig-gini-entropy}

“Redundancy when panels are concatenated” means the following. High
Gini markers are specific, but not automatically non-redundant: a gene
with high Gini for type $A$ and another with high Gini for type $B$
need not be collinear if they separate *different* contrasts, yet they
**can** be nearly collinear as rows of $\boldsymbol{\mu}$ when both
types co-express correlated programmes (shared lineage, activation
axis). Redundancy arises mainly when multiple markers reflect the same
contrast. Concatenating the top-$k$ genes per type therefore often
inflates $|\mathcal{G}|$ without improving the geometry of
$\boldsymbol{\mu}_{\mathcal{G}}$: $\kappa_2(\boldsymbol{\mu}_{\mathcal{G}})$
stays large and NNLS / SVR / GLS remain unstable
[@newmanRobustEnumerationCell2015]. Likewise, edges in a
partial-correlation graph flag candidate influences, not assured
independence. That is why @sec-univ-one-vs-others must be followed by
an all-versus-all stage (see @sec-univ-all-vs-all,
@sec-multi-all-vs-all).

::: {#nte-ovr-concat .callout-warning}
## :warning: Limitation of one-versus-rest concatenation

Union of per-type markers often yields a multicollinear
$\boldsymbol{\mu}_{\mathcal{G}}$ that is still too large for stable
NNLS / SVR / GLS
[@newmanRobustEnumerationCell2015;
@avilacobosComputationalDeconvolutionTranscriptomics2018]. A global
all-versus-all refinement is therefore required
(see @sec-multi-all-vs-all).
:::

## All versus all {#sec-univ-all-vs-all}

### Genetic algorithms (AutoGeneS)

AutoGeneS searches gene subsets with a multi-objective genetic
algorithm that **minimises inter-population correlation** while
**maximising centroid distance**
[@alieeAutoGeneSAutomaticGene2021]. Related multi-objective genetic
searches appear in network module discovery (e.g. MOGAMUN)
[@novoa-del-toroMultiobjectiveGeneticAlgorithm2021]. AutoGeneS is
explicitly all-versus-all: the loss is defined on the full $J$-column
signature rather than on independent one-versus-rest lists.

DeCovarT replaces that dual objective with a single **global overlap**
criterion on the Gaussian mixture
$\{(p_{j},\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j})\}$ (see
@eq-overlap and @sec-overlap).

# Multivariate approaches {#sec-multivariate}

Multivariate selectors use joint dependence—covariance, precision, or
information matrices—so that a gene is scored by how it improves
identifiability of $\boldsymbol{p}$ under @eq-mixture.

## Limitations of mean-only condition numbers {#sec-kappa-limits}

Condition-number pruning (see @eq-condition-number) operates almost
exclusively on the **mean** signature $\boldsymbol{\mu}_{\mathcal{G}}$.
It does not see covariance rewiring: two closely related types can
share nearly collinear means yet differ in their precision graphs
$\boldsymbol{\Omega}_{j}=\boldsymbol{\Sigma}_{j}^{-1}$. In
**continua**—gastruloids, differentiation trajectories, activation
gradients—neighbouring states are transcriptionally similar by
construction, so $\kappa(\boldsymbol{\mu})$ is intrinsically large.
Aggressive $\kappa$-minimisation then tends to keep a few dominant
lineage representatives and drop genes that separate neighbours only
through **network rewiring** rather than large mean fold-changes
[@tsoucasAccurateEstimationCelltype2019;
@wangBulkTissueCell2019].

@fig-kappa-gastruloid illustrates this failure mode for a gastruloid
lineage continuum: mean-conditioned DWLS-style selection can
under-represent adjacent or low-abundance states relative to methods
that retain softer, covariance-aware allocation.

![Condition-number pruning on correlated developmental cell types
(gastruloid continuum). Mean-only $\kappa$ filtering can concentrate
signal in dominant collinear lineages and shrink closely related
states.](figures/condition_number_poor_performance_correlated_cell_types.png){#fig-kappa-gastruloid}

::: {#nte-kappa-means .callout-important}
## :warning: When $\kappa(\boldsymbol{\mu})$ is not enough

Use $\kappa$ as a **diagnostic** of mean geometry, not as the sole
selector, whenever cell states form a continuum or when
discriminatory signal lives in $\boldsymbol{\Sigma}_{j}$ (see
@sec-indeed and @sec-optimal-design). Pair it with overlap
(see @eq-overlap), Fisher information (see @eq-fisher), or differential-network
candidates.
:::

## One versus others {#sec-multi-one-vs-others}

Univariate DGE ignores gene–gene interactions and therefore requires
aggressive multiple-testing correction without modelling network
rewiring. Differential-network and GGM methods address that gap.

### Differential networks (INDEED and related) {#sec-indeed}

INDEED combines mean (fold-change) evidence with tests of changes in
**partial correlation**, ranking genes that both shift in mean **and**
rewire neighbourhood structure
[@zuoINDEEDIntegratedDifferential2016]. @fig-indeed summarises the
graphical abstract and computational pipeline.

::: {#fig-indeed layout-ncol=2}

![INDEED graphical abstract: parallel differential expression and
differential-network tracks merged into an activity-ranked biomarker
list (adapted from @zuoINDEEDIntegratedDifferential2016).](figures/INDEED_graphical_abstract.png){#fig-indeed-abstract}

![INDEED pipeline: group-wise graphical lasso, differential partial
correlations $\Delta\mathrm{pc}$ with permutation filtering, DE
$p$-values, and activity-score prioritisation (adapted from
@zuoINDEEDIntegratedDifferential2016).](figures/INDEED_pipeline.png){#fig-indeed-pipeline}

INDEED integrates differential expression with differential network
structure to prioritise biomolecules.
:::

For groups $k\in\{1,2\}$ (e.g. focal cell type versus the rest), let
$\boldsymbol{\Omega}^{(k)}$ be a sparse precision estimate from
graphical lasso. Partial correlations are

$$
\mathrm{pc}_{gg'}^{(k)}
=
-\frac{\omega_{gg'}^{(k)}}
{\sqrt{\omega_{gg}^{(k)}\,\omega_{g'g'}^{(k)}}},
\qquad
\Delta\mathrm{pc}_{gg'}
=
\mathrm{pc}_{gg'}^{(1)}-\mathrm{pc}_{gg'}^{(2)}.
$$ {#eq-indeed-dpc}

Significant edges are retained after a permutation (or bootstrap) null
on $\Delta\mathrm{pc}_{gg'}$. With a differential-expression $p$-value
$p_{g}^{\mathrm{DE}}$ for gene $g$, INDEED forms an **activity score**
that multiplies mean evidence by differential connectivity, schematically

$$
\mathrm{AS}_{g}
=
\bigl(-\log p_{g}^{\mathrm{DE}}\bigr)
\sum_{g'\neq g}
\bigl\lvert\Delta\mathrm{pc}_{gg'}\bigr\rvert
\,
\mathbf{1}\!\left\{\Delta\mathrm{pc}_{gg'}\ \text{significant}\right\},
$$ {#eq-indeed-as}

and ranks genes by $\mathrm{AS}_{g}$
[@zuoINDEEDIntegratedDifferential2016]. Relative to pure DGE, this
favours genes whose local **undirected** neighbourhood (Markov boundary
in the GGM sense; @sec-markov-seeds) differs between the focal type and
the rest. Related ideas appear in weighted graphical-lasso pipelines
with PPI priors and differential-network scores (DNS).

### Sparse GGMs and cooperative penalties {#sec-sparse-ggm}

Cell-type precision matrices
$\boldsymbol{\Omega}_{j}=\boldsymbol{\Sigma}_{j}^{-1}$ estimated by
graphical lasso are shrunk; non-zero partial correlations are typically
underestimated. A practical mitigation is to take the gLasso
**support** as topological constraints and re-estimate by constrained
MLE when the graph is chordal (otherwise the mapping is non-trivial)
[@dahlMaximumLikelihoodEstimation2005]. Cooperative / sign-coherent
group lasso penalties can encode pathway modules and PPI-informed edge
weights [@chiquetSparsitySigncoherentGroups2012]. PPI priors also
shrink the combinatorial space of undirected graphs explored per cell
type.

::: {#nte-network-candidates .callout-warning}
## :warning: From one-versus-others networks to a panel

Network-based one-versus-others lists still concatenate into a
redundant design (see @nte-ovr-concat). Treat them as **candidate
generators** for the all-versus-all stage in @sec-multi-all-vs-all.
:::

## All versus all {#sec-multi-all-vs-all}

### Variance-weighted condition number {#sec-weighted-kappa}

With gene-specific uncertainties $\tau_{g}^{2}$ (biological + sampling
+ optional sc–bulk discordance), standardise rows and evaluate
conditioning in the $(J-1)$-dimensional contrast space induced by
$\sum_{j}p_{j}=1$:

$$
\widetilde{\mu}_{gj}
=
\frac{\mu_{gj}-\bar{\mu}_{g}}{\tau_{g}},
\qquad
\kappa\!\left(
\boldsymbol{W}_{\mathcal{G}}^{1/2}\boldsymbol{\mu}_{\mathcal{G}}\boldsymbol{C}
\right),
$$ {#eq-weighted-kappa}

where $\boldsymbol{W}_{\mathcal{G}}=\mathrm{diag}(1/\tau_{g}^{2})$ and
$\boldsymbol{C}$ removes the common-expression direction. Unweighted
$\kappa(\boldsymbol{\mu}_{\mathcal{G}})$ alone can favour quiet but
biased genes [@gongDeconRNASeqStatisticalFramework2013]; see also
@sec-kappa-limits.

### Sparse classifiers as candidate generators {#sec-sparse-classifiers}

Beyond geometric scores on $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}$,
multiclass models that treat cell type as a categorical response are
useful **candidate generators**—not as final deconvolution solvers.
Embed them in **stability selection**: resample at the donor level,
tune penalties inside each resample, and keep genes selected in a high
fraction of replicates
[@meinshausenStabilitySelection2010].

- **Multinomial logistic regression with elastic net.** Combined
  $\ell_1$ / $\ell_2$ penalties yield sparse class weights while
  shrinking correlated genes jointly
  [@zouRegularizationVariableSelection2005; @R-glmnet]. Suitable when
  the number of donor-level pseudobulks is moderate to large.
- **Sparse / penalised LDA.** Maximises a between- / within-class
  variance ratio with an $\ell_1$ penalty on loadings, favouring a
  compact, non-redundant discriminant gene set
  [@wittenPenalizedClassificationUsing2011].
- **Group and tree-guided lasso.** An $\ell_1/\ell_2$ group penalty
  selects or drops whole genes or pathways together
  [@yuanModelSelectionEstimation2006]; hierarchical / tree-guided
  variants share strength across related cell-type labels when a
  lineage taxonomy is known.

Treat non-zero coefficients as proposals for @sec-optimal-design and
@sec-overlap, then refine with a wrapper (@sec-wrapper).

```{r}
#| label: tbl-classifier-candidates
#| tbl-cap: "Sparse multiclass methods as panel candidate generators."
#| echo: false

classifiers <- data.frame(
  Method = c(
    "Multinomial elastic net",
    "Sparse LDA",
    "Group / tree lasso",
    "Stability selection"
  ),
  Role = c(
    "Joint classification; L1 sparsity + L2 grouping of correlates",
    "Direct class-separation criterion with sparse loadings",
    "Pathway- or hierarchy-aware block selection",
    "Donor-level resampling wrapper around any of the above"
  ),
  Caveat = c(
    "May still retain collinear markers unless L2 is strong",
    "Needs more observations than pure logistic paths",
    "Requires predefined groups or a taxonomy tree",
    "Computationally heavy; prefer as the aggregation layer"
  )
)
knitr::kable(classifiers)
```

### D- / E- / A-optimal design {#sec-optimal-design}

Under the DeCovarT mixture (@eq-mixture), write the bulk covariance as

$$
\boldsymbol{V}(\boldsymbol{p})
=
\boldsymbol{\Sigma}_{\mathrm{meas}}
+
\sum_{j=1}^{J}p_{j}^{2}\boldsymbol{\Sigma}_{j}.
$$ {#eq-bulk-covariance}

Unlike ordinary linear regression, $\boldsymbol{V}$ depends on
$\boldsymbol{p}$. With mean map
$\boldsymbol{M}(\boldsymbol{p})=\boldsymbol{\mu}\boldsymbol{p}$ and
Jacobian columns $\boldsymbol{\mu}_{.j}-\boldsymbol{\mu}_{.J}$ in the
$(J-1)$-dimensional simplex contrast space, the Gaussian Fisher
information is

$$
\mathcal{I}(\boldsymbol{p})
=
\boldsymbol{M}'^{\top}
\boldsymbol{V}(\boldsymbol{p})^{-1}
\boldsymbol{M}'
+
\frac{1}{2}
\sum_{a,b}
\operatorname{tr}\!\left(
\frac{\partial\boldsymbol{V}}{\partial p_{a}}
\boldsymbol{V}(\boldsymbol{p})^{-1}
\frac{\partial\boldsymbol{V}}{\partial p_{b}}
\boldsymbol{V}(\boldsymbol{p})^{-1}
\right),
$$ {#eq-fisher}

where $\partial\boldsymbol{V}/\partial p_{j}=2p_{j}\boldsymbol{\Sigma}_{j}$.
The first term is mean information; the second captures information
from the changing covariance. For a gene panel $\mathcal{G}$, restrict
rows of $\boldsymbol{\mu}$ and $\boldsymbol{\Sigma}_{j}$ accordingly.
When $\boldsymbol{V}$ is treated as fixed (or diagonal / shrinkage on
$\mathcal{G}$), @eq-fisher collapses to the familiar whitened Gram

$$
\mathcal{I}_{\mathcal{G}}
\approx
\boldsymbol{\mu}_{\mathcal{G}}^{\top}
\boldsymbol{\Sigma}_{\mathcal{G}}^{-1}
\boldsymbol{\mu}_{\mathcal{G}}.
$$

Evaluate candidate panels over a **composition grid**
$\mathcal{P}\subset\Delta^{J-1}$ (balanced, rare-cell, sparse,
near-collinear scenarios) and report worst-case and average criteria,
so a design that only works at one corner of the simplex is rejected.

Classical alphabetical criteria
[@atkinsonOptimumExperimentalDesigns1992;
@pukelsheimOptimalDesignExperiments2006] are

$$
\max_{\mathcal{G}}\min_{\boldsymbol{p}\in\mathcal{P}}
\log\det\bigl(\mathcal{I}_{\mathcal{G}}(\boldsymbol{p})+\epsilon\boldsymbol{I}\bigr)
\quad\text{(robust D)},
\qquad
\max_{\mathcal{G}}\lambda_{\min}(\mathcal{I}_{\mathcal{G}})
\quad\text{(E)},
\qquad
\min_{\mathcal{G}}\operatorname{tr}(\mathcal{I}_{\mathcal{G}}^{-1})
\quad\text{(A)}.
$$ {#eq-optimal-design}

@fig-dea-design recalls the three criteria geometrically.

![D-, E- and A-optimal design for gene-panel selection: shared Fisher
information $\mathcal{I}_{\mathcal{G}}$, confidence-ellipse intuition,
and links to condition number / Gram volume.](figures/dea_optimal_design_comparison.png){#fig-dea-design}

::: {#nte-dea-proscons .callout-note}
## :information_source: Pros and cons of D / E / A

- **D-optimality** ($\log\det$): shrinks the volume of the asymptotic
  confidence ellipsoid for $\boldsymbol{p}$. Efficient in an average
  sense; can still leave one weak contrast if other directions improve.
- **E-optimality** ($\lambda_{\min}$): protects the **worst** contrast.
  Closest in spirit to minimising $\kappa_2(\mathcal{I}_{\mathcal{G}})$.
- **A-optimality** ($\operatorname{tr}\mathcal{I}^{-1}$): minimises
  average parameter variance under the Gaussian approximation.

Prefer shrinkage covariances and a small ridge $\epsilon\boldsymbol{I}$
when $|\mathcal{G}|$ is moderate. For combinatorial search, use a
greedy / Fedorov exchange as the auditable baseline, and a genetic
algorithm only when the objective mixes covariance, overlap, or
mechanistic constraints (@sec-mechanistic).
:::

**Relation to $\kappa$ and the Gram volume.** If
$\boldsymbol{\Sigma}_{\mathcal{G}}=\boldsymbol{I}$, then
$\mathcal{I}_{\mathcal{G}}=\boldsymbol{\mu}_{\mathcal{G}}^{\top}\boldsymbol{\mu}_{\mathcal{G}}$
and D-optimality maximises the log **Gram volume** of the mean columns.
E-optimality maximises $\lambda_{\min}$ of that Gram matrix and is a
cousin of minimising $\kappa_2(\boldsymbol{\mu}_{\mathcal{G}})$
(see @eq-condition-number). The
[synthetic scenarios](synthetic-scenarios.html#sec-geometric-score)
vignette discusses the same objects as AutoGeneS-style scalar
summaries of $\boldsymbol{\mu}$.

#### Software for D-optimal search in R

For **standard** D-optimal experiments on a candidate factor grid,
`skpr` is a strong general-purpose choice (D / A / E / G / I, blocking,
diagnostics). `AlgDesign::optFederov()` remains a lightweight Fedorov
exchange over a finite candidate set. Neither package natively
implements the proportion-dependent FIM in @eq-fisher, so for DeCovarT
panels we recommend a **custom objective** plus greedy / exchange
search, with `GA` only as optional refinement.

```{r}
#| label: ex-skpr-d-optimal
#| echo: true
#| eval: false
# Conventional D-optimal design on a candidate grid (illustration only).
# Requires the skpr package (CRAN); not used at knit time.
candidates <- expand.grid(
  x1 = seq(-1, 1, length.out = 11),
  x2 = seq(-1, 1, length.out = 11)
)
design <- skpr::gen_design(
  candidateset = candidates,
  model = ~ x1 + x2 + I(x1^2) + I(x2^2) + x1:x2,
  trials = 20,
  optimality = "D",
  repeats = 100
)
skpr::get_optimality(design, "D")
```

```{r}
#| label: ex-algdesign-federov
#| echo: true
#| eval: false
candidates <- expand.grid(
  x1 = seq(-1, 1, length.out = 11),
  x2 = seq(-1, 1, length.out = 11)
)
result <- AlgDesign::optFederov(
  formula = ~ x1 + x2 + I(x1^2) + I(x2^2) + x1:x2,
  data = candidates,
  nTrials = 20,
  criterion = "D",
  nRepeats = 100
)
result$D
```

```{r}
#| label: ex-robust-d-panel
#| echo: true
#| eval: false
# Robust D-score for a gene panel over composition scenarios.
# `information_matrices[[s]]` is I_G(p_s) for scenario s (J-1 dim.).
d_objective <- function(selected, information_matrices, ridge = 1e-8) {
  scores <- vapply(information_matrices, function(info_full) {
    info_s <- info_full[selected, selected, drop = FALSE]
    info_s <- info_s + diag(ridge, nrow(info_s))
    as.numeric(determinant(info_s, logarithm = TRUE)$modulus)
  }, numeric(1))
  min(scores)
}
# Optimise the binary selection with a custom exchange / GenSA / GA.
```

### Marker-separation metrics {#sec-overlap}

AutoGeneS optimises correlation and centroid distance separately
[@alieeAutoGeneSAutomaticGene2021]. Probabilistic alternatives fold
mean **and** covariance structure into a single separation score for
the purified laws
$\mathcal{N}(\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j})$.

**MixSim pairwise overlap.** For densities $f_{j}$ and $f_{\ell}$, the
one-sided mass
$\omega_{j\mid\ell}=\Pr_{X\sim f_{j}}(f_{\ell}(X)>f_{j}(X))$ yields
the pairwise overlap

$$
\omega_{j\ell}
=
\omega_{j\mid\ell}+\omega_{\ell\mid j}
=
\int\min\bigl(f_{j}(x),f_{\ell}(x)\bigr)\,dx
\in[0,1],
$$ {#eq-pairwise-overlap}

with $0$ for perfectly separable Gaussians and $1$ for identical laws
[@melnykovMixSimPackageSimulating2012; @R-MixSim]. Mixture proportions
enter the MAP rule inside MixSim; they should not be multiplied into
$\omega_{j\mid\ell}$ a second time when reporting pairwise overlap.
The unweighted average
$\overline{\omega}=\tfrac{2}{J(J-1)}\sum_{j<\ell}\omega_{j\ell}$ is a
natural panel monitor.

Package helper `compute_average_overlap()` instead returns a
**proportion-weighted** average of MixSim's directional masses,

$$
\overline{\Omega}
=
\frac{2}{J(J-1)}
\sum_{1\le j<\ell\le J}
\bigl(p_{j}\Omega_{j\ell}+p_{\ell}\Omega_{\ell j}\bigr),
$$ {#eq-overlap}

which emphasises abundant types in the mixture prior. Lower values of
either summary correspond to better-separated purified profiles.

**Jeffreys (symmetric KL) divergence.** With
$J(f_{j},f_{\ell})=D_{\mathrm{KL}}(f_{j}\parallel f_{\ell})+D_{\mathrm{KL}}(f_{\ell}\parallel f_{j})$
and multivariate normals,

$$
\begin{aligned}
D_{\mathrm{KL}}(\mathcal{N}_{j}\parallel\mathcal{N}_{\ell})
&=
\tfrac{1}{2}\Bigl[
\ln\frac{\lvert\boldsymbol{\Sigma}_{\ell}\rvert}{\lvert\boldsymbol{\Sigma}_{j}\rvert}
-d
+\operatorname{tr}(\boldsymbol{\Sigma}_{\ell}^{-1}\boldsymbol{\Sigma}_{j})
+(\boldsymbol{\mu}_{\ell}-\boldsymbol{\mu}_{j})^{\top}
\boldsymbol{\Sigma}_{\ell}^{-1}
(\boldsymbol{\mu}_{\ell}-\boldsymbol{\mu}_{j})
\Bigr],
\\
J(f_{j},f_{\ell})
&=
\tfrac{1}{2}\operatorname{tr}\bigl(
\boldsymbol{\Sigma}_{\ell}^{-1}\boldsymbol{\Sigma}_{j}
+\boldsymbol{\Sigma}_{j}^{-1}\boldsymbol{\Sigma}_{\ell}
-2\boldsymbol{I}
\bigr)
\\
&\quad
+\tfrac{1}{2}(\boldsymbol{\mu}_{j}-\boldsymbol{\mu}_{\ell})^{\top}
\bigl(\boldsymbol{\Sigma}_{j}^{-1}+\boldsymbol{\Sigma}_{\ell}^{-1}\bigr)
(\boldsymbol{\mu}_{j}-\boldsymbol{\mu}_{\ell}).
\end{aligned}
$$ {#eq-jeffreys}

$J\ge 0$ is unbounded; large values flag strong mean or covariance
mismatch.

**Chernoff information.** The Chernoff information

$$
C(f_{j},f_{\ell})
=
-\min_{t\in[0,1]}
\ln\int f_{j}(x)^{t}f_{\ell}(x)^{1-t}\,dx
$$ {#eq-chernoff}

is the best exponential error exponent for distinguishing the two laws.
For Gaussians the integral is closed in $t$ after substituting
$\boldsymbol{\Sigma}_{t}=t\boldsymbol{\Sigma}_{\ell}+(1-t)\boldsymbol{\Sigma}_{j}$,
but $t^{\star}$ is found numerically. Large $C$ implies strong
separability ($P_{\mathrm{error}}\le e^{-C}$).

**2-Wasserstein / Bures geometry.** The squared $W_{2}$ distance
between Gaussians is

$$
W_{2}^{2}\bigl(
\mathcal{N}(\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j}),
\mathcal{N}(\boldsymbol{\mu}_{.\ell},\boldsymbol{\Sigma}_{\ell})
\bigr)
=
\lVert\boldsymbol{\mu}_{.j}-\boldsymbol{\mu}_{.\ell}\rVert_{2}^{2}
+
\mathfrak{B}^{2}(\boldsymbol{\Sigma}_{j},\boldsymbol{\Sigma}_{\ell}),
$$ {#eq-wasserstein}

with $\mathfrak{B}^{2}$ the Bures metric (matrix square-root term). It
is a true metric and reduces to Euclidean mean distance when
covariances coincide.

```{r}
#| label: tbl-separation-metrics
#| tbl-cap: "Comparison of purified-Gaussian separation metrics for panel monitoring."
#| echo: false

metrics <- data.frame(
  Metric = c(
    "MixSim overlap",
    "Jeffreys (sym. KL)",
    "Chernoff information",
    "2-Wasserstein"
  ),
  Range = c(
    "[0, 1]",
    "[0, Inf)",
    "[0, Inf)",
    "[0, Inf)"
  ),
  Focus = c(
    "Bayes misclassification mass",
    "Average log-likelihood gap",
    "Optimal error exponent",
    "Geometric mean + covariance shape"
  ),
  Notes = c(
    "Monte Carlo in MixSim; intuitive",
    "Closed form for Gaussians",
    "Needs a 1-D line search in t",
    "Matrix square-root cost O(d^3)"
  )
)
knitr::kable(metrics)
```

D/E/A (@eq-optimal-design) optimise **estimation** of $\boldsymbol{p}$
under @eq-mixture; overlap / Jeffreys / Chernoff / $W_{2}$ monitor
**class separation**. They need not agree. Prefer overlap for an
interpretable error rate, Jeffreys for a closed-form divergence,
Chernoff when a single discriminability index is desired, and $W_{2}$
when covariance shape matters
[@R-MixSim; @alieeAutoGeneSAutomaticGene2021].

```{r}
#| label: ex-separation-metrics-r
#| echo: true
#| eval: false
# Jeffreys divergence and Gaussian W2^2 between two purified types.
jeffreys_gaussian <- function(mu1, S1, mu2, S2) {
  inv1 <- solve(S1)
  inv2 <- solve(S2)
  d <- mu1 - mu2
  term_tr <- sum(diag(inv2 %*% S1 + inv1 %*% S2 - 2 * diag(length(mu1))))
  term_mu <- as.numeric(t(d) %*% (inv1 + inv2) %*% d)
  0.5 * (term_tr + term_mu)
}

wasserstein2_sq <- function(mu1, S1, mu2, S2) {
  # Requires a SPD square-root helper (e.g. expm::sqrtm).
  sqrtS1 <- expm::sqrtm(S1)
  inner <- expm::sqrtm(sqrtS1 %*% S2 %*% sqrtS1)
  sum((mu1 - mu2)^2) + sum(diag(S1 + S2 - 2 * inner))
}
```

# Wrapper optimisation on pseudo-bulks and matched bulks {#sec-wrapper}

When matched sc/bulk aliquots exist (e.g. Kassandra-style benchmarks)
[@zaitsevPreciseReconstructionTME2022], evaluate candidate panels by
running the **intended** deconvolution algorithm on held-out mixtures
and scoring RMSE / worst-type error / rare-cell recall under nested
**donor-level** cross-validation. Use the one-standard-error rule to
prefer the smallest panel statistically indistinguishable from the
best. Wrappers overfit the mixture simulator if real matched bulks are
omitted
[@sturmComprehensiveEvaluationTranscriptomebased2019;
@jinBenchmarkRNAseqDeconvolution2021].

```{r}
#| label: tbl-strategy-summary
#| tbl-cap: "Compact map from selection strategy to typical role in a reference-based pipeline."

strategies <- data.frame(
  Stage = c(
    "Univariate 1 vs others",
    "Univariate all vs all",
    "Multivariate 1 vs others",
    "Multivariate all vs all",
    "Wrapper"
  ),
  Examples = c(
    "DE / F-test / Gini entropy",
    "AutoGeneS",
    "INDEED; sparse GGM",
    "Weighted κ; D/E/A-opt; classifiers; overlap",
    "Nested CV on matched bulks"
  ),
  Role = c(
    "Candidate generation",
    "Panel compaction",
    "Network-aware candidates",
    "Identifiability of p",
    "Empirical size selection"
  )
)

knitr::kable(strategies)
```

# Recommended pipeline {#sec-recommended}

- ① **Reference moments** (see @sec-preprocess). Donor-level
  pseudobulks; estimate $\boldsymbol{\mu}$ and variance components on a
  linear scale; optional SCTransform / top-$2{,}000$ HVGs
  (see @sec-sctransform) **without** regressing out cell type.
- ② **Reliability filter.** Drop unstable / undetectable / discordant
  genes; optionally inflate $\tau_{g}^{2}$ by observed sc–bulk
  discrepancy.
- ③ **Candidates.** Union of stable one-versus-others DE
  (see @sec-univ-one-vs-others), high Gini / low entropy
  (see @sec-gini-entropy), differential-network hits
  (see @sec-indeed), and optional sparse classifiers with donor-level
  stability selection (see @sec-sparse-classifiers)
  → roughly $10^{2}$–$10^{3}$ genes.
- ④ **Separation monitors.** Rank or filter with MixSim overlap,
  Jeffreys, Chernoff, and/or $W_{2}$ (see @sec-overlap).
- ⑤ **Information-optimal subset.** Greedy / exchange D- or E-optimal
  search over a composition grid (see @eq-optimal-design,
  @eq-fisher), with weighted $\kappa_2$ (see @eq-weighted-kappa) as a
  diagnostic; heed @sec-kappa-limits on continua.
- ⑥ **Wrapper prune** (see @sec-wrapper). Nested donor CV and matched
  bulks; keep the smallest competitive panel.
- ⑦ **Mechanistic annotation** (see @sec-mechanistic). Optional
  CARNIVAL / CORNETO pathway hypotheses as soft coverage constraints.

```{mermaid}
%%| label: fig-recommended-pipeline
%%| fig-cap: "Recommended reference-based feature-selection pipeline from donor pseudobulks to validated, mechanistically annotated panels."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  A["Donor split + pseudobulk"] --> B["SCTransform HVGs\n(no cell-type regression)"]
  B --> C["Reliability / dropout filter"]
  C --> D["Candidates:\nDE + Gini + INDEED +\nsparse classifiers"]
  D --> E["Stability selection"]
  E --> F["Overlap / Jeffreys /\nChernoff / W2 monitors"]
  F --> G["Robust D/E/A design\nover composition grid"]
  G --> H["Greedy / exchange search"]
  H --> I["Optional GA refinement"]
  I --> J["Held-out wrapper validation"]
  J --> K["CARNIVAL / CORNETO\nannotation"]
  K --> L["Report panel +\nuncertainty + pathways"]
```

The distinctive contribution for DeCovarT-style second-generation
models is a **bulk-calibrated, variance-weighted information criterion
plus overlap**, exploiting paired sc/bulk designs when available.

# Perspectives {#sec-perspectives}

## Mechanistic priors {#sec-mechanistic}

PPI / TF networks and pathway modules can regularise both GGM
estimation and cooperative penalties
[@chiquetSparsitySigncoherentGroups2012;
@zuoINDEEDIntegratedDifferential2016], shrinking the search toward
biologically plausible supports rather than purely statistical markers.

**CARNIVAL** contextualises a prior-knowledge network (e.g. OmniPath)
from gene-expression footprints via integer linear programming: TF /
pathway activity scores (DoRothEA and related) are matched to upstream
regulators, yielding directed subnetworks and key mediators for each
cell-type contrast [@liuFromExpressionFootprints2019]. For panel
design, annotate candidate markers by membership in (or upstream of)
inferred pathways, or softly prioritise genes that lie on divergent
axes between types. Outputs are hypotheses: they depend on the PKN and
activity discretisation, may be non-unique, and do not prove causality.

**CORNETO** casts multi-sample network inference as a joint
mixed-integer programme with structured sparsity, recovering a shared
backbone plus sample- or cell-type-specific edges across directed,
undirected, signed, or hypergraph priors
[@rodriguezMierUnifyingMultisample2025]. Cell-specific edges highlight
discriminative modules; shared edges flag core programmes. After a
statistically chosen panel, require coverage of distinct CORNETO
modules, or add a soft term that rewards selecting genes from different
subnetworks. Limitations mirror CARNIVAL: solver cost, PKN dependence,
and false-positive / missing edges.

Use CARNIVAL / CORNETO for **annotation and soft constraints**, not as
a replacement for overlap (@eq-overlap) or Fisher design
(@eq-optimal-design).

## Continuous states, potency and archetypes {#sec-continua}

Many tissues are better described by **continua** (differentiation
trajectories, activation gradients) than by discrete labels. In that
regime, a finite $J$-column signature is a piecewise approximation of
an archetype geometry [@hartInferringBiologicalTasks2015] or a potency
axis [@gulatiSinglecellTranscriptionalDiversity2020]: genes should
separate extreme vertices **and** resolve interior mixtures along
trajectories. Open directions include replacing hard one-versus-others
contrasts with scores along inferred latent axes, and evaluating
overlap / Fisher information on continuous rather than categorical
components. Until such methods mature, hierarchical discrete panels
(lineage → subtype) remain a pragmatic compromise
[@wangBulkTissueCell2019]. See also @sec-kappa-limits.

## Seeds and undirected Markov boundaries {#sec-markov-seeds}

A complementary panel-expansion strategy starts from a small set of
**seeds**—genes that are already highly discriminatory for
$\boldsymbol{p}$ under @eq-mixture or @eq-indeed-as—then grows each
seed by its **Markov boundary** in a cell-type (or differential) GGM.

In an undirected graphical model, a Markov blanket of gene $g$ is any
set $S$ such that $g$ is conditionally independent of all other genes
given $S$; a **Markov boundary** is a minimal such set
[@pearlProbabilisticReasoningIntelligent1988]. For a Gaussian Markov
random field / GGM, the Markov boundary of $g$ coincides with its
graph neighbours: the indices $g'$ with $\omega_{gg'}\neq 0$ in
$\boldsymbol{\Omega}$
[@bolinExplicitLinkGraphical2026]. Expanding seeds by these
neighbours therefore adds genes that are **not** redundant once the
seed is conditioned on—they carry residual information about the
local dependency structure.

::: {#nte-u-vs-dsep .callout-important}
## :warning: Undirected ($U$) separation, not $d$-separation

This construction uses the **undirected** Markov property (conditioning
on neighbours in $\boldsymbol{\Omega}$). Classical $d$-separation
applies to **directed** acyclic Gaussian Bayesian networks and is the
wrong language for GGM / glasso supports. Do not treat directed path
blocking rules as interchangeable with undirected Markov boundaries
[@pearlProbabilisticReasoningIntelligent1988;
@bolinExplicitLinkGraphical2026].
:::

Operational sketch:

- ① Select seeds by mean separation, INDEED activity (see @eq-indeed-as),
  or high contribution to $\mathcal{I}_{\mathcal{G}}$
  (see @eq-fisher).
- ② Estimate $\boldsymbol{\Omega}_{j}$ (or a differential precision) on
  the HVG universe (see @sec-sctransform).
- ③ For each seed $g$, add its undirected neighbours
  $\{g':\omega_{gg'}\neq 0\}$ (the Markov boundary).
- ④ Re-run all-versus-all compaction (see @sec-optimal-design,
  @sec-overlap) so that boundary expansion does not reintroduce
  mean-collinear bulk (see @nte-ovr-concat).

# Software notes {#sec-software}

- Variance stabilisation / Pearson residuals: `sctransform` /
  `Seurat::SCTransform`
  [@hafemeisterNormalizationVarianceStabilization2019].
- Signature QC / compaction patterns: DeconvExplorer
  (`removePercentZeros`, `removeUnspecificGenes`,
  `selectGenesByScore` with Gini or entropy).
- Sparse multiclass candidates: `glmnet` multinomial elastic net
  [@R-glmnet; @zouRegularizationVariableSelection2005]; group penalties
  following [@yuanModelSelectionEstimation2006]; stability selection
  [@meinshausenStabilitySelection2010].
- Separation monitors: `MixSim::overlap()` [@R-MixSim;
  @melnykovMixSimPackageSimulating2012].
- Conventional D-optimal grids: `skpr::gen_design()`,
  `AlgDesign::optFederov()` [@R-AlgDesign] (see @sec-optimal-design);
  custom FIM + exchange / `GenSA` / `GA` [@R-GenSA; @R-GA] for
  proportion-dependent DeCovarT panels.
- Mechanistic annotation: CARNIVAL / CORNETO
  [@liuFromExpressionFootprints2019;
  @rodriguezMierUnifyingMultisample2025].
- Gold-standard reference-based estimators for downstream validation:
  CIBERSORT / CIBERSORTx, EPIC, quanTIseq, ABIS, MuSiC, DWLS
  [@newmanRobustEnumerationCell2015; @newmanDeterminingCellType2019;
  @racleSimultaneousEnumerationCancer2017; @finotello_etal19;
  @monacoRNASeqSignaturesNormalized2019; @wangBulkTissueCell2019;
  @tsoucasAccurateEstimationCelltype2019].
- Optimal-design background:
  [@atkinsonOptimumExperimentalDesigns1992;
  @pukelsheimOptimalDesignExperiments2006].
