---
title: "Feature selection for reference-based deconvolution"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Feature selection for reference-based deconvolution}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  eq-title: "Equation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)
```

# Scope {#sec-scope}

This vignette surveys **feature selection for reference-based
(signature-based) deconvolution**—methods that recover cellular
proportions $\boldsymbol{p}$ from a bulk mixture using a purified
expression design matrix
[@gaspard-boulincCelltypeDeconvolutionMethods2025;
@avilacobosComputationalDeconvolutionTranscriptomics2018;
@sturmComprehensiveEvaluationTranscriptomebased2019]. Feature selection
itself is a classical pre-modelling filter
[@guyonIntroductionVariableFeature2003]. We deliberately exclude
**marker-based** abundance scoring (enrichment / GSEA-style modules),
which returns proxies of presence rather than a consensual gene panel
suited to linear or probabilistic mixture models
[@aranXCellDigitallyPortraying2017; @zhang_etal17].

The focus is **second-generation** signatures built from single-cell
pseudobulks (CIBERSORTx, MuSiC, DWLS, Kassandra and related pipelines)
[@newmanDeterminingCellType2019; @wangBulkTissueCell2019;
@tsoucasAccurateEstimationCelltype2019;
@zaitsevPreciseReconstructionTME2022], where the selected genes must
discriminate **all** cell types in the signature jointly, not only
one-versus-rest markers. Correlation-aware compact panels such as
DUBStepR illustrate the same goal from a single-cell feature-selection
angle [@ranjanDUBStepRScalableCorrelationbased2021].

Interactive signature refinement utilities (zero filtering,
specificity binning, Gini / entropy compaction) are implemented in
[DeconvExplorer](https://github.com/omnideconv/deconvExplorer)
(omnideconv ecosystem; cite as software once added to
`packages.bib`).

```{mermaid}
%%| label: fig-feature-selection-layout
%%| fig-cap: "Layout of this vignette: univariate and multivariate selection, each split into one-versus-others then all-versus-all, followed by perspectives for continuous cell states."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  A["Reference μ, Σ from sc / bulk"] --> B["Univariate"]
  A --> C["Multivariate"]
  B --> B1["One vs others"]
  B --> B2["All vs all"]
  C --> C1["One vs others\n(GGM / DiffNet)"]
  C --> C2["All vs all\n(κ, D/E-opt, overlap)"]
  B2 --> D["Consensual panel G"]
  C2 --> D
  D --> E["Perspectives:\nmechanistic / potency / archetypes"]
```

Sections are cross-linked below: @sec-preprocess builds moments;
@sec-univariate and @sec-multivariate score genes; @sec-wrapper and
@sec-recommended assemble a pipeline; @sec-perspectives covers
continua and Markov blankets.

# Notation {#sec-notation}

We align notation with the DeCovarT generative model. Let $G$ genes and
$J$ cell types index the purified means
$\boldsymbol{\mu}=(\boldsymbol{\mu}_{.j})_{j=1}^{J}\in\mathbb{R}^{G\times J}$
and (optional) covariances $\{\boldsymbol{\Sigma}_{j}\}_{j=1}^{J}$. A bulk
observation obeys

$$
\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})
\sim
\mathcal{N}_{G}\!\left(
\boldsymbol{\mu}\boldsymbol{p},\,
\sum_{j=1}^{J}p_{j}^{2}\boldsymbol{\Sigma}_{j}
\right),
\quad
\boldsymbol{p}\in\Delta^{J-1},
$$ {#eq-mixture}

with $\boldsymbol{\zeta}=(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$
and simplex
$\Delta^{J-1}=\{\boldsymbol{p}:p_{j}\ge 0,\sum_{j}p_{j}=1\}$. For a
gene subset $\mathcal{G}\subseteq\{1,\ldots,G\}$, write
$\boldsymbol{\mu}_{\mathcal{G}}$ for the corresponding rows. The
condition number of a design matrix $\boldsymbol{X}$ is

$$
\kappa(\boldsymbol{X})
=
\frac{\lambda_{\max}(\boldsymbol{X})}{\lambda_{\min}(\boldsymbol{X})},
$$ {#eq-condition-number}

with large $\kappa$ indicating multicollinearity and numerical
instability [@gongDeconRNASeqStatisticalFramework2013;
@newmanRobustEnumerationCell2015]. Singular-value form
$\kappa_2(\boldsymbol{X})=\sigma_{\max}(\boldsymbol{X})/\sigma_{\min}(\boldsymbol{X})$
is used interchangeably when $\boldsymbol{X}$ is rectangular (see also
the
[synthetic scenarios](synthetic-scenarios.html#sec-geometric-score)
vignette).

# Preprocessing shared by all strategies {#sec-preprocess}

Before scoring genes:

- ① Aggregate **donor-level** cell-type pseudobulks (not individual
  cells as independent replicates). Pseudobulking reduces sparsity and
  avoids treating cells as independent biological replicates in
  differential testing [@squairConfrontingFalseDiscoveries2021].
- ② Estimate gene–cell-type means $\mu_{gj}$ and variance components on
  a **linear** scale (counts, CPM, or TPM). Log-space breaks the
  mixture linearity in @eq-mixture
  [@sturmComprehensiveEvaluationTranscriptomebased2019;
  @jinBenchmarkRNAseqDeconvolution2021;
  @hoffmannRobustComputationalReconstitution2006]. Variance
  stabilisation for *single-cell* QC is handled separately below
  (see @sec-sctransform).
- ③ Filter genes with poor bulk detectability, extreme zero inflation
  after pseudobulking, strong donor instability, or systematic
  sc–bulk discordance.
- ④ When the biological target is **cell number** rather than RNA mass,
  correct transcriptome-size differences across cell types
  [@luTranscriptomeSizeMatters2025].
- ⑤ Apply the **same** normalisation to bulk and signature
  [@jinBenchmarkRNAseqDeconvolution2021;
  @racleSimultaneousEnumerationCancer2017].

## Variance stabilisation and highly variable genes {#sec-sctransform}

Single-cell count matrices mix **technical** depth noise with
**biological** cell-type variability. Regularised negative-binomial
regression as in `Seurat::SCTransform` / `sctransform` separates the
two via **Pearson residuals**
[@hafemeisterNormalizationVarianceStabilization2019]:

1. Infer a depth-aware mean $\hat{\mu}_{ig}$ for cell $i$ and gene $g$
   (library size enters as an offset; optional covariates can be
   included in the design).
2. Infer a gene-wise dispersion $\hat{\theta}_{g}$ with empirical-Bayes
   shrinkage and a local mean–dispersion trend across genes.
3. Form Pearson residuals
   $$
   r_{ig}
   =
   \frac{y_{ig}-\hat{\mu}_{ig}}
   {\sqrt{\hat{\mu}_{ig}+\hat{\mu}_{ig}^{2}/\hat{\theta}_{g}}},
   $$ {#eq-pearson-residual}
   which are approximately homoscedastic and closer to Gaussian than
   raw counts—useful for PCA / HVG ranking, **not** as the linear
   mixture scale in @eq-mixture.

@fig-sc-transform summarises the negative-binomial observation model,
depth regression, residualisation, and empirical-Bayes shrinkage.

![SCTransform workflow: negative-binomial mean–variance model, depth
regression, Pearson residuals, and empirical-Bayes regularisation
across genes.](figures/sc_transform.png){#fig-sc-transform}

After residualisation, retain a compact **highly variable gene (HVG)**
universe before signature scoring. A practical design that emphasises
cell-type structure (rather than a global intercept) is

$$
\sim\ \texttt{CellType}
$$

**without intercept**, so the estimated coefficients contrast each
labelled type against the **global mean expression across cell types
and samples**. Rank genes by residual variance (or the corresponding
model $F$ / deviance explained) and keep the top $2{,}000$ HVGs as the
working universe for the selectors in @sec-univariate and
@sec-multivariate. Downstream deconvolution still uses
**linear-scale** $\mu_{gj}$ on that gene set
[@sturmComprehensiveEvaluationTranscriptomebased2019].

::: {#nte-sctransform-check .callout-tip}
## :information_source: Sanity check after SCTransform

Per gene, residual histograms and mean–variance plots of $r_{ig}$
should look roughly Gaussian and flat. Strong remaining depth trends
or heavy tails indicate that the offset / covariate design is
misspecified before any marker ranking.
:::

# Univariate approaches {#sec-univariate}

Univariate selectors score each gene marginally (or with a simple
contrast) and ignore partial correlations among genes.

## One versus others {#sec-univ-one-vs-others}

Classical signature pipelines score gene $g$ for cell type $j$ against
the remaining types, then concatenate per-type shortlists
[@abbasDeconvolutionBloodMicroarray2009;
@newmanRobustEnumerationCell2015; @finotello_etal19].

### Differential expression and fold-change ranks

Limma–voom, edgeR and DESeq2 remain the workhorses for
one-versus-rest contrasts
[@ritchieLimmaPowersDifferential2015a;
@robinsonEdgeRBioconductorPackage2010;
@loveModeratedEstimationFold2014a]. CIBERSORT retains genes with
adjusted $p\le 0.3$ ordered by decreasing fold-change before a global
condition-number prune [@newmanRobustEnumerationCell2015]. Abbas et
al. keep genes that exceed the second- and third-ranked cell types,
then choose the panel with smallest $\kappa$
[@abbasDeconvolutionBloodMicroarray2009]. quanTIseq / xCell-style
filters further drop non-haematopoietic contaminants
[@finotello_etal19; @aranXCellDigitallyPortraying2017].

### ANOVA $F$-test {#sec-anova-f}

Wang et al. rank genes by the nested-model $F$ statistic comparing a
cell-type factor to an intercept-only model
[@wangSilicoEstimatesTissue2010]. For gene $g$ with expression
$x_{gi}$ in sample $i$ and cell-type labels $z_{i}\in\{1,\ldots,J\}$,

$$
F_{g}
=
\frac{
\bigl(\mathrm{SSR}_{0}-\mathrm{SSR}_{1}\bigr)/(J-1)
}{
\mathrm{SSR}_{1}/(n-J)
},
$$ {#eq-anova-f}

where $\mathrm{SSR}_{0}=\sum_{i}(x_{gi}-\bar{x}_{g})^{2}$ and
$\mathrm{SSR}_{1}=\sum_{j}\sum_{i:z_{i}=j}(x_{gi}-\bar{x}_{gj})^{2}$.
Large $F_{g}$ indicates that cell-type means explain a substantial
share of gene variance.

Classical ANOVA inference for @eq-anova-f assumes, among other things:

- independent observations (hence donor-level pseudobulks rather than
  cells as replicates; @sec-preprocess);
- roughly **Gaussian conditional residuals**
  $x_{gi}\mid z_{i}=j\sim\mathcal{N}(\mu_{gj},\sigma_{g}^{2})$ with
  common within-type variance (homoscedasticity);
- no strong outliers or unmodelled batch structure.

When residuals are heavy-tailed or variance scales with the mean,
prefer a GLM / limma–voom pipeline
[@ritchieLimmaPowersDifferential2015a] and treat $F_{g}$ as a ranking
score rather than a calibrated $p$-value.

### Gini and entropy specificity {#sec-gini-entropy}

BioQC and DeconvExplorer compact signatures with Gini or entropy
scores over the row $\boldsymbol{\mu}_{g\cdot}$ [@zhang_etal17]. Write
non-negative mean expression as relative **shares**

$$
q_{gj}
=
\frac{\mu_{gj}}{\sum_{\ell=1}^{J}\mu_{g\ell}},
\qquad
\sum_{j}q_{gj}=1,
$$ {#eq-shares}

so $\boldsymbol{q}_{g}$ is a discrete distribution over cell types.
The **Gini coefficient** of inequality among ordered shares
$q_{g(1)}\le\cdots\le q_{g(J)}$ and the **Shannon entropy** are

$$
\mathrm{Gini}(g)
=
\frac{J}{J-1}
\left(
1-2\sum_{j=1}^{J}
\frac{J-j+\tfrac{1}{2}}{J}\,q_{g(j)}
\right),
\quad
H(g)
=
-\sum_{j=1}^{J}q_{gj}\log q_{gj}.
$$ {#eq-gini-entropy}

High Gini (or low entropy) flags genes whose mass concentrates on few
types—useful for one-versus-rest compaction. @fig-gini-entropy
contrasts the two scores.

![Gini versus Shannon entropy for cell-type specificity: formulas,
simplified share plots, and panel-wise pros/cons for one-versus-rest
compaction.](figures/gini_vs_entropy_specificity.png){#fig-gini-entropy}

“Redundancy when panels are concatenated” means the following. A gene
with high Gini for type $A$ and another with high Gini for type $B$
may still be **nearly collinear** as rows of $\boldsymbol{\mu}$ if both
types co-express correlated programmes (shared lineage, activation
axis). Concatenating the top-$k$ genes per type therefore inflates
$|\mathcal{G}|$ without improving the geometry of
$\boldsymbol{\mu}_{\mathcal{G}}$: $\kappa(\boldsymbol{\mu}_{\mathcal{G}})$
stays large and NNLS / SVR / GLS remain unstable
[@newmanRobustEnumerationCell2015]. That is why
@sec-univ-one-vs-others must be followed by an all-versus-all stage
(see @sec-univ-all-vs-all, @sec-multi-all-vs-all).

::: {#nte-ovr-concat .callout-warning}
## :warning: Limitation of one-versus-rest concatenation

Union of per-type markers often yields a multicollinear
$\boldsymbol{\mu}_{\mathcal{G}}$ that is still too large for stable
NNLS / SVR / GLS
[@newmanRobustEnumerationCell2015;
@avilacobosComputationalDeconvolutionTranscriptomics2018]. A global
all-versus-all refinement is therefore required
(see @sec-multi-all-vs-all).
:::

## All versus all {#sec-univ-all-vs-all}

### Genetic algorithms (AutoGeneS)

AutoGeneS searches gene subsets with a multi-objective genetic
algorithm that **minimises inter-population correlation** while
**maximising centroid distance**
[@alieeAutoGeneSAutomaticGene2021]. Related multi-objective genetic
searches appear in network module discovery (e.g. MOGAMUN)
[@novoa-del-toroMultiobjectiveGeneticAlgorithm2021]. AutoGeneS is
explicitly all-versus-all: the loss is defined on the full $J$-column
signature rather than on independent one-versus-rest lists.

DeCovarT replaces that dual objective with a single **global overlap**
criterion on the Gaussian mixture
$\{(p_{j},\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j})\}$ (see
@eq-overlap and @sec-overlap).

# Multivariate approaches {#sec-multivariate}

Multivariate selectors use joint dependence—covariance, precision, or
information matrices—so that a gene is scored by how it improves
identifiability of $\boldsymbol{p}$ under @eq-mixture.

## Limitations of mean-only condition numbers {#sec-kappa-limits}

Condition-number pruning (see @eq-condition-number) operates almost
exclusively on the **mean** signature $\boldsymbol{\mu}_{\mathcal{G}}$.
It does not see covariance rewiring: two closely related types can
share nearly collinear means yet differ in their precision graphs
$\boldsymbol{\Omega}_{j}=\boldsymbol{\Sigma}_{j}^{-1}$. In
**continua**—gastruloids, differentiation trajectories, activation
gradients—neighbouring states are transcriptionally similar by
construction, so $\kappa(\boldsymbol{\mu})$ is intrinsically large.
Aggressive $\kappa$-minimisation then tends to keep a few dominant
lineage representatives and drop genes that separate neighbours only
through **network rewiring** rather than large mean fold-changes
[@tsoucasAccurateEstimationCelltype2019;
@wangBulkTissueCell2019].

@fig-kappa-gastruloid illustrates this failure mode for a gastruloid
lineage continuum: mean-conditioned DWLS-style selection can
under-represent adjacent or low-abundance states relative to methods
that retain softer, covariance-aware allocation.

![Condition-number pruning on correlated developmental cell types
(gastruloid continuum). Mean-only $\kappa$ filtering can concentrate
signal in dominant collinear lineages and shrink closely related
states.](figures/condition_number_poor_performance_correlated_cell_types.png){#fig-kappa-gastruloid}

::: {#nte-kappa-means .callout-important}
## :warning: When $\kappa(\boldsymbol{\mu})$ is not enough

Use $\kappa$ as a **diagnostic** of mean geometry, not as the sole
selector, whenever cell states form a continuum or when
discriminatory signal lives in $\boldsymbol{\Sigma}_{j}$ (see
@sec-indeed and @sec-optimal-design). Pair it with overlap
(see @eq-overlap), Fisher information (see @eq-fisher), or differential-network
candidates.
:::

## One versus others {#sec-multi-one-vs-others}

Univariate DGE ignores gene–gene interactions and therefore requires
aggressive multiple-testing correction without modelling network
rewiring. Differential-network and GGM methods address that gap.

### Differential networks (INDEED and related) {#sec-indeed}

INDEED combines mean (fold-change) evidence with tests of changes in
**partial correlation**, ranking genes that both shift in mean **and**
rewire neighbourhood structure
[@zuoINDEEDIntegratedDifferential2016]. @fig-indeed summarises the
graphical abstract and computational pipeline.

::: {#fig-indeed layout-ncol=2}

![INDEED graphical abstract: parallel differential expression and
differential-network tracks merged into an activity-ranked biomarker
list (adapted from @zuoINDEEDIntegratedDifferential2016).](figures/INDEED_graphical_abstract.png){#fig-indeed-abstract}

![INDEED pipeline: group-wise graphical lasso, differential partial
correlations $\Delta\mathrm{pc}$ with permutation filtering, DE
$p$-values, and activity-score prioritisation (adapted from
@zuoINDEEDIntegratedDifferential2016).](figures/INDEED_pipeline.png){#fig-indeed-pipeline}

INDEED integrates differential expression with differential network
structure to prioritise biomolecules.
:::

For groups $k\in\{1,2\}$ (e.g. focal cell type versus the rest), let
$\boldsymbol{\Omega}^{(k)}$ be a sparse precision estimate from
graphical lasso. Partial correlations are

$$
\mathrm{pc}_{gg'}^{(k)}
=
-\frac{\omega_{gg'}^{(k)}}
{\sqrt{\omega_{gg}^{(k)}\,\omega_{g'g'}^{(k)}}},
\qquad
\Delta\mathrm{pc}_{gg'}
=
\mathrm{pc}_{gg'}^{(1)}-\mathrm{pc}_{gg'}^{(2)}.
$$ {#eq-indeed-dpc}

Significant edges are retained after a permutation (or bootstrap) null
on $\Delta\mathrm{pc}_{gg'}$. With a differential-expression $p$-value
$p_{g}^{\mathrm{DE}}$ for gene $g$, INDEED forms an **activity score**
that multiplies mean evidence by differential connectivity, schematically

$$
\mathrm{AS}_{g}
=
\bigl(-\log p_{g}^{\mathrm{DE}}\bigr)
\sum_{g'\neq g}
\bigl\lvert\Delta\mathrm{pc}_{gg'}\bigr\rvert
\,
\mathbf{1}\!\left\{\Delta\mathrm{pc}_{gg'}\ \text{significant}\right\},
$$ {#eq-indeed-as}

and ranks genes by $\mathrm{AS}_{g}$
[@zuoINDEEDIntegratedDifferential2016]. Relative to pure DGE, this
favours genes whose local **undirected** neighbourhood (Markov boundary
in the GGM sense; @sec-markov-seeds) differs between the focal type and
the rest. Related ideas appear in weighted graphical-lasso pipelines
with PPI priors and differential-network scores (DNS).

### Sparse GGMs and cooperative penalties {#sec-sparse-ggm}

Cell-type precision matrices
$\boldsymbol{\Omega}_{j}=\boldsymbol{\Sigma}_{j}^{-1}$ estimated by
graphical lasso are shrunk; non-zero partial correlations are typically
underestimated. A practical mitigation is to take the gLasso
**support** as topological constraints and re-estimate by constrained
MLE when the graph is chordal (otherwise the mapping is non-trivial)
[@dahlMaximumLikelihoodEstimation2005]. Cooperative / sign-coherent
group lasso penalties can encode pathway modules and PPI-informed edge
weights [@chiquetSparsitySigncoherentGroups2012]. PPI priors also
shrink the combinatorial space of undirected graphs explored per cell
type.

::: {#nte-network-candidates .callout-warning}
## :warning: From one-versus-others networks to a panel

Network-based one-versus-others lists still concatenate into a
redundant design (see @nte-ovr-concat). Treat them as **candidate
generators** for the all-versus-all stage in @sec-multi-all-vs-all.
:::

## All versus all {#sec-multi-all-vs-all}

### Variance-weighted condition number {#sec-weighted-kappa}

With gene-specific uncertainties $\tau_{g}^{2}$ (biological + sampling
+ optional sc–bulk discordance), standardise rows and evaluate
conditioning in the $(J-1)$-dimensional contrast space induced by
$\sum_{j}p_{j}=1$:

$$
\widetilde{\mu}_{gj}
=
\frac{\mu_{gj}-\bar{\mu}_{g}}{\tau_{g}},
\qquad
\kappa\!\left(
\boldsymbol{W}_{\mathcal{G}}^{1/2}\boldsymbol{\mu}_{\mathcal{G}}\boldsymbol{C}
\right),
$$ {#eq-weighted-kappa}

where $\boldsymbol{W}_{\mathcal{G}}=\mathrm{diag}(1/\tau_{g}^{2})$ and
$\boldsymbol{C}$ removes the common-expression direction. Unweighted
$\kappa(\boldsymbol{\mu}_{\mathcal{G}})$ alone can favour quiet but
biased genes [@gongDeconRNASeqStatisticalFramework2013]; see also
@sec-kappa-limits.

### D- / E- / A-optimal design {#sec-optimal-design}

Under approximately Gaussian errors
$\boldsymbol{y}_{\mathcal{G}}\mid\boldsymbol{p}
\sim\mathcal{N}(\boldsymbol{\mu}_{\mathcal{G}}\boldsymbol{p},
\boldsymbol{\Sigma}_{\mathcal{G}})$, the **Fisher information** for the
proportions (after removing the simplex null direction) is the
$J\times J$ Gram matrix of the whitened means

$$
\mathcal{I}_{\mathcal{G}}
=
\boldsymbol{\mu}_{\mathcal{G}}^{\top}
\boldsymbol{\Sigma}_{\mathcal{G}}^{-1}
\boldsymbol{\mu}_{\mathcal{G}}
\in\mathbb{R}^{J\times J}.
$$ {#eq-fisher}

Here $\boldsymbol{\Sigma}_{\mathcal{G}}$ is the bulk (or
pseudobulk) covariance on genes $\mathcal{G}$—diagonal or shrinkage
estimators are typical in practice. Large eigenvalues of
$\mathcal{I}_{\mathcal{G}}$ correspond to well-informed contrasts of
$\boldsymbol{p}$; a vanishing $\lambda_{\min}(\mathcal{I}_{\mathcal{G}})$
means at least one cell-type contrast is unidentifiable from the panel.

Classical alphabetical criteria
[@atkinsonOptimumExperimentalDesigns1992;
@pukelsheimOptimalDesignExperiments2006] are

$$
\max_{\mathcal{G}}\log\det(\mathcal{I}_{\mathcal{G}}+\lambda\boldsymbol{I})
\quad\text{(D)},
\qquad
\max_{\mathcal{G}}\lambda_{\min}(\mathcal{I}_{\mathcal{G}})
\quad\text{(E)},
\qquad
\min_{\mathcal{G}}\operatorname{tr}(\mathcal{I}_{\mathcal{G}}^{-1})
\quad\text{(A)}.
$$ {#eq-optimal-design}

@fig-dea-design recalls the three criteria geometrically.

![D-, E- and A-optimal design for gene-panel selection: shared Fisher
information $\mathcal{I}_{\mathcal{G}}$, confidence-ellipse intuition,
and links to condition number / Gram volume.](figures/dea_optimal_design_comparison.png){#fig-dea-design}

::: {#nte-dea-proscons .callout-note}
## :information_source: Pros and cons of D / E / A

- **D-optimality** ($\log\det$): shrinks the volume of the asymptotic
  confidence ellipsoid for $\boldsymbol{p}$. Efficient in an average
  sense; can still leave one weak contrast if other directions improve.
- **E-optimality** ($\lambda_{\min}$): protects the **worst** contrast.
  Closest in spirit to minimising
  $\kappa(\mathcal{I}_{\mathcal{G}})$ or maximising
  $\lambda_{\min}/\lambda_{\max}$. Conservative when several contrasts
  are already strong.
- **A-optimality** ($\operatorname{tr}\mathcal{I}^{-1}$): minimises
  average parameter variance (MSE under the Gaussian approximation).
  Interpretable, but less geometric than D/E.

In practice prefer shrinkage $\boldsymbol{\Sigma}_{\mathcal{G}}$ and a
ridge $\lambda\boldsymbol{I}$ for D when $|\mathcal{G}|$ is moderate and
$J$ is not tiny.
:::

**Relation to $\kappa$ and the Gram volume.** If one pretends
$\boldsymbol{\Sigma}_{\mathcal{G}}=\boldsymbol{I}$, then
$\mathcal{I}_{\mathcal{G}}=\boldsymbol{\mu}_{\mathcal{G}}^{\top}\boldsymbol{\mu}_{\mathcal{G}}$
and D-optimality maximises $\log\det(\boldsymbol{\mu}_{\mathcal{G}}^{\top}\boldsymbol{\mu}_{\mathcal{G}})$,
i.e. the log **Gram volume** of the mean columns. E-optimality maximises
$\lambda_{\min}$ of that Gram matrix and is therefore a direct cousin of
minimising $\kappa_2(\boldsymbol{\mu}_{\mathcal{G}})$
(see @eq-condition-number). The
[synthetic scenarios](synthetic-scenarios.html#sec-geometric-score)
vignette discusses the same objects as AutoGeneS-style scalar
summaries of $\boldsymbol{\mu}$ (Gram determinant $V(\boldsymbol{\mu})$
and reciprocal condition number $\kappa_2(\boldsymbol{\mu})^{-1}$).
Weighted / information-optimal criteria here extend those mean-only
scores by inserting $\boldsymbol{\Sigma}_{\mathcal{G}}^{-1}$.

**Relation to overlap and Wasserstein geometry.** The MixSim-style
**average overlap** $\overline{\Omega}$ (see @eq-overlap) is a
*probabilistic* separation score for the purified Gaussians
$\mathcal{N}(\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j})$: it answers
how often a MAP rule confuses type $j$ with $\ell$. D/E/A instead
optimise **estimation** of $\boldsymbol{p}$ under a single bulk
covariance. They need not agree: a panel can have small confidence
ellipsoids for $\boldsymbol{p}$ yet still high pairwise overlap if
priors $p_{j}$ or class-conditional covariances differ. A related
geometric alternative is the **2-Wasserstein** distance between
purified Gaussians,

$$
W_{2}^{2}\bigl(
\mathcal{N}(\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j}),
\mathcal{N}(\boldsymbol{\mu}_{.\ell},\boldsymbol{\Sigma}_{\ell})
\bigr)
=
\lVert\boldsymbol{\mu}_{.j}-\boldsymbol{\mu}_{.\ell}\rVert_{2}^{2}
+
\mathfrak{B}^{2}(\boldsymbol{\Sigma}_{j},\boldsymbol{\Sigma}_{\ell}),
$$ {#eq-wasserstein}

with $\mathfrak{B}^{2}$ the Bures metric on covariances. $W_{2}$ (and
overlap) treat each purified law symmetrically; Fisher information
treats the mixture regression for $\boldsymbol{p}$. Use D/E/A for
panel design under @eq-mixture, and overlap / $W_{2}$ as
interpretable monitors of class separation
[@R-MixSim; @alieeAutoGeneSAutomaticGene2021].

### Global overlap (DeCovarT proposal) {#sec-overlap}

AutoGeneS optimises correlation and centroid distance separately
[@alieeAutoGeneSAutomaticGene2021]. We instead minimise the **average
pairwise overlap** of the purified Gaussians, which folds mean
separation **and** covariance structure into one interpretable
probability: the chance of mis-assigning a draw under a MAP rule. With
MixSim-style pairwise misclassification masses $\Omega_{jl}$
[@R-MixSim],

$$
\overline{\Omega}
=
\frac{2}{J(J-1)}
\sum_{1\le j<\ell\le J}
\bigl(p_{j}\Omega_{j\ell}+p_{\ell}\Omega_{\ell j}\bigr).
$$ {#eq-overlap}

Package helper `compute_average_overlap()` implements @eq-overlap.
Lower $\overline{\Omega}$ corresponds to better-separated purified
profiles and, empirically, to more accurate recovery of
$\boldsymbol{p}$.

# Wrapper optimisation on pseudo-bulks and matched bulks {#sec-wrapper}

When matched sc/bulk aliquots exist (e.g. Kassandra-style benchmarks)
[@zaitsevPreciseReconstructionTME2022], evaluate candidate panels by
running the **intended** deconvolution algorithm on held-out mixtures
and scoring RMSE / worst-type error / rare-cell recall under nested
**donor-level** cross-validation. Use the one-standard-error rule to
prefer the smallest panel statistically indistinguishable from the
best. Wrappers overfit the mixture simulator if real matched bulks are
omitted
[@sturmComprehensiveEvaluationTranscriptomebased2019;
@jinBenchmarkRNAseqDeconvolution2021].

```{r}
#| label: tbl-strategy-summary
#| tbl-cap: "Compact map from selection strategy to typical role in a reference-based pipeline."

strategies <- data.frame(
  Stage = c(
    "Univariate 1 vs others",
    "Univariate all vs all",
    "Multivariate 1 vs others",
    "Multivariate all vs all",
    "Wrapper"
  ),
  Examples = c(
    "DE / F-test / Gini entropy",
    "AutoGeneS",
    "INDEED; sparse GGM",
    "Weighted <U+03BA>; D/E/A-opt; overlap",
    "Nested CV on matched bulks"
  ),
  Role = c(
    "Candidate generation",
    "Panel compaction",
    "Network-aware candidates",
    "Identifiability of p",
    "Empirical size selection"
  )
)

knitr::kable(strategies)
```

# Recommended pipeline {#sec-recommended}

- ① **Reference moments** (see @sec-preprocess). Donor-level
  pseudobulks; estimate $\boldsymbol{\mu}$ and variance components on a
  linear scale; optional SCTransform / top-$2{,}000$ HVGs
  (see @sec-sctransform) to define the gene universe.
- ② **Reliability filter.** Drop unstable / undetectable / discordant
  genes; optionally inflate $\tau_{g}^{2}$ by observed sc–bulk
  discrepancy.
- ③ **Candidates.** Union of stable one-versus-others DE
  (see @sec-univ-one-vs-others), high Gini / low entropy
  (see @sec-gini-entropy), and (optionally) differential-network hits
  (see @sec-indeed) → roughly $10^{2}$–$10^{3}$ genes.
- ④ **Information-optimal subset.** Greedy D- or E-optimal selection
  (see @eq-optimal-design) or weighted $\kappa$ (see @eq-weighted-kappa), with
  overlap (see @eq-overlap) as a secondary monitor; heed
  @sec-kappa-limits on continua.
- ⑤ **Wrapper prune** (see @sec-wrapper). Evaluate panel sizes along the
  path under nested donor CV and matched bulks; keep the smallest
  competitive panel.

The distinctive contribution for DeCovarT-style second-generation
models is a **bulk-calibrated, variance-weighted information criterion
plus overlap**, exploiting paired sc/bulk designs when available.

# Perspectives {#sec-perspectives}

## Mechanistic priors {#sec-mechanistic}

PPI / TF networks and pathway modules can regularise both GGM
estimation and cooperative penalties
[@chiquetSparsitySigncoherentGroups2012;
@zuoINDEEDIntegratedDifferential2016], shrinking the search toward
biologically plausible supports rather than purely statistical markers.

## Continuous states, potency and archetypes {#sec-continua}

Many tissues are better described by **continua** (differentiation
trajectories, activation gradients) than by discrete labels. In that
regime, a finite $J$-column signature is a piecewise approximation of
an archetype geometry [@hartInferringBiologicalTasks2015] or a potency
axis [@gulatiSinglecellTranscriptionalDiversity2020]: genes should
separate extreme vertices **and** resolve interior mixtures along
trajectories. Open directions include replacing hard one-versus-others
contrasts with scores along inferred latent axes, and evaluating
overlap / Fisher information on continuous rather than categorical
components. Until such methods mature, hierarchical discrete panels
(lineage → subtype) remain a pragmatic compromise
[@wangBulkTissueCell2019]. See also @sec-kappa-limits.

## Seeds and undirected Markov boundaries {#sec-markov-seeds}

A complementary panel-expansion strategy starts from a small set of
**seeds**—genes that are already highly discriminatory for
$\boldsymbol{p}$ under @eq-mixture or @eq-indeed-as—then grows each
seed by its **Markov boundary** in a cell-type (or differential) GGM.

In an undirected graphical model, a Markov blanket of gene $g$ is any
set $S$ such that $g$ is conditionally independent of all other genes
given $S$; a **Markov boundary** is a minimal such set
[@pearlProbabilisticReasoningIntelligent1988]. For a Gaussian Markov
random field / GGM, the Markov boundary of $g$ coincides with its
graph neighbours: the indices $g'$ with $\omega_{gg'}\neq 0$ in
$\boldsymbol{\Omega}$
[@bolinExplicitLinkGraphical2026]. Expanding seeds by these
neighbours therefore adds genes that are **not** redundant once the
seed is conditioned on—they carry residual information about the
local dependency structure.

::: {#nte-u-vs-dsep .callout-important}
## :warning: Undirected ($U$) separation, not $d$-separation

This construction uses the **undirected** Markov property (conditioning
on neighbours in $\boldsymbol{\Omega}$). Classical $d$-separation
applies to **directed** acyclic Gaussian Bayesian networks and is the
wrong language for GGM / glasso supports. Do not treat directed path
blocking rules as interchangeable with undirected Markov boundaries
[@pearlProbabilisticReasoningIntelligent1988;
@bolinExplicitLinkGraphical2026].
:::

Operational sketch:

- ① Select seeds by mean separation, INDEED activity (see @eq-indeed-as),
  or high contribution to $\mathcal{I}_{\mathcal{G}}$
  (see @eq-fisher).
- ② Estimate $\boldsymbol{\Omega}_{j}$ (or a differential precision) on
  the HVG universe (see @sec-sctransform).
- ③ For each seed $g$, add its undirected neighbours
  $\{g':\omega_{gg'}\neq 0\}$ (the Markov boundary).
- ④ Re-run all-versus-all compaction (see @sec-optimal-design,
  @sec-overlap) so that boundary expansion does not reintroduce
  mean-collinear bulk (see @nte-ovr-concat).

# Software notes {#sec-software}

- Variance stabilisation / Pearson residuals: `sctransform` /
  `Seurat::SCTransform`
  [@hafemeisterNormalizationVarianceStabilization2019].
- Signature QC / compaction patterns: DeconvExplorer
  (`removePercentZeros`, `removeUnspecificGenes`,
  `selectGenesByScore` with Gini or entropy).
- Gold-standard reference-based estimators for downstream validation:
  CIBERSORT / CIBERSORTx, EPIC, quanTIseq, ABIS, MuSiC, DWLS
  [@newmanRobustEnumerationCell2015; @newmanDeterminingCellType2019;
  @racleSimultaneousEnumerationCancer2017; @finotello_etal19;
  @monacoRNASeqSignaturesNormalized2019; @wangBulkTissueCell2019;
  @tsoucasAccurateEstimationCelltype2019].
- Optimal-design background:
  [@atkinsonOptimumExperimentalDesigns1992;
  @pukelsheimOptimalDesignExperiments2006].
