---
title: "Feature selection for reference-based deconvolution"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Feature selection for reference-based deconvolution}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
    number-sections: true
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  eq-title: "Equation"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)
```

# Scope

This vignette surveys **feature selection for reference-based
(signature-based) deconvolution**—methods that recover cellular
proportions $\boldsymbol{p}$ from a bulk mixture using a purified
expression design matrix
[@gaspard-boulincCelltypeDeconvolutionMethods2025;
@avilacobosComputationalDeconvolutionTranscriptomics2018;
@sturmComprehensiveEvaluationTranscriptomebased2019]. Feature selection
itself is a classical pre-modelling filter
[@guyonIntroductionVariableFeature2003]. We deliberately exclude
**marker-based** abundance scoring (enrichment / GSEA-style modules),
which returns proxies of presence rather than a consensual gene panel
suited to linear or probabilistic mixture models
[@aranXCellDigitallyPortraying2017; @zhang_etal17].

The focus is **second-generation** signatures built from single-cell
pseudobulks (CIBERSORTx, MuSiC, DWLS, Kassandra and related pipelines)
[@newmanDeterminingCellType2019; @wangBulkTissueCell2019;
@tsoucasAccurateEstimationCelltype2019;
@zaitsevPreciseReconstructionTME2022], where the selected genes must
discriminate **all** cell types in the signature jointly, not only
one-versus-rest markers. Correlation-aware compact panels such as
DUBStepR illustrate the same goal from a single-cell feature-selection
angle [@ranjanDUBStepRScalableCorrelationbased2021].

Interactive signature refinement utilities (zero filtering,
specificity binning, Gini / entropy compaction) are implemented in
[DeconvExplorer](https://github.com/omnideconv/deconvExplorer)
(omnideconv ecosystem; cite as software once added to
`packages.bib`).

```{mermaid}
%%| label: fig-feature-selection-layout
%%| fig-cap: "Layout of this vignette: univariate and multivariate selection, each split into one-versus-others then all-versus-all, followed by perspectives for continuous cell states."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  A["Reference μ, Σ from sc / bulk"] --> B["Univariate"]
  A --> C["Multivariate"]
  B --> B1["One vs others"]
  B --> B2["All vs all"]
  C --> C1["One vs others\n(GGM / DiffNet)"]
  C --> C2["All vs all\n(κ, D/E-opt, overlap)"]
  B2 --> D["Consensual panel G"]
  C2 --> D
  D --> E["Perspectives:\nmechanistic / potency / archetypes"]
```

# Notation

We align notation with the DeCovarT generative model. Let
$G$ genes and $J$ cell types index the purified means
$\boldsymbol{\mu}=(\boldsymbol{\mu}_{.j})_{j=1}^{J}\in\mathbb{R}^{G\times J}$
and (optional) covariances
$\{\boldsymbol{\Sigma}_{j}\}_{j=1}^{J}$. A bulk observation obeys

$$
\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})
\sim
\mathcal{N}_{G}\!\left(
\boldsymbol{\mu}\boldsymbol{p},\,
\sum_{j=1}^{J}p_{j}^{2}\boldsymbol{\Sigma}_{j}
\right),
\quad
\boldsymbol{p}\in\Delta^{J-1},
$$ {#eq-mixture}

with $\boldsymbol{\zeta}=(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$
and simplex
$\Delta^{J-1}=\{\boldsymbol{p}:p_{j}\ge 0,\sum_{j}p_{j}=1\}$. For a
gene subset $\mathcal{G}\subseteq\{1,\ldots,G\}$, write
$\boldsymbol{\mu}_{\mathcal{G}}$ for the corresponding rows. The
condition number of a design matrix $\boldsymbol{X}$ is

$$
\kappa(\boldsymbol{X})
=
\frac{\lambda_{\max}(\boldsymbol{X})}{\lambda_{\min}(\boldsymbol{X})},
$$ {#eq-condition-number}

with large $\kappa$ indicating multicollinearity and numerical
instability [@gongDeconRNASeqStatisticalFramework2013;
@newmanRobustEnumerationCell2015].

# Preprocessing shared by all strategies

Before scoring genes:

1. Aggregate **donor-level** cell-type pseudobulks (not individual
   cells as independent replicates). Pseudobulking reduces sparsity and
   avoids treating cells as independent biological replicates in
   differential testing [@squairConfrontingFalseDiscoveries2021].
2. Estimate gene–cell-type means $\mu_{gj}$ and variance components on
   a **linear** scale (counts, CPM, or TPM). Log-space breaks the
   mixture linearity in @eq-mixture
   [@sturmComprehensiveEvaluationTranscriptomebased2019;
   @jinBenchmarkRNAseqDeconvolution2021;
   @hoffmannRobustComputationalReconstitution2006].
3. Filter genes with poor bulk detectability, extreme zero inflation
   after pseudobulking, strong donor instability, or systematic
   sc–bulk discordance.
4. When the biological target is **cell number** rather than RNA mass,
   correct transcriptome-size differences across cell types
   [@luTranscriptomeSizeMatters2025].
5. Apply the **same** normalisation to bulk and signature
   [@jinBenchmarkRNAseqDeconvolution2021; @racleSimultaneousEnumerationCancer2017].

# Univariate approaches

Univariate selectors score each gene marginally (or with a simple
contrast) and ignore partial correlations among genes.

## One versus others

Classical signature pipelines score gene $g$ for cell type $j$ against
the remaining types, then concatenate per-type shortlists
[@abbasDeconvolutionBloodMicroarray2009;
@newmanRobustEnumerationCell2015; @finotello_etal19].

### Differential expression and fold-change ranks

Limma–voom, edgeR and DESeq2 remain the workhorses for
one-versus-rest contrasts
[@ritchieLimmaPowersDifferential2015a;
@robinsonEdgeRBioconductorPackage2010;
@loveModeratedEstimationFold2014a]. CIBERSORT retains genes with
adjusted $p\le 0.3$ ordered by decreasing fold-change before a global
condition-number prune [@newmanRobustEnumerationCell2015]. Abbas et
al. keep genes that exceed the second- and third-ranked cell types,
then choose the panel with smallest $\kappa$
[@abbasDeconvolutionBloodMicroarray2009]. quanTIseq / xCell-style
filters further drop non-haematopoietic contaminants
[@finotello_etal19; @aranXCellDigitallyPortraying2017].

### ANOVA $F$-test

Wang et al. rank genes by the nested-model $F$ statistic comparing a
cell-type factor to an intercept-only model
[@wangSilicoEstimatesTissue2010]. For gene $g$ with expression
$x_{gi}$ in sample $i$ and cell-type labels $z_{i}\in\{1,\ldots,J\}$,

$$
F_{g}
=
\frac{
\bigl(\mathrm{SSR}_{0}-\mathrm{SSR}_{1}\bigr)/(J-1)
}{
\mathrm{SSR}_{1}/(n-J)
},
$$ {#eq-anova-f}

where $\mathrm{SSR}_{0}=\sum_{i}(x_{gi}-\bar{x}_{g})^{2}$ and
$\mathrm{SSR}_{1}=\sum_{j}\sum_{i:z_{i}=j}(x_{gi}-\bar{x}_{gj})^{2}$.
Large $F_{g}$ indicates that cell-type means explain a substantial
share of gene variance.

### Gini and entropy specificity

BioQC and DeconvExplorer compact signatures with Gini or entropy
scores over the row $\boldsymbol{\mu}_{g\cdot}$
[@zhang_etal17]. With relative shares
$q_{gj}=\mu_{gj}/\sum_{\ell}\mu_{g\ell}$,

$$
\mathrm{Gini}(g)
=
\frac{J}{J-1}
\left(
1-2\sum_{j=1}^{J}
\frac{J-j+\tfrac{1}{2}}{J}\,q_{g(j)}
\right),
\quad
H(g)
=
-\sum_{j=1}^{J}q_{gj}\log q_{gj},
$$ {#eq-gini-entropy}

where $q_{g(j)}$ are the ordered shares. High Gini (or low entropy)
flags genes concentrated in few cell types—useful for one-versus-rest
compaction, but prone to redundancy when panels are concatenated.

::: {.callout-note}
## Limitation of one-versus-rest concatenation
Union of per-type markers often yields a multicollinear
$\boldsymbol{\mu}_{\mathcal{G}}$ that is still too large for stable
NNLS / SVR / GLS [@newmanRobustEnumerationCell2015;
@avilacobosComputationalDeconvolutionTranscriptomics2018]. A global
all-versus-all refinement is therefore required.
:::

## All versus all

### Condition-number pruning

After candidate generation, regression-based methods select
$\mathcal{G}$ to minimise $\kappa(\boldsymbol{\mu}_{\mathcal{G}})$
(@eq-condition-number), optionally after projecting out the
sum-to-one direction
[@abbasDeconvolutionBloodMicroarray2009;
@gongDeconRNASeqStatisticalFramework2013;
@newmanRobustEnumerationCell2015;
@boldinaA2SignAgnosticAlgorithms2022]. This is the historical
gold-standard metric for LM22-style panels and remains the default in
many second-generation pipelines.

### Genetic algorithms (AutoGeneS)

AutoGeneS searches gene subsets with a multi-objective genetic
algorithm that **minimises inter-population correlation** while
**maximising centroid distance**
[@alieeAutoGeneSAutomaticGene2021]. Related multi-objective genetic
searches appear in network module discovery (e.g. MOGAMUN)
[@novoa-del-toroMultiobjectiveGeneticAlgorithm2021]. AutoGeneS is
explicitly all-versus-all: the loss is defined on the full $J$-column
signature rather than on independent one-versus-rest lists.

DeCovarT replaces that dual objective with a single **global overlap**
criterion on the Gaussian mixture
$\{(p_{j},\boldsymbol{\mu}_{.j},\boldsymbol{\Sigma}_{j})\}$ (see
@eq-overlap below and the synthetic-scenarios vignette).

# Multivariate approaches

Multivariate selectors use joint dependence—covariance, precision, or
information matrices—so that a gene is scored by how it improves
identifiability of $\boldsymbol{p}$ under @eq-mixture.

## One versus others

Univariate DGE ignores gene–gene interactions and therefore requires
aggressive multiple-testing correction without modelling network
rewiring. Differential-network and GGM methods address that gap.

### Differential networks (INDEED and related)

INDEED combines mean (fold-change) evidence with bootstrap tests of
changes in partial correlation, ranking genes that both shift in mean
**and** rewire neighbourhood structure
[@zuoINDEEDIntegratedDifferential2016]. Related ideas appear in
weighted graphical lasso pipelines that incorporate PPI priors and
differential-network scores (DNS). Relative to pure DGE, these methods
favour genes whose local Markov blanket differs between a focal cell
type and the rest.

### Sparse GGMs and cooperative penalties

Cell-type precision matrices $\boldsymbol{\Omega}_{j}=\boldsymbol{\Sigma}_{j}^{-1}$
estimated by graphical lasso are shrunk; non-zero partial correlations
are typically underestimated. A practical mitigation is to take the
gLasso **support** as topological constraints and re-estimate by
constrained MLE when the graph is chordal (otherwise the mapping is
non-trivial)
[@dahlMaximumLikelihoodEstimation2005]. Cooperative / sign-coherent
group lasso penalties can encode pathway modules and PPI-informed edge
weights [@chiquetSparsitySigncoherentGroups2012]. PPI priors also
shrink the combinatorial space of undirected graphs explored per cell
type.

::: {.callout-warning}
## From one-versus-others networks to a panel
Network-based one-versus-others lists still concatenate into a
redundant design. They are best treated as **candidate generators** for
the all-versus-all stage below.
:::

## All versus all

### Variance-weighted condition number

With gene-specific uncertainties $\tau_{g}^{2}$ (biological + sampling +
optional sc–bulk discordance), standardise rows and evaluate
conditioning in the $(J-1)$-dimensional contrast space induced by
$\sum_{j}p_{j}=1$:

$$
\widetilde{\mu}_{gj}
=
\frac{\mu_{gj}-\bar{\mu}_{g}}{\tau_{g}},
\qquad
\kappa\!\left(
\boldsymbol{W}_{\mathcal{G}}^{1/2}\boldsymbol{\mu}_{\mathcal{G}}\boldsymbol{C}
\right),
$$ {#eq-weighted-kappa}

where
$\boldsymbol{W}_{\mathcal{G}}=\mathrm{diag}(1/\tau_{g}^{2})$ and
$\boldsymbol{C}$ removes the common-expression direction. Unweighted
$\kappa(\boldsymbol{\mu}_{\mathcal{G}})$ alone can favour quiet but
biased genes [@gongDeconRNASeqStatisticalFramework2013].

### D- / E- / A-optimal design

Under approximately Gaussian errors
$\boldsymbol{y}_{\mathcal{G}}\mid\boldsymbol{p}\sim\mathcal{N}(\boldsymbol{\mu}_{\mathcal{G}}\boldsymbol{p},\boldsymbol{\Sigma}_{\mathcal{G}})$,
the information for $\boldsymbol{p}$ (after the simplex constraint) is

$$
\mathcal{I}_{\mathcal{G}}
=
\boldsymbol{\mu}_{\mathcal{G}}^{\top}
\boldsymbol{\Sigma}_{\mathcal{G}}^{-1}
\boldsymbol{\mu}_{\mathcal{G}}.
$$ {#eq-fisher}

Common criteria are

$$
\max_{\mathcal{G}}\log\det(\mathcal{I}_{\mathcal{G}}+\lambda\boldsymbol{I})
\quad\text{(D)},
\qquad
\max_{\mathcal{G}}\lambda_{\min}(\mathcal{I}_{\mathcal{G}})
\quad\text{(E)},
\qquad
\min_{\mathcal{G}}\operatorname{tr}(\mathcal{I}_{\mathcal{G}}^{-1})
\quad\text{(A)}.
$$ {#eq-optimal-design}

E-optimality is closely related to protecting the hardest cell-type
contrast; unweighted condition-number minimisation is a rough E-proxy
without an explicit noise model. In practice use diagonal or shrinkage
$\boldsymbol{\Sigma}_{\mathcal{G}}$.

### Pairwise Gaussian separation with redundancy control

A cheap multivariate screen scores each gene by standardised pairwise
separation

$$
d_{g,jl}^{2}
=
\frac{(\mu_{gj}-\mu_{gl})^{2}}
{\sigma_{gj}^{2}+\sigma_{gl}^{2}+\varepsilon}
$$ {#eq-pairwise-sep}

and solves a set-cover-like objective with diminishing returns and a
redundancy penalty. Recursive / hierarchical selection (broad lineage
then subtypes) mirrors MuSiC’s motivation for correlated cell types
[@wangBulkTissueCell2019]. Follow with @eq-fisher or @eq-weighted-kappa
for the final panel.

### Global overlap (DeCovarT proposal)

AutoGeneS optimises correlation and centroid distance separately
[@alieeAutoGeneSAutomaticGene2021]. We instead minimise the **average
pairwise overlap** of the purified Gaussians, which folds mean
separation **and** covariance structure into one interpretable
probability: the chance of mis-assigning a draw under a MAP rule.
With MixSim-style pairwise misclassification masses
$\Omega_{jl}$ [@R-MixSim],

$$
\overline{\Omega}
=
\frac{2}{J(J-1)}
\sum_{1\le j<l\le J}
\bigl(p_{j}\Omega_{jl}+p_{l}\Omega_{lj}\bigr).
$$ {#eq-overlap}

Package helper `compute_average_overlap()` implements @eq-overlap.
Lower $\overline{\Omega}$ corresponds to better-separated purified
profiles and, empirically, to more accurate recovery of
$\boldsymbol{p}$.

### Wrapper optimisation on pseudo-bulks and matched bulks

When matched sc/bulk aliquots exist (e.g. Kassandra-style benchmarks)
[@zaitsevPreciseReconstructionTME2022], evaluate candidate panels by
running the **intended** deconvolution algorithm on held-out mixtures
and scoring RMSE / worst-type error / rare-cell recall under nested
**donor-level** cross-validation. Use the one-standard-error rule to
prefer the smallest panel statistically indistinguishable from the
best. Wrappers overfit the mixture simulator if real matched bulks are
omitted [@sturmComprehensiveEvaluationTranscriptomebased2019;
@jinBenchmarkRNAseqDeconvolution2021].

```{r}
#| label: tbl-strategy-summary
#| tbl-cap: "Compact map from selection strategy to typical role in a reference-based pipeline."

strategies <- data.frame(
  Stage = c(
    "Univariate · 1 vs others",
    "Univariate · all vs all",
    "Multivariate · 1 vs others",
    "Multivariate · all vs all",
    "Wrapper"
  ),
  Examples = c(
    "DE / F-test / Gini–entropy",
    "κ pruning; AutoGeneS",
    "INDEED; sparse GGM",
    "Weighted κ; D/E-opt; overlap",
    "Nested CV on matched bulks"
  ),
  Role = c(
    "Candidate generation",
    "Panel compaction",
    "Network-aware candidates",
    "Identifiability of p",
    "Empirical size selection"
  )
)

knitr::kable(strategies)
```

# Recommended pipeline

1. **Reference moments.** Donor-level pseudobulks; estimate
   $\boldsymbol{\mu}$ and variance components on a linear scale.
2. **Reliability filter.** Drop unstable / undetectable / discordant
   genes; optionally inflate $\tau_{g}^{2}$ by observed sc–bulk
   discrepancy.
3. **Candidates.** Union of stable one-versus-others DE, high
   $d_{g,jl}^{2}$, and (optionally) differential-network hits → roughly
   $10^{2}$–$10^{3}$ genes.
4. **Information-optimal subset.** Greedy D- or E-optimal selection
   (@eq-optimal-design) or weighted $\kappa$ (@eq-weighted-kappa), with
   overlap (@eq-overlap) as a secondary monitor.
5. **Wrapper prune.** Evaluate panel sizes along the path under nested
   donor CV and matched bulks; keep the smallest competitive panel.

The distinctive contribution for DeCovarT-style second-generation
models is a **bulk-calibrated, variance-weighted information criterion
plus overlap**, exploiting paired sc/bulk designs when available.

# Perspectives

## Mechanistic priors

PPI / TF networks and pathway modules can regularise both GGM
estimation and cooperative penalties
[@chiquetSparsitySigncoherentGroups2012;
@zuoINDEEDIntegratedDifferential2016], shrinking the search toward
biologically plausible supports rather than purely statistical markers.

## Continuous states, potency and archetypes

Many tissues are better described by **continua** (differentiation
trajectories, activation gradients) than by discrete labels. In that
regime, a finite $J$-column signature is a piecewise approximation of
an archetype geometry [@hartInferringBiologicalTasks2015] or a potency
axis [@gulatiSinglecellTranscriptionalDiversity2020]: genes should
separate extreme vertices **and** resolve interior mixtures along
trajectories. Open directions include replacing hard one-versus-others
contrasts with scores along inferred latent axes, and evaluating
overlap / Fisher information on continuous rather than categorical
components. Until such methods mature, hierarchical discrete panels
(lineage → subtype) remain a pragmatic compromise
[@wangBulkTissueCell2019].

# Software notes

- DeCovarT helpers: `generate_mean_signature_matrix()`,
  `compute_average_overlap()`, `simulate_hierarchical_grn_moments()`.
- Signature QC / compaction patterns: DeconvExplorer
  (`removePercentZeros`, `removeUnspecificGenes`,
  `selectGenesByScore` with Gini or entropy).
- Gold-standard reference-based estimators for downstream validation:
  CIBERSORT / CIBERSORTx, EPIC, quanTIseq, ABIS, MuSiC, DWLS
  [@newmanRobustEnumerationCell2015; @newmanDeterminingCellType2019;
  @racleSimultaneousEnumerationCancer2017; @finotello_etal19;
  @monacoRNASeqSignaturesNormalized2019; @wangBulkTissueCell2019;
  @tsoucasAccurateEstimationCelltype2019].
