---
title: "§2.2 Covariance-driven scenario (G = 20, J = 3)"
subtitle: "Full script: `scripts/fig03_variance_driven.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{§2.2 Covariance-driven scenario}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  fig-prefix: "Fig."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
---

```{r}
#| label: setup
#| eval: false
library(DeCovarT)
```

The bivariate toy confirms the theoretical benefit of covariance
modelling in a transparent setting but cannot demonstrate realistic
gene-regulatory network (GRN) topology. This scenario uses $G = 20$
genes and $J = 3$ cell types in which types 1 and 2 are **mean-collinear**
($\cos(\boldsymbol{\mu}_{\cdot 1},\boldsymbol{\mu}_{\cdot 2})=0.9$): the
network topology encoded in $\boldsymbol{\Sigma}_j$ is the remaining
discriminative signal. Construction, ADEMP benchmark, and the static
network figure live in a **single** script,
`scripts/fig03_variance_driven.R` (seed `20260807`).

> **Script:** from the repository root,
> `mkdir -p logs && nohup Rscript --no-save --no-restore scripts/fig03_variance_driven.R > "logs/fig03_$(date +%F)_covariance_driven.log" 2>&1 &`
> (full, $n = 50$).
> Figures: `output/fig03/`. ADEMP reporting:
> [how to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html#sec-ademp).
> Scenario geometry:
> [scenario descriptors](theory-synthetic-scenarios-mean-covariance.html#sec-scenario-descriptors).

```{mermaid}
%%| label: fig-covariance-design
%%| eval: true
%%| fig-cap: "Covariance-driven factorial design: three Shannon compositions, a 2-by-2 graph assignment on the mean-collinear cell types (scale-free versus cluster SBM), type 3 held at scale-free, three precision cushions, and five solvers (4 x 3 x 3 = 36 scenarios)."
flowchart TD
  P["3 compositions<br/>H* = 1 / 0.5 / 0.1"] --> S["36 scenarios"]
  G["2 x 2 graph assignments<br/>CT1, CT2: scale-free / cluster SBM"] --> S
  C["CT3 graph held at scale-free"] --> S
  K["3 precision cushions u<br/>well 0.5 / moderate 0.2 / ill 0.1"] --> S
  S --> A["5 deconvolution solvers"]
```

# Generative model {#sec-scenario-grid}

One fixed mean signature $\boldsymbol{\mu}\in\mathbb{R}^{G\times J}$ is
built with `generate_mean_signature_matrix()` and an explicit Gram
`target_gram` ([mean signature
construction](theory-synthetic-scenarios-mean-covariance.html#sec-lr-block-mu)).
There is no marker block and no null (`equal_all`) block: uninformative
genes would be removed before deconvolution.

$$
R
=
\begin{pmatrix}
1 & 0.9 & 0.1 \\
0.9 & 1 & 0.1 \\
0.1 & 0.1 & 1
\end{pmatrix},
\qquad
\boldsymbol{\mu}
=
s\mathbf{Q}R^{1/2},
\quad s=10.
$$ {#eq-fig03-gram}

Because $\mathbf{Q}^{\mathsf{T}}\mathbf{Q}=\mathbf{I}_J$, the pairwise
cosines of the columns of $\boldsymbol{\mu}$ equal $R$ exactly (up to
rounding). Euclidean gaps follow
$\lVert\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\rVert
=s\sqrt{2(1-R_{jk})}$. The call uses `nonnegative = TRUE` (disjoint
gene-block frame) so that $\boldsymbol{\mu}\ge 0$, as required by
`deconvolute_ratios()`.

| pair | cosine | Euclidean |
|:-----|-------:|----------:|
| celltype_1 vs celltype_2 | 0.9 | $10\sqrt{0.2}\approx 4.472$ |
| celltype_1 vs celltype_3 | 0.1 | $10\sqrt{1.8}\approx 13.416$ |
| celltype_2 vs celltype_3 | 0.1 | $10\sqrt{1.8}\approx 13.416$ |

: Exact pairwise geometry of the $20\times 3$ mean signature (`target_gram`, `mean_scale` $=10$). {#tbl-mean-geometry}

Each of the mean-collinear types (CT1, CT2) independently draws an
undirected skeleton from one of two generators, then completes a signed
precision with
`build_normalised_precision()`
([GGM networks](theory-synthetic-scenarios-mean-covariance.html#sec-ggm-networks);
topology families in
[the topology table](theory-synthetic-scenarios-mean-covariance.html#tbl-topologies)):

- **Scale-free** (`graph_model = "scale_free"`): Barabási–Albert
  preferential attachment (`power = 1`, `edges_per_node = 1`).
- **Cluster / stochastic block model**
  (`graph_model = "stochastic_block_model"`): three blocks with
  probabilities $(0.5,0.25,0.25)$, $p_{\mathrm{within}}=0.25$,
  $p_{\mathrm{between}}=0.01$.

Erdős–Rényi is not used. Type 3, already separated by the Gram
($R_{13}=R_{23}=0.1$), is held at scale-free, so the topology factorial
is $2^{2}=4$ assignments on CT1 and CT2. Those generator names and
parameter lists are stored on every row of `hybrid_config.rds` as
`graph_ct1` / `graph_ct2` / `graph_ct3` and the list-column
`graph_generators`, and again inside each `true_theta`.

The spectral cushion $u$ (`precision_shift`) sets
$\lambda_{\min}(\boldsymbol{\Omega}_j)=u$ after the shift, so smaller $u$
worsens $\kappa(\boldsymbol{\Omega}_j)=\lambda_{\max}/u$ and, after
inversion and convolution, $\kappa\{\boldsymbol{\Sigma}(\boldsymbol{p})\}$.

| Factor | Levels |
|---|---|
| Graph model (CT1, CT2) | scale-free; cluster SBM |
| Graph model (CT3) | scale-free (fixed) |
| Precision shift $u$ | well ($0.5$); moderate ($0.2$); ill ($0.1$) |
| Topology assignments | $2^{2}=\mathbf{4}$ |
| Covariance draws | $4\times 3=\mathbf{12}$ |

: Covariance factorial (graphs $\times$ precision cushion). {#tbl-cov-factorial}

`describe_simulation_scenario()` records the realised
$\kappa\{\boldsymbol{\Sigma}(\boldsymbol{p})\}$ and the covariance-information
fraction $f_{\mathrm{cov}}$ on every row (see the
[descriptors section](theory-synthetic-scenarios-mean-covariance.html#sec-scenario-descriptors),
especially the tangent-Fisher callout). Those columns are
**measured**, not additional design factors: $u$ is the knob;
$f_{\mathrm{cov}}$ reports how much of the Fisher information is
second-order.

![Two generator examples on $G=20$ nodes (scale-free, cluster SBM).
Mean-collinear cell types 1 and 2 are each assigned one of these
families; type 3 is held at scale-free.](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

Two generators were considered but not adopted:

- **Toeplitz covariances**: encode only local (banded) dependencies;
  cannot model cell-type-specific sparse precision.
- **`MixSim` overlap** [@melnykovMixSimPackageSimulating2012]: designed
  for isotropic or full-rank Gaussian clusters; no cell-type-specific
  precision input.

# Inference {#sec-hybrid-deconvolution}

Compositions are one-dominant vectors from
`composition_from_entropy()` targeting normalised Shannon entropy
$H^{\star}\in\{1,0.5,0.1\}$ (balanced; moderately unbalanced; highly
unbalanced). All DeCovarT solvers start at the barycentre
(`initial_p = "barycentre"`). Other initialisation strategies are
reserved for separate scenarios.

| Factor | Levels |
|---|---|
| Composition $\boldsymbol{p}$ | $H^{\star}=1$; $H^{\star}=0.5$; $H^{\star}=0.1$ |
| Algorithms | DeconRNASeq (`lsei`), CIBERSORT, L-BFGS-B, Newton–Raphson, Marquardt–Levenberg |
| **Total scenarios** | $4\times 3\times 3=\mathbf{36}$ |

$N = 50$ Monte Carlo replicates per scenario.

| Algorithm | Function | Constraint |
|:----------|:---------|:-----------|
| QP (`DeconRNASeq`-style) | `deconvolute_ratios_deconrnaseq` | simplex equality / inequality |
| CIBERSORT-style $\nu$-SVR | `deconvolute_ratios_cibersort` | non-negative then `repair_simplex()` |
| L-BFGS-B | `deconvolute_ratios_L_BFGS_B` | ILR $\to$ open simplex |
| Newton–Raphson | `deconvolute_ratios_Newton_Raphson` | ILR / `nlminb` |
| Marquardt–Levenberg | `deconvolute_ratios_Marquardt_Levenberg` | ILR $\to$ open simplex |

: Solvers used in the covariance-driven comparison. {#tbl-solver-hybrid}

Mean-only baselines (LSEI, CIBERSORT) ignore $\boldsymbol{\Sigma}_j$.
The three DeCovarT maps use the same convolution likelihood. The shipped
catalogue (robust linear model, GLS, BFGS, simulated annealing) is
listed in
[`deconvolute_ratios()`](../reference/deconvolute_ratios.html).

```{r}
#| label: inference
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "50"))
ALGORITHMS <- c(
  "lsei",
  "cibersort",
  "LBFGS",
  "Newton-Raphson",
  "Marquardt-Levenberg"
)
# Full pipeline: Rscript scripts/fig03_variance_driven.R
```

# Visualisations {#sec-hybrid-viz}

| Output | Description |
|---|---|
| `output/fig03/fig03_raincloud.pdf` | Raincloud of Monte Carlo errors, faceted by composition and $\kappa$ label |
| `output/fig03/fig03_forest.pdf` | ADEMP forest plot |
| `output/fig03/fig03_metric_dots.pdf` | Metric dot plot (RMSE, MAE, coverage) |

### Expected findings

Because CT 1 and CT 2 share cosine $0.9$, LSEI and CIBERSORT cannot
separate them from the mean signature alone. DeCovarT (L-BFGS-B,
Newton–Raphson, Marquardt–Levenberg) uses $\boldsymbol{\Sigma}_j$; its
advantage should grow where $f_{\mathrm{cov}}$ is large, where $u$ is
small (ill-conditioned but still informative second-order structure),
and where the composition is highly unbalanced ($H^{\star}=0.1$).

## See also

- Bivariate toy: [§2.1](fig02-bivariate-toy.html)
- Moment generator:
  [How to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html)
- Feature-selection on this design: [Appendix S6](supp-S6-feature-selection.html)
- Identifiability of collinear means:
  [Appendix S1](supp-S1-identifiability.html)

## References
