---
title: "§2.1 Bivariate toy model (G = 2 genes, J = 2 cell types)"
subtitle: "Full script: `scripts/fig02_bivariate_toy.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{§2.1 Bivariate toy model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  fig-prefix: "Fig."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
---

```{r}
#| label: setup
#| include: false
#| eval: true
options(tinytable_html_mathjax = TRUE)
mu_small <- matrix(
  c(20, 22, 22, 20),
  nrow = 2,
  dimnames = list(c("gene_1", "gene_2"), c("ct1", "ct2"))
)
p_bal <- c(0.5, 0.5)
```

This article is the bivariate Gaussian-convolution toy of the methods
paper ($G=2$, $J=2$). It isolates gene–gene correlation from mean
separation. The factorial grid and ADEMP pipeline live in a **single**
script, `scripts/fig02_bivariate_toy.R`. Scenario builders
(`build_bivariate_scenario_config()`,
`bivariate_toy_deconvolution_functions()`) sit at the top of that file
so tests can source them with `DECOVART_SOURCE_HELPERS=1` without
running 648 scenarios.

[`run_simulation_benchmark()`](../reference/run_simulation_benchmark.html)
wraps `simulate_bulk_mixture()`, `deconvolute_ratios()`, and
`compute_benchmark_metrics()`. Scenario rows are sequential; sample-level
workers live only in `deconvolute_ratios()`. Reporting conventions
(ADEMP, Nature Methods, raincloud / forest plots) are in
[how to build synthetic scenarios](how-to-build-synthetic-scenarios-mean-covariance.html#sec-ademp).

> **Script:** `Rscript scripts/fig02_bivariate_toy.R` (full, $n = 500$) or
> `N_REPLICATES=2 Rscript scripts/fig02_bivariate_toy.R` (smoke test).
> Outputs land in `output/fig02/`.

# Generative model {#sec-bivariate-toy}

With $p_1+p_2=1$, only one free ILR coordinate is estimated (see
[derivatives under simplex transforms](theory-decovart-generative-model.html)).
Even when mean profiles are similar, gene–gene correlation can degrade
mean-only solvers, and is partly recovered once
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$
enters the likelihood.

::: {#fig-toy-design layout-ncol=2}

![Hyperparameter grid used in the bivariate manuscript study: two
centroid geometries, balanced versus unbalanced $\boldsymbol{p}$, a
gene–gene correlation sweep, and the first-generation plus DeCovarT
solvers.](figures/fig_toy_model_parameters.svg){#fig-toy-hyperparams}

![AI-generated 2D density of the two-gene mixture while only the
pairwise correlation changes, at fixed means, marginal variances, and
composition. The ggplot panel below recomputes the same contrast and
annotates MixSim overlap.](figures/fig_toy_2d_density.svg){#fig-toy-density-ai}

Factorial design (left) and correlation-only 2D densities (right) for
the bivariate toy. {#fig-toy-design}
:::

## Factorial design (648 scenarios)

| Factor | Levels |
|---|---|
| Composition $\boldsymbol{p}$ | balanced $(1/2,\,1/2)$; highly unbalanced $(19/20,\,1/20)$ |
| Mean separation (CLD) | small: $\boldsymbol{\mu}_{\cdot 1}=(20,22)$, $\boldsymbol{\mu}_{\cdot 2}=(22,20)$; large: $(20,40)$, $(40,20)$ |
| Gene–gene corr. CT 1 ($\rho_1$) | $\{-0.8,\,-0.6,\,\ldots,\,0.8\}$ — 9 levels |
| Gene–gene corr. CT 2 ($\rho_2$) | $\{-0.8,\,-0.6,\,\ldots,\,0.8\}$ — 9 levels |
| Variance structure | homoscedastic $\sigma^2 = (1,1)$; heteroscedastic $\sigma^2 = (1,2)$ |
| **Total scenarios** | $2 \times 2 \times 9 \times 9 \times 2 = \mathbf{648}$ |

$N = 500$ Monte Carlo replicates per scenario (smoke test: $n = 2$).

```{r}
#| label: gen-model
library(DeCovarT)
Sys.setenv(DECOVART_SOURCE_HELPERS = "1")
source("scripts/fig02_bivariate_toy.R")

bivariate_config <- build_bivariate_scenario_config(
  proportions = list(
    "balanced" = c(0.5, 0.5),
    "highly_unbalanced" = c(19 / 20, 1 / 20)
  ),
  signature_matrices = list(
    "small_CLD" = matrix(c(20, 22, 22, 20), nrow = 2),
    "large_CLD" = matrix(c(20, 40, 40, 20), nrow = 2)
  ),
  corr_sequence = seq(-0.8, 0.8, by = 0.2),
  diagonal_terms = list(
    "homoscedastic" = c(1, 1),
    "heteroscedastic" = c(1, 2)
  )
)
nrow(bivariate_config)
```

```{r}
#| label: fig-toy-density-ggplot
#| fig-cap: >-
#|   Two-gene mixture densities at fixed means
#|   $\boldsymbol{\mu}_{\cdot 1}=(20,22)$, $\boldsymbol{\mu}_{\cdot 2}=(22,20)$,
#|   $\boldsymbol{p}=(1/2,1/2)$ and unit diagonals. Only the common
#|   gene–gene correlation $\rho$ changes. Labels report MixSim
#|   $\overline{\omega}$ (`compute_average_overlap()`).
#| fig-width: 8.4
#| fig-height: 4.4
#| echo: false
set.seed(20260828)
rho_grid <- c(-0.6, 0, 0.6)
n_draw <- 400L
dens_df <- list()
ovl_df <- list()
for (rho in rho_grid) {
  R <- matrix(c(1, rho, rho, 1), nrow = 2)
  Sigma <- array(c(R, R), dim = c(2, 2, 2))
  dimnames(Sigma) <- list(
    rownames(mu_small),
    rownames(mu_small),
    colnames(mu_small)
  )
  sim <- DeCovarT::simulate_bulk_mixture(
    mu_small,
    Sigma,
    p = p_bal,
    n = n_draw
  )
  dens_df[[length(dens_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    gene_1 = sim$Y[1, ],
    gene_2 = sim$Y[2, ]
  )
  theta <- list(p = p_bal, mu = mu_small, sigma = Sigma)
  ovl_df[[length(ovl_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    ovl = DeCovarT::compute_average_overlap(theta)
  )
}
dens_df <- do.call(rbind, dens_df)
ovl_df <- do.call(rbind, ovl_df)
ovl_df$lab <- sprintf("overlap = %.2f", ovl_df$ovl)
ggplot2::ggplot(dens_df, ggplot2::aes(x = gene_1, y = gene_2)) +
  ggplot2::stat_density_2d_filled(contour_var = "ndensity", bins = 12) +
  ggplot2::geom_label(
    data = ovl_df,
    ggplot2::aes(x = Inf, y = Inf, label = lab),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.1,
    fill = "#fffbe9",
    label.size = 0.15
  ) +
  ggplot2::facet_wrap(~rho, nrow = 1) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Expression of gene 1",
    y = "Expression of gene 2"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

## Log-likelihood along the simplex

For $J=2$ the free coordinate is $p_1\in(0,1)$. The listing below
evaluates
$\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}(p_1,1-p_1)$ on one bulk draw.
It is not executed during `R CMD build`: Quarto's Windows CLI starts a
new R process that cannot see the temporary library
([quarto-r#217](https://github.com/quarto-dev/quarto-r/issues/217)).
The same likelihood path is checked in `tests/testthat/`.

```{r}
#| label: fig-toy-loglik
set.seed(20260828)
rho <- 0.6
R <- matrix(c(1, rho, rho, 1), nrow = 2)
Sigma <- array(c(R, R), dim = c(2, 2, 2))
dimnames(Sigma) <- list(
  rownames(mu_small),
  rownames(mu_small),
  colnames(mu_small)
)
y_one <- DeCovarT::simulate_bulk_mixture(
  mu_small,
  Sigma,
  p = p_bal,
  n = 1
)$Y[, 1]
p1_grid <- seq(0.02, 0.98, by = 0.01)
ll_grid <- vapply(
  p1_grid,
  function(p1) {
    DeCovarT::loglik_multivariate(c(p1, 1 - p1), y_one, mu_small, Sigma)
  },
  numeric(1)
)
ll_df <- data.frame(p1 = p1_grid, loglik = ll_grid)
ggplot2::ggplot(ll_df, ggplot2::aes(x = p1, y = loglik)) +
  ggplot2::geom_line(colour = "#1B4F72", linewidth = 0.8) +
  ggplot2::geom_vline(
    xintercept = 0.5,
    linetype = "dashed",
    colour = "grey30"
  ) +
  ggplot2::labs(
    x = "p1 (p2 = 1 - p1)",
    y = "Log-likelihood"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

# Inference {#sec-bivariate-inference}

Seven solvers are evaluated on each scenario:

| Solver | Description |
|---|---|
| `NNLS` | Non-negative least squares (mean-only baseline) |
| `DeconRNASeq` | NNLS via `lsei` [@gongDeconRNASeqStatisticalFramework2013] |
| `L-BFGS-B` | Quasi-Newton on box-constrained simplex |
| `gradient` | Projected gradient descent in ILR |
| `Newton-Raphson` | Second-order in ILR |
| `Marquardt-Levenberg` | Damped Newton via `marqLevAlg` |
| `SA` | Simulated annealing |

Performance is summarised using the ADEMP framework
[@morrisUsingSimulationStudies2019]: **bias**, **RMSE**, **coverage**
(Wilson intervals), and optimiser **failure rate**.

```{r}
#| label: inference
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "500"))
deconvolution_functions <- bivariate_toy_deconvolution_functions(
  itmax = 200L,
  epsilon = 1e-4
)
bivariate <- run_simulation_benchmark(
  scenario_config = bivariate_config,
  deconvolution_functions = deconvolution_functions,
  n = N_REPLICATES,
  cores = 1L
)
```

`config` stores Shannon entropy of $\boldsymbol{p}$ and MixSim overlap;
`optimisation` stores per-sample $\hat{\boldsymbol{p}}$, elapsed time, and
memory; `regression` and `monte_carlo` are the composition and ADEMP
blocks from `compute_benchmark_metrics()`.

# Visualisations {#sec-bivariate-viz}

When $n \ge 10$ replicates, four figures are written to `output/fig02/`:
MSE heatmap on the $\rho_1 \times \rho_2$ grid, raincloud of Monte Carlo
errors, ADEMP forest plot, and algorithm-similarity tile.

### Expected findings

With **large mean separation** and $\rho_j = 0$ both groups of algorithms
should perform well. As $|\rho|$ increases and the means move closer, NNLS
and DeconRNASeq degrade faster than DeCovarT solvers because they ignore
covariance information. The MSE heatmap should display a characteristic
saddle: lowest error on the diagonal $\rho_1 = \rho_2$ and highest near
zero mean separation.

## See also

- Variance-driven hybrid ($G=50$, $J=3$):
  [§2.2](fig03-variance-driven.html)
- Moment generator and ADEMP reporting:
  [How to build synthetic scenarios](how-to-build-synthetic-scenarios-mean-covariance.html)
- Regular-case MLE checks:
  [Appendix S1](supp-S1-identifiability.html)

## References
