---
title: "§2.2 Variance-driven hybrid scenario (G = 50, J = 3)"
subtitle: "Full script: `scripts/fig03_variance_driven.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{§2.2 Variance-driven hybrid scenario}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  fig-prefix: "Fig."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
---

```{r}
#| label: setup
#| eval: false
library(DeCovarT)
```

The bivariate toy confirms the theoretical benefit of covariance
modelling in a transparent setting but cannot demonstrate realistic
gene-regulatory network (GRN) topology. This scenario uses $G = 50$
genes and $J = 3$ cell types in which types 1 and 2 are **near
mean-collinear** ($\rho_{12} \approx 0.97$): the network topology encoded
in $\boldsymbol{\Sigma}_j$ is the discriminative signal. Construction,
ADEMP benchmark, and the static network figure live in a **single**
script, `scripts/fig03_variance_driven.R` (seed `20260807`).

> **Script:** `Rscript scripts/fig03_variance_driven.R` (full, $n = 200$)
> or `N_REPLICATES=2 Rscript scripts/fig03_variance_driven.R` (smoke).
> Moments: `data/synthetic_networks/true_grn_moments.rds`. Figures:
> `output/fig03/`. ADEMP reporting:
> [how to build synthetic scenarios](how-to-build-synthetic-scenarios-mean-covariance.html#sec-ademp).

# Generative model {#sec-scenario-grid}

Types 1 and 2 share a high local cosine on 30 genes and differ only
through $\boldsymbol{\Omega}$; type 3 has a compact marker block; 10 genes
are a null (`equal_all`) block.

$$
\underbrace{\mathcal{G}_{12}}_{30\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{3}}_{10\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{\mathrm{eq}}}_{10\text{ genes}},
\qquad
\boldsymbol{\Omega}_j
=
\mathrm{blockdiag}\bigl(
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{12})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{3})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{\mathrm{eq}})}
\bigr).
$$

$\boldsymbol{\mu}$ uses two calls to `generate_mean_signature_matrix()`
([mean signature construction](how-to-build-synthetic-scenarios-mean-covariance.html#sec-lr-block-mu)):
$\rho_{12}=0.95$ on $\mathcal{G}_{12}$, then $\rho_{3}=0.05$ on a
marker pool for type 3. Local topologies follow the BLGGM two-scale
pattern (@tbl-hybrid-topology-design;
[GGM networks](how-to-build-synthetic-scenarios-mean-covariance.html#sec-ggm-networks)).

| Gene block | # genes | Cell type 1 | Cell type 2 | Cell type 3 |
|---|---|---|---|---|
| `shared_12_vs_3` | 30 | SBM (hub-like modules) | star (one driver) | Erdős–Rényi |
| `marker_3` | 10 | Erdős–Rényi | Erdős–Rényi | scale-free |
| `equal_all` | 10 | Erdős–Rényi | Erdős–Rényi | Erdős–Rényi |

: Local topology per (gene block, cell type); block-diagonal precision support. {#tbl-hybrid-topology-design}

| pair | cosine | Euclidean |
|:-----|-------:|----------:|
| celltype_1 vs celltype_2 | 0.973 | 2.765 |
| celltype_1 vs celltype_3 | 0.562 | 11.219 |
| celltype_2 vs celltype_3 | 0.562 | 11.219 |

: Realised pairwise geometry of the $50\times 3$ mean signature (seed 20260807). {#tbl-mean-geometry}

| cell type | topology | $\lambda_{\min}$ | $\lambda_{\max}$ | $\kappa(\Omega)$ | prop. inhib. |
|:----------|:---------|-----------------:|-----------------:|-----------------:|-----------:|
| celltype_1 | SBM | 0.100 | 1.685 | 16.846 | 0.500 |
| celltype_2 | star | 0.100 | 3.331 | 33.311 | 0.489 |
| celltype_3 | scale-free | 0.100 | 2.096 | 20.961 | 0.500 |

: Precision spectra of each cell type's $\boldsymbol{\Omega}_j$ (seed 20260807). {#tbl-topology-diagnostics}

Types 1 and 2 remain near-collinear on the full $G=50$ vector (cosine
$\approx 0.97$). Type 3 is only *moderately* separated (cosine
$\approx 0.56$) because the shared $\mathcal{G}_{12}$ baseline and the
flat `equal_all` block inflate every pairwise inner product—the
finite-$J$ cosine bias of
[asymptotic cosine control](how-to-build-synthetic-scenarios-mean-covariance.html#sec-asymptotic-cosine).
@tbl-topology-diagnostics shows that types 1 and 2 still differ in
$\boldsymbol{\Omega}$.

![Cell-type-specific block topologies: SBM/hub-like modules (type 1) and
a star (type 2) on the 30 `shared_12_vs_3` genes; scale-free wiring
(type 3) on `marker_3`. The `equal_all` block is Erdős–Rényi in every
type. Edge colour: precision sign (red inhibitory, teal activatory).](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

::: {#nte-related-dd .callout-info}
## Related literature (`muscat` / `scDD`)

Pseudobulk tools such as `muscat`
[@crowellMuscatDetectsSubpopulationspecific2020] aggregate single-cell
counts across samples. `scDD`
[@korthauerStatisticalApproachIdentifying2016] partitions genes into
EE, DE, DP, DM, and DB patterns. The hybrid scenario is a
network-level analogue of three of these (@fig-dd-taxonomy):
`equal_all` is EE; `marker_3` is DE for type 3; `shared_12_vs_3` has
**no mean shift between types 1 and 2**, only a precision-topology
contrast (DM-like for deconvolution, not scDD's original single-gene
modality test).
:::

![Schematic EE / DE / DM-like taxonomy for the three gene
blocks.](figures/fig_dd_taxonomy_ee_de_dm.jpg){#fig-dd-taxonomy width="100%"}

Two generators were considered but not adopted:

- **Toeplitz covariances**: encode only local (banded) dependencies;
  cannot model cell-type-specific sparse precision.
- **`MixSim` overlap** [@melnykovMixSimPackageSimulating2012]: designed
  for isotropic or full-rank Gaussian clusters; no cell-type-specific
  precision input.

# Inference {#sec-hybrid-deconvolution}

| Factor | Levels |
|---|---|
| Composition $\boldsymbol{p}$ | balanced $(1/3,1/3,1/3)$; moderately unbalanced $(0.5,0.3,0.2)$; highly unbalanced $(0.7,0.2,0.1)$ |
| Algorithms | NNLS, DeconRNASeq, Marquardt–Levenberg |
| **Total scenarios** | $3 \times 3 = \mathbf{9}$ (per fixed GRN realisation) |

$N = 200$ Monte Carlo replicates per scenario (smoke test: $n = 2$).
All solvers start from the open simplex via `starting_simplex()`
(barycentre by default; Dirichlet or QP via `initial_p`).

| Algorithm | Function | Constraint |
|:----------|:---------|:-----------|
| NNLS | `deconvolute_ratios_nnls` | `repair_simplex()` |
| QP (`DeconRNASeq`-style) | `deconvolute_ratios_deconrnaseq` | simplex equality / inequality |
| Marquardt–Levenberg | `deconvolute_ratios_Marquardt_Levenberg` | ILR $\to$ open simplex |

: Solvers used in the variance-driven comparison. {#tbl-solver-hybrid}

The shipped catalogue (CIBERSORT-style $\nu$-SVR, `MASS::rlm`, GLS,
Newton–Raphson, L-BFGS-B, BFGS, simulated annealing) is listed in
[`deconvolute_ratios()`](../reference/deconvolute_ratios.html). A
temporary solver smoke test on this GRN lives in
`tests/manual/deconvolute_simulated_scenario.R` (Rbuildignored).

```{r}
#| label: inference
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "200"))
ALGORITHMS <- c("NNLS", "DeconRNASeq", "Marquardt-Levenberg")
# Full pipeline: Rscript scripts/fig03_variance_driven.R
```

# Visualisations {#sec-hybrid-viz}

| Output | Description |
|---|---|
| `output/fig03/fig03_raincloud.pdf` | Raincloud of RMSE by algorithm and composition |
| `output/fig03/fig03_forest.pdf` | ADEMP forest plot |
| `output/fig03/fig03_metric_dots.pdf` | Metric dot plot across compositions |

### Expected findings

Because CT 1 and CT 2 share near-identical mean profiles
($\rho_{12} \approx 0.97$), NNLS and DeconRNASeq cannot separate them —
their RMSE for that pair should sit near the variance floor. DeCovarT
(Marquardt–Levenberg) uses the block-diagonal precision structure;
its advantage grows as the composition becomes unbalanced and the rare
type (CT 3 at 10%) requires second-order information.

## See also

- Bivariate toy: [§2.1](fig02-bivariate-toy.html)
- Moment generator:
  [How to build synthetic scenarios](how-to-build-synthetic-scenarios-mean-covariance.html)
- Feature-selection on this design: [Appendix S6](supp-S6-feature-selection.html)
- Identifiability of collinear means:
  [Appendix S1](supp-S1-identifiability.html)

## References
