---
title: "2.2 Covariance-driven scenario (G = 20, J = 3)"
subtitle: "Full script: `scripts/fig03_variance_driven.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{§2.2 Covariance-driven scenario}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  fig-prefix: "Fig."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
---

```{r}
#| label: setup
#| eval: false
library(DeCovarT)
```

The bivariate toy confirms the theoretical benefit of covariance
modelling in a transparent setting but cannot demonstrate realistic
gene-regulatory network (GRN) topology. This scenario uses $G = 20$
genes and $J = 3$ cell types in which types 1 and 2 are **mean-collinear**
($\cos(\boldsymbol{\mu}_{\cdot 1},\boldsymbol{\mu}_{\cdot 2})=0.9$): the
network topology encoded in $\boldsymbol{\Sigma}_j$ is the remaining
discriminative signal. Construction, ADEMP benchmark, and the static
network figure live in a **single** script,
`scripts/fig03_variance_driven.R` (seed `20260807`).

```{mermaid}
%%| label: fig-covariance-design
%%| eval: true
%%| fig-cap: "Covariance-driven factorial: one fixed Gram (cosines 0.9 / 0.1), a 2-by-2 graph assignment on the mean-collinear types (scale-free versus cluster SBM; type 3 held at scale-free), three target MixSim average overlaps, three Shannon compositions, and five solvers (4 x 3 x 3 = 36 scenarios)."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  P["3 compositions<br/>H* = 1 / 0.5 / 0.1"] --> S["36 scenarios"]
  G["2 x 2 graph assignments<br/>CT1, CT2: scale-free / cluster SBM"] --> S
  C["CT3 graph held at scale-free"] --> S
  K["3 target BarOmega levels<br/>low / moderate / high"] --> S
  S --> A["5 deconvolution solvers"]
```

```{mermaid}
%%| label: fig-covariance-design-tree
%%| eval: true
%%| fig-cap: "Hierarchical reading of the covariance factorial: 2-by-2 graph assignment on mean-collinear cell types 1 and 2 (type 3 held at scale-free), then low / moderate / high target MixSim BarOmega (4 x 3 = 12 covariance draws). Each leaf is crossed with the three Shannon compositions H* = 1 / 0.5 / 0.1."
---
config:
  layout: elk
  theme: sandstone
---
flowchart TB
  ROOT["Covariance-driven G=20, J=3"]
  ROOT --> SfSf["CT1 SF, CT2 SF"]
  ROOT --> SfSb["CT1 SF, CT2 SBM"]
  ROOT --> SbSf["CT1 SBM, CT2 SF"]
  ROOT --> SbSb["CT1 SBM, CT2 SBM"]
  SfSf --> SfSfLo["Low overlap"]
  SfSf --> SfSfMo["Moderate overlap"]
  SfSf --> SfSfHi["High overlap"]
  SfSb --> SfSbLo["Low overlap"]
  SfSb --> SfSbMo["Moderate overlap"]
  SfSb --> SfSbHi["High overlap"]
  SbSf --> SbSfLo["Low overlap"]
  SbSf --> SbSfMo["Moderate overlap"]
  SbSf --> SbSfHi["High overlap"]
  SbSb --> SbSbLo["Low overlap"]
  SbSb --> SbSbMo["Moderate overlap"]
  SbSb --> SbSbHi["High overlap"]
```

# Generative model {#sec-scenario-grid}

One fixed mean signature $\boldsymbol{\mu}\in\mathbb{R}^{G\times J}$ is
built with `generate_mean_signature_matrix()` and an explicit Gram
`target_gram` ([mean signature
construction](theory-synthetic-scenarios-mean-covariance.html#sec-lr-block-mu)).
There is no marker block and no null (`equal_all`) block: uninformative
genes would be removed before deconvolution.

$$
R
=
\begin{pmatrix}
1 & 0.9 & 0.1 \\
0.9 & 1 & 0.1 \\
0.1 & 0.1 & 1
\end{pmatrix},
\qquad
\boldsymbol{\mu}
=
s\mathbf{Q}R^{1/2},
\quad s=10.
$$ {#eq-fig03-gram}

Because $\mathbf{Q}^{\mathsf{T}}\mathbf{Q}=\mathbf{I}_J$, the pairwise
cosines of the columns of $\boldsymbol{\mu}$ equal $R$ exactly (up to
rounding). Euclidean gaps follow
$\lVert\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\rVert
=s\sqrt{2(1-R_{jk})}$. The call uses `nonnegative = TRUE` (disjoint
gene-block frame) so that $\boldsymbol{\mu}\ge 0$, as required by
`deconvolute_ratios()`.

| pair | cosine | Euclidean |
|:-----|-------:|----------:|
| celltype_1 vs celltype_2 | 0.9 | $10\sqrt{0.2}\approx 4.472$ |
| celltype_1 vs celltype_3 | 0.1 | $10\sqrt{1.8}\approx 13.416$ |
| celltype_2 vs celltype_3 | 0.1 | $10\sqrt{1.8}\approx 13.416$ |

: Exact pairwise geometry of the $20\times 3$ mean signature (`target_gram`, `mean_scale` $=10$). {#tbl-mean-geometry}

Each of the mean-collinear types (CT1, CT2) independently draws an
undirected skeleton from one of two generators, then completes a signed
precision with
`build_normalised_precision()`
([GGM networks](theory-synthetic-scenarios-mean-covariance.html#sec-ggm-networks);
topology families in
[the topology table](theory-synthetic-scenarios-mean-covariance.html#tbl-topologies)):

- **Scale-free** (`graph_model = "scale_free"`): Barabási–Albert
  preferential attachment (`power = 1`, `edges_per_node = 1`).
- **Cluster / stochastic block model**
  (`graph_model = "stochastic_block_model"`): three blocks with
  probabilities $(0.5,0.25,0.25)$, $p_{\mathrm{within}}=0.25$,
  $p_{\mathrm{between}}=0.01$.

Erdős–Rényi is not used. Type 3, already separated by the Gram
($R_{13}=R_{23}=0.1$), is held at scale-free, so the topology factorial
is $2^{2}=4$ assignments on CT1 and CT2. Those generator names and
parameter lists are stored on every row of `hybrid_config.rds` as
`graph_ct1` / `graph_ct2` / `graph_ct3` and the list-column
`graph_generators`, and again inside each `true_theta`.

The spectral cushion $u$ used at completion is now a **fixed**
engineering constant (`PRECISION_SHIFT_BASE = 0.2`) so that each
graph has one SPD precision. The experimental knob is a global
covariance scale $s$: $\boldsymbol{\Sigma}_j\mapsto s\boldsymbol{\Sigma}_j$
keeps every structural zero of $\boldsymbol{\Omega}_j$ because
$\boldsymbol{\Omega}_j(s)=\boldsymbol{\Omega}_j/s$.
`scale_covariances_to_overlap()` searches $s$ so that MixSim
`BarOmega` matches a target in $\{0.02,0.08,0.20\}$ (low / moderate /
high), using Sobol Monte Carlo for $G=20$
([overlap note](theory-distance-covariance.html#sec-fig03-scale)).
This is the graph-constrained analogue of MixSim / FSDA average-overlap
simulation
([CRAN MixSim](https://cran.r-project.org/web/packages/MixSim/refman/MixSim.html#MixSim);
[@melnykovMixSimPackageSimulating2012; @rianiSimulatingMixturesMultivariate2015]):
we do **not** call `MixSim()` as a generator, because it cannot take
a precision support as input.

Larger $s$ also raises the covariance-information fraction
$f_{\mathrm{cov}}$ (mean Fisher scales as $1/s$; the covariance block
is scale-invariant). See the
[tangent-Fisher callout](theory-synthetic-scenarios-mean-covariance.html#nte-desc-fisher).

| Factor | Levels |
|---|---|
| Graph model (CT1, CT2) | scale-free; cluster SBM |
| Graph model (CT3) | scale-free (fixed) |
| Target `BarOmega` | low ($0.02$); moderate ($0.08$); high ($0.20$) |
| Topology assignments | $2^{2}=\mathbf{4}$ |
| Covariance draws | $4\times 3=\mathbf{12}$ |

: Covariance factorial (graphs $\times$ target average overlap). {#tbl-cov-factorial}

`describe_simulation_scenario()` records the realised `BarOmega`,
$f_{\mathrm{cov}}$, $\kappa\{\boldsymbol{\Sigma}(\boldsymbol{p})\}$, and
pairwise AIRM distance of the $\boldsymbol{\Sigma}_j$ on every row.
The overlap target is matched at the **balanced** composition;
unbalanced $\boldsymbol{p}$ change MixSim's MAP weights, so realised
`BarOmega` is higher there. AIRM is invariant to the global scale
$s$ (it scores topology / shape). $f_{\mathrm{cov}}$ rises with $s$.

![Two generator examples on $G=20$ nodes (scale-free, cluster SBM).
Mean-collinear cell types 1 and 2 are each assigned one of these
families; type 3 is held at scale-free.](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

Two generators were considered but not adopted:

- **Toeplitz covariances**: encode only local (banded) dependencies;
  cannot model cell-type-specific sparse precision.
- **`MixSim()` as a mixture generator**: draws unconstrained means and
  covariances for a target average overlap; no graph-constrained
  precision input. DeCovarT keeps the graphs and matches `BarOmega`
  by scaling $\boldsymbol{\Sigma}_j$ instead.

# Inference {#sec-hybrid-deconvolution}

Compositions are one-dominant vectors from
`composition_from_entropy()` targeting normalised Shannon entropy
$H^{\star}\in\{1,0.5,0.1\}$ (balanced; moderately unbalanced; highly
unbalanced). All DeCovarT solvers start at the barycentre
(`initial_p = "barycentre"`). Other initialisation strategies are
reserved for separate scenarios.

| Factor | Levels |
|---|---|
| Composition $\boldsymbol{p}$ | $H^{\star}=1$; $H^{\star}=0.5$; $H^{\star}=0.1$ |
| Algorithms | DeconRNASeq (`lsei`), CIBERSORT, L-BFGS-B, Newton–Raphson, Marquardt–Levenberg |
| **Total scenarios** | $4\times 3\times 3=\mathbf{36}$ |

$N = 50$ Monte Carlo replicates per scenario.

| Algorithm | Function | Constraint |
|:----------|:---------|:-----------|
| QP (`DeconRNASeq`-style) | `deconvolute_ratios_deconrnaseq` | simplex equality / inequality |
| CIBERSORT-style $\nu$-SVR | `deconvolute_ratios_cibersort` | non-negative then `repair_simplex()` |
| L-BFGS-B | `deconvolute_ratios_L_BFGS_B` | ILR $\to$ open simplex |
| Newton–Raphson | `deconvolute_ratios_Newton_Raphson` | ILR / `nlminb` |
| Marquardt–Levenberg | `deconvolute_ratios_Marquardt_Levenberg` | ILR $\to$ open simplex |

: Solvers used in the covariance-driven comparison. {#tbl-solver-hybrid}

Mean-only baselines (LSEI, CIBERSORT) ignore $\boldsymbol{\Sigma}_j$.
The three DeCovarT maps use the same convolution likelihood. The shipped
catalogue (robust linear model, GLS, BFGS, simulated annealing) is
listed in
[`deconvolute_ratios()`](../reference/deconvolute_ratios.html).

```{r}
#| label: inference
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "50"))
ALGORITHMS <- c(
  "lsei",
  "cibersort",
  "LBFGS",
  "Newton-Raphson",
  "Marquardt-Levenberg"
)
# Full pipeline: Rscript scripts/fig03_variance_driven.R
```

# Visualisations {#sec-hybrid-viz}

| Output | Description |
|---|---|
| `output/fig03/density_visualisations/mean_signature.pdf` | Unscaled \(\boldsymbol{\mu}\) heatmap |
| `output/fig03/density_visualisations/latent_projections.pdf` | 16-page 2D latent-space book (Thomson / MCFA / `MclustDR`) |
| `output/fig03/density_visualisations/loglik_rgl.pdf` | 3D ALR log-likelihood (`rgl`, \(\rho_1,\rho_2,\ell\)) |
| `output/fig03/density_visualisations/network_topologies.pdf` | Graph skeletons, one page per MixSim overlap |
| `output/fig03/performance_visualisations/heatmap_rmse.pdf` | RMSE tiles |
| `output/fig03/performance_visualisations/heatmap_aitchison.pdf` | Aitchison tiles |
| `output/fig03/performance_visualisations/raincloud.pdf` | Monte Carlo \(\hat{\boldsymbol{p}}\) rainclouds |
| `output/fig03/performance_visualisations/forest.pdf` | ADEMP Wald forest |
| `output/fig03/performance_visualisations/runtime.pdf` | Solver wall-clock (log10 seconds) |
| `output/fig03/performance_visualisations/memory.pdf` | Peak memory (log10 MiB) |

Redraw with `FIG03_POSTPROCESS_ONLY=1 Rscript scripts/fig03_variance_driven.R`
(no ADEMP refit). ggplot `data` tables go under
`output/fig03/ggplot_rds/`.

## Two-dimensional projections of a Gaussian convolution {#sec-hybrid-2d-proj}

With \(G=20\) the purified Gaussians and the bulk convolution are no
longer drawable in the gene plane. The 16-page book
`latent_projections.pdf` therefore maps each selected scenario to a
2D latent display. Pages keep the **lowest and highest** MixSim
overlap (low / high; the moderate arm is omitted) and the **highest
and lowest** Shannon entropy of \(\boldsymbol{p}\) (balanced versus
highly unbalanced), crossed with the four CT1/CT2 graph assignments
(\(2\times 2\times 4=16\)). Each page is a \(3\times 2\) of
scatterplots with 95% Gaussian ellipses, cell-type colour, axis
labels that report the share of variability on the first two
coordinates, a transparent cowplot scree inset of the first ten
scaled eigenvalues, and a six-line footnote for panels A--F. Square
panels use a unit aspect ratio (not `coord_equal`), so unequal
factor-score ranges no longer squash the shared-plane row.

The three routes below all return an ordered set of eigenvalues
that summarise how much variability sits in the plotted plane. They
are **not** interchangeable with the DeCovarT likelihood: the bulk
profile is a **convolution** of Gaussians, not a Gaussian mixture.

::: {#nte-conv-not-gmm .callout-warning}
A $J$-type bulk draw is
$Y\mid\boldsymbol{p}\sim\mathcal{N}\bigl(\boldsymbol{\mu}(\boldsymbol{p}),\boldsymbol{\Sigma}(\boldsymbol{p})\bigr)$
with
$\boldsymbol{\mu}(\boldsymbol{p})=\sum_j p_j\boldsymbol{\mu}_{\cdot j}$ and
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$.
Every observation is a weighted sum of **all** types
([`simulate_bulk_mixture()`](../reference/simulate_bulk_mixture.html)).
A finite mixture instead draws a latent label and then one
component,
$f(\boldsymbol{y})=\sum_k\pi_k\,\phi(\boldsymbol{y};\boldsymbol{\mu}_k,\boldsymbol{\Sigma}_k)$,
with mixing weights \(\boldsymbol{p}\) (not \(\boldsymbol{p}^2\)).
`MclustDR` and mixtures of factor analysers target that mixture.
Their \(\pi_k\) are not cell-type proportions, and their component
densities are not the convolution law used for deconvolution.
:::

### Independent factor analyses per cell type

The top row fits a two-factor Gaussian factor model **separately**
to purified draws from each type (`stats::factanal()`, Thomson /
regression scores
[@thomsonMethodsEstimatingMental1938]). The orthogonal-factor model
is
$\boldsymbol{x}=\boldsymbol{\mu}+\Lambda\boldsymbol{f}+\boldsymbol{\varepsilon}$
with $\boldsymbol{f}\sim N(\boldsymbol{0},I)$ and diagonal uniquenesses
$\Psi$, so
$\boldsymbol{\Sigma}=\Lambda\Lambda^\top+\Psi$.
Thomson scores are the posterior mean
$\hat{\boldsymbol{f}}_R=E[\boldsymbol{f}\mid\boldsymbol{x}]=\Lambda^\top\boldsymbol{\Sigma}^{-1}(\boldsymbol{x}-\boldsymbol{\mu})$.
That is the natural probabilistic answer to “where does this
20-gene profile sit in the plane?”. Bartlett scores (uniqueness-
weighted GLS) remain a `factanal(..., scores = "Bartlett")`
sensitivity check and are not plotted. Gene-wise marginal variances
are comparable after the global MixSim scale, so uniquenesses are
less heterogeneous than in a typical psychometric battery.
Independent per-type fits do **not** share a loading matrix: the
three planes are not superimposed.

```{r}
#| label: fa-thomson-sketch
#| eval: false
fit <- stats::factanal(
  X,
  factors = 2,
  rotation = "none",
  scores = "regression"
)
scores <- as.data.frame(fit$scores)
ggplot2::ggplot(scores, ggplot2::aes(Factor1, Factor2)) +
  ggplot2::geom_point() +
  ggplot2::coord_equal() +
  ggplot2::theme_classic()
```

### Shared MCFA plane (`EMMIXmfa`)

The bottom-left and bottom-centre panels use mixtures of common
factor analysers [@MixturesFactorAnalyzers2000; @R-EMMIXmfa]: a
shared loading matrix $A$ with
$\boldsymbol{\mu}_i=A\boldsymbol{\xi}_i$ and
$\boldsymbol{\Sigma}_i=A\Omega_i A^\top+D$, so that posterior factor
means live in **one** $q=2$ plane. The eigenvalues of
$A\bar{\Omega}A^\top+D$ are the scree inset.

- **Convolution (panel D).** Bulk columns come from
  `simulate_bulk_mixture()` (covariance weights $p_j^2$). MCFA is
  fitted to those unlabelled bulk draws, then the *latent* purified
  slices \(\mathcal{X}\) from the same simulation are scored into
  that shared plane and coloured by true type.
- **Unsupervised mixture (panel E).** Each draw is from **one**
  component with probability $p_j$ (not $p_j^2$). MCFA does not
  use the labels; colour is the generating type, so the plot shows
  how well the shared factors recover overlap.

### Supervised `MclustDR` plane

The bottom-right panel takes the same labelled mixture draws and
fits `mclust::MclustDA(..., modelType = "EDDA")` then
`mclust::MclustDR()`
[@R-mclust; @scruccaDimensionReductionModelbased2010a;
@scruccaGraphicalToolsModelbased2015].
See the [`MclustDR` reference](https://mclust-org.github.io/mclust/reference/MclustDR.html).
The kernel blends between-class mean separation with covariance
differences (`lambda = 1` emphasises separation). The generalised
eigenvalues (`evalues`) are the scree inset: they are
cluster-relevant variability, not ordinary PCA variance. This is
the right 2D plot **if** class labels are known. On bulk
convolution draws a one-component Gaussian is the truth, so
`Mclust` would split a single ellipsoid; that misspecified fit is
not shown.

## Solver runtime and memory {#sec-hybrid-resources}

`runtime.pdf` and `memory.pdf` use the same composition order as
the other performance books (balanced, then moderately unbalanced,
then highly unbalanced): **one page per** \(\boldsymbol{p}\). Each page
facets the three MixSim overlap levels, puts CT1/CT2 topology on
the x-axis (`SF/SF`, `SF/SBM`, `SBM/SF`, `SBM/SBM`), and maps
solver to colour, fill, and grouping. The geometry is a vertical
ggdist half-eye (raincloud) of the per-sample
`optimisation$elapsed_sec` or `memory_bytes` (converted to MiB),
with a left-side `geom_rug()` of the raw Monte Carlo draws and a
log10 y-axis.

## Expected findings

Because CT 1 and CT 2 share cosine $0.9$, LSEI and CIBERSORT cannot
separate them from the mean signature alone. DeCovarT (L-BFGS-B,
Newton--Raphson, Marquardt--Levenberg) uses $\boldsymbol{\Sigma}_j$.
**Topology** (scale-free versus cluster SBM) still matters, but
**average overlap** is the stronger predictor of error and of Wald
uncertainty: high `BarOmega` scenarios remain hard even when the
graphs differ, whereas low overlap with a large $f_{\mathrm{cov}}$
is where covariance-aware solvers should pull ahead of mean-only
baselines, especially at $H^{\star}=0.1$.

## See also

- Bivariate toy: [§2.1](fig02-bivariate-toy.html)
- Moment generator:
  [How to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html)
- Overlap and SPD distances:
  [Distances, overlap, and covariance information](theory-distance-covariance.html)
- Feature-selection on this design: [Appendix S6](supp-S6-feature-selection.html)
- Identifiability of collinear means:
  [Appendix S1](supp-S1-identifiability.html)

## References
