---
title: "§2.1 Bivariate toy model (G = 2 genes, J = 2 cell types)"
subtitle: "Full script: `scripts/fig02_bivariate_toy.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{§2.1 Bivariate toy model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  fig-prefix: "Fig."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
---

```{r}
#| label: setup
#| include: false
#| eval: true
options(tinytable_html_mathjax = TRUE)
mu_small <- matrix(
  c(20, 22, 22, 20),
  nrow = 2,
  dimnames = list(c("gene_1", "gene_2"), c("ct1", "ct2"))
)
p_bal <- c(0.5, 0.5)
```

This article is the bivariate Gaussian-convolution toy of the methods
paper ($G=2$, $J=2$). It isolates gene–gene correlation from mean
separation. The factorial grid and ADEMP pipeline live in a **single**
script, `scripts/fig02_bivariate_toy.R`. Scenario builders
(`build_bivariate_scenario_config()`,
`bivariate_toy_deconvolution_functions()`) sit at the top of that file.
Sourcing the script from a vignette or from a temp copy in tests
defines those functions without launching the 972-scenario pipeline.

[`run_simulation_benchmark()`](../reference/run_simulation_benchmark.html)
wraps `simulate_bulk_mixture()`, `deconvolute_ratios()`, and
`compute_benchmark_metrics()`. Scenario rows are sequential; sample-level
workers live only in `deconvolute_ratios()`. Reporting conventions
(ADEMP, Nature Methods, raincloud / forest plots) are in
[how to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html#sec-ademp).

> **Script:** from the repository root,
> `mkdir -p logs && nohup Rscript --no-save --no-restore scripts/fig02_bivariate_toy.R > "logs/fig02_$(date +%F)_bivariate_toy.log" 2>&1 &`
> (full, $n = 500$).
> To redraw figures from a finished ADEMP RDS without refitting,
> `FIG02_POSTPROCESS_ONLY=1 Rscript --no-save --no-restore scripts/fig02_bivariate_toy.R`.
> Outputs land in `output/fig02/` (RDS artefacts),
> `output/fig02/density_visualisations/`,
> `output/fig02/performance_visualisations/`, and
> `output/fig02/ggplot_rds/`.

```{mermaid}
%%| label: fig-bivariate-design
%%| eval: true
%%| fig-cap: "Bivariate toy factorial design: three compositions, two variance structures, two CLD mean separations, a 9-by-9 gene-gene correlation grid, and seven solvers (3 x 2 x 2 x 9 x 9 = 972 scenarios)."
%%{init: {"theme": "sandstone"}}%%
flowchart TD
  P["3 compositions<br/>Ba / Mo / Hi"] --> S["972 scenarios"]
  V["2 variance structures<br/>Ho homoscedastic / He heteroscedastic"] --> S
  C["2 CLD mean separations<br/>Sm small / Lg large"] --> S
  R["9 x 9 gene-gene correlations<br/>rho in -0.8 to 0.8, step 0.2"] --> S
  S --> A["7 deconvolution solvers"]
```

```{mermaid}
%%| label: fig-bivariate-design-tree
%%| eval: true
%%| fig-cap: "Hierarchical reading of the same factorial: small versus large CLD, then homoscedastic versus heteroscedastic variances, then balanced / moderately unbalanced / highly unbalanced compositions (2 x 2 x 3 = 12 meta-scenarios). Each leaf is crossed with the 9-by-9 gene-gene correlation grid."
---
config:
  layout: elk
  theme: sandstone
---
flowchart TB
  ROOT["Bivariate toy"]
  ROOT --> Sm["Small CLD"]
  ROOT --> Lg["Large CLD"]
  Sm --> SmHo["Homoscedastic"]
  Sm --> SmHe["Heteroscedastic"]
  Lg --> LgHo["Homoscedastic"]
  Lg --> LgHe["Heteroscedastic"]
  SmHo --> SmHoBa["Balanced"]
  SmHo --> SmHoMo["Moderately unbalanced"]
  SmHo --> SmHoHi["Highly unbalanced"]
  SmHe --> SmHeBa["Balanced"]
  SmHe --> SmHeMo["Moderately unbalanced"]
  SmHe --> SmHeHi["Highly unbalanced"]
  LgHo --> LgHoBa["Balanced"]
  LgHo --> LgHoMo["Moderately unbalanced"]
  LgHo --> LgHoHi["Highly unbalanced"]
  LgHe --> LgHeBa["Balanced"]
  LgHe --> LgHeMo["Moderately unbalanced"]
  LgHe --> LgHeHi["Highly unbalanced"]
```

# Generative model {#sec-bivariate-toy}

With $p_1+p_2=1$, only one free ILR coordinate is estimated (see
[derivatives under simplex transforms](theory-decovart-generative-model.html)).
Even when mean profiles are similar, gene–gene correlation can degrade
mean-only solvers, and is partly recovered once
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$
enters the likelihood.

::: {#fig-toy-design layout-ncol=2}

![Hyperparameter grid used in the bivariate manuscript study: two
centroid geometries, balanced versus unbalanced $\boldsymbol{p}$, a
gene–gene correlation sweep, and the first-generation plus DeCovarT
solvers.](figures/fig_toy_model_parameters.svg){#fig-toy-hyperparams}

![AI-generated 2D density of the two-gene mixture while only the
pairwise correlation changes, at fixed means, marginal variances, and
composition. The ggplot panel below recomputes the same contrast and
annotates MixSim overlap.](figures/fig_toy_2d_density.svg){#fig-toy-density-ai}

Factorial design (left) and correlation-only 2D densities (right) for
the bivariate toy. {#fig-toy-design}
:::

## Factorial design (972 scenarios)

| Factor | Levels |
|---|---|
| Composition $\boldsymbol{p}$ | balanced $(1/2,\,1/2)$; moderately unbalanced $(17/20,\,3/20)$; highly unbalanced $(99/100,\,1/100)$ |
| Mean separation (CLD) | small: $\boldsymbol{\mu}_{\cdot 1}=(20,22)$, $\boldsymbol{\mu}_{\cdot 2}=(22,20)$; large: $(20,40)$, $(40,20)$ |
| Gene–gene corr. CT 1 ($\rho_1$) | $\{-0.8,\,-0.6,\,\ldots,\,0.8\}$ — 9 levels |
| Gene–gene corr. CT 2 ($\rho_2$) | $\{-0.8,\,-0.6,\,\ldots,\,0.8\}$ — 9 levels |
| Variance structure | homoscedastic $\sigma^2 = (1,1)$; heteroscedastic $\sigma^2 = (1,2)$ |
| **Total scenarios** | $3 \times 2 \times 9 \times 9 \times 2 = \mathbf{972}$ |

$N = 500$ Monte Carlo replicates per scenario.

**CLD** is centroid (mean-profile) Euclidean separation
$\lVert\boldsymbol{\mu}_{\cdot 1}-\boldsymbol{\mu}_{\cdot 2}\rVert_2$,
not an Aitchison distance; see
[the descriptor notes](theory-synthetic-scenarios-mean-covariance.html#nte-desc-cld).
Small CLD uses $\boldsymbol{\mu}_{\cdot 1}=(20,22)$,
$\boldsymbol{\mu}_{\cdot 2}=(22,20)$ ($d=2\sqrt{2}\approx 2.83$,
cosine $\approx 0.995$). Large CLD uses $(20,40)$ and $(40,20)$
($d=20\sqrt{2}\approx 28.3$, cosine $0.8$).

### Scenario ID

Each row is labelled `B{index}_{variance}_{composition}_{CLD}`. The
index is the factorial row number ($1$–$972$). Example:
`B1_Ho_Ba_Sm` is the first row, homoscedastic, balanced proportions,
small CLD.

| Token | Meaning |
|---|---|
| `B` | Bivariate toy ($G=2$, $J=2$) |
| `1`–`972` | Factorial row index |
| `Ho` / `He` | Homoscedastic / heteroscedastic marginal variances |
| `Ba` / `Mo` / `Hi` | Balanced / moderately unbalanced / highly unbalanced $\boldsymbol{p}$ |
| `Sm` / `Lg` | Small / large CLD (Euclidean centroid gap) |

The slim design grid is `output/fig02/bivariate_config.rds` (no
`true_theta`). Geometry, MixSim `BarOmega`, Hellinger, and Jeffreys
live in `bivariate_descriptors.rds`. Convolution parameters are
`bivariate_theta.rds`. ADEMP metrics (`regression`, `monte_carlo`,
`optimisation`, `call`) are `bivariate_benchmark.rds`. All four files
share the `ID` key. `read_simulation_artefacts(dir, "bivariate", assemble = TRUE)`
rejoins them for plotting.

The Monte Carlo $\hat{\boldsymbol{p}}$ draws themselves are the cell-type
columns of `optimisation` (one row per replicate $\times$ algorithm
$\times$ scenario). `pivot_mc_estimates()` stacks those columns into
long form. Interior Wald intervals for convolution-likelihood
solvers use [`vcov_ilr_delta()`](../reference/vcov_ilr_delta.html)
(expected Fisher information through the ILR chart), matching
[`confint.decovart_fit()`](../reference/fit_decovart.html). Coverage,
`mean_model_se`, and `se_sd_ratio` (mean model SE / empirical SD) are
then finite. Mean-only solvers (NNLS, LSEI) leave those Wald columns
missing: they do not maximise the convolution likelihood, so an ILR
interval would not apply.

```{r}
#| label: gen-model
library(DeCovarT)
source("scripts/fig02_bivariate_toy.R")

bivariate_config <- build_bivariate_scenario_config()
nrow(bivariate_config)
```

```{r}
#| label: fig-toy-density-ggplot
#| fig-cap: >-
#|   Two-gene mixture densities at fixed means
#|   $\boldsymbol{\mu}_{\cdot 1}=(20,22)$, $\boldsymbol{\mu}_{\cdot 2}=(22,20)$,
#|   $\boldsymbol{p}=(1/2,1/2)$ and unit diagonals. Only the common
#|   gene–gene correlation $\rho$ changes. Labels report MixSim
#|   $\overline{\omega}$ (`compute_average_overlap()`).
#| fig-width: 8.4
#| fig-height: 4.4
#| echo: false
set.seed(20260828)
rho_grid <- c(-0.6, 0, 0.6)
n_draw <- 400L
dens_df <- list()
ovl_df <- list()
for (rho in rho_grid) {
  R <- matrix(c(1, rho, rho, 1), nrow = 2)
  Sigma <- array(c(R, R), dim = c(2, 2, 2))
  dimnames(Sigma) <- list(
    rownames(mu_small),
    rownames(mu_small),
    colnames(mu_small)
  )
  sim <- DeCovarT::simulate_bulk_mixture(
    mu_small,
    Sigma,
    p = p_bal,
    n = n_draw
  )
  dens_df[[length(dens_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    gene_1 = sim$Y[1, ],
    gene_2 = sim$Y[2, ]
  )
  theta <- list(p = p_bal, mu = mu_small, sigma = Sigma)
  ovl_df[[length(ovl_df) + 1L]] <- data.frame(
    rho = sprintf("rho = %s", format(rho, nsmall = 1)),
    ovl = DeCovarT::compute_average_overlap(theta)
  )
}
dens_df <- do.call(rbind, dens_df)
ovl_df <- do.call(rbind, ovl_df)
ovl_df$lab <- sprintf("overlap = %.2f", ovl_df$ovl)
ggplot2::ggplot(dens_df, ggplot2::aes(x = gene_1, y = gene_2)) +
  ggplot2::stat_density_2d_filled(contour_var = "ndensity", bins = 12) +
  ggplot2::geom_label(
    data = ovl_df,
    ggplot2::aes(x = Inf, y = Inf, label = lab),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.1,
    fill = "#fffbe9",
    label.size = 0.15
  ) +
  ggplot2::facet_wrap(~rho, nrow = 1) +
  ggplot2::guides(fill = "none") +
  ggplot2::labs(
    x = "Expression of gene 1",
    y = "Expression of gene 2"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

## Log-likelihood along the simplex

For $J=2$ the free coordinate is $p_1\in(0,1)$. The listing below
evaluates
$\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}(p_1,1-p_1)$ on one bulk draw.
It is not executed during `R CMD build`: Quarto's Windows CLI starts a
new R process that cannot see the temporary library
([quarto-r#217](https://github.com/quarto-dev/quarto-r/issues/217)).
The same likelihood path is checked in `tests/testthat/`.

```{r}
#| label: fig-toy-loglik
set.seed(20260828)
rho <- 0.6
R <- matrix(c(1, rho, rho, 1), nrow = 2)
Sigma <- array(c(R, R), dim = c(2, 2, 2))
dimnames(Sigma) <- list(
  rownames(mu_small),
  rownames(mu_small),
  colnames(mu_small)
)
y_one <- DeCovarT::simulate_bulk_mixture(
  mu_small,
  Sigma,
  p = p_bal,
  n = 1
)$Y[, 1]
p1_grid <- seq(0.02, 0.98, by = 0.01)
ll_grid <- vapply(
  p1_grid,
  function(p1) {
    DeCovarT::loglik_multivariate(c(p1, 1 - p1), y_one, mu_small, Sigma)
  },
  numeric(1)
)
ll_df <- data.frame(p1 = p1_grid, loglik = ll_grid)
ggplot2::ggplot(ll_df, ggplot2::aes(x = p1, y = loglik)) +
  ggplot2::geom_line(colour = "#1B4F72", linewidth = 0.8) +
  ggplot2::geom_vline(
    xintercept = 0.5,
    linetype = "dashed",
    colour = "grey30"
  ) +
  ggplot2::labs(
    x = "p1 (p2 = 1 - p1)",
    y = "Log-likelihood"
  ) +
  ggplot2::theme_bw(base_size = 11)
```

# Inference {#sec-bivariate-inference}

Seven solvers are evaluated on each scenario:

| Solver | Description |
|---|---|
| `NNLS` | Non-negative least squares (mean-only baseline) |
| `DeconRNASeq` | NNLS via `lsei` [@gongDeconRNASeqStatisticalFramework2013] |
| `L-BFGS-B` | Quasi-Newton on box-constrained simplex |
| `gradient` | Projected gradient descent in ILR |
| `Newton-Raphson` | Second-order in ILR |
| `Marquardt-Levenberg` | Damped Newton via `marqLevAlg` |
| `SA` | Simulated annealing |

Performance is summarised using the ADEMP framework
[@morrisUsingSimulationStudies2019]: **bias**, **RMSE**, **coverage**
of ILR Wald intervals (Wilson intervals on the coverage *rate*),
**SE/SD**, and optimiser **failure rate**. Mean-only baselines have
no convolution Wald SE.

```{r}
#| label: inference
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "500"))
deconvolution_functions <- bivariate_toy_deconvolution_functions(
  itmax = 200L,
  epsilon = 1e-4
)
bivariate <- run_simulation_benchmark(
  scenario_config = bivariate_config,
  deconvolution_functions = deconvolution_functions,
  n = N_REPLICATES,
  cores = 1L
)
```

`config` on the in-memory benchmark still carries Shannon entropy of
$\boldsymbol{p}$ and MixSim overlap for tests. Saved artefacts split that
information: `optimisation` stores per-sample $\hat{\boldsymbol{p}}$,
elapsed time, and memory; `regression` and `monte_carlo` are the
composition and ADEMP blocks from `compute_benchmark_metrics()`.

# Visualisations {#sec-bivariate-viz}

Figures are written from the split RDS (no ADEMP refit). Density books
go to `output/fig02/density_visualisations/`; RMSE / MAE / Aitchison
tiles, raincloud, forest, similarity, and solver-dot books go to
`output/fig02/performance_visualisations/`. Tile heatmaps
(`ggplot2::geom_tile()`) show RMSE, MAE, and
Aitchison distance on the $(\rho_1,\rho_2)$ plane: one PDF per metric,
one page per meta-scenario (CLD \(\times\) variance \(\times\)
composition), seven solver panels. Scenario factors are relevelled
(small then large CLD; homoscedastic then heteroscedastic; balanced,
moderately unbalanced, highly unbalanced) when artefacts are read, and
solvers follow `nnls`, `lsei`, `SA`, `gradient`, `LBFGS`,
`Newton-Raphson`, `Marquardt-Levenberg`. Density and performance books
share the four correlation corners
$(\rho_1,\rho_2)\in\{(0,0),(-0.8,-0.8),(0.8,0.8),(-0.8,0.8)\}$
($2\times 2\times 3=12$ pages). Similarity heatmaps compute Pearson
\(r\) of paired \(\hat{\boldsymbol{p}}\) **within each corner** (not pooled
across the 9-by-9 grid), with average-linkage clustering of \(1-r\) and
a dendrogram on the right of the tiles. Wald forests drop NNLS, LSEI,
and SA; solid whiskers are \(1.96\) times the expected-Fisher Wald SE
at the true composition, dashed whiskers use the empirical Monte Carlo
SD, and bold two-line labels (one per solver, centred between cell
types) report RMSE and coverage, which are the same for \(p_1\) and
\(p_2\) on the simplex. Rainclouds plot \(\hat p\) (not the error) with
vertical lines at the two true proportions. Solver-dot pages use the
same 9-by-9 \((\rho_1,\rho_2)\) grid as the RMSE heatmaps: colour is
mean RMSE and size is mean Aitchison distance, with horizontal y-axis
labels. ggplot `data` tables are saved under `output/fig02/ggplot_rds/`.

### Expected findings

With **large mean separation** and $\rho_j = 0$ both groups of algorithms
should perform well. As $|\rho|$ increases and the means move closer, NNLS
and DeconRNASeq degrade faster than DeCovarT solvers because they ignore
covariance information. The MSE heatmap should display a characteristic
saddle: lowest error on the diagonal $\rho_1 = \rho_2$ and highest near
zero mean separation.

## See also

- Variance-driven hybrid ($G=20$, $J=3$):
  [§2.2](fig03-covariance-driven.html)
- Moment generator and ADEMP reporting:
  [How to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html)
- Regular-case MLE checks:
  [Appendix S1](supp-S1-identifiability.html)

## References
