---
title: "Derivatives of the DeCovarT generative model under simplex transforms"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Derivatives of the DeCovarT generative model under simplex transforms}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  lst-prefix: "Listing"
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup
#| include: false
#| eval: true
# CRAN vignettes use Quarto Markdown tables (no tinytable) so vignette
# rebuilds do not depend on Suggests packages that change S4 APIs.
NULL
```

DeCovarT maximises a Gaussian-convolution log-likelihood for cellular
ratios $\boldsymbol{p}\in\Delta^{J-1}$. This vignette derives the
analytic first- and second-order derivatives used by the frequentist
solvers, first with respect to unconstrained ambient proportions
$\boldsymbol{p}$, then after the additive log-ratio (ALR) map that
enforces the simplex. It closes with the expected Fisher information and
the ALR delta-method covariance that feed
[`vcov.decovart_fit()`](../reference/fit_decovart.html) and
[`confint.decovart_fit()`](../reference/fit_decovart.html).
Package entry points are
[`gradient_loglik_unconstrained()`](../reference/gradient_loglik_unconstrained.html),
[`hessian_loglik_unconstrained()`](../reference/hessian_loglik_unconstrained.html),
[`gradient_loglik_constrained()`](../reference/gradient_loglik_constrained.html),
[`hessian_loglik_constrained()`](../reference/hessian_loglik_constrained.html),
[`expected_fisher_unconstrained()`](../reference/expected_fisher_unconstrained.html)
and
[`vcov_alr_delta()`](../reference/vcov_alr_delta.html).

::: {#nte-notation .callout-note}
## Notation

- $\boldsymbol{\mu}\in\mathbb{R}^{G\times J}$: mean signature.
- $\boldsymbol{\Sigma}_j\in\mathrm{SPD}(G)$: cell-type covariance;
  $\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^{2}\boldsymbol{\Sigma}_j$
  and
  $\boldsymbol{\Theta}(\boldsymbol{p})=\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}$.
- $\boldsymbol{p}$: proportions on the open simplex (ambient score
  equations treat them as free coordinates in $(0,1)^{J}$).
- $\boldsymbol{\rho}\in\mathbb{R}^{J-1}$: ALR coordinates with reference
  part $J$; $\boldsymbol{p}=\boldsymbol{\psi}(\boldsymbol{\rho})$.
:::

# Matrix calculus reminders {#sec-matrix-calculus}

The score and Hessian of the convolution log-likelihood follow from a
handful of matrix-derivative identities. The statements below mirror the
matrix-calculus remarks in the manuscript Methods appendix.

::: {#rem-first-order-matrix}
## First-order matrix calculus

Let $\boldsymbol{A}=\boldsymbol{A}(p)$ and $\boldsymbol{B}=\boldsymbol{B}(p)$
be invertible matrix-valued functions of a scalar $p$, and let
$\boldsymbol{U}$, $\boldsymbol{V}$ be constant matrices of compatible size.
Then

$$
\begin{aligned}
\frac{\partial\det(\boldsymbol{A})}{\partial p}
&=
\det(\boldsymbol{A})\,
\mathrm{tr}\!\left(
  \boldsymbol{A}^{-1}
  \frac{\partial\boldsymbol{A}}{\partial p}
\right),
\\
\frac{\partial(\boldsymbol{U}\boldsymbol{A}\boldsymbol{V})}{\partial p}
&=
\boldsymbol{U}
\frac{\partial\boldsymbol{A}}{\partial p}
\boldsymbol{V},
\\
\frac{\partial\boldsymbol{A}^{-1}}{\partial p}
&=
-
\boldsymbol{A}^{-1}
\frac{\partial\boldsymbol{A}}{\partial p}
\boldsymbol{A}^{-1}.
\end{aligned}
$$ {#eq-matrix-first-order}

The chain rule on $\log\det$ yields

$$
\frac{\partial\log\det(\boldsymbol{A})}{\partial p}
=
\mathrm{tr}\!\left(
  \boldsymbol{A}^{-1}
  \frac{\partial\boldsymbol{A}}{\partial p}
\right),
\qquad
\frac{\partial\log\det(\boldsymbol{A}^{-1})}{\partial p}
=
-
\mathrm{tr}\!\left(
  \boldsymbol{A}^{-1}
  \frac{\partial\boldsymbol{A}}{\partial p}
\right).
$$ {#eq-logdet}

For a residual quadratic form with symmetric precision
$\boldsymbol{\Theta}$ and mean map linear in $p$,

$$
\frac{\partial}{\partial p}
\bigl(
  (\boldsymbol{y}-\boldsymbol{x}p)^{\top}
  \boldsymbol{\Theta}
  (\boldsymbol{y}-\boldsymbol{x}p)
\bigr)
=
-2\,
\boldsymbol{x}^{\top}
\boldsymbol{\Theta}
(\boldsymbol{y}-\boldsymbol{x}p).
$$ {#eq-quad-form-deriv}
:::

::: {#rem-second-order-matrix}
## Second-order matrix calculus

Differentiating the inverse once more gives

$$
\frac{\partial^{2}\boldsymbol{A}^{-1}}{\partial p_{i}\partial p_{j}}
=
\boldsymbol{A}^{-1}
\Biggl(
  \frac{\partial\boldsymbol{A}}{\partial p_{i}}
  \boldsymbol{A}^{-1}
  \frac{\partial\boldsymbol{A}}{\partial p_{j}}
  -
  \frac{\partial^{2}\boldsymbol{A}}{\partial p_{i}\partial p_{j}}
  +
  \frac{\partial\boldsymbol{A}}{\partial p_{j}}
  \boldsymbol{A}^{-1}
  \frac{\partial\boldsymbol{A}}{\partial p_{i}}
\Biggr)
\boldsymbol{A}^{-1},
$$ {#eq-matrix-second-order}

and the trace is linear,
$\partial\mathrm{tr}(\boldsymbol{A})/\partial p_{i}
=\mathrm{tr}(\partial\boldsymbol{A}/\partial p_{i})$. Combining
@eq-matrix-first-order with that linearity yields

$$
\frac{\partial^{2}\log\det(\boldsymbol{A}^{-1})}{\partial p_{i}\partial p_{j}}
=
-
\mathrm{tr}\!\left[
  \boldsymbol{A}^{-1}
  \frac{\partial^{2}\boldsymbol{A}}{\partial p_{i}\partial p_{j}}
\right]
+
\mathrm{tr}\!\left[
  \left(
    \boldsymbol{A}^{-1}
    \frac{\partial\boldsymbol{A}}{\partial p_{i}}
  \right)
  \left(
    \boldsymbol{A}^{-1}
    \frac{\partial\boldsymbol{A}}{\partial p_{j}}
  \right)
\right].
$$ {#eq-logdet-hess}
:::

::: {#nte-inner-products .callout-tip}
## Inner products in the score

Several terms are $\boldsymbol{\Theta}$-inner products of signature
columns,
$\boldsymbol{\mu}_{\cdot j}^{\top}\boldsymbol{\Theta}\boldsymbol{\mu}_{\cdot k}$.
Package helper [`.inner_product()`](../reference/dot-inner_product.html)
evaluates that bilinear form; see also the squared Mahalanobis distance
helper in the same family.
:::

# Unconstrained score and Hessian {#sec-unconstrained}

Conditional on the reference moments
$\boldsymbol{\zeta}=(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$, the
bulk follows
$\boldsymbol{y}\mid\boldsymbol{\zeta},\boldsymbol{p}
\sim\mathcal{N}_{G}\bigl(\boldsymbol{\mu}\boldsymbol{p},
\boldsymbol{\Sigma}(\boldsymbol{p})\bigr)$. Up to an additive constant
independent of $\boldsymbol{p}$,

$$
\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}(\boldsymbol{p})
=
\tfrac{1}{2}\log\det\boldsymbol{\Theta}(\boldsymbol{p})
-
\tfrac{1}{2}
(\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
\boldsymbol{\Theta}(\boldsymbol{p})
(\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p}).
$$ {#eq-loglik}

::: {#nte-halfdet .callout-important}
## The determinant term carries a factor $\tfrac{1}{2}$

Both terms of @eq-loglik are halved, because
$\log\det\boldsymbol{\Theta}=-\log\det\boldsymbol{\Sigma}$ enters the
Gaussian log-density as
$-\tfrac{1}{2}\log\det\boldsymbol{\Sigma}(\boldsymbol{p})$. DeCovarT
releases before 2.3.0 implemented
$-\log\det\boldsymbol{\Sigma}(\boldsymbol{p})$, which doubled the
determinant contribution. The objective was then *not* the stated
Gaussian model, and it disagreed with the expected Fisher information of
@sec-fisher-wald, which had always been the one implied by the correct
density. With the factor restored,
`loglik_multivariate()` matches `mvtnorm::dmvnorm(..., log = TRUE)` up to
the constant $\tfrac{G}{2}\log(2\pi)$, and
$\mathbb{E}[-\mathbf{H}]=I(\boldsymbol{p})$ holds exactly
(see @nte-obs-vs-exp). Point estimates move only slightly, because the
two terms trade off differently, but likelihood **values**, AIC,
likelihood-ratio statistics, and any curvature-based standard error do
change.
:::

::: {#nte-unconstrained-meaning .callout-warning}
## What “unconstrained” means here

@sec-unconstrained treats each $p_{j}$ as an ambient coordinate in
$(0,1)$ and ignores $\sum_{j}p_{j}=1$. The resulting gradient and
Hessian are the building blocks of the ALR chain rule in
@sec-constrained. They are **not** the coordinates used by
Marquardt–Levenberg or Newton–Raphson in the package (those optimise in
$\boldsymbol{\rho}$-space). The box-constrained L-BFGS-B path is the
exception: it works directly in $\boldsymbol{p}$ with $[0,1]^{J}$ boxes
and does not enforce the simplex exactly.
:::

Applying @eq-logdet and @eq-quad-form-deriv with
$\partial\boldsymbol{\Sigma}/\partial p_{j}=2p_{j}\boldsymbol{\Sigma}_{j}$
gives the ambient score. Colour coding matches the manuscript: purple
determinant / precision terms, blue mean residual, orange covariance
quadratic.

::: {#thm-score-unconstrained}
## Unconstrained gradient

$$
\begin{aligned}
\frac{\partial\ell}{\partial p_{j}}
&=
\underbrace{-p_{j}\,\mathrm{tr}\bigl(
  \boldsymbol{\Theta}\boldsymbol{\Sigma}_{j}
\bigr)}_{\text{purple}}
+
\underbrace{
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot j}
}_{\text{blue}}
\\
&\quad+
\underbrace{
  p_{j}\,
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{j}
  \boldsymbol{\Theta}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})
}_{\text{orange}}.
\end{aligned}
$$ {#eq-grad-unconstrained}
:::

Differentiating @eq-grad-unconstrained once more yields the ambient
Hessian $\mathbf{H}\in\mathcal{M}_{J\times J}$. Diagonal and off-diagonal
blocks keep the same colour pairing.

::: {#thm-hess-unconstrained}
## Unconstrained Hessian

For $i\in\{1,\ldots,J\}$,

$$
\begin{aligned}
\mathbf{H}_{ii}
&=
\frac{\partial^{2}\ell}{\partial p_{i}^{2}}
=
\underbrace{
  -\,\mathrm{tr}(\boldsymbol{\Theta}\boldsymbol{\Sigma}_{i})
  +
  2p_{i}^{2}\,\mathrm{tr}\bigl(
    (\boldsymbol{\Theta}\boldsymbol{\Sigma}_{i})^{2}
  \bigr)
}_{\text{purple}}
\\
&\quad
\underbrace{
  -2p_{i}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{i}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot i}
  -
  \boldsymbol{\mu}_{\cdot i}^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot i}
}_{\text{blue}}
\\
&\quad
\underbrace{
  -2p_{i}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{i}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot i}
  -
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \bigl(
    4p_{i}^{2}\boldsymbol{\Sigma}_{i}\boldsymbol{\Theta}\boldsymbol{\Sigma}_{i}
    -
    \boldsymbol{\Sigma}_{i}
  \bigr)
  \boldsymbol{\Theta}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})
}_{\text{orange}},
\end{aligned}
$$ {#eq-hess-diag}

and for $i\neq j$,

$$
\begin{aligned}
\mathbf{H}_{ij}
&=
\frac{\partial^{2}\ell}{\partial p_{i}\partial p_{j}}
=
\underbrace{
  2p_{i}p_{j}\,
  \mathrm{tr}\bigl(
    \boldsymbol{\Theta}\boldsymbol{\Sigma}_{j}
    \boldsymbol{\Theta}\boldsymbol{\Sigma}_{i}
  \bigr)
}_{\text{purple}}
\\
&\quad
\underbrace{
  -2p_{i}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{i}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot j}
  -
  \boldsymbol{\mu}_{\cdot i}^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot j}
}_{\text{blue}}
\\
&\quad
\underbrace{
  -2p_{j}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{j}
  \boldsymbol{\Theta}
  \boldsymbol{\mu}_{\cdot i}
  -
  4p_{i}p_{j}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})^{\top}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{i}
  \boldsymbol{\Theta}
  \boldsymbol{\Sigma}_{j}
  \boldsymbol{\Theta}
  (\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p})
}_{\text{orange}}.
\end{aligned}
$$ {#eq-hess-off}
:::

The stationary equation $\nabla\ell(\boldsymbol{p})=\boldsymbol{0}$ has
no closed root, so DeCovarT uses iterative solvers. Analytic gradients
and Hessians are checked against Richardson extrapolation in
`numDeriv` [@R-numDeriv].

# Constrained optimisation on the simplex {#sec-constrained}

Constrained optimisation may eliminate constraints by reparametrisation,
enforce them softly with penalties or barriers, or enforce them exactly
with projections / KKT methods. DeCovarT uses a $C^{2}$ diffeomorphism
from free Euclidean coordinates $\boldsymbol{\rho}\in\mathbb{R}^{J-1}$
onto the open simplex, so every Newton step stays feasible by
construction.

## Log-sum-exp and softmax {#sec-lse}

The softmax of scores $\boldsymbol{z}\in\mathbb{R}^{K}$ is evaluated
through the log-sum-exp (LSE) to avoid overflow,

$$
s_{i}
=
\frac{e^{z_{i}}}{\sum_{k}e^{z_{k}}}
=
e^{z_{i}-\operatorname{LSE}(\boldsymbol{z})},
\qquad
\operatorname{LSE}(\boldsymbol{z})
=
m+\log\sum_{k}e^{z_{k}-m},
\quad
m=\max_{k}z_{k}.
$$ {#eq-softmax}

Its Jacobian is
$\mathbf{J}_{\mathrm{softmax}}=\operatorname{diag}(\boldsymbol{s})
-\boldsymbol{s}\boldsymbol{s}^{\top}$, which is also
$\nabla^{2}\operatorname{LSE}(\boldsymbol{z})$. The second derivative is
the third-order tensor

$$
H_{ijk}
=
s_{i}\bigl[
  (\delta_{ij}-s_{j})(\delta_{ik}-s_{k})
  -
  s_{j}(\delta_{jk}-s_{k})
\bigr].
$$ {#eq-softmax-hess}

## Additive log-ratio map {#sec-alr}

Pin the last category as reference ($\rho_{J}\equiv 0$). The *additive
logistic* transform $\boldsymbol{\psi}$ (inverse ALR) and its inverse are
[@aitchisonStatisticalAnalysisCompositional1982]

$$
p_{j}
=
\frac{e^{\rho_{j}}}{\sum_{k<J}e^{\rho_{k}}+1}\ (j<J),
\qquad
p_{J}
=
\frac{1}{\sum_{k<J}e^{\rho_{k}}+1},
\qquad
\rho_{j}
=
\log\!\left(\frac{p_{j}}{p_{J}}\right).
$$ {#eq-alr-maps}

Equivalently,
$\boldsymbol{p}=\operatorname{softmax}(\rho_{1},\ldots,\rho_{J-1},0)$.
Package helpers
[`additive_logistic()`](../reference/additive_logistic.html) and
[`additive_log_ratio()`](../reference/additive_log_ratio.html) implement
@eq-alr-maps; they match `compositions::alrInv()` / `alr()` with the last
part as reference [@R-compositions].

::: {#prp-alr-diffeo}
## ALR is a $C^{2}$ diffeomorphism

$\boldsymbol{\psi}:\mathbb{R}^{J-1}\to(0,1)^{J}$ with
$\sum_{j}p_{j}=1$ is bijective and twice continuously differentiable, so
every critical point in $\boldsymbol{\rho}$-space maps to a unique open
simplex composition and conversely.
:::

## Efficient derivatives of $\boldsymbol{\psi}$ {#sec-alr-deriv}

Write $A=\sum_{j'<J}e^{\rho_{j'}}+1$ and $B_{i}=A-e^{\rho_{i}}$. The
Jacobian
$\mathbf{J}_{\boldsymbol{\psi}}\in\mathcal{M}_{J\times(J-1)}$ is the
softmax Jacobian with the reference column dropped,

$$
\frac{\partial p_{i}}{\partial\rho_{j}}
=
\begin{cases}
e^{\rho_{i}}B_{i}/A^{2}, & i=j,\ i<J,\\
-e^{\rho_{j}}e^{\rho_{i}}/A^{2}, & i\neq j,\ i<J,\\
-e^{\rho_{j}}/A^{2}, & i=J,
\end{cases}
$$ {#eq-alr-jac}

or compactly
$\partial p_{i}/\partial\rho_{a}=p_{i}(\delta_{ia}-p_{a})$ for
$a=1,\ldots,J-1$. The Hessian is a third-order tensor of shape
$(J-1)\times(J-1)\times J$; explicit case distinctions appear in the
manuscript appendix and in
[`hessian_additive_logistic()`](../reference/hessian_additive_logistic.html).
In practice DeCovarT evaluates the Jacobian / Hessian from the softmax
restriction rather than expanding every Boolean case by hand
([`jacobian_additive_logistic()`](../reference/jacobian_additive_logistic.html)).

## Change of variables: ambient $\boldsymbol{p}$ to ALR $\boldsymbol{\rho}$ {#sec-chain-rule}

Let $\ell_{\boldsymbol{\rho}}(\boldsymbol{\rho})
=\ell_{\boldsymbol{y}\mid\boldsymbol{\zeta}}
\bigl(\boldsymbol{\psi}(\boldsymbol{\rho})\bigr)$. The chain rule maps the
unconstrained derivatives of @sec-unconstrained into free Euclidean
coordinates.

::: {#thm-chain-rule}
## Constrained score and Hessian

$$
\frac{\partial\ell_{\boldsymbol{\rho}}}{\partial\rho_{j}}
=
\sum_{i=1}^{J}
\frac{\partial\ell}{\partial p_{i}}
\frac{\partial p_{i}}{\partial\rho_{j}},
$$ {#eq-chain-first}

$$
\begin{aligned}
\frac{\partial^{2}\ell_{\boldsymbol{\rho}}}{\partial\rho_{k}\partial\rho_{j}}
&=
\sum_{i=1}^{J}
\sum_{l=1}^{J}
\frac{\partial p_{i}}{\partial\rho_{j}}
\frac{\partial^{2}\ell}{\partial p_{i}\partial p_{l}}
\frac{\partial p_{l}}{\partial\rho_{k}}
\\
&\quad+
\sum_{i=1}^{J}
\frac{\partial\ell}{\partial p_{i}}
\frac{\partial^{2}p_{i}}{\partial\rho_{k}\partial\rho_{j}}.
\end{aligned}
$$ {#eq-chain-second}
:::

::: {#nte-chain-impl .callout-important}
## Implementation map

| Object | Package function |
|:--|:--|
| Ambient score / Hessian | `gradient_loglik_unconstrained()`, `hessian_loglik_unconstrained()` |
| Map Jacobian / Hessian | `jacobian_additive_logistic()`, `hessian_additive_logistic()` |
| ALR score / Hessian | `gradient_loglik_constrained()`, `hessian_loglik_constrained()` |

@eq-chain-second is the sum of a covariant pull-back of $\mathbf{H}$ and
a contraction of the ambient score against the map Hessian. Omitting the
second summand is a common bug when porting first-order code to Newton
methods.
:::

Extensions on the same footing include **ILR** or **CLR** coordinates [@pawlowsky-glahnCompositionalDataAnalysis2011] and Bayesian formulations: a Dirichlet or logistic-normal prior on $\boldsymbol{p}$ pairs naturally with the Gaussian bulk likelihood (the experimental MAP `CTS` path in `R/03_04_DeCovarT_estimate_CTS_MAP_Bayesian.R` is one starting point). Regardless of coordinate system, the workflow is the same: optimise in an unconstrained parameterisation, then map back to the simplex for interpretation.

# Expected Fisher information and Wald inference {#sec-fisher-wald}

Point estimation alone does not report uncertainty. Under a regular
large-sample regime the MLE is asymptotically normal with covariance
given by the inverse Fisher information. DeCovarT builds Wald standard
errors from the *expected* Fisher information of the multivariate-normal
convolution, then maps that information through the ALR chart.

## Unconstrained expected Fisher information {#sec-fisher-p}

For
$\boldsymbol{y}\sim\mathcal{N}_{G}\bigl(\boldsymbol{\mu}\boldsymbol{p},
\boldsymbol{\Sigma}(\boldsymbol{p})\bigr)$ the expected Fisher information
has entries

$$
I(\boldsymbol{p})_{jk}
=
\boldsymbol{\mu}_{\cdot j}^{\top}
\boldsymbol{\Theta}(\boldsymbol{p})
\boldsymbol{\mu}_{\cdot k}
+
2p_{j}p_{k}\,
\mathrm{tr}\bigl(
  \boldsymbol{\Theta}(\boldsymbol{p})\boldsymbol{\Sigma}_{j}
  \boldsymbol{\Theta}(\boldsymbol{p})\boldsymbol{\Sigma}_{k}
\bigr).
$$ {#eq-fisher-p}

The first summand is the mean contribution (a $\boldsymbol{\Theta}$-inner
product of signature columns); the second is the covariance contribution
of the quadratic map
$\boldsymbol{p}\mapsto\boldsymbol{\Sigma}(\boldsymbol{p})$. See the
multivariate-normal formula on
[Wikipedia: Fisher information](https://en.wikipedia.org/wiki/Fisher_information#Multivariate_normal_distribution).
[`expected_fisher_unconstrained()`](../reference/expected_fisher_unconstrained.html)
implements @eq-fisher-p.

::: {#nte-obs-vs-exp .callout-note}
## Observed versus expected information

The analytic Hessian of @sec-unconstrained is the *observed* information
(up to sign). Wald intervals in `decovart_fit` use the *expected*
information @eq-fisher-p, which does not depend on the residual
$\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p}$ and is positive definite under
the usual full-rank assumptions on $(\boldsymbol{\mu},\{\boldsymbol{\Sigma}_{j}\})$.

The two are consistent by construction. Substituting
$\mathbb{E}[\boldsymbol{r}]=\boldsymbol{0}$ and
$\mathbb{E}[\boldsymbol{r}\boldsymbol{r}^{\top}]
=\boldsymbol{\Sigma}(\boldsymbol{p})$ into @eq-hess-diag and
@eq-hess-off cancels the determinant traces against the residual traces
and returns @eq-fisher-p exactly, i.e.
$\mathbb{E}[-\mathbf{H}]=I(\boldsymbol{p})$. That identity only holds with
the factor $\tfrac{1}{2}$ of @nte-halfdet in place.
:::

## Constrained case: ALR delta method {#sec-fisher-rho}

Fisher information transforms as a covariant quadratic form under the
ALR chart. With
$\mathbf{J}_{\boldsymbol{\psi}}
=\partial\boldsymbol{\psi}/\partial\boldsymbol{\rho}^{\top}$,

$$
I_{\boldsymbol{\rho}}
=
\mathbf{J}_{\boldsymbol{\psi}}^{\top}
I(\boldsymbol{p})
\mathbf{J}_{\boldsymbol{\psi}}.
$$ {#eq-fisher-rho}

Under regularity,
$\hat{\boldsymbol{\rho}}
\overset{a}{\sim}
\mathcal{N}(\boldsymbol{\rho}_{0},I_{\boldsymbol{\rho}}^{-1})$
[@vaartAsymptoticStatistics2000].
The first-order delta method [@oehlertNoteDeltaMethod1992] then returns
the simplex covariance used by
[`vcov.decovart_fit()`](../reference/fit_decovart.html),

$$
\mathrm{Var}(\hat{\boldsymbol{p}})
\approx
\mathbf{J}_{\boldsymbol{\psi}}
I_{\boldsymbol{\rho}}^{-1}
\mathbf{J}_{\boldsymbol{\psi}}^{\top}.
$$ {#eq-delta-p}

Diagonal square roots are asymptotic standard errors; Wald intervals at
level $1-\alpha$ are
$\hat{p}_{j}\pm z_{1-\alpha/2}\,\mathrm{SE}_{j}$ with
$z_{q}=\Phi^{-1}(q)$, as in
[`confint.decovart_fit()`](../reference/fit_decovart.html).
[`vcov_alr_delta()`](../reference/vcov_alr_delta.html) implements
@eq-fisher-rho--@eq-delta-p. The construction is undefined on the
simplex boundary (the ALR chart blows up); the helper then returns `NA`
with a warning. The ALR chart itself follows Aitchison's compositional
geometry [@aitchisonStatisticalAnalysisCompositional1982].

::: {#nte-wald-limits .callout-warning}
## Wald is the cheapest, not the safest, interval

@eq-delta-p linearises the ALR chart at $\hat{\boldsymbol{p}}$, so it is
neither invariant to reparametrisation nor confined to $[0,1]$, and it
degenerates entirely when a proportion reaches a simplex face. The
companion vignette
[MLE properties and asymptotic inference](https://bastienchassagnol.github.io/DeCovarT/articles/DeCovarT-MLE-properties.html)
derives the alternatives: profile likelihood-ratio intervals, the
chi-bar-square calibration of Chernoff and of Self and Liang for nulls on
the boundary, a parametric bootstrap, and a reference-sample bootstrap
of donors or cells (or a Dirichlet composition sweep). It also shows why
these limits need replicate samples that share one composition.
Shuffling genes or cell-type names of an averaged signature is not a
bootstrap: the maximised likelihood is equivariant under relabelling.
:::

::: {.callout-warning}
## Numerical speed-ups and solver safeguards {#sec-numerical-speedups}

The analytic maps above are only half of a usable optimiser. Practical
bottlenecks that showed up on the hybrid scenario in
[Manuscript synthetic simulation scenarios](DeCovarT-manuscript-scenarios.html#sec-hybrid-deconvolution)
live in `R/03_03_DeCovarT_estimate_ratios_frequentist.R`.

---

### Cache a Cholesky factorisation of $\boldsymbol{\Sigma}(\boldsymbol{p})$

Within one iteration, objective / gradient / Hessian callbacks hit the
same trial $\boldsymbol{p}$. The helper `.sigma_p_factorisation()` caches
a single Cholesky factor
$\boldsymbol{R}$ with
$\boldsymbol{\Sigma}(\boldsymbol{p})=\boldsymbol{R}^{\mathsf{T}}\boldsymbol{R}$
and returns
$\log\det\boldsymbol{\Sigma}(\boldsymbol{p})=2\sum_g\log R_{gg}$
together with the precision
$\boldsymbol{\Theta}(\boldsymbol{p})=\boldsymbol{R}^{-1}\boldsymbol{R}^{-\mathsf{T}}$
(via `chol2inv`) without repeating an $O(G^{3})$ factorisation.

[`loglik_multivariate()`](../reference/loglik_multivariate.html) then
evaluates the Mahalanobis term exactly as `mvtnorm::dmvnorm`
[@R-mvtnorm]: it solves $\boldsymbol{R}^{\mathsf{T}}\boldsymbol{z}
=\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p}$ by
`backsolve(..., transpose = TRUE)` and takes
$\lVert\boldsymbol{z}\rVert^{2}$, rather than forming
$\boldsymbol{\Theta}(\boldsymbol{p})$ and a dense quadratic form. The two
routes agree to machine precision; the backsolve path is the one used
for the objective, while the cached inverse is retained for the analytic
score and Hessian, which need
$\boldsymbol{\Theta}(\boldsymbol{p})$ in several trace and inner-product
terms. A QR factorisation of the covariance itself would recover the
same log-determinant and quadratic form at a larger $O(G^{3})$ constant
and is not used: Cholesky is the natural factorisation of a symmetric
positive-definite matrix.

The implemented log-density omits the additive
$-\tfrac{G}{2}\log(2\pi)$ of the full Gaussian, which does not depend
on $\boldsymbol{p}$ and therefore cannot change the MLE, the score, or
the Hessian. Tests check that subtracting that constant recovers
`dmvnorm(..., log = TRUE)` exactly.

---

### Guard the box-constrained L-BFGS-B path

`deconvolute_ratios_L_BFGS_B()` optimises in $\boldsymbol{p}$ with boxes
$[0,1]^{J}$ that do not enforce $\mathbf{1}^{\top}\boldsymbol{p}=1$. Near
$\sum_{j}p_{j}\approx 0$ (or on a failed `chol()`), safeguarded wrappers
return a finite penalty and a zero gradient rather than aborting
`optim()`.

---

### `marqLevAlg` Hessian sign under `minimize = FALSE`

[`marqLevAlg`](https://cran.r-project.org/package=marqLevAlg)
[@R-marqLevAlg] implements Marquardt–Levenberg with the
relative-distance-to-minimum (RDM) rule of Commenges et al.
[@commengesNewtonLikeAlgorithmLikelihood2006]. With
`minimize = FALSE`, the package negates only `fn` and `gr`; the analytic
`hess` is passed through unchanged while internal Cholesky routines
assume a **positive-definite** matrix at the optimum. When maximising a
log-likelihood, pass
`hess = function(...) -hessian_loglik_constrained(...)`. Reported
upstream as
[VivianePhilipps/marqLevAlgParallel#3](https://github.com/VivianePhilipps/marqLevAlgParallel/issues/3).
:::

# Model objects and S3 accessors {#sec-s3}

[`fit_decovart()`](../reference/fit_decovart.html) returns a `decovart_fit`
object with standard S3 methods for coefficients, fitted values, residuals,
and (under regularity) a delta-method covariance. The chunk below mirrors
the appendix equivariance setup:

```{r}
#| label: s3-accessors
#| warning: false
#| eval: true
library(DeCovarT)
genes <- paste0("g", 1:3)
cts <- paste0("ct", 1:2)
mu <- matrix(
  c(20, 40, 25, 35, 30, 22),
  nrow = 3,
  dimnames = list(genes, cts)
)
Sigma <- array(0, dim = c(3, 3, 2), dimnames = list(genes, genes, cts))
Sigma[, , 1] <- diag(c(1.0, 1.2, 0.8))
Sigma[, , 2] <- diag(c(1.5, 0.9, 1.1))
p_true <- c(0.65, 0.35)
y <- matrix(
  drop(mu %*% p_true + rnorm(3, sd = 0.05)),
  ncol = 1L,
  dimnames = list(genes, "s1")
)

fit <- fit_decovart(
  mu,
  y,
  Sigma = Sigma,
  method = "Marquardt-Levenberg",
  itmax = 80
)
coef(fit)
nobs(fit)
c(
  n_genes = attr(nobs(fit), "n_genes"),
  n_celltypes = attr(nobs(fit), "n_celltypes")
)
fitted(fit)
residuals(fit)
vcov(fit)
```

`formula()` and `predict()` are not implemented: DeCovarT does not
forecast bulk expression.

# Numerical consistency of the likelihood and its derivatives {#sec-numerical-consistency}

Analytic formulae can be differentiated correctly *and* still implement
the wrong objective. The checks below separate those two questions.
They are the same identities that
`tests/testthat/test-03_03_DeCovarT.R` and
`tests/testthat/test-03_06_decovart_inference.R` enforce on every
package build; the chunks here are a readable, small-\(G\) illustration.

```{r}
#| label: consistency-setup
#| eval: true
#| include: false
library(DeCovarT)
genes <- paste0("g", 1:3)
cts <- paste0("ct", 1:3)
mu_chk <- matrix(
  c(20, 40, 15, 40, 20, 25, 25, 30, 35),
  nrow = 3,
  dimnames = list(genes, cts)
)
Sigma_chk <- array(0, dim = c(3, 3, 3), dimnames = list(genes, genes, cts))
Sigma_chk[, , 1] <- diag(c(1, 1.5, 1.2))
Sigma_chk[, , 2] <- diag(c(1.1, 0.9, 1.3))
Sigma_chk[, , 3] <- diag(c(0.8, 1.2, 1.0))
p_chk <- c(0.5, 0.3, 0.2)
set.seed(1)
y_chk <- drop(mu_chk %*% p_chk) + c(0.4, -0.2, 0.1)
cov_chk <- .compute_global_variance(p_chk, Sigma_chk)
```

## Density against `mvtnorm`

The first check is against an independent implementation of the
multivariate Gaussian log-density, not against our own inverse. With
the $\tfrac{1}{2}\log\det$ factor restored, DeCovarT and
`mvtnorm::dmvnorm` [@R-mvtnorm] differ by exactly
$\tfrac{G}{2}\log(2\pi)$ (DeCovarT omits the negative constant):

```{r}
#| label: consistency-dmvnorm
#| eval: true
n_genes <- length(y_chk)
ours <- loglik_multivariate(p_chk, y_chk, mu_chk, Sigma_chk)
reference <- if (requireNamespace("mvtnorm", quietly = TRUE)) {
  as.numeric(mvtnorm::dmvnorm(
    y_chk,
    mean = drop(mu_chk %*% p_chk),
    sigma = cov_chk,
    log = TRUE
  ))
} else {
  residual <- y_chk - drop(mu_chk %*% p_chk)
  as.numeric(
    -0.5 * determinant(cov_chk, logarithm = TRUE)$modulus -
      0.5 * drop(crossprod(residual, solve(cov_chk, residual))) -
      0.5 * n_genes * log(2 * pi)
  )
}
c(
  loglik_multivariate = ours,
  dmvnorm = reference,
  difference = ours - 0.5 * n_genes * log(2 * pi) - reference
)
```

A finite-difference check of the *same* objective cannot detect a shared
factor-of-two error in the log-likelihood and its derivatives. That is
why this density comparison is independent of `numDeriv`.

## Richardson extrapolation of the score and Hessian

Once the objective is the intended Gaussian, the analytic score and
Hessian are checked against Richardson extrapolation in `numDeriv`
[@R-numDeriv]. Richardson's tableau extrapolates central differences to
cancel successive even powers of the step size, which is more accurate
than a single two-point stencil at the same $O(\varepsilon^{-1})$ cost
in function evaluations.

```{r}
#| label: consistency-numderiv
#| eval: true
if (!requireNamespace("numDeriv", quietly = TRUE)) {
  numderiv_report <- c(gradient_rel = NA_real_, hessian_rel = NA_real_)
} else {
  grad_n <- numDeriv::grad(
    loglik_multivariate,
    p_chk,
    method = "Richardson",
    method.args = list(eps = 1e-4, r = 6),
    y = y_chk,
    mean_signature_matrix = mu_chk,
    Sigma = Sigma_chk
  )
  hess_n <- numDeriv::hessian(
    loglik_multivariate,
    p_chk,
    method = "Richardson",
    method.args = list(eps = 1e-4, r = 4),
    y = y_chk,
    mean_signature_matrix = mu_chk,
    Sigma = Sigma_chk
  )
  grad_a <- gradient_loglik_unconstrained(
    p_chk, y_chk, mu_chk, Sigma_chk
  )
  hess_a <- hessian_loglik_unconstrained(
    p_chk, y_chk, mu_chk, Sigma_chk
  )
  numderiv_report <- c(
    gradient_rel = max(abs(grad_a - grad_n)) / max(abs(grad_n)),
    hessian_rel = max(abs(hess_a - hess_n)) / max(abs(hess_n))
  )
}
numderiv_report
```

Relative discrepancies of order \(10^{-8}\) are typical on this
three-gene toy; that is the same tolerance used in the test suite.

## Expected information versus Monte Carlo \(\mathbb{E}[-\mathbf{H}]\)

Matching `numDeriv` shows that the derivatives belong to the implemented
objective. Matching the *expected* Fisher information
[@eq-fisher-p] to a Monte Carlo average of \(-\mathbf{H}(\boldsymbol{p};
\boldsymbol{Y})\) shows that the objective is the Gaussian whose
information matrix we use for Wald intervals. Draw
\(\boldsymbol{Y}^{(b)}\sim\mathcal{N}_G(\boldsymbol{\mu}\boldsymbol{p},
\boldsymbol{\Sigma}(\boldsymbol{p}))\) at the *true* \(\boldsymbol{p}\)
(the parametric bootstrap of
[`bootstrap_decovart()`](../reference/bootstrap_decovart.html), used here
as a diagnostic rather than as an inferential procedure):

```{r}
#| label: consistency-fisher-mc
#| eval: true
set.seed(2)
n_mc <- 40L
info <- expected_fisher_unconstrained(p_chk, mu_chk, Sigma_chk)
neg_hess <- array(NA_real_, c(3L, 3L, n_mc))
for (b in seq_len(n_mc)) {
  y_b <- MASS::mvrnorm(
    n = 1L,
    mu = drop(mu_chk %*% p_chk),
    Sigma = cov_chk
  )
  neg_hess[, , b] <- -hessian_loglik_unconstrained(
    p_chk, y_b, mu_chk, Sigma_chk
  )
}
mc_info <- apply(neg_hess, c(1L, 2L), mean)
c(
  rel_frobenius = sqrt(sum((mc_info - info)^2)) / sqrt(sum(info^2))
)
```

The same identity is what closed the factor-of-two inconsistency: with
the old \(-\log\det\) objective, \(\mathbb{E}[-\mathbf{H}]\) could not
equal \(I(\boldsymbol{p})\). A few dozen draws already bring the relative
Frobenius gap below \(10^{-2}\) on this toy; the test suite uses a
tighter Monte Carlo.

A related, weaker check compares the empirical covariance of
parametric-bootstrap MLEs to the Cramér–Rao map
[`vcov_alr_delta()`](../reference/vcov_alr_delta.html). That comparison
is noisy at small replication (see
[MLE properties](https://bastienchassagnol.github.io/DeCovarT/articles/DeCovarT-MLE-properties.html#sec-replication)
for why one bulk column is not an asymptotic sample) and is therefore
left to the inference vignette and to
[`bootstrap_decovart()`](../reference/bootstrap_decovart.html).

# References
