---
title: "Simulating high-dimensional Gaussian Markov regulatory networks"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating high-dimensional Gaussian Markov regulatory networks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

A Gaussian Markov network simulation has four conceptually separate layers (@eq-pipeline). For non-Gaussian expression data, a fifth observation layer can be added (e.g. Poisson–log-normal counts or zero-inflated measurements).

$$
G \longrightarrow W(G) \longrightarrow \Omega \succ 0
\longrightarrow X_i \sim \mathcal{N}_p(\mu_i,\Omega^{-1})
$$ {#eq-pipeline}

Here $G$ is an **undirected** graph, $W(G)$ a symmetric weighted matrix with the same off-diagonal support, $\Omega$ a strictly positive-definite precision, and $X_i$ an expression profile. The **graph** and the **precision** must not be conflated: a binary adjacency is almost never itself positive definite, so a separate weighting and regularisation step is required.

This vignette covers **undirected** Gaussian Markov / precision-graph simulation only.

DeCovarT’s synthetic helpers follow the same split ([synthetic scenarios vignette](synthetic-scenarios.html)): `generate_random_network_skeleton()` draws $G$, and `build_normalised_precision()` completes it to $\Omega\succ 0$ by an affine spectral shift.

```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "Undirected simulation pipeline from topology to observations (optional observation layer dashed)."
flowchart LR
  G["Undirected graph G"] --> W["Signed weights W(G)"]
  W --> Omega["SPD precision Ω"]
  Omega --> Latent["Latent Gaussian X"]
  Mu["Mean design μ"] --> Latent
  Latent --> Obs["Observation layer"]
  Latent -.-> Obs
```

# Graph structure generators

## Evidence from the literature

@tbl-studies summarises how recent simulation studies combine topology, precision construction, and mean / observation models.

| Source | Graph structures | Precision construction | Mean / observation | Purpose |
|---------------|---------------|---------------|---------------|---------------|
| PLNnetwork [@chiquetVariationalInferenceSparse2018] | Erdős–Rényi, preferential attachment, affiliation | $\Omega = vG + \operatorname{diag}(\lvert\lambda_{\min}(vG)\rvert + u)$ | Latent log-abundance $XB$; compositional / multinomial counts | Count-network estimators under compositionality |
| SILGGM [@zhangSILGGMExtensivePackage2018] | Band, hub, Erdős–Rényi, scale-free via `huge` | Package covariance / precision simulation | Centred Gaussian; focus on precision entries | Large-$p$ inferential and computational stress tests |
| BLGGM [@wuEstimatingHeterogeneousGene2022] | Block mixtures of dense, circle, star, signed modules | Module matrices with structured signed off-diagonals | Cell-type-specific $\mu_k$, $\Omega_k$; expression-dependent dropout | Joint clustering, heterogeneous networks, zero inflation |
| Structure learning / modular GRNs [@federicoStructureLearningGene2023] | Modular and related hierarchical graphs | GGM linked to each graph | Shared structure across related networks; $n\ll p$ | Modular constraints and information sharing |

: Literature survey of undirected graph → precision → observation designs. {#tbl-studies}

Several conclusions follow.

1.  **Band** is not intrinsically a random-graph model: its support is usually deterministic once the bandwidth is fixed (edge weights and signs may still be random). Hub constructions are often partly deterministic as well. Erdős–Rényi, preferential attachment, and stochastic block models are genuinely stochastic.
2.  A positive off-diagonal entry in $\Omega$ encodes a **negative** partial correlation (@eq-partial-cor). Uniformly positive precision weights therefore induce uniformly negative edgewise partial correlations; random or signed biological weights are needed when activation-like and inhibition-like associations matter.
3.  These models target **undirected** conditional dependence. Federico and colleagues prefer a Markov-network representation when observational data do not justify edge directionality [@federicoStructureLearningGene2023].
4.  **BLGGM as a hybrid topology design.** Wu and Luo’s simulation can be read as two nested layers [@wuEstimatingHeterogeneousGene2022]. At the **global** scale, genes are partitioned into blocks (a stochastic-block–like partition). Within each block, a **finer** local topology is specified—dense clique-like modules, circles, stars / hubs, or signed dense substructures—so that different blocks can stand for distinct regulatory regimes (e.g. dense co-expression programmes, cyclic signalling motifs, master-regulator hubs). Signs can also differ across modules or cell types. This is richer than a flat ER draw, yet still fully undirected.

$$
\rho_{jk\mid -\{j,k\}}
=
-\frac{\Omega_{jk}}{\sqrt{\Omega_{jj}\Omega_{kk}}}
$$ {#eq-partial-cor}


@tbl-topologies compares the main **undirected** topology families used in GGM benchmarks, with R entry points. @fig-topologies sketches the six families most often used in random-like network structure generation.

| Structure | Definition and controls | R entry points | Advantages | Limitations |
|---------------|---------------|---------------|---------------|---------------|
| **Erdős–Rényi** | Each edge present independently with probability $q$; $q=d_{\mathrm{avg}}/(p-1)$ targets average degree | `huge::huge.generator(graph="random")` [@R-huge]; `BDgraph::graph.sim(graph="random")` [@R-BDgraph] | Exact sparsity control in expectation; few nuisance parameters | Narrow degree distribution; little modular organisation |
| **Band / AR** | Edge $j$–$k$ if $1\le\lvert j-k\rvert\le b$ | `huge.generator(graph="band")` [@R-huge]; `BDgraph::bdgraph.sim(graph="AR(1)")` / `"AR(2)"` [@R-BDgraph] | Local dependence and controlled max degree | Artificial gene ordering; no hubs or modules |
| **Hub / star** | $g$ groups; one centre linked to remaining members | `huge.generator(graph="hub")` [@R-huge]; `graph.sim(graph="hub")` / `"star"` [@R-BDgraph] | Stress-tests high-degree regulators | Pure stars omit cross-talk |
| **Scale-free** | Preferential attachment (e.g. Barabási–Albert with $m$ edges per new node) | `huge.generator(graph="scale-free")` [@R-huge]; DeCovarT `graph_model = "scale_free"` | Heterogeneous degrees | Weak modularity under plain BA growth |
| **Cluster / SBM** | Within-block edge probability exceeds between-block | `huge.generator(graph="cluster")` [@R-huge]; DeCovarT `graph_model = "stochastic_block_model"` | Pathway / module structure | Equal blocks can be unrealistically regular |
| **Small-world / lattice / circle** | High clustering with short paths, or fixed neighbourhoods | DeCovarT `graph_model = "small_world"` (Watts–Strogatz); `BDgraph::graph.sim(graph="smallworld")` [@R-BDgraph] | Local modules and spatial organisation | Narrow degrees; needs a defensible ordering |

: Topology families for undirected GGM simulation (R-focused). {#tbl-topologies}

## Biological relevance of the six families

Not every topology is equally useful as a **gene-regulatory** prior. The three
simulation studies in @tbl-studies, together with classical network biology,
suggest the following reading.

| Family | Biological reading | Why it matters for DeCovarT |
|--------|--------------------|-----------------------------|
| **Star / hub** | Master-regulator TF with many targets; one-to-many signalling | Stress-tests recovery of a few high-degree hubs that dominate partial correlations [@wuEstimatingHeterogeneousGene2022; @federicoStructureLearningGene2023] |
| **Scale-free (BA)** | Preferential attachment produces a heavy-tailed degree distribution and several hubs [@barabasiEmergenceScalingRandom1999] | Standard high-dimensional GGM stress test used by SILGGM via `huge` [@zhangSILGGMExtensivePackage2018]; useful for testing degree heterogeneity, but not a default empirical model of gene regulation |
| **Cluster / SBM** | Co-regulated modules / pathways with dense within-block links | Matches pathway organisation; BLGGM’s global block partition is SBM-like before finer within-block motifs [@wuEstimatingHeterogeneousGene2022] |
| **Small-world** | High local clustering with short paths (ring + shortcuts) | Local co-expression neighbourhoods plus long-range regulatory shortcuts; useful when modularity is soft rather than hard blocks [@federicoStructureLearningGene2023] |
| **Band / AR** | Ordered local dependence only | Useful numerical control, but gene indices rarely encode a natural order—limited biological fidelity [@zhangSILGGMExtensivePackage2018] |
| **Erdős–Rényi** | Homogeneous random wiring | Null / baseline sparsity; little pathway or hub structure [@zhangSILGGMExtensivePackage2018] |

: Biological interpretability of undirected topology families. {#tbl-bio-topologies}

### The scale-free hypothesis: a useful stress test, not a biological default

The phrase *scale-free network* usually means that the degree distribution
follows a power law, $\Pr(K=k)\propto k^{-\alpha}$, at least above a lower
cut-off. This statistical statement is distinct from the Barabási–Albert
(BA) growth mechanism. Preferential attachment can generate a power law, but
observing a heavy-tailed degree distribution does not establish preferential
attachment; other mechanisms can produce similar tails, and variants of
preferential attachment need not produce a pure power law
[@lima-mendezPowerfulLawPower2009;
@broidoScalefreeNetworksAre2019].

Whether biological networks are scale-free has therefore remained contested.
Early claims often relied on approximately straight log–log degree plots.
Such plots are not goodness-of-fit tests: binning, the plotting scale, a small
number of hubs, and incomplete or biased sampling can all make non-power-law
distributions appear linear. The underlying graph representation matters as
well. For example, pool metabolites can become artificial hubs in metabolic
graphs, while protein-interaction networks depend strongly on the assay and
sampling scheme [@lima-mendezPowerfulLawPower2009]. A power law should instead
be fitted to the discrete degree data, tested for plausibility, and compared
on the same support with alternatives such as log-normal, exponential,
stretched-exponential, and power-law-with-cut-off distributions.

The evidence is not uniformly negative. Using adjacency-spectrum
distributions rather than degree alone, Takahashi and colleagues classified
eight protein–protein interaction networks as scale-free
[@takahashiDiscriminatingDifferentClasses2012]. Importantly, this was a
relative model-selection result: BA-like networks fitted better than the two
candidate alternatives, Erdős–Rényi and Watts–Strogatz networks. It did not
show that a power law was an adequate absolute model, that preferential
attachment generated the observed networks, or that unobserved interactions
would preserve the classification. The authors explicitly noted both the
restricted candidate set and possible sampling artefacts.

A broader test of 928 real-world network data sets reached a more cautious
conclusion [@broidoScalefreeNetworksAre2019]. Across all domains, only 4% met
the strongest scale-free criterion, whereas a log-normal distribution fitted
as well as or better than a power law for 88% of degree distributions. Among
the biological networks, 63% showed neither direct nor indirect evidence of
scale-free structure, although 6% met the strongest criterion, principally
metabolic networks. These proportions depend on the network corpus and the
operational definition of scale-freeness, but they reject a universal
scale-free law rather than excluding scale-free structure from every
biological system.

For DeCovarT, a BA graph should consequently be interpreted as a deliberately
demanding **hub-heterogeneity benchmark**. It tests whether covariance
estimation and deconvolution recover a few high-degree vertices without
claiming that a gene-regulatory network grew by preferential attachment.
Results should be compared with density-matched SBM, small-world, star, and
Erdős–Rényi scenarios. Conclusions that depend only on the BA scenario should
be labelled topology-specific; hub recovery, robustness to vertex removal,
and biological realism do not follow from a fitted heavy tail alone.

**BLGGM as a hybrid.** Wu and Luo combine an SBM-like **global** gene partition
with **local** dense, circle, star, or signed modules inside blocks
[@wuEstimatingHeterogeneousGene2022]. Distinct local motifs can stand for
distinct processes (e.g. master-regulator stars vs dense co-expression
programmes). DeCovarT’s exported generators currently expose the three
families most useful as standalone benchmarks: `scale_free`,
`stochastic_block_model`, and `small_world`.

## Package roles (R)

-   **`igraph`** [@R-igraph] — DeCovarT draws the three undirected skeletons used in simulation (`scale_free` via `sample_pa()`, `stochastic_block_model` via `sample_sbm()`, `small_world` via `sample_smallworld()`).
-   **`huge`** [@R-huge] — compact end-to-end GGM benchmark via `huge.generator()`: topology, SPD precision, covariance, and $\mathcal{N}_p$ samples. Supports random, hub, cluster, band, and scale-free graphs [@zhangSILGGMExtensivePackage2018]; useful as a literature comparator, not the DeCovarT skeleton backend.
-   **`BDgraph`** [@R-BDgraph] — `graph.sim()` / `bdgraph.sim()` cover overlapping families plus small-world and scale-free; `rgwish()` samples G-Wishart precisions on a fixed graph.
-   **`bnlearn`** [@R-bnlearn] — directed / DAG and some undirected structure learning; helpful when the scientific story is causal or when comparing directed versus undirected ground truths.
-   Prefer keeping the **same** topology → weight → SPD → mean pipeline when swapping generators, so differences reflect the graph family rather than an incidental change of SPD or sampling code.

![Six undirected topology families used in GGM simulation: Erdős–Rényi, Barabási–Albert (preferential attachment / scale-free), stochastic block (cluster), band / AR, star / hub, and small-world (Watts–Strogatz).\
\
](figures/fig_network_topologies_six.png){#fig-topologies}

# Weight design and random signs

Topology generators return a binary undirected support $E$. **Signs and magnitudes** are a separate design layer $W(G)$ in @eq-pipeline: they are assigned *after* (or jointly with) the skeleton, then completed to an SPD precision. Remember that

$$
\operatorname{sign}(\rho_{jk\mid -\{j,k\}})
=
-\operatorname{sign}(\Omega_{jk})
$$ {#eq-sign-flip}

(@eq-partial-cor), so a positive precision weight is an inhibitory partial correlation. @tbl-weights contrasts three standard undirected strategies.

| Approach | What is randomised | How $\Omega$ is obtained | When to use |
|------------------|------------------|------------------|------------------|
| **i.i.d. edge signs** | For each $\{j,k\}\in E$, draw $s_{jk}\in\{-1,+1\}$ (e.g. fair coin or $\mathbb{P}(+)=\pi$) and magnitude $m_{jk}>0$; set $W_{jk}=W_{kj}=s_{jk}m_{jk}$ | Support-preserving SPD map of $W$ (spectral shift, diagonal dominance, …) | Simple signed ER / hub / SBM benchmarks; explicit control of the inhibitory fraction |
| **Partial-correlation scaling** | Fill a symmetric matrix $R$ with $R_{jk}=0$ off $E$ and $R_{jk}\in(-1,1)$ (signed) on $E$; ensure $\rho(R)<1$ | $\Omega = D^{1/2}(I-R)D^{1/2}$ (@eq-partial-scale) | Direct control of partial-correlation signs and strengths |
| **G-Wishart** (`BDgraph::rgwish()`) | Continuous random $\Omega$ on the fixed graph zeros | $\Omega\sim W_G(b,D)$ already SPD [@R-BDgraph; @BDgraph2019] | Heterogeneous random weights without hand-tuned magnitudes; Bayesian-style replicates |

: Three undirected strategies for signed / random edge weights after a fixed skeleton. {#tbl-weights}

## i.i.d. signs on a fixed skeleton

Draw $G$ (ER, band, hub, …). Independently for each undirected edge,

$$
W_{jk}=W_{kj}=s_{jk}\,m_{jk},
\qquad
s_{jk}\in\{-1,+1\},
\quad m_{jk}>0,
$$ {#eq-iid-signs}

then apply a support-preserving SPD completion (next section). Uniform positive loadings $W=vG$ are the special case $s_{jk}\equiv +1$ used by PLN-style and many `huge` simulations [@chiquetVariationalInferenceSparse2018]—all partial correlations then share the same sign.

## Partial-correlation scaling and why $I-R$

In a Gaussian Markov model the **matrix of partial correlations** (with ones on the diagonal) relates to the precision by a diagonal congruence. Write the off-diagonal partial correlations as a symmetric matrix $R$ with

$$
R_{jj}=0,
\qquad
R_{jk}
=
\begin{cases}
\rho_{jk\mid -\{j,k\}} & \{j,k\}\in E,\\
0 & \{j,k\}\notin E,
\end{cases}
$$ {#eq-R-support}

and choose magnitudes so that the spectral radius satisfies $\rho(R)<1$. Then

$$
\Omega
=
D^{1/2}(I-R)D^{1/2}
$$ {#eq-partial-scale}

for a positive diagonal $D$ (often $D=I$ after standardising margins).

**Why subtract from the identity?** The precision encodes *conditional* dependence. For a correlation-type matrix, $I-R$ places **ones on the diagonal** (self-precision after scaling) and **negatives of the partial correlations** off the diagonal, matching @eq-sign-flip: if $R_{jk}>0$ (positively partially correlated variables), then $(I-R)_{jk}<0$, hence $\Omega_{jk}<0$.

## G-Wishart draws on a fixed graph

Given adjacency $G$, `BDgraph::rgwish()` samples $\Omega\sim W_G(b,D)$ already positive definite and with $\Omega_{jk}=0$ whenever $(j,k)\notin E$ [@R-BDgraph; @BDgraph2019]. Signs and magnitudes arise from the continuous distribution rather than from an explicit coin-flip layer. This is the most “automatic” signed-weight generator once the skeleton is fixed; the trade-off is less direct control of the inhibitory fraction than @eq-iid-signs or @eq-partial-scale.

# Positive-definiteness strategies

@tbl-pd contrasts constructions that turn a weighted support $W$ (or a partial-correlation design) into $\Omega\succ 0$.

| Construction | Formula | SPD? | Preserves exact support? | Behaviour |
|---------------|---------------|--------------:|--------------:|---------------|
| Uniform spectral shift | $\Omega=W+\max(0,\varepsilon-\lambda_{\min}(W))I$ | Yes | Yes | Same shift on every eigenvalue; off-diagonals (hence signs) unchanged |
| `huge` / PLN-style loading | Diagonal load from $\lvert\lambda_{\min}\rvert$, baseline, and $u$ | Yes | Yes | $v$ sets edge magnitude; loading sets conditioning [@chiquetVariationalInferenceSparse2018] |
| Partial-correlation scaling | $\Omega=D^{1/2}(I-R)D^{1/2}$ with $\rho(R)<1$ | Yes | Yes | Signs and strengths set on the partial-correlation scale (@eq-partial-scale) |
| Graph Laplacian / M-matrix | $\Omega=L_G+\varepsilon I$ | Yes | Yes | Off-diagonals non-positive ⇒ all edgewise partial correlations positive |
| G-Wishart | $\Omega\sim W_G(b,D)$ | Yes | Yes | Heterogeneous random weights; `BDgraph::rgwish()` [@R-BDgraph] |
| Eigenvalue clipping | Clip $\Lambda$ then reconstruct $Q\Lambda Q^{\top}$ | Yes | **No** | Fills structural zeros |
| Nearest PD projection | e.g. `Matrix::nearPD()` [@R-Matrix] | Yes | **No** (in general) | Repair tool, not a support-preserving truth |

: Positive-definiteness constructions for graph-constrained precisions. {#tbl-pd}

## Why the uniform spectral shift is attractive

If $W=Q\Lambda Q^{\top}$, then (@eq-spectral-shift)

$$
W+cI = Q(\Lambda+cI)Q^{\top}.
$$ {#eq-spectral-shift}

Every eigenvalue increases by $c$, while every off-diagonal of $W$ is unchanged, so the edge set

$$
E=\{(j,k):j<k,\ \Omega_{jk}\neq 0\}
$$ {#eq-edge-set}

is identical before and after loading—and **signs of** $W$ are preserved.

The floor $\varepsilon$ also controls the condition number

$$
\kappa(\Omega)=\frac{\lambda_{\max}(\Omega)}{\lambda_{\min}(\Omega)}.
$$ {#eq-kappa}

Large loading yields a well-conditioned but weakly dependent network; loading only slightly above $-\lambda_{\min}$ strengthens dependence but can make estimation fragile. Benchmarks should report partial-correlation distributions and $\kappa(\Omega)$, not only the adjacency.

# Conclusions

> DeCovarT draws `scale_free`, `stochastic_block_model`, and `small_world` skeletons with \pkg{igraph}, then i.i.d. signed weights (`prop_inhibitory`) and a spectral shift to guarantee $\Omega\succ 0$.

## Further high-dimensional designs

Beyond the families in @tbl-topologies:

-   **Degree-prescribed / configuration** graphs separate degree heterogeneity from preferential-attachment *mechanisms*.
-   **Spatial / geometric** graphs suit spatial transcriptomics or chromatin neighbourhoods.
-   **Differential networks:**
    -   Start from a shared base $G_0$
    -   Control the differential shift $\Omega_k=\Omega_0+\Delta_k$
    -   Ensure the SPD step
    -   Alternatively, vary the *support, weight, and sign* separately [@federicoStructureLearningGene2023; @wuEstimatingHeterogeneousGene2022].
-   **Other distributions,** using the GGM layer as a parameter:
    -   **Latent-variable GGMs** — sparse $\Omega_S$ plus low-rank confounding $Bf_i$.
    -   **Nonparanormal / Gaussian-copula** margins on a latent Gaussian graph.
    -   **Poisson–log-normal / compositional** observation layers on latent log-abundances [@chiquetVariationalInferenceSparse2018].
    -   **Zero-inflated mixture GGMs** with cell-type-specific $(\mu_k,\Omega_k)$ and expression-dependent dropout [@wuEstimatingHeterogeneousGene2022].

## When mean variation should not be omitted

For $X\sim\mathcal{N}_p(\mu,\Omega^{-1})$,

$$
X_j \perp X_k \mid X_{-\{j,k\}}
\quad\Longleftrightarrow\quad
\Omega_{jk}=0.
$$ {#eq-ci}

The mean does not appear in @eq-ci, so pure support-recovery studies often set $\mu=0$ after centring [@zhangSILGGMExtensivePackage2018]. That is inadequate when the simulation should mimic expression: pooling groups with distinct $\mu_{g(i)}$ induces between-group covariance

$$
\operatorname{Cov}(X)
=
\operatorname{E}\{\operatorname{Cov}(X\mid g)\}
+
\operatorname{Cov}\{\operatorname{E}(X\mid g)\},
$$ {#eq-law-total-cov}

which can be mistaken for network structure. PLNnetwork therefore includes covariates / offsets, and BLGGM jointly models cell-type means, precisions, and dropout [@chiquetVariationalInferenceSparse2018; @wuEstimatingHeterogeneousGene2022].

@tbl-means crosses mean variations with network changes:

| Mean scenario | Construction | Question tested |
|------------------------|----------------------|---------------------------|
| Null mean | $\mu_i=0$ | Pure graph / precision recovery |
| Sparse DE | Few marker coordinates differ by group | Robustness to phenotype shifts |
| Dense low-rank | $\mu_i = Bz_i$ | Batch / cell-state confounding |
| Mean **and** network heterogeneity | Both $\mu_k$ and $\Omega_k$ vary | Separating DE from differential connectivity |

: Mean designs to cross with topology/precision factors. {#tbl-means}

## Recommended benchmark specification

Use one pipeline for every topology (@eq-benchmark-pipe, @fig-pipeline):

$$
\text{topology}
\rightarrow
\text{signed weights}
\rightarrow
\text{SPD precision}
\rightarrow
\text{mean design}
\rightarrow
\text{latent Gaussian}
\rightarrow
\text{observation model}
$$ {#eq-benchmark-pipe}

| Axis | Suggested levels |
|-----------------------|-------------------------------------------------|
| Dimension, namely number of variables (here, genes) | $p \in \{100,500,1000\}$ |
| Topology | ER, band, hub, scale-free, SBM, small-world |
| Expected degree | $\approx 2,4,8$ (keep graphs sparse) |
| Edge weights | Constant; i.i.d. signed; partial-correlation scaling; G-Wishart |
| Conditioning | Report $\lambda_{\min}$, $\lambda_{\max}$, $\kappa(\Omega)$ |
| Means | Zero; sparse group shifts, driven by cell-type; joint $\mu$–$\Omega$ heterogeneity, for example playing on the *cosine similarity*, or the *condition number*, of the mean profile |

: Compact factorial axes for a reproducible undirected GGM simulation study. {#tbl-benchmark}

**Main insights.** Prefer the uniform spectral shift (@eq-spectral-shift) or partial-correlation scaling (@eq-partial-scale) when support and signs must be exact; add the mean layer **independently** of graph generation.

# Nine factorial simulation scenarios

The script `scripts/generate_random_network_skeleton.R` builds a $3\times 3$
grid of undirected Markov-network moments. Notation matches the manuscript:
$G$ genes and $N$ bulk samples for $\boldsymbol{Y}\in\mathcal{M}_{G\times N}$.

$$
\begin{aligned}
\text{topology}
&\in
\{\text{scale-free (A)},\
\text{SBM (B)},\
\text{small-world (C)}\},
\\
\rho
&\in
\{0.05,\,0.35,\,0.70\}
\quad\text{(cosine levels 1–3)},
\\
\text{prop. inhibitory edges}
&=
\tfrac12,
\qquad
\text{target }\kappa(\boldsymbol{\Omega})
\approx
50.
\end{aligned}
$$ {#eq-scenario-grid}

Scenario IDs are $\mathtt{A1}$–$\mathtt{C3}$ (topology letter × cosine level).
@tbl-scenarios reports realised diagnostics for each run (seeded).

```{r}
#| label: tbl-scenarios
#| tbl-cap: "Nine undirected Markov-network scenarios (3 topologies × 3 cosine levels)."
#| warning: false
#| message: false

if (requireNamespace("DeCovarT", quietly = TRUE)) {
  library(DeCovarT)
} else {
  stop("Package 'DeCovarT' is not installed.")
}

G <- 60L
N <- 80L
n_celltypes <- 3L
mean_scale <- 10
precision_scale <- 0.3
prop_inhibitory <- 0.5
target_kappa <- 50
cosine_levels <- c(0.05, 0.35, 0.70)

topology_specs <- list(
  A = list(
    topology = "scale_free",
    graph_model = "scale_free",
    graph_params = list()
  ),
  B = list(
    topology = "stochastic_block_model",
    graph_model = "stochastic_block_model",
    graph_params = list(block_prob = c(0.5, 0.25, 0.25))
  ),
  C = list(
    topology = "small_world",
    graph_model = "small_world",
    graph_params = list(nei = 2L, p = 0.05)
  )
)

calibrate_u <- function(adjacency) {
  shifts <- c(0.05, 0.1, 0.2, 0.5, 1, 2, 5)
  best_u <- shifts[[1]]
  best_gap <- Inf
  for (u in shifts) {
    W <- DeCovarT:::assign_iid_signed_weights(
      adjacency,
      prop_inhibitory = prop_inhibitory,
      weight_magnitude = precision_scale
    )
    Omega <- DeCovarT:::build_normalised_precision(W, u)
    gap <- abs(log(kappa(Omega, exact = TRUE)) - log(target_kappa))
    if (gap < best_gap) {
      best_gap <- gap
      best_u <- u
    }
  }
  best_u
}

rows <- list()
seed0 <- 100L
for (t_idx in seq_along(topology_specs)) {
  t_code <- names(topology_specs)[[t_idx]]
  spec <- topology_specs[[t_code]]
  for (c_idx in seq_along(cosine_levels)) {
    scenario_id <- paste0(t_code, c_idx)
    set.seed(seed0 + 10L * t_idx + c_idx)
    A <- DeCovarT:::generate_random_network_skeleton(
      n_genes = G,
      graph_model = spec$graph_model,
      graph_params = spec$graph_params
    )
    u <- calibrate_u(A)
    set.seed(seed0 + 10L * t_idx + c_idx)
    moments <- simulate_hierarchical_grn_moments(
      n_genes = G,
      n_celltypes = n_celltypes,
      mean_scale = mean_scale,
      target_cosine = cosine_levels[[c_idx]],
      precision_shift = u,
      precision_scale = precision_scale,
      prop_inhibitory = prop_inhibitory,
      graph_model = spec$graph_model,
      graph_params = spec$graph_params
    )
    Omega <- moments$graph_structure$normalised_precision
    W <- moments$graph_structure$weighted_adjacency
    mu <- moments$mean_profiles
    evals <- eigen(Omega, symmetric = TRUE, only.values = TRUE)$values
    upper_w <- W[upper.tri(W) & A == 1]
    rows[[scenario_id]] <- data.frame(
      ID = scenario_id,
      topology = spec$topology,
      rho = cosine_levels[[c_idx]],
      lambda_min = min(evals),
      lambda_max = max(evals),
      kappa_Omega = max(evals) / min(evals),
      prop_inhib = if (length(upper_w)) mean(upper_w > 0) else NA_real_,
      mean_abs_cos = moments$objectives$mean_abs_cosine,
      recip_kappa_mu = 1 / kappa(mu, exact = TRUE),
      gram_det = sqrt(max(0, det(crossprod(mu)))),
      stringsAsFactors = FALSE
    )
  }
}

scenario_df <- do.call(rbind, rows)
rownames(scenario_df) <- NULL

if (requireNamespace("tinytable", quietly = TRUE)) {
  scenario_df |>
    tinytable::tt(digits = 3) |>
    tinytable::group_tt(
      i = c(
        rep("A <U+2014> scale-free (Barab<U+00E1>si<U+2013>Albert)", 3L),
        rep("B <U+2014> stochastic block model", 3L),
        rep("C <U+2014> small-world (Watts<U+2013>Strogatz)", 3L)
      )
    ) |>
    tinytable::format_tt(
      j = c(
        "rho",
        "lambda_min",
        "lambda_max",
        "kappa_Omega",
        "prop_inhib",
        "mean_abs_cos",
        "recip_kappa_mu",
        "gram_det"
      ),
      digits = 3
    )
} else {
  knitr::kable(scenario_df, digits = 3)
}
```

Column guide: $\lambda_{\min}$, $\lambda_{\max}$, and
$\kappa(\boldsymbol{\Omega})=\lambda_{\max}/\lambda_{\min}$ summarise the
precision spectrum; `prop_inhib` is the realised fraction of inhibitory
precision edges; `mean_abs_cos` is the AutoGeneS cosine objective on
$\boldsymbol{\mu}$; `recip_kappa_mu` is
$\kappa_2(\boldsymbol{\mu})^{-1}$; `gram_det` is
$\sqrt{\det(\boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu})}$.

# References
