---
title: "Simulation reporting: ADEMP metrics and Nature Methods checks"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulation reporting: ADEMP metrics and Nature Methods checks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup-reporting
#| include: false
#| eval: true
options(tinytable_html_mathjax = TRUE)
```

# Purpose

This article is a reporting checklist for DeCovarT simulation studies,
not a new experiment. It maps the package outputs onto the ADEMP
structure of Morris, White and Crowther
[@morrisUsingSimulationStudies2019] and onto the Nature Methods
requirements for data, code and statistical reporting
([reporting standards](https://www.nature.com/nmeth/editorial-policies/reporting-standards)).

There is **no composite global score**. Global composition distances,
cell-type Pearson / F1, and Monte Carlo operating characteristics answer
different questions and must stay in separate tables.

# ADEMP mapping {#sec-ademp}

The estimand is the composition $\boldsymbol{p}$ on the simplex. The
data-generating mechanism is the Gaussian convolution
$\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},\sum_j p_j^2\boldsymbol{\Sigma}_j)$.
Methods are the solvers passed to `deconvolute_ratios()`.

```{r}
#| label: tbl-ademp
#| echo: false
#| eval: true
#| tbl-cap: "ADEMP performance measures implemented in `compute_benchmark_metrics()` and `coverage_mc_interval()`."
ademp <- data.frame(
  ADEMP = c(
    "Aims",
    "Data-generating mechanism",
    "Estimand",
    "Methods",
    "Bias",
    "Empirical SE",
    "Model SE",
    "Relative error in SE",
    "RMSE",
    "Coverage",
    "Mean interval width",
    "Monte Carlo SE of coverage",
    "Convergence",
    "Type I / power"
  ),
  Package = c(
    "Compare solvers on a known convolution",
    "simulate_bulk_mixture() + describe_simulation_scenario()",
    "p (simplex); cell-type and global scores kept separate",
    "deconvolute_ratios() / fit_decovart()",
    "monte_carlo$bias",
    "monte_carlo$empirical_sd",
    "monte_carlo$mean_model_se (and mean_model_sd)",
    "monte_carlo$se_sd_ratio",
    "monte_carlo$rmse (cell type) and regression$global$rmse",
    "monte_carlo$coverage with Wilson interval by default",
    "monte_carlo$mean_interval_width",
    "monte_carlo$mcse_coverage",
    "optimisation$numerical_converged, theoretical_converged, kkt_residual",
    "Not a primary target (composition estimation, not a null test)"
  ),
  Status = c(
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept",
    "Kept (Wilson default; Wald / Agresti--Coull optional)",
    "Kept",
    "Kept",
    "Kept",
    "Out of scope"
  ),
  stringsAsFactors = FALSE
)
tinytable::tt(ademp, width = 1)
```

Wilson is the default interval around the *estimated coverage rate*
$\hat\pi=X/N$, not the interval for $\hat p_j$
[@wilsonProbableInferenceLaw1927]. Wald and Agresti--Coull
[@agrestiApproximateBetterExact1998] are available through
`coverage_interval`. Bias-eliminated coverage (covering
$\bar{\hat\theta}$ rather than $\theta$) is not implemented; report
ordinary coverage and bias side by side instead.

# Nature Methods reporting {#sec-nmeth}

Nature Methods requires a reporting summary, a data-availability
statement, and promptly available code and protocols. The mapping below
is the package counterpart of that editorial checklist (and of the
machine-learning reporting summary used at the journal).

| Requirement | Where it lives in DeCovarT |
|:--|:--|
| Data-generating mechanism fully specified | `theta_true` (`p`, `mu`, `sigma`) plus `descriptors` and `call` |
| Estimand and metrics pre-declared | This table; `compute_benchmark_metrics()` blocks |
| Software versions | `sessioninfo::session_info()` in analysis scripts |
| Random-number streams | `furrr_options(seed = TRUE)`: L'Ecuyer-CMRG per worker |
| Code availability | GitHub repository; package functions, not one-off scripts |
| No undisclosed composite score | Global and cell-type tables remain separate |
| Sample size / Monte Carlo error | `n` and `mcse_coverage` (and Wilson bounds) |

: Nature Methods reporting items covered by the simulation API. {#tbl-nmeth}

# NeurIPS code completeness {#sec-neurips}

The [ML Code Completeness Checklist](https://github.com/paperswithcode/releasing-research-code)
used at NeurIPS is five items. Four are already covered above or in
the package metadata; this section only records the mapping.

| Checklist item | Where it lives |
|:--|:--|
| Specification of dependencies | `DESCRIPTION` (Imports / Suggests with versions) |
| Fitting (“training”) code | `fit_decovart()`, `deconvolute_ratios()` |
| Evaluation code | `compute_benchmark_metrics()`, `run_simulation_benchmark()` |
| Pre-trained models | Not applicable: DeCovarT is an estimator, not a stored neural net. Toy convolution fixtures live in `inst/extdata/` |
| README table of results plus commands | README / vignette chunks (this article; manuscript scenarios) |

: NeurIPS / Papers with Code completeness mapped onto the package. {#tbl-neurips}

Random-number streams and Monte Carlo error are in @sec-rng and
@tbl-nmeth, not duplicated here.

# rOpenSci statistical standards {#sec-srr}

rOpenSci [statistical software standards](https://stats-devguide.ropensci.org/standards.html)
are tagged in `R/srr-stats-standards.R` (srrstats, srrstatsNA,
srrstatsTODO) and on the exported helpers. The load-bearing blocks
are:

- **General / testing (G\*):** input checks and missing-data policy on
  `deconvolute_ratios()`; Monte Carlo tests with
  `simulate_bulk_mixture()` (G5.0, G5.5, G5.6).
- **Regression (RE\*):** `fit_decovart()` return value, `coef` /
  `vcov` / `confint`, convergence, plots (RE4, RE6, RE7).
- **Probability distributions (PD\*):** Gaussian convolution
  log-likelihood, score, and Hessian.

Do not add a second copy of those tags in vignettes. Runtime scaling
of solvers versus \(G\) and \(J\) remains srrstatsTODO (G5.7,
RE5.0). ADEMP columns themselves are @sec-ademp.

# Parallel random numbers {#sec-rng}

The sample loop in `deconvolute_ratios()` uses
`furrr::future_map()` with `seed = TRUE`. That is the R 4.1 / `future`
implementation of independent L'Ecuyer-CMRG streams, which is the
construction Morris, White and Crowther recommend for parallel Monte
Carlo [@morrisUsingSimulationStudies2019]. Scenario rows in
`run_simulation_benchmark()` stay sequential, so streams are never
nested. Do not use `set.seed(seed + i)` on workers. For an explicit
cluster script, set `RNGkind("L'Ecuyer-CMRG")` once and advance with
`parallel::nextRNGStream`. Prefer that `furrr` path over `mirai` for
published benchmarks; `mirai` is for low-latency APIs.

# What is deliberately absent

- Cholesky fill-in or sparse-solver speed diagnostics (the user must
  declare the covariance structure; otherwise the dense Cholesky factor
  of $\Sigma(p)$ is used).
- Type I error and power as default columns (the target is an estimand,
  not a null hypothesis).

See also
[synthetic scenarios](https://bastienchassagnol.github.io/DeCovarT/articles/synthetic-scenarios.html#sec-scenario-descriptors)
for the geometry table attached to every bulk draw.
