---
title: "Simplex maps and numerically stable softmax / ALR derivatives"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simplex maps and numerically stable softmax / ALR derivatives}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    eval: false
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  tbl-prefix: "Table"
  lst-prefix: "Listing"
format:
  html:
    embed-resources: true
    toc: true
---

```{r}
#| label: setup
#| include: false
#| eval: true
options(tinytable_html_mathjax = TRUE)
```

DeCovarT estimates cellular ratios $\boldsymbol{p}\in\Delta^{J-1}$ by
maximising a Gaussian-convolution log-likelihood over an *unconstrained*
parameter $\boldsymbol{\rho}\in\mathbb{R}^{J-1}$. This vignette collects the
simplex coordinate maps, the matrix-algebra vocabulary used in the analytic
gradients, and the closed-form softmax / additive log-ratio derivatives
needed for efficient optimisation. For every map we give the explicit tensor
formula, a base-`R` implementation (checked against **numDeriv**
[@R-numDeriv]), and a batched `PyTorch` counterpart. Naming follows the
DeCovarT API: `additive_logistic()` for
$\boldsymbol{\rho}\mapsto\boldsymbol{p}$ and `additive_log_ratio()` for
$\boldsymbol{p}\mapsto\boldsymbol{\rho}$.

# Simplex coordinate maps

The forward map [`additive_logistic()`](../reference/additive_logistic.html)
sends unconstrained coordinates $\boldsymbol{\rho}$ to the open simplex,

$$
p_j=\frac{e^{\rho_j}}{\sum_{k<J}e^{\rho_k}+1}\ (j<J),\qquad
p_J=\frac{1}{\sum_{k<J}e^{\rho_k}+1}.
$$ {#eq-additive-logistic-intro}

@eq-additive-logistic-intro is a softmax in which the last category $J$ is
pinned as a reference ($\rho_J\equiv 0$); in Aitchison's framework it is the
*additive logistic transform*, i.e. the inverse additive log-ratio map
($\mathrm{alr}^{-1}$).

The inverse map [`additive_log_ratio()`](../reference/additive_logistic.html)
sends a composition back to $\mathbb{R}^{J-1}$,

$$
\rho_j=\log\!\left(\frac{p_j}{p_J}\right),\qquad j=1,\ldots,J-1.
$$ {#eq-additive-log-ratio-intro}

@eq-additive-log-ratio-intro is the *additive log-ratio* ($\mathrm{alr}$)
transform with reference part $p_J$, equivalently the multinomial-logit link
with reference category $J$.

@tbl-coordinate-maps summarises the two helpers and their equivalents across
statistical, machine-learning, and compositional-data-analysis literatures.

| DeCovarT function | Direction | Transform | Equivalent formulations |
|:--|:--|:--|:--|
| `additive_logistic()` | $\boldsymbol{\rho}\mapsto\boldsymbol{p}$ (to $\Delta^{J-1}$) | inverse additive log-ratio ($\mathrm{alr}^{-1}$) | softmax with reference category $J$; `compositions::alrInv()`; `skbio.stats.composition.alr_inv()` |
| `additive_log_ratio()` | $\boldsymbol{p}\mapsto\boldsymbol{\rho}$ (to $\mathbb{R}^{J-1}$) | additive log-ratio ($\mathrm{alr}$) | multinomial-logit link (reference $J$); `compositions::alr()`; `skbio.stats.composition.alr()` |

: DeCovarT simplex coordinate maps and their equivalents. {#tbl-coordinate-maps}

# Matrix-induced scalar products {#sec-bilinear-forms}

The DeCovarT log-likelihood and its derivatives evaluate several
matrix-induced scalar products involving a symmetric positive-definite
covariance or precision matrix $\boldsymbol{A}$ (non-degenerate Gaussian
laws). @tbl-bilinear-forms records the standard terminology; a quadratic form
is the bilinear form evaluated on the same vector twice,
$q(\boldsymbol{x})=b(\boldsymbol{x},\boldsymbol{x})$.

```{r}
#| label: tbl-bilinear-forms
#| tbl-cap: "Standard algebraic operations induced by a matrix $\\boldsymbol{A}$."
#| echo: false
#| eval: true
tinytable::tt(
  data.frame(
    Expression = c(
      "$\\boldsymbol{x}^{\\mathsf{T}}\\boldsymbol{y}$",
      "$\\boldsymbol{x}^{\\mathsf{T}}\\boldsymbol{A}\\boldsymbol{y}$",
      "$\\boldsymbol{x}^{\\mathsf{T}}\\boldsymbol{A}\\boldsymbol{x}$",
      "$\\boldsymbol{x}^{\\mathsf{T}}\\boldsymbol{A}^{-1}\\boldsymbol{x}$",
      paste0(
        "$(\\boldsymbol{x}-\\boldsymbol{\\mu})^{\\mathsf{T}}",
        "\\boldsymbol{\\Sigma}^{-1}",
        "(\\boldsymbol{x}-\\boldsymbol{\\mu})$"
      ),
      paste0(
        "$\\sqrt{(\\boldsymbol{x}-\\boldsymbol{\\mu})^{\\mathsf{T}}",
        "\\boldsymbol{\\Sigma}^{-1}",
        "(\\boldsymbol{x}-\\boldsymbol{\\mu})}$"
      )
    ),
    Terminology = c(
      "Dot product / Euclidean inner product",
      paste0(
        "Bilinear form (inner product if $\\boldsymbol{A}$ is SPD)"
      ),
      "Quadratic form",
      "Quadratic form induced by $\\boldsymbol{A}^{-1}$",
      "Squared Mahalanobis distance",
      "Mahalanobis distance"
    )
  )
) |>
  tinytable::style_tt(j = 1:2, align = "c") |>
  tinytable::format_tt(j = 1:2, markdown = TRUE)
```

Package helpers `.bilinear_form()` and `.squared_mahalanobis_distance()`
implement the middle rows of @tbl-bilinear-forms. The second solves
$\boldsymbol{\Sigma}\boldsymbol{z}=\boldsymbol{\delta}$ rather than forming
$\boldsymbol{\Sigma}^{-1}$ explicitly. Calling
$\boldsymbol{x}^{\mathsf{T}}\boldsymbol{A}\boldsymbol{y}$ a "dot product" is
misleading unless $\boldsymbol{A}=\boldsymbol{I}$; see also the
[dot product](https://en.wikipedia.org/wiki/Dot_product) and
[inner product space](https://en.wikipedia.org/wiki/Inner_product_space)
articles for the Euclidean special case
$\langle\boldsymbol{x},\boldsymbol{y}\rangle=\boldsymbol{x}^{\mathsf{T}}\boldsymbol{y}$.

# Log-sum-exp and softmax

The softmax of a score vector $\boldsymbol{z}\in\mathbb{R}^{K}$ is evaluated
through the log-sum-exp (LSE) to avoid overflow,

$$
s_i=\operatorname{softmax}(\boldsymbol{z})_i=\frac{e^{z_i}}{\sum_{k}e^{z_k}}
=e^{z_i-\operatorname{LSE}(\boldsymbol{z})},
\qquad
\operatorname{LSE}(\boldsymbol{z})=m+\log\!\sum_{k}e^{z_k-m},
\quad m=\max_k z_k.
$$ {#eq-softmax}

```{#lst-lse-r .r lst-cap="Stable log-sum-exp and softmax in R" eval=true}
log_sum_exp <- function(z) {
  m <- max(z)
  m + log(sum(exp(z - m)))
}

softmax <- function(z) {
  exp(z - log_sum_exp(z))
}
```

```{#lst-softmax-torch .python lst-cap="Softmax and log-sum-exp in PyTorch"}
import torch
import torch.nn.functional as F


def softmax(z: torch.Tensor, dim: int = -1) -> torch.Tensor:
    return F.softmax(z, dim=dim)


def log_sum_exp(z: torch.Tensor, dim: int = -1) -> torch.Tensor:
    return torch.logsumexp(z, dim=dim)
```

## Softmax Jacobian and Hessian

The softmax Jacobian is the difference between a diagonal and a rank-one term,

$$
\frac{\partial s_i}{\partial z_j}=s_i(\delta_{ij}-s_j),
\qquad
\mathbf{J}_{\mathrm{softmax}}=\operatorname{diag}(\boldsymbol{s})
-\boldsymbol{s}\boldsymbol{s}^{\top}.
$$ {#eq-softmax-jac}

This matrix also equals the Hessian of the log-sum-exp,
$\nabla^2\operatorname{LSE}(\boldsymbol{z})=\operatorname{diag}(\boldsymbol{s})
-\boldsymbol{s}\boldsymbol{s}^{\top}$. Because softmax is vector-valued, its
second derivative is a third-order tensor
$H_{ijk}=\partial^2 s_i/(\partial z_j\partial z_k)$,

$$
H_{ijk}=s_i\bigl[(\delta_{ij}-s_j)(\delta_{ik}-s_k)-s_j(\delta_{jk}-s_k)\bigr],
$$ {#eq-softmax-hess}

which, slice by slice, is the rank-one correction

$$
\mathbf{H}^{(i)}=s_i\Bigl[(\boldsymbol{e}_i-\boldsymbol{s})
(\boldsymbol{e}_i-\boldsymbol{s})^{\top}
-\bigl(\operatorname{diag}(\boldsymbol{s})-\boldsymbol{s}\boldsymbol{s}^{\top}\bigr)
\Bigr],
$$ {#eq-softmax-hess-slice}

with $\boldsymbol{e}_i$ the $i$-th canonical basis vector. @lst-softmax-deriv-r
materialises @eq-softmax-jac and @eq-softmax-hess-slice in $O(K^2)$ per slice.

```{#lst-softmax-deriv-r .r lst-cap="Softmax Jacobian and Hessian tensor in R" eval=true}
softmax_jacobian <- function(z) {
  s <- softmax(z)
  diag(s) - tcrossprod(s)
}

softmax_hessian <- function(z) {
  s <- softmax(z)
  k <- length(s)
  jac <- diag(s) - tcrossprod(s)
  hessian <- array(0, dim = c(k, k, k))
  for (i in seq_len(k)) {
    di <- (seq_len(k) == i) - s # e_i - s
    hessian[i, , ] <- s[i] * (tcrossprod(di) - jac)
  }
  hessian
}
```

```{#lst-softmax-deriv-torch .python lst-cap="Batched softmax Jacobian and Hessian in PyTorch"}
def softmax_jacobian(z: torch.Tensor) -> torch.Tensor:
    """z: [..., K] -> [..., K, K]."""
    s = softmax(z, dim=-1)
    return torch.diag_embed(s) - s.unsqueeze(-1) * s.unsqueeze(-2)


def softmax_hessian(z: torch.Tensor) -> torch.Tensor:
    """z: [K] -> H[i, j, k] of shape [K, K, K]."""
    s = softmax(z, dim=0)
    eye = torch.eye(s.numel(), dtype=s.dtype, device=s.device)
    diff = eye - s  # (e_i - s) stacked over i
    outer = diff.unsqueeze(-1) * diff.unsqueeze(-2)
    jac = torch.diag(s) - torch.outer(s, s)
    return s[:, None, None] * (outer - jac.unsqueeze(0))
```

# Additive log-ratio coordinates

Let $\boldsymbol{p}=(p_1,\ldots,p_J)$ be a composition with $p_j>0$ and
$\sum_j p_j=1$, and take the last part $J$ as reference. The *additive log-ratio*
(alr) transform sends the composition to $\mathbb{R}^{J-1}$,

$$
\rho_i=\log\!\Bigl(\frac{p_i}{p_J}\Bigr),\qquad i=1,\ldots,J-1,
$$ {#eq-alr}

and its inverse, the *additive logistic* transform ($\mathrm{alr}^{-1}$), pins a
reference logit at zero, $\tilde{\boldsymbol{\rho}}=(\rho_1,\ldots,\rho_{J-1},0)$,
and applies a softmax,

$$
\boldsymbol{p}=\operatorname{softmax}(\tilde{\boldsymbol{\rho}}),
\qquad
p_i=\frac{e^{\rho_i}}{1+\sum_{j<J}e^{\rho_j}}\ (i<J),
\qquad
p_J=\frac{1}{1+\sum_{j<J}e^{\rho_j}}.
$$ {#eq-alr-inv}

```{#lst-alr-maps-r .r lst-cap="Additive logistic and additive log-ratio maps in R" eval=true}
# rho -> p  (additive logistic, alr^{-1}); reference logit appended as 0
additive_logistic <- function(rho) {
  softmax(c(rho, 0))
}

# p -> rho  (additive log-ratio, alr); last part as reference
additive_log_ratio <- function(p) {
  j <- length(p)
  log(p[-j] / p[j])
}
```

```{#lst-alr-maps-torch .python lst-cap="Additive logistic and additive log-ratio maps in PyTorch"}
def additive_logistic(rho: torch.Tensor) -> torch.Tensor:
    """rho: [..., J-1] -> p: [..., J] (reference category = last)."""
    reference = torch.zeros_like(rho[..., :1])
    return torch.softmax(torch.cat([rho, reference], dim=-1), dim=-1)


def additive_log_ratio(p: torch.Tensor) -> torch.Tensor:
    """p: [..., J] -> rho: [..., J-1]."""
    log_p = torch.log(p)
    return log_p[..., :-1] - log_p[..., -1:]
```

## Derivatives of the additive logistic map ($\boldsymbol{\rho}\mapsto\boldsymbol{p}$)

The additive logistic map is the softmax restricted to the first $J-1$ inputs,
so its Jacobian $\mathbf{J}_{\boldsymbol{\psi}}\in\mathcal{M}_{J\times(J-1)}$ is
the softmax Jacobian with the reference column dropped,

$$
\frac{\partial p_i}{\partial\rho_a}=p_i(\delta_{ia}-p_a),
\qquad
\mathbf{J}_{\boldsymbol{\psi}}
=\bigl[\operatorname{diag}(\boldsymbol{p})
-\boldsymbol{p}\boldsymbol{p}^{\top}\bigr]_{:,\,1:(J-1)},
\qquad a=1,\ldots,J-1.
$$ {#eq-alr-inv-jac}

Its Hessian is a tensor
$\mathbf{H}_{\boldsymbol{\psi}}\in\mathcal{M}_{J\times(J-1)\times(J-1)}$, the
softmax Hessian @eq-softmax-hess-slice restricted to the first $J-1$ input
indices,

$$
\frac{\partial^2 p_i}{\partial\rho_a\partial\rho_b}
=p_i\bigl[(\delta_{ia}-p_a)(\delta_{ib}-p_b)-p_a(\delta_{ab}-p_b)\bigr],
\qquad a,b=1,\ldots,J-1.
$$ {#eq-alr-inv-hess}

```{#lst-alr-inv-deriv-r .r lst-cap="Jacobian and Hessian of the additive logistic map in R" eval=true}
additive_logistic_jacobian <- function(rho) {
  p <- additive_logistic(rho)
  j <- length(p)
  (diag(p) - tcrossprod(p))[, -j, drop = FALSE]
}

additive_logistic_hessian <- function(rho) {
  p <- additive_logistic(rho)
  j <- length(p)
  jac <- diag(p) - tcrossprod(p)
  hessian <- array(0, dim = c(j, j - 1, j - 1))
  for (i in seq_len(j)) {
    di <- (seq_len(j) == i) - p
    slice <- p[i] * (tcrossprod(di) - jac)
    hessian[i, , ] <- slice[-j, -j]
  }
  hessian
}
```

```{#lst-alr-inv-deriv-torch .python lst-cap="Jacobian and Hessian of the additive logistic map in PyTorch"}
def additive_logistic_jacobian(rho: torch.Tensor) -> torch.Tensor:
    """rho: [J-1] -> Jacobian [J, J-1]."""
    p = additive_logistic(rho)
    return (torch.diag(p) - torch.outer(p, p))[:, :-1]


def additive_logistic_hessian(rho: torch.Tensor) -> torch.Tensor:
    """rho: [J-1] -> Hessian [J, J-1, J-1]."""
    reference = torch.zeros_like(rho[:1])
    augmented = torch.cat([rho, reference])
    return softmax_hessian(augmented)[:, :-1, :-1]
```

## Derivatives of the additive log-ratio map ($\boldsymbol{p}\mapsto\boldsymbol{\rho}$)

Treating $\boldsymbol{p}$ as ambient coordinates in $\mathbb{R}^{J}$ (admissible
tangent directions satisfy $\sum_i \mathrm{d}p_i=0$), the alr Jacobian
$\mathbf{J}_{\mathrm{alr}}\in\mathcal{M}_{(J-1)\times J}$ is sparse,

$$
\frac{\partial\rho_i}{\partial p_j}=\frac{\delta_{ij}}{p_i}-\frac{\delta_{jJ}}{p_J},
\qquad
\mathbf{J}_{\mathrm{alr}}
=\Bigl[\operatorname{diag}\!\bigl(p_1^{-1},\ldots,p_{J-1}^{-1}\bigr)
\ \big|\ -p_J^{-1}\boldsymbol{1}_{J-1}\Bigr],
$$ {#eq-alr-fwd-jac}

and its Hessian
$\mathbf{H}_{\mathrm{alr}}\in\mathcal{M}_{(J-1)\times J\times J}$ is diagonal in
the last two modes,

$$
\frac{\partial^2\rho_i}{\partial p_j\partial p_k}
=-\frac{\delta_{ij}\delta_{ik}}{p_i^{2}}
+\frac{\delta_{jJ}\delta_{kJ}}{p_J^{2}}.
$$ {#eq-alr-fwd-hess}

```{#lst-alr-fwd-deriv-r .r lst-cap="Jacobian and Hessian of the additive log-ratio map in R" eval=true}
additive_log_ratio_jacobian <- function(p) {
  j <- length(p)
  idx <- seq_len(j - 1)
  jac <- matrix(0, j - 1, j)
  jac[cbind(idx, idx)] <- 1 / p[idx]
  jac[, j] <- -1 / p[j]
  jac
}

additive_log_ratio_hessian <- function(p) {
  j <- length(p)
  hessian <- array(0, dim = c(j - 1, j, j))
  for (i in seq_len(j - 1)) {
    hessian[i, i, i] <- -1 / p[i]^2
    hessian[i, j, j] <- 1 / p[j]^2
  }
  hessian
}
```

```{#lst-alr-fwd-deriv-torch .python lst-cap="Jacobian and Hessian of the additive log-ratio map in PyTorch"}
def additive_log_ratio_jacobian(p: torch.Tensor) -> torch.Tensor:
    """p: [J] -> Jacobian [J-1, J]."""
    j = p.numel()
    jac = torch.zeros(j - 1, j, dtype=p.dtype, device=p.device)
    idx = torch.arange(j - 1, device=p.device)
    jac[idx, idx] = 1.0 / p[:-1]
    jac[:, -1] = -1.0 / p[-1]
    return jac


def additive_log_ratio_hessian(p: torch.Tensor) -> torch.Tensor:
    """p: [J] -> Hessian [J-1, J, J]."""
    j = p.numel()
    hessian = torch.zeros(j - 1, j, j, dtype=p.dtype, device=p.device)
    idx = torch.arange(j - 1, device=p.device)
    hessian[idx, idx, idx] = -1.0 / p[:-1].square()
    hessian[:, -1, -1] = 1.0 / p[-1].square()
    return hessian
```

# The `compositions` package and closed-form checks {#sec-compositions-checks}

The **compositions** package [@R-compositions] provides a general toolbox for
Aitchison geometry on the simplex. Beyond the additive log-ratio pair
`alr()` / `alrInv()` used here, it implements the centred log-ratio
`clr()` / `clrInv()` and the isometric log-ratio `ilr()` / `ilrInv()`,
together with composition classes (`acomp()`, `rcomp()`) and the
perturbation and powering operations of the Aitchison simplex. DeCovarT
relies only on the additive log-ratio pair because it yields the sparsest
$(J-1)$-dimensional parametrisation with an interpretable reference
category, but any of the alternative bases would define an equally valid
unconstrained coordinate system. DeCovarT's `additive_logistic()` and
`additive_log_ratio()` match `alrInv()` / `alr()` with the last part fixed as
reference (@lst-compositions-r).

```{#lst-compositions-r .r lst-cap="Cross-check against the compositions package" eval=true}
p <- c(0.2, 0.3, 0.5)
rho <- additive_log_ratio(p)

if (requireNamespace("compositions", quietly = TRUE)) {
  rho_pkg <- as.numeric(compositions::alr(compositions::acomp(p)))
  cat("alr agreement:", isTRUE(all.equal(as.numeric(rho), rho_pkg)), "\n")
  p_pkg <- as.numeric(compositions::alrInv(rho_pkg))
  cat("alrInv round-trip:", isTRUE(all.equal(p, p_pkg)), "\n")
}
```

The analytic Jacobians and Hessians are validated against Richardson
extrapolation in `R` (@lst-check-r).

```{#lst-check-r .r lst-cap="Validation with numDeriv" eval=true}
rho <- c(0.4, -1.1)
p <- c(0.2, 0.3, 0.5)

# additive logistic: Jacobian and per-output Hessian
jac_ok <- all.equal(
  numDeriv::jacobian(additive_logistic, rho),
  additive_logistic_jacobian(rho)
)
hess_ok <- all.equal(
  numDeriv::hessian(function(r) additive_logistic(r)[1], rho),
  additive_logistic_hessian(rho)[1, , ]
)

# additive log-ratio: ambient Jacobian
alr_jac_ok <- all.equal(
  numDeriv::jacobian(additive_log_ratio, p),
  additive_log_ratio_jacobian(p)
)

c(logistic_jacobian = isTRUE(jac_ok),
  logistic_hessian = isTRUE(hess_ok),
  alr_jacobian = isTRUE(alr_jac_ok))
```

# Numerical speed-ups and solver safeguards {#sec-numerical-speedups}

The analytic maps above are only half of a usable optimiser: each Newton-type
iteration also evaluates the Gaussian-convolution log-likelihood and its
derivatives with respect to $\boldsymbol{p}$ (or $\boldsymbol{\rho}$). Several
practical bottlenecks and numerical inconsistencies showed up when running the
hybrid scenario in [Deconvolution use cases](DeCoVart-use-cases.html#sec-hybrid-deconvolution); the
changes below live in `R/03_03_DeCovarT_estimate_ratios_frequentist.R`.

## Raise the Newton–Raphson evaluation budget

`deconvolute_ratios_Newton_Raphson()` wraps `stats::nlminb()`. An earlier
control list set `eval.max = 1`, which caps the **total** number of objective
evaluations for the whole run (not per iteration). The solver therefore
evaluated the log-likelihood once at the equi-balanced start and stopped,
returning the untouched initial guess on every sample. Removing that entry
(keeping `iter.max`, `rel.tol`, \ldots) restores genuine Newton steps; see
the before/after contrast (`Newton_Raphson` versus `Newton_Raphson_fixed`) in
[Table solver diagnostics](DeCoVart-use-cases.html#tbl-solver-diagnostics).

## Cache a Cholesky factorisation of $\boldsymbol{\Sigma}(\boldsymbol{p})$

Within one iteration, `optim()` / `nlminb()` / `marqLevAlg()` treat the
log-likelihood, gradient and Hessian as independent callbacks, yet they all
hit the **same** trial $\boldsymbol{p}$. Assembling
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_j p_j^2\boldsymbol{\Sigma}_j$ and
factorising it separately in each callback (and again inside the constrained
Hessian chain rule) paid for an $O(G^3)$ factorisation up to four times per
iteration. The internal helper `.sigma_p_factorisation()` caches a single
Cholesky factor keyed on exact equality of `(p, Sigma)` and returns

$$
\boldsymbol{\Sigma}(\boldsymbol{p})
=
\mathbf{R}^{\mathsf{T}}\mathbf{R},
\qquad
\log\det\boldsymbol{\Sigma}(\boldsymbol{p})
=
2\sum_{g=1}^{G}\log R_{gg},
\qquad
\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}
=
\mathbf{R}^{-1}(\mathbf{R}^{\mathsf{T}})^{-1}
$$ {#eq-chol-sigma-p}

via `chol()` / `chol2inv()` (@eq-chol-sigma-p). The unconstrained log-likelihood
then uses the cached log-determinant and inverse,

$$
\ell_{\boldsymbol{y}\,|\,\boldsymbol{\zeta}}(\boldsymbol{p})
=
-\log\det\boldsymbol{\Sigma}(\boldsymbol{p})
-\tfrac{1}{2}
\boldsymbol{r}^{\mathsf{T}}
\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}
\boldsymbol{r},
\qquad
\boldsymbol{r}
=
\boldsymbol{y}-\boldsymbol{\mu}\boldsymbol{p},
$$ {#eq-loglik-chol}

and both `gradient_loglik_unconstrained()` and
`hessian_loglik_unconstrained()` reuse the same
$\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}$ (@eq-loglik-chol). Analytic gradients
and Hessians still match `numDeriv` to $\sim 10^{-8}$–$10^{-9}$ after the
refactor.

## Guard the box-constrained L-BFGS-B path

`deconvolute_ratios_L_BFGS_B()` optimises directly in $\boldsymbol{p}$ with
box constraints $[0,1]^J$. Those boxes alone do **not** enforce
$\mathbf{1}^{\mathsf{T}}\boldsymbol{p}=1$, so line searches can drive
$\boldsymbol{\Sigma}(\boldsymbol{p})$ singular. Both the objective and the
analytic gradient are wrapped: near $\sum_j p_j\approx 0$ (or on a failed
`chol()`), the log-likelihood returns a finite penalty and the gradient a
zero vector rather than aborting `optim()`:

```{r}
#| label: lst-safe-lbfgs
#| eval: false
#| echo: true
safe_loglik <- function(p, y, mean_signature_matrix, Sigma) {
  if (sum(p) < 1e-8) {
    return(-1e12)
  }
  tryCatch(
    loglik_multivariate(p, y, mean_signature_matrix, Sigma),
    error = function(e) -1e12
  )
}
safe_gradient <- function(p, y, mean_signature_matrix, Sigma) {
  if (sum(p) < 1e-8) {
    return(rep(0, length(p)))
  }
  tryCatch(
    gradient_loglik_unconstrained(p, y, mean_signature_matrix, Sigma),
    error = function(e) rep(0, length(p))
  )
}
```

## Supply the analytic gradient to L-BFGS-B

The same solver previously relied on finite-difference gradients. Passing
`gr = safe_gradient` (the unconstrained analytic score) removes that cost and
keeps the safeguarded behaviour above.

## `marqLevAlg` Hessian sign under `minimize = FALSE`

::: {#nte-marqlevalg-hess .callout-caution}
## `marqLevAlg(minimize = FALSE)` does not flip `hess`

[`marqLevAlg`](https://cran.r-project.org/package=marqLevAlg) [@R-marqLevAlg]
implements the Marquardt–Levenberg algorithm with the relative-distance-to-minimum
(RDM) stopping rule of Commenges et al. [@commengesNewtonLikeAlgorithmLikelihood2006]:

$$
C_k
\approx
m^{-1}
\boldsymbol{U}(\boldsymbol{\theta}_k)^{\mathsf{T}}
\mathbf{G}(\boldsymbol{\theta}_k)^{-1}
\boldsymbol{U}(\boldsymbol{\theta}_k).
$$ {#eq-rdm-commenges}

With `minimize = FALSE`, the package **only** negates the user-supplied `fn`
and `gr`. The analytic `hess` is passed through **unchanged**, while the
internal Cholesky / inversion routines (`dsinv`, `dchole`) assume a
**positive-definite** matrix at the optimum (minimisation convention). When
maximising a log-likelihood, `hessian_loglik_constrained()` is
negative-definite at the MLE, so `dsinv` fails (`ier = -1`) on every
iteration, `rdm` sticks at the unevaluated sentinel `epsd + 1` (e.g.
`1.0001`), and the algorithm always exhausts `maxiter` (`istop = 2`) even
when the iterate is already at the maximum.

**Workaround used in DeCovarT:** pass
`hess = function(...) -hessian_loglik_constrained(...)` and keep
`minimize = FALSE`. Reported upstream as
[VivianePhilipps/marqLevAlgParallel#3](https://github.com/VivianePhilipps/marqLevAlgParallel/issues/3).
:::

@tbl-marqlevalg-hess summarises the empirical confirmation on the hybrid
$J=3$ scenario ($m=J-1=2$ free ALR coordinates). Point estimates barely
moved (the wrongly scaled Newton step was often rescued by the internal line
search), but iterations dropped roughly $20$–$40\times$ and `istop` / `rdm`
became honest relative to @eq-rdm-commenges.

```{r}
#| label: tbl-marqlevalg-hess
#| tbl-cap: "Effect of negating the analytic Hessian passed to marqLevAlg when maximize = TRUE (minimize = FALSE)."
#| echo: false
#| eval: true
marq_hess <- data.frame(
  `Hessian sign` = c(
    "Wrong (log-likelihood Hessian as-is)",
    "Negated (`-hess`)"
  ),
  istop = c("2 (hit maxiter)", "1 (converged)"),
  iterations = c("200 / 200", "4–11"),
  rdm = c("1.0001 (sentinel)", "~1e-13 to 1e-21"),
  check.names = FALSE
)
tinytable::tt(marq_hess) |>
  tinytable::style_tt(j = 2:4, align = "c")
```
