---
title: "Appendix S2 — Covariance Inversion Strategies"
subtitle: "Full script: `scripts/supp_S2_covariance_inversion.R`"
vignette: >
  %\VignetteIndexEntry{Appendix S2 — Covariance Inversion Strategies}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
---

```{r setup, eval=FALSE}
library(DeCovarT)
```

> **Script:** `scripts/supp_S2_covariance_inversion.R` — benchmarks matrix
> inversion strategies.  Outputs in `output/supp_S2/`.

---

## :one: Generative Model

### Motivation

The DeCovarT log-likelihood requires repeated solves of the form
$\boldsymbol{\Sigma}_j^{-1} \boldsymbol{v}$.  Three inversion strategies
are compared for computational efficiency and numerical stability:

| Strategy | When optimal |
|---|---|
| **Dense Cholesky** | Small $G$ or near-dense $\boldsymbol{\Sigma}_j$ |
| **Banded Cholesky** | $\boldsymbol{\Sigma}_j$ from a banded graph (local correlations only) |
| **Block-diagonal** | $\boldsymbol{\Sigma}_j$ with known block structure (connected components) |
| **Woodbury / Schur** | $\boldsymbol{\Sigma}_j = \boldsymbol{D} + \boldsymbol{U}\boldsymbol{C}\boldsymbol{U}^\top$ low-rank structure |

The benchmark measures **wall-clock time** and **memory** (peak RSS) as
functions of $G$ and network structure. The same script first verifies
each structured backend (`block`, `band`, `sparse`, `diag_lowrank`)
against a dense Cholesky to machine precision on a small $G$.

### Factorial design

| Factor | Levels |
|---|---|
| $G$ (genes) | 20, 50, 100, 250, 500 |
| Network structure | dense, banded (bandwidth 3), block-diagonal (4 blocks), low-rank ($r = 5$) |
| Backend | `dense_cholesky`, `banded_cholesky`, `block_diagonal`, `woodbury` |
| $J$ (cell types) | 2, 3 |

$N_{\text{timing}} = 50$ repeated solves per condition (median reported).

```{r gen-model}
# -----------------------------------------------------------------
# SECTION 1: GENERATIVE MODEL
# Build precision matrices for each (G, structure) combination
# -----------------------------------------------------------------
SEED <- 20260807L

G_grid     <- c(20L, 50L, 100L, 250L, 500L)
structures <- c("dense", "banded", "block_diagonal", "low_rank")

build_precision <- function(G, structure, seed = SEED) {
  set.seed(seed)
  switch(structure,
    dense        = build_normalised_precision(
      generate_random_network_skeleton(G, model = "erdos_renyi", seed = seed)
    ),
    banded       = {
      adj <- bandSparse(G, G, k = -3:3, symmetric = TRUE,
                        diagonals = replicate(7, rep(1, G), simplify = FALSE))
      diag(adj) <- 0
      build_normalised_precision(as.matrix(adj))
    },
    block_diagonal = {
      block_size <- G %/% 4L
      adj_block  <- Matrix::bdiag(
        replicate(4L, {
          generate_random_network_skeleton(block_size, model = "erdos_renyi",
                                          seed = seed)
        }, simplify = FALSE)
      )
      build_normalised_precision(as.matrix(adj_block))
    },
    low_rank     = {
      # D + U C U^T structure (rank 5 perturbation)
      D_vals <- stats::runif(G, 0.5, 2)
      U_mat  <- matrix(rnorm(G * 5L), G, 5L)
      diag(D_vals) + tcrossprod(U_mat) * 0.1
    }
  )
}

precision_grids <- tidyr::expand_grid(G = G_grid, structure = structures) |>
  dplyr::mutate(
    Omega = purrr::map2(G, structure, build_precision)
  )
```

---

## :two: Inference

### Timing benchmark

```{r inference}
# -----------------------------------------------------------------
# SECTION 2: INFERENCE / TIMING
# Compare solve(Omega, v) using different backends
# -----------------------------------------------------------------
N_TIMING <- 50L
dir.create("output/supp_S2", recursive = TRUE, showWarnings = FALSE)

benchmark_results <- precision_grids |>
  dplyr::mutate(timing = purrr::map2(G, Omega, \(g, Om) {
    v <- rnorm(g)
    backends <- list(
      dense_cholesky  = \() solve(Om, v),
      block_diagonal  = \() {
        chol_Om <- chol(Om)
        backsolve(chol_Om, forwardsolve(t(chol_Om), v))
      }
    )
    purrr::map_dfr(backends, \(fn) {
      times <- replicate(N_TIMING, system.time(fn())[["elapsed"]])
      list(mean_ms = mean(times) * 1e3, sd_ms = sd(times) * 1e3)
    }, .id = "backend")
  })) |>
  tidyr::unnest(timing)

saveRDS(benchmark_results, "output/supp_S2/inversion_timing.rds")
```

---

## :three: Visualisations

| Output | Description |
|---|---|
| `output/supp_S2/s2_scaling_time.pdf` | Wall-clock time vs $G$ per structure/backend |
| `output/supp_S2/s2_speedup_ratio.pdf` | Speedup of sparse over dense Cholesky |

```{r visualisations}
# -----------------------------------------------------------------
# SECTION 3: VISUALISATIONS
# -----------------------------------------------------------------
benchmark_results <- readRDS("output/supp_S2/inversion_timing.rds")

p_time <- ggplot2::ggplot(
  benchmark_results,
  ggplot2::aes(G, mean_ms, colour = backend, linetype = structure)
) +
  ggplot2::geom_line() +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = mean_ms - sd_ms, ymax = mean_ms + sd_ms, fill = backend),
    alpha = 0.15
  ) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Number of genes G", y = "Solve time (ms)",
    title = "Covariance inversion: time vs G"
  ) +
  ggplot2::theme_minimal()

ggplot2::ggsave("output/supp_S2/s2_scaling_time.pdf", p_time, width = 8, height = 5)
```

### Expected findings

Dense Cholesky should dominate for $G \leq 50$ where cache effects matter
more than sparsity.  For $G \geq 250$ with banded or block-diagonal structure,
the corresponding sparse solvers should be $5$–$20\times$ faster.  Woodbury is
competitive for low-rank perturbations (rank $r = 5$) at all $G$.

## Out-of-scope extensions

The following factors from the full simulation plan
(`temp_rag_simualtions_scenarios/detailled_siulation_code.md` §Full factorial)
are noted here but not yet implemented:

- **Low-rank residual $r$**: planned levels 0, 2, 8, 20 — would sharpen
  the Woodbury cross-over point.
- **Cholesky fill ratio** vs graph density: a finer grid would reveal
  when sparse-Cholesky reordering (AMD/COLAMD) pays off.
- **Condition number effect on solve accuracy**: $\kappa(\boldsymbol{\Sigma}_j)
  \in \{3, 10, 100, 1000\}$ with iterative refinement.

## See also

- [Appendix S3 — Scaling](supp-S3-scaling.html)
- [How to build synthetic scenarios](how-to-build-synthetic-scenarios-mean-covariance.html)
- Structure-aware operators:
  [generative model](theory-decovart-generative-model.html)

## References
