---
title: "Appendix S2 — Covariance Inversion Strategies"
subtitle: "Full script: `scripts/supp_S2_covariance_inversion.R`"
vignette: >
  %\VignetteIndexEntry{Appendix S2 — Covariance Inversion Strategies}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
---

```{r setup, eval=FALSE}
library(DeCovarT)
```

> **Script:** `scripts/supp_S2_covariance_inversion.R` — benchmarks matrix
> inversion strategies.  Outputs in `output/supp_S2/`.

---

## :one: Generative Model

### Motivation

The DeCovarT log-likelihood requires repeated solves of the form
$\boldsymbol{\Sigma}_j^{-1} \boldsymbol{v}$.  Three inversion strategies
are compared for computational efficiency and numerical stability:

| Strategy | When optimal |
|---|---|
| **Dense Cholesky** | Small $G$ or near-dense $\boldsymbol{\Sigma}_j$ |
| **Banded Cholesky** | $\boldsymbol{\Sigma}_j$ from a banded graph (local correlations only) |
| **Block-diagonal** | $\boldsymbol{\Sigma}_j$ with known block structure (connected components) |
| **Woodbury / Schur** | $\boldsymbol{\Sigma}_j = \boldsymbol{D} + \boldsymbol{U}\boldsymbol{C}\boldsymbol{U}^\top$ low-rank structure |

The benchmark measures **wall-clock time** and **memory** (peak RSS) as
functions of $G$ and network structure. The same script first verifies
each structured backend (`block`, `band`, `sparse`, `diag_lowrank`)
against a dense Cholesky to machine precision on a small $G$.

### Factorial design

| Factor | Levels |
|---|---|
| $G$ (genes) | 20, 50, 100, 250, 500 |
| Network structure | dense, banded (bandwidth 3), block-diagonal (4 blocks), low-rank ($r = 5$) |
| Backend | `dense_cholesky`, `banded_cholesky`, `block_diagonal`, `woodbury` |
| $J$ (cell types) | 2, 3 |

$N_{\text{timing}} = 50$ repeated solves per condition (median reported).

```{r gen-model}
# -----------------------------------------------------------------
# SECTION 1: GENERATIVE MODEL
# Build precision matrices for each (G, structure) combination
# -----------------------------------------------------------------
SEED <- 20260807L

G_grid     <- c(20L, 50L, 100L, 250L, 500L)
structures <- c("dense", "banded", "block_diagonal", "low_rank")

build_precision <- function(G, structure, seed = SEED) {
  set.seed(seed)
  switch(structure,
    dense        = build_normalised_precision(
      generate_random_network_skeleton(G, model = "erdos_renyi", seed = seed)
    ),
    banded       = {
      adj <- bandSparse(G, G, k = -3:3, symmetric = TRUE,
                        diagonals = replicate(7, rep(1, G), simplify = FALSE))
      diag(adj) <- 0
      build_normalised_precision(as.matrix(adj))
    },
    block_diagonal = {
      block_size <- G %/% 4L
      adj_block  <- Matrix::bdiag(
        replicate(4L, {
          generate_random_network_skeleton(block_size, model = "erdos_renyi",
                                          seed = seed)
        }, simplify = FALSE)
      )
      build_normalised_precision(as.matrix(adj_block))
    },
    low_rank     = {
      # D + U C U^T structure (rank 5 perturbation)
      D_vals <- stats::runif(G, 0.5, 2)
      U_mat  <- matrix(rnorm(G * 5L), G, 5L)
      diag(D_vals) + tcrossprod(U_mat) * 0.1
    }
  )
}

precision_grids <- tidyr::expand_grid(G = G_grid, structure = structures) |>
  dplyr::mutate(
    Omega = purrr::map2(G, structure, build_precision)
  )
```

---

## :two: Inference

### Timing benchmark

```{r inference}
# -----------------------------------------------------------------
# SECTION 2: INFERENCE / TIMING
# Compare solve(Omega, v) using different backends
# -----------------------------------------------------------------
N_TIMING <- 50L
dir.create("output/supp_S2", recursive = TRUE, showWarnings = FALSE)

benchmark_results <- precision_grids |>
  dplyr::mutate(timing = purrr::map2(G, Omega, \(g, Om) {
    v <- rnorm(g)
    backends <- list(
      dense_cholesky  = \() solve(Om, v),
      block_diagonal  = \() {
        chol_Om <- chol(Om)
        backsolve(chol_Om, forwardsolve(t(chol_Om), v))
      }
    )
    purrr::map_dfr(backends, \(fn) {
      times <- replicate(N_TIMING, system.time(fn())[["elapsed"]])
      list(mean_ms = mean(times) * 1e3, sd_ms = sd(times) * 1e3)
    }, .id = "backend")
  })) |>
  tidyr::unnest(timing)

saveRDS(benchmark_results, "output/supp_S2/inversion_timing.rds")
```

---

## :three: Visualisations

The four panels below split into two questions. @sec-s2-inversion-panels
asks how the **linear algebra** behind
$\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}\boldsymbol{v}$ scales with $G$ and
graph sparsity. @sec-s2-solver-panels asks how the **iterative MLE
solvers** spend that linear algebra: iterations to a tolerance, wall-clock
time, memory, simplex / KKT feasibility, and agreement of analytic
derivatives with Richardson extrapolation in `numDeriv` [@R-numDeriv].
The layout follows the DICEPro supplementary solver and penalty panels
(their Figures S1, S2 and S5): log-scaled trajectories for cost, then a
separate check that first-order information and constraints are
numerically sound. The $G$--sparsity--time surface is the analogue of
VIMixR supplementary Figure S4.

| Output | Description |
|---|---|
| `output/supp_S2/s2_inversion_time.pdf` | Panel A: wall-clock solve time vs $G$ |
| `output/supp_S2/s2_inversion_complexity.pdf` | Panel B: time and peak RSS vs $G$ and sparsity |
| `output/supp_S2/s2_solver_cost.pdf` | Panel C: Marquardt, L-BFGS-B, Newton, gradient |
| `output/supp_S2/s2_kkt_gradient.pdf` | Panel D: KKT / simplex residual and score check |

### Covariance inversion: wall-clock time and complexity {#sec-s2-inversion-panels}

Each likelihood, score, and Hessian evaluation factorises the mixture
covariance
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_{j}p_{j}^{2}\boldsymbol{\Sigma}_{j}$
once and then solves against one or more right-hand sides
([generative model](theory-decovart-generative-model.html#sec-numerical-speedups)).
Assembling $\boldsymbol{\Sigma}(\boldsymbol{p})$ costs $O(J G^{2})$. A dense
Cholesky factorisation of the $G\times G$ result costs $O(G^{3})$ and
dominates when $G$ is large. Structured backends (banded, block-diagonal,
Woodbury) replace that cubic with a cost that tracks bandwidth, block
size, or the low-rank residual $r$, which is why @fig-s2-inversion
must show **structure** and **backend** together, not $G$ alone.

**Panel A (running time).** Plot median wall-clock time per
`sigma_solve()` against $G$ on **log--log** axes, with one colour per
backend and one linetype per network structure. The log scale is the
same device as DICEPro Figure S1A: early-$G$ cache effects stay visible
while the $G=500$--$1000$ regime does not collapse the faster solvers onto
the axis. A dense Cholesky should approach slope $3$ on this plot. Banded
and block-diagonal backends should peel away below that line as soon as
fill-in stays cheaper than a full $G^{3}$ factorisation. Ribbons are
Monte Carlo SD across the $N_{\text{timing}}$ repeated solves in
Section 2, not confidence intervals for a statistical estimand.

**Panel B (complexity and memory).** Following VIMixR Figure S4, treat
sparsity as a second experimental factor rather than a colour annotation.
Facet or group by edge density (or bandwidth / number of blocks) and
show both **seconds per solve** and **peak RSS**. The intended reading
is a three-way relationship: larger $G$ inflates time cubically for
dense $\boldsymbol{\Sigma}_{j}$; a sparser Markov network reduces fill-in
and peak memory; the structured backend only pays off once that fill-in
advantage exceeds the overhead of the specialised kernel. At small $G$
($G\le 50$) dense Cholesky can still win because of cache locality, as
in the expected findings below.

```{r visualisations-inversion}
# -----------------------------------------------------------------
# SECTION 3a: INVERSION PANELS (A, B)
# -----------------------------------------------------------------
benchmark_results <- readRDS("output/supp_S2/inversion_timing.rds")

p_time <- ggplot2::ggplot(
  benchmark_results,
  ggplot2::aes(G, mean_ms, colour = backend, linetype = structure)
) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = mean_ms - sd_ms, ymax = mean_ms + sd_ms, fill = backend),
    alpha = 0.15,
    colour = NA
  ) +
  ggplot2::scale_x_log10() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Number of genes G",
    y = "Solve time (ms, log10)"
  ) +
  ggplot2::theme_bw()

ggplot2::ggsave(
  "output/supp_S2/s2_inversion_time.pdf",
  p_time,
  width = 8,
  height = 5
)
```

::: {#fig-s2-inversion}
![Panel A: median wall-clock time per structured solve versus $G$ on log--log axes, coloured by backend and lined by network structure.](../output/supp_S2/s2_inversion_time.pdf){#fig-s2-inversion-time}

![Panel B: solve time and peak RSS versus $G$ at several sparsity levels (bandwidth, number of blocks, or edge density), the VIMixR-style complexity surface.](../output/supp_S2/s2_inversion_complexity.pdf){#fig-s2-inversion-complexity}

Wall-clock cost of $\boldsymbol{\Sigma}(\boldsymbol{p})^{-1}\boldsymbol{v}$ as a function of gene count $G$ and Markov-network sparsity. Panel A compares inversion backends; panel B relates the same times (and peak memory) to graph sparsity. Assembly of $\boldsymbol{\Sigma}(\boldsymbol{p})$ is linear in $J$ and quadratic in $G$; dense Cholesky of the mixture covariance is cubic in $G$.
:::

### Constrained MLE solvers: cost, feasibility, and gradient checks {#sec-s2-solver-panels}

Covariance inversion is one callback. The MLE loop repeats that callback
until a convergence test fires. `fit_decovart()` exposes three second-order
or quasi-Newton paths
([`fit_decovart()`](../reference/fit_decovart.html)): Marquardt--Levenberg
via `marqLevAlg` [@R-marqLevAlg; @commengesNewtonLikeAlgorithmLikelihood2006],
box-constrained L-BFGS-B in $\boldsymbol{p}$ via `stats::optim()`, and
Newton--Raphson / `nlminb` in ILR coordinates. The projected gradient
in ILR is the fourth comparator (the same quartet as the bivariate toy
vignette, excluding mean-only NNLS / `lsei` and simulated annealing).
Wall time for one sample therefore scales roughly as
$(\text{iterations})\times\bigl(O(J G^{2})+O(G^{3})\bigr)$ for a dense
backend: **cubic in $G$**, and only **linear in $J$** through assembly
of $\boldsymbol{\Sigma}(\boldsymbol{p})$ and the $J$-vector score, not through
an extra matrix inverse per cell type.

**Panel C (solvers).** Three aligned metrics, in the spirit of DICEPro
Figure S1A--C:

- iterations (or function evaluations) until the package tolerance;
- total wall-clock time per sample;
- peak memory.

Do not map two of those scores to colour and size on one scatter. Facet
by metric, as in `plot_mc_metric_dots()`. Marquardt and Newton should
reach the threshold in few iterations because they use curvature;
L-BFGS-B is cheaper per step but may need more evaluations; projected
gradient is the slow baseline (DICEPro's Adam / gradient-descent
comparison). Report median and SD across Monte Carlo samples, and keep
$J$ as a second $x$-axis or facet so the claimed linear-$J$ assembly
cost is visible beside the cubic-$G$ inversion cost.

**Panel D (simplex adherence and analytic gradients).** Two stacked
checks, combining DICEPro Figure S5 (unit-sum / reconstruction trade-off)
with their Figure S2A (analytical versus finite-difference gradients).

The ILR / softmax chart is a $C^{2}$ diffeomorphism onto the open simplex
([constrained optimisation](theory-decovart-generative-model.html#sec-constrained)),
so Marquardt, Newton, and the projected gradient keep
$\sum_{j}\hat p_{j}=1$ by construction. L-BFGS-B only boxes
$[0,1]^{J}$ and renormalises after `optim()`; during the line search
$\lvert 1-\sum_{j}p_{j}\rvert$ can leave zero, which is why that path
needs the guarded objective in
[solver safeguards](theory-decovart-generative-model.html#sec-numerical-speedups).
The ADEMP field `optimisation$kkt_residual`
([synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html#sec-ademp))
is the projected-score residual
$\lVert\Pi_{\Delta}(\hat{\boldsymbol{p}}+\alpha\nabla\ell)-\hat{\boldsymbol{p}}\rVert_{2}/\alpha$.
It is the right analogue of DICEPro's unit-sum violation for an
interior (or bound-constrained) stationary point: small residual means
the score has no feasible first-order improvement. Plot it on a log
scale against iteration or against reconstruction error, with a Pareto
outline only if a penalty-weight grid is actually run.

The companion scatter is analytic $\nabla\ell$ (and, where stored, the
Hessian) against Richardson extrapolation in `numDeriv` [@R-numDeriv],
as already used in
[the generative-model vignette](theory-decovart-generative-model.html)
and in `tests/testthat/test-03_03_DeCovarT.R`. DICEPro reported relative
errors below $6\times 10^{-6}$ with central differences
$\varepsilon=10^{-5}$. DeCovarT prefers Richardson's tableau rather than
a single-step central difference, because the score mixes traces of
$\boldsymbol{\Theta}(\boldsymbol{p})$ with quadratic forms that have a wide
dynamic range. Points should fall on the identity; stratify by parameter
block (ILR coordinates versus unconstrained $\boldsymbol{p}$ on the
L-BFGS-B path).

```{r visualisations-solvers}
# -----------------------------------------------------------------
# SECTION 3b: SOLVER PANELS (C, D)
# -----------------------------------------------------------------
solver_results <- readRDS("output/supp_S2/solver_timing.rds")

p_solver <- ggplot2::ggplot(
  solver_results,
  ggplot2::aes(G, value, colour = solver)
) +
  ggplot2::geom_line() +
  ggplot2::geom_point() +
  ggplot2::facet_grid(metric ~ J, scales = "free_y", labeller = ggplot2::label_both) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Number of genes G",
    y = "Metric (log10)",
    colour = "Solver"
  ) +
  ggplot2::theme_bw()

ggplot2::ggsave(
  "output/supp_S2/s2_solver_cost.pdf",
  p_solver,
  width = 9,
  height = 6
)
```

::: {#fig-s2-solvers}
![Panel C: iterations to tolerance, wall-clock time, and peak memory for Marquardt--Levenberg, L-BFGS-B, Newton--Raphson (`nlminb`), and projected gradient, versus $G$ and faceted by $J$.](../output/supp_S2/s2_solver_cost.pdf){#fig-s2-solver-cost}

![Panel D: log-scale KKT / simplex residual (feasibility of $\hat{\boldsymbol{p}}$) and analytic score versus Richardson `numDeriv` gradient, stratified by parameter block.](../output/supp_S2/s2_kkt_gradient.pdf){#fig-s2-kkt-gradient}

Computational profile of the four DeCovarT MLE solvers and first-order diagnostics. Panel C separates speed of convergence from wall-clock time and memory. Panel D checks unit-simplex / KKT adherence and confirms that analytic derivatives match Richardson extrapolation.
:::

### Expected findings

Dense Cholesky should dominate for $G \le 50$ where cache effects matter
more than sparsity. For $G \ge 250$ with banded or block-diagonal
structure, the corresponding sparse solvers should be $5$--$20\times$
faster. Woodbury is competitive for low-rank perturbations (rank
$r = 5$) at all $G$. On log--log axes the dense inversion curve should
approach slope $3$ in $G$; increasing $J$ should shift times almost
linearly through assembly of $\boldsymbol{\Sigma}(\boldsymbol{p})$, not cubically.

Marquardt--Levenberg and Newton--Raphson should need far fewer
iterations than projected gradient, at a higher cost per step.
L-BFGS-B should sit between them in wall-clock time, with a larger
KKT / unit-sum residual before the final renormalisation. Analytic and
Richardson gradients should agree to relative error $\ll 10^{-6}$ on
interior points; discrepancies flag a backend or ILR-chart bug, not
optimiser tuning.

## Out-of-scope extensions

The following factors from the full simulation plan
(`temp_rag_simualtions_scenarios/detailled_siulation_code.md` §Full factorial)
are noted here but not yet implemented:

- **Low-rank residual $r$**: planned levels 0, 2, 8, 20 — would sharpen
  the Woodbury cross-over point.
- **Cholesky fill ratio** vs graph density: a finer grid would reveal
  when sparse-Cholesky reordering (AMD/COLAMD) pays off.
- **Condition number effect on solve accuracy**: $\kappa(\boldsymbol{\Sigma}_j)
  \in \{3, 10, 100, 1000\}$ with iterative refinement.

## See also

- [Appendix S3 — Scaling](https://bastienchassagnol.github.io/DeCovarT/articles/supp-S3-scaling.html)
  (RMSE and difficulty axes; this appendix is inversion and solver cost)
- [How to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html#sec-ademp)
- Structure-aware operators and solver safeguards:
  [generative model](theory-decovart-generative-model.html#sec-numerical-speedups)

## References
