---
title: "Appendix S1 — Regular-case MLE properties and identifiability"
subtitle: "Full script: `scripts/supp_S1_identifiability.R`"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Appendix S1 — Regular-case MLE properties and identifiability}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
crossref:
  eq-prefix: "Eq."
  sec-prefix: "Sec."
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
---

```{r}
#| label: setup
#| include: false
options(tinytable_html_mathjax = TRUE)
```

This appendix is the numerical counterpart of
[MLE properties and asymptotic inference](theory-DeCovarT-MLE-properties.html).
That article states what the Gaussian-convolution MLE is entitled to
claim. The chunks here check those claims on small, fully specified
scenarios.

Two regimes are kept distinct:

1. **Regular (interior) case.** True $\boldsymbol{p}$ sits in the open
   simplex. Affine equivariance, small bulk perturbations, and random
   ILR starts probe estimator invariance and solver stability when the
   Wilks $\chi^2$ calibration of
   [interior inference](theory-DeCovarT-MLE-properties.html#sec-interior)
   applies.
2. **Identifiability edge cases.** Collinear or identical means,
   boundary zeros, and near-flat likelihood ridges, matching
   [identifiability](theory-DeCovarT-MLE-properties.html#sec-identifiability)
   and
   [boundary calibration](theory-DeCovarT-MLE-properties.html#sec-boundary).

Heavy Monte Carlo factorial runs live in
`scripts/supp_S1_identifiability.R` (`N_REPLICATES=2` for a smoke test;
outputs in `output/supp_S1/`). The live chunks below use $G\le 3$ and
are meant to be rebuilt with the vignette.

```{r}
#| label: setup-s1
#| warning: false
library(DeCovarT)

set.seed(1)
genes <- paste0("g", 1:3)
cts <- paste0("ct", 1:2)
mu <- matrix(
  c(20, 40, 25, 35, 30, 22),
  nrow = 3,
  dimnames = list(genes, cts)
)
Sigma <- array(0, dim = c(3, 3, 2), dimnames = list(genes, genes, cts))
Sigma[, , 1] <- diag(c(1.0, 1.2, 0.8))
Sigma[, , 2] <- diag(c(1.5, 0.9, 1.1))
p_true <- c(0.65, 0.35)
y <- matrix(
  drop(mu %*% p_true + rnorm(3, sd = 0.05)),
  ncol = 1,
  dimnames = list(genes, "s1")
)
```

# Regular case: interior of the simplex {#sec-interior-regular}

The generating composition below is strictly positive. That is the
setting in which `fit_decovart()` Wald intervals are defined and the
score equations of
[the convolution log-likelihood](theory-DeCovarT-MLE-properties.html#eq-loglik)
are regular.

## Gene-wise affine equivariance {#sec-affine}

Let $D=\mathrm{diag}(s_1,\ldots,s_G)$ with $s_g\neq 0$ and
$\boldsymbol{m}\in\mathbb{R}^{G}$. The same gene-wise map applied to bulk,
means and covariances,
$$
\boldsymbol{y}^\star_{\cdot i}
= D^{-1}(\boldsymbol{y}_{\cdot i}-\boldsymbol{m}),
\quad
\boldsymbol{\mu}^\star
= D^{-1}(\boldsymbol{\mu}-\boldsymbol{m}\boldsymbol{1}_J^\top),
\quad
\boldsymbol{\Sigma}_j^\star
= D^{-1}\boldsymbol{\Sigma}_j D^{-1},
$$
preserves the linear convolution because
$\boldsymbol{1}_J^\top\boldsymbol{p}_{\cdot i}=1$. The Mahalanobis residual
is invariant, so
$\hat{\boldsymbol{p}}(\boldsymbol{Y},\boldsymbol{\mu},\boldsymbol{\Sigma})
=\hat{\boldsymbol{p}}(\boldsymbol{Y}^\star,\boldsymbol{\mu}^\star,
\boldsymbol{\Sigma}^\star)$ in exact arithmetic. A z-score computed
**once per gene from $\boldsymbol{\mu}$** (not per cell type, not per bulk
sample) is that map; see
[equivariance in the MLE article](theory-DeCovarT-MLE-properties.html#sec-mle-properties).

A logarithm is not affine. It is concave, so Jensen's inequality
changes first and second moments, and
$\log(\boldsymbol{\mu}\boldsymbol{p})\neq(\log\boldsymbol{\mu})\boldsymbol{p}$.
`CIBERSORT` therefore requires non-negative expression, no missing
values, and a non-log linear scale
[@newmanRobustEnumerationCell2015].

```{r}
#| label: affine-invariance
#| warning: false
fit_raw <- fit_decovart(
  mu,
  y,
  Sigma = Sigma,
  method = "Marquardt-Levenberg",
  itmax = 80
)

centre <- c(2, -1, 0.5)
scale <- c(2, 0.5, 4)
D_inv <- 1 / scale
y_star <- (y - centre) / scale
mu_star <- (mu - centre) / scale
Sigma_star <- Sigma
for (j in seq_len(2)) {
  Sigma_star[, , j] <- Sigma[, , j] * tcrossprod(D_inv)
}
p_star <- deconvolute_ratios_Marquardt_Levenberg(
  drop(y_star),
  mu_star,
  Sigma_star,
  itmax = 80
)
fit_z <- fit_decovart(
  mu,
  y,
  Sigma = Sigma,
  method = "Marquardt-Levenberg",
  itmax = 80,
  standardise = TRUE
)

rbind(
  raw = drop(coef(fit_raw)),
  general_affine = p_star,
  gene_wise_z = drop(coef(fit_z))
)
```

The two rows should agree up to optimiser noise: `standardise = TRUE`
is a gene-wise affine map. Cell-type-wise `scale()` of $\boldsymbol{\mu}$
would estimate a different $D$ per column and is not supported.

## Small perturbations of the bulk {#sec-perturb}

With known interior $\boldsymbol{p}$, a small Gaussian perturbation of
$\boldsymbol{y}_{\cdot i}$ should move Marquardt--Levenberg estimates only
slightly. Three covariance *models* are compared:

1. Full cell-type tensor $\boldsymbol{\Sigma}_j$ (DeCovarT convolution).
2. Cell-type diagonal: off-diagonals of each $\boldsymbol{\Sigma}_j$ set to
   zero (gene-wise weighted convolution).
3. Fixed global diagonal GLS: $W=\operatorname{diag}\{\Sigma(p^{\star})\}$
   held **constant**, fitted by
   [`deconvolute_ratios_gls()`](../reference/deconvolute_ratios_gls.html)
   (`MASS::lm.gls`). Do **not** copy that $W$ into every tensor slice:
   $\sum_j p_j^2 W=\|p\|_2^2 W$ still depends on $p$.

```{r}
#| label: perturbation
#| warning: false
set.seed(2)
p_true <- c(0.6, 0.4)
y0 <- matrix(drop(mu %*% p_true), ncol = 1, dimnames = list(genes, "s1"))
y_pert <- y0 + rnorm(3, sd = 0.2)

Sigma_diag <- Sigma
for (j in 1:2) {
  Sigma_diag[, , j] <- diag(diag(Sigma[, , j]))
}
w_gls <- fixed_gls_covariance(Sigma, p = p_true, diagonal = TRUE)

fit_spec <- function(y, S) {
  drop(coef(fit_decovart(
    mu,
    y,
    Sigma = S,
    method = "Marquardt-Levenberg",
    itmax = 80
  )))
}

est <- cbind(
  full = fit_spec(y_pert, Sigma),
  celltype_diagonal = fit_spec(y_pert, Sigma_diag),
  gls_fixed = deconvolute_ratios_gls(
    drop(y_pert),
    mu,
    w_gls
  )
)
rmse <- apply(est, 2, function(p_hat) sqrt(mean((p_hat - p_true)^2)))
list(estimates = est, rmse_to_truth = rmse)
```

## Random ILR starts {#sec-init-seeds}

Marquardt--Levenberg and Newton--Raphson maximise the convolution
log-likelihood with analytic Hessians. That is cheaper per step than a
first-order quasi-Newton map, but a start near a simplex face can stall
on a plateau. Three Dirichlet$(1,1)$ draws (`starting_simplex(...,
"dirichlet")`, seeds 11, 22 and 33) are passed as `initial_p` on the
same perturbed bulk. Concentration $\alpha=1$ is uniform on the simplex;
$\alpha>1$ is centre-biased and $\alpha<1$ is face-biased. Sequence
several draws (or
[`multistart_decovart()`](../reference/multistart_decovart.html))
rather than a single random start. A QP warm start is
`initial_p = "qp"`. On this well-conditioned interior toy the
second-order estimates should agree; a spread of several percentage
points is a warning that the start has trapped the solver.

```{r}
#| label: initial-seeds
#| warning: false
seeds <- c(11L, 22L, 33L)
starts <- lapply(seeds, function(s) {
  set.seed(s)
  starting_simplex(2L, "dirichlet", nms = cts)
})
names(starts) <- paste0("seed_", seeds)

fit_from_start <- function(fun, p0) {
  fun(drop(y_pert), mu, Sigma, itmax = 80, initial_p = p0)
}

ml <- vapply(
  starts,
  function(p0) fit_from_start(deconvolute_ratios_Marquardt_Levenberg, p0),
  numeric(2)
)
nr <- vapply(
  starts,
  function(p0) fit_from_start(deconvolute_ratios_Newton_Raphson, p0),
  numeric(2)
)
list(
  starts = starts,
  marquardt = ml,
  newton_raphson = nr
)
```

A fuller comparison of normalisation (CPM, log2), tolerance, and
dimension $(G,J)$ belongs in
[Appendix S3](supp-S3-scaling.html).

# Identifiability edge cases {#sec-identifiability-edges}

Standard deconvolution theory assumes that the mean signature
$\boldsymbol{\mu}$ has full column rank. That is **sufficient** for
identifiability of $\boldsymbol{p}$ but **not necessary**: the
convolution law also depends on
$\boldsymbol{\Sigma}(\boldsymbol{p})=\sum_{j}p_{j}^{2}\boldsymbol{\Sigma}_{j}$
([proposition](theory-DeCovarT-MLE-properties.html#prp-identifiability)).
Identical means with distinct covariances remain identifiable; the
negative control is identical means *and* proportional covariances.

## Collinear mean columns {#sec-collinearity}

`fit_decovart()` flags collinear mean columns as a numerical hazard
even when the covariance tensor separates the types.

```{r}
#| label: collinear-two
#| warning: true
mu2 <- mu
mu2[, 2] <- mu2[, 1]
tryCatch(
  fit_decovart(
    mu2,
    y,
    Sigma = Sigma,
    method = "Marquardt-Levenberg",
    itmax = 20
  ),
  warning = function(w) conditionMessage(w),
  error = function(e) conditionMessage(e)
)
```

```{r}
#| label: collinear-parent
#| warning: true
genes3 <- paste0("g", 1:3)
cts3 <- c("child_1", "child_2", "parent")
mu3 <- cbind(
  child_1 = c(20, 40, 15),
  child_2 = c(40, 20, 25),
  parent = 0.5 * c(20, 40, 15) + 0.5 * c(40, 20, 25)
)
rownames(mu3) <- genes3
Sigma3 <- array(0, dim = c(3, 3, 3), dimnames = list(genes3, genes3, cts3))
for (j in 1:3) {
  Sigma3[, , j] <- diag(3)
}
y3 <- matrix(mu3 %*% c(0.2, 0.3, 0.5), ncol = 1)
dimnames(y3) <- list(genes3, "s1")
qr(mu3)$rank
tryCatch(
  fit_decovart(
    mu3,
    y3,
    Sigma = Sigma3,
    method = "Marquardt-Levenberg",
    itmax = 20
  ),
  warning = function(w) conditionMessage(w),
  error = function(e) conditionMessage(e)
)
```

Missing values are not a separate predictor / response policy (RE2.2):
quantified bulk RNA-seq matrices are complete, and the API errors on
`NA` / `NaN` / `Inf`.

## Factorial Monte Carlo (script) {#sec-s1-factorial}

The remaining three sub-scenarios are too large for a vignette rebuild.
Run `Rscript scripts/supp_S1_identifiability.R` (or
`N_REPLICATES=2` for a smoke test).

| Sub-scenario | Factor | Levels |
|---|---|---|
| **S1a** Boundary | $p_3 = 0$ | Wald / profile / bootstrap CI on $J=3$, $G=10$ |
| **S1b** Same means | Covariance separation $\rho_{12}$ | $0$, $0.5$, $0.9$ ($J=2$, $G=5$) |
| **S1c** Multimodal | Gene–gene correlation $\rho$ | $-0.8$, $0$, $0.8$ (low CLD, $J=2$, $G=2$) |

$N = 1000$ Monte Carlo replicates per sub-scenario in a full run
(smoke test: $n = 2$). Solvers: Marquardt--Levenberg for S1b/S1c;
S1a also records interval coverage near a simplex face.

```{r}
#| label: s1-factorial
#| eval: false
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "1000"))
# See scripts/supp_S1_identifiability.R for config_1a / config_1b /
# config_1c and the run_simulation_benchmark() calls.
```

### Expected findings

- **S1a (boundary):** NNLS clips to zero by construction. Wald intervals
  from `confint.decovart_fit()` are undefined on a face; use
  [`lrt_decovart()`](../reference/lrt_decovart.html) /
  [`confint_profile_decovart()`](../reference/confint_profile_decovart.html)
  with chi-bar-square calibration
  ([boundary nulls](theory-DeCovarT-MLE-properties.html#sec-boundary)).
- **S1b (same means):** RMSE decreases as covariance separation grows.
  Below $\rho_{12}=0$ (identical diagonal $\boldsymbol{\Sigma}_j$) the
  pair is a negative control. DeCovarT's advantage is largest when
  means are uninformative.
- **S1c (multimodal):** The fraction of multistarts recovering the
  global mode decreases as the mean columns approach each other;
  Marquardt--Levenberg should explore more of the basin than a
  single L-BFGS-B start.

## See also

- [MLE properties](theory-DeCovarT-MLE-properties.html)
- [Appendix S2 — Covariance inversion](supp-S2-covariance-inversion.html)
- [Appendix S3 — Solver scalability](supp-S3-scaling.html)
- [How to build synthetic scenarios](theory-synthetic-scenarios-mean-covariance.html)

## References
