---
title: "Appendix S3 — Solver Scalability"
subtitle: "Full script: `scripts/supp_S3_scaling.R`"
vignette: >
  %\VignetteIndexEntry{Appendix S3 — Solver Scalability}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    toc-depth: 3
execute:
  echo: true
  eval: false
  warning: false
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
---

```{r setup, eval=FALSE}
library(DeCovarT)
```

> **Script:** `scripts/supp_S3_scaling.R` — scalability grid.
> Outputs in `output/supp_S3/`.

---

## :one: Generative Model

### Motivation

This appendix maps DeCovarT performance across the five main
statistical difficulty axes identified in the full factorial plan
(`temp_rag_simualtions_scenarios/detailled_siulation_code.md`):

| Axis | Range |
|---|---|
| Cell types $J$ | 2, 4, 8 |
| Genes $G$ | 20, 50, 100, 250 |
| Mean cosine $\rho_\mu$ | 0, 0.5, 0.9 |
| Precision condition $\kappa$ | 3, 10, 100 |
| Composition entropy $H / \log J$ | $\approx 1.0,\, 0.75,\, 0.40$ |

### Full factorial

$$
3 \times 4 \times 3 \times 3 \times 3 = 324 \text{ scenarios}
$$

$N = 100$ Monte Carlo replicates per scenario (smoke test: $n = 2$).

### Realised difficulty descriptors recorded

Each generated problem stores the following descriptors for post-hoc
analysis (see `detailled_siulation_code.md` §Difficulty descriptors):

| Descriptor | Formula |
|---|---|
| Normalised Shannon entropy | $H/\log J$ |
| Mean pairwise cosine | $\bar{\rho}_\mu$ |
| Contrast smallest singular value | $\sigma_{\min}(\boldsymbol{M})$ |
| Pairwise Bhattacharyya distance | closed-form Gaussian |
| Covariance heterogeneity | $\operatorname{mean}_F$ distance |
| Precision average degree | $2|E|/G$ |
| $\kappa(\boldsymbol{\Sigma}_j)$ | spectral condition number |

```{r gen-model}
# -----------------------------------------------------------------
# SECTION 1: GENERATIVE MODEL
# Build the 324-scenario grid
# -----------------------------------------------------------------
SEED <- 20260807L

scaling_grid <- tidyr::expand_grid(
  J           = c(2L, 4L, 8L),
  G           = c(20L, 50L, 100L, 250L),
  cosine_mu   = c(0.0, 0.5, 0.9),
  kappa_target = c(3, 10, 100),
  entropy_norm = c(1.0, 0.75, 0.40)   # approx Dirichlet concentration
) |>
  dplyr::mutate(
    scenario_id = dplyr::row_number(),
    mu = purrr::pmap(list(J, G, cosine_mu), \(j, g, rho) {
      generate_mean_signature_matrix(
        n_cell_types      = j,
        n_genes           = g,
        cosine_similarity = rho,
        mean_scale        = 100,
        seed              = SEED + j * g
      )
    }),
    proportions = purrr::map2(J, entropy_norm, \(j, h) {
      # concentration parameter from normalised entropy
      alpha <- exp(h * log(j)) / j
      as.vector(gtools::rdirichlet(1L, rep(alpha, j)))
    }),
    covariance = purrr::pmap(list(J, G, kappa_target), \(j, g, kappa) {
      purrr::map(seq_len(j), \(jj) {
        net <- generate_random_network_skeleton(g, model = "erdos_renyi",
                                               seed = SEED + jj)
        build_normalised_precision(net, target_kappa = kappa) |> solve()
      })
    })
  )
```

---

## :two: Inference

```{r inference}
# -----------------------------------------------------------------
# SECTION 2: INFERENCE
# -----------------------------------------------------------------
N_REPLICATES <- as.integer(Sys.getenv("N_REPLICATES", unset = "100"))
ALGORITHMS   <- c("NNLS", "Marquardt-Levenberg")
dir.create("output/supp_S3", recursive = TRUE, showWarnings = FALSE)

scaling_results <- purrr::pmap(
  dplyr::select(scaling_grid, proportions, mu, covariance),
  \(proportions, mu, covariance) {
    run_simulation_benchmark(
      n_replicates          = N_REPLICATES,
      true_proportions      = proportions,
      mean_signature_matrix = mu,
      covariance_matrices   = covariance,
      algorithm_names       = ALGORITHMS
    )
  },
  .progress = TRUE
)

saveRDS(scaling_results, "output/supp_S3/scaling_benchmark.rds")
saveRDS(scaling_grid,    "output/supp_S3/scaling_grid.rds")
```

---

## :three: Visualisations

| Output | Description |
|---|---|
| `output/supp_S3/s3_rmse_by_G.pdf` | RMSE vs $G$ per cosine × kappa |
| `output/supp_S3/s3_rmse_by_J.pdf` | RMSE vs $J$ per entropy level |
| `output/supp_S3/s3_convergence_rate.pdf` | Optimiser convergence rate heatmap |

```{r visualisations}
# -----------------------------------------------------------------
# SECTION 3: VISUALISATIONS
# -----------------------------------------------------------------
scaling_results <- readRDS("output/supp_S3/scaling_benchmark.rds")
scaling_grid    <- readRDS("output/supp_S3/scaling_grid.rds")

if (N_REPLICATES >= 10) {
  # Attach RMSE to grid
  scaling_summary <- scaling_grid |>
    dplyr::mutate(
      rmse = purrr::map_dbl(scaling_results, \(res) {
        mean(res[["Marquardt-Levenberg"]][["RMSE"]], na.rm = TRUE)
      })
    )

  p_g <- ggplot2::ggplot(
    scaling_summary,
    ggplot2::aes(G, rmse, colour = factor(cosine_mu), linetype = factor(kappa_target))
  ) +
    ggplot2::geom_line() +
    ggplot2::facet_wrap(~J, labeller = ggplot2::label_both) +
    ggplot2::labs(x = "G (genes)", y = "Mean RMSE", colour = "Cosine μ",
                  linetype = "κ target") +
    ggplot2::theme_minimal()

  ggplot2::ggsave("output/supp_S3/s3_rmse_by_G.pdf", p_g, width = 9, height = 5)
}
```

## Out-of-scope factors (full factorial plan)

The following factors from the full simulation plan are noted here but not
yet implemented as they require either additional function support or
substantial wall-clock time:

- **Low-rank residual $r$** (0, 2, 8, 20): Woodbury inversion advantage.
- **Reference covariance sample size** (oracle; 20; 50; 200 donors/cells):
  tests estimation noise rather than assuming a perfect $\boldsymbol{\Sigma}_j$.
  This is a critical real-data concern — estimated covariances from small
  single-cell cohorts add noise that is not modelled in the current simulation.
- **Connected-component structure** (1, 2, 4, $J$-linked): predicts exact
  block-factorisation gains (see Appendix S2).
- **BFGS-in-ALR optimiser**: a fair comparison requires implementing gradient
  descent in the same $J-1$ dimensional ALR chart as Marquardt–Levenberg; see
  `gradient_loglik_constrained()`.

## See also

- [Appendix S2 — Covariance Inversion](https://bastienchassagnol.github.io/DeCovarT/articles/supp-S2-covariance-inversion.html)
- [Appendix S4 — Covariance Modelling](https://bastienchassagnol.github.io/DeCovarT/articles/supp-S4-covariance-modeling.html)

## References
