---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator `simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture simulation. Mean profiles follow an AutoGeneS-inspired construction [@alieeAutoGeneSAutomaticGene2021] in which a single dial $\rho$ (`target_cosine`) sets pairwise collinearity between cell-type columns of $\boldsymbol{\mu}$, while a fixed scale $s$ (`mean_scale`) sets column norms. Prefer varying $\rho$ across scenarios when precision weights already control second-order dependence. Covariances share a graph-constrained signed precision $\boldsymbol{\Omega}$, mapped to $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that can be passed to `simulate_bulk_mixture()` and then to any deconvolution routine exposed by `deconvolute_ratios()`.

```{r pipeline-mermaid, echo=FALSE, results='asis'}
mmd_candidates <- c(
  "figures/synthetic_grn_pipeline.mmd",
  file.path("vignettes", "figures", "synthetic_grn_pipeline.mmd"),
  system.file(
    "vignettes",
    "figures",
    "synthetic_grn_pipeline.mmd",
    package = "DeCovarT"
  )
)
mmd_path <- mmd_candidates[file.exists(mmd_candidates)][1]
if (is.na(mmd_path) || !nzchar(mmd_path)) {
  cat(
    "```mermaid\n",
    "flowchart LR\n",
    "  moments[simulate_hierarchical_grn_moments] --> bulk[simulate_bulk_mixture]\n",
    "  bulk --> deconv[deconvolute_ratios]\n",
    "```\n",
    sep = ""
  )
} else {
  mmd <- paste(readLines(mmd_path, warn = FALSE), collapse = "\n")
  cat("```mermaid\n", mmd, "\n```\n", sep = "")
}
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  prop_inhibitory = 0.5,
  graph_model = "scale_free"
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is insufficient for stable deconvolution [@alieeAutoGeneSAutomaticGene2021] when centroids also collapse toward the origin. In the scenarios of this package we therefore **hold `mean_scale` fixed** and dial only `target_cosine`, so that Euclidean norms (and second-order precision weights) stay comparable across runs.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()`

Notation: $G$ genes, $J$ cell types indexed by $j=1,\ldots,J$, and
columns $\boldsymbol{\mu}_{\cdot j}\in\mathbb{R}^{G}$ of the mean
signature. Fix the column scale $s$ (`mean_scale`, default $10$ in the
nine-scenario grid) and dial only the target cosine $\rho\in[0,1]$
(`target_cosine`).

Let $\boldsymbol{u}=G^{-1/2}\mathbf{1}_{G}$ be the shared unit direction
and let $\boldsymbol{v}_{j}$ be the $\ell_2$-normalised indicator of the
gene block assigned to type $j$ (disjoint partition of
$\{1,\ldots,G\}$). Then $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$
for $j\neq k$. Each column is the normalised blend

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}
=
\sqrt{\rho}\,\boldsymbol{u}
+\sqrt{1-\rho}\,\boldsymbol{v}_{j},
\qquad
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{\tilde{\boldsymbol{\mu}}_{\cdot j}}{
  \|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2
},
\qquad
j=1,\ldots,J.
$$ {#eq-mean-cosine}

Expanding the un-normalised inner product for $j\neq k$ gives the exact
identity

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
=
\rho
+
\sqrt{\rho(1-\rho)}\,
\bigl(
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}
  +
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{k}
\bigr),
$$ {#eq-mean-inner}

because $\boldsymbol{u}^{\mathsf{T}}\boldsymbol{u}=1$ and
$\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$. The cross terms
$\boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}$ equal the (positive) mass of
the shared direction on block $j$; they vanish in the idealised case
where each $\boldsymbol{v}_{j}$ is taken orthogonal to $\boldsymbol{u}$,
in which @eq-mean-inner collapses to $\rho$ and, since
$\|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2=1$, the cosine of
$\boldsymbol{\mu}$ equals $\rho$ exactly. With block indicators the
cross terms are of order $J^{-1/2}$, so after the column
renormalisation in @eq-mean-cosine the realised pairwise cosine tracks
$\rho$ closely once each block contains many genes.

The scale $s$ sets column norms without changing angles:
$\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2\propto s$ at
fixed $\rho$. We keep $s=10$ across scenarios and only vary $\rho$.

### Toy illustration: $J=2$ cell types, $G=2$ genes

Take equal blocks so
$\boldsymbol{v}_{1}=(1,0)^{\mathsf{T}}$,
$\boldsymbol{v}_{2}=(0,1)^{\mathsf{T}}$, and
$\boldsymbol{u}=2^{-1/2}(1,1)^{\mathsf{T}}$. Then
$\boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}=2^{-1/2}$ and @eq-mean-inner
becomes

$$
\tilde{\boldsymbol{\mu}}_{\cdot 1}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot 2}
=
\rho
+
\sqrt{2\rho(1-\rho)}.
$$

After renormalisation the cosine of $(\boldsymbol{\mu}_{\cdot 1},
\boldsymbol{\mu}_{\cdot 2})$ still rises monotonically with $\rho$. The
chunk below reports the realised cosine at the three operating points
used in the nine-scenario grid ($\rho\in\{0,\,0.3,\,1\}$; intermediate
collinearity near $0.3$):

```{r cosine-toy-2x2}
rhos <- c(low = 0, mid = 0.3, high = 1)
toy <- lapply(rhos, function(rho) {
  mu <- DeCovarT:::generate_mean_signature_matrix(
    n_genes = 2L,
    n_celltypes = 2L,
    mean_scale = 10,
    target_cosine = rho,
    gene_names = c("g1", "g2"),
    celltype_names = c("j=1", "j=2")
  )
  cos_12 <- sum(mu[, 1] * mu[, 2]) /
    sqrt(sum(mu[, 1]^2) * sum(mu[, 2]^2))
  data.frame(
    rho_target = rho,
    cos_realised = cos_12,
    mu_j1_g1 = mu[1, 1],
    mu_j1_g2 = mu[2, 1],
    mu_j2_g1 = mu[1, 2],
    mu_j2_g2 = mu[2, 2]
  )
})
do.call(rbind, toy)
```

The two-gene toy above is useful for geometry, but it overstates finite-$G$
effects. In this construction the shared direction contributes through an
arithmetic average across genes, so concentration improves quickly as $G$
grows under gene-wise independence.

```{r}
#| label: tbl-cosine-concentration
#| tbl-cap: "Realised cosine approaches the target as the number of genes increases."
#| echo: false
rhos_check <- c(0, 0.3, 0.6, 0.9, 1)
G_values <- c(2, 80, 200)

cosine_concentration <- do.call(
  rbind,
  lapply(G_values, function(G) {
    do.call(
      rbind,
      lapply(rhos_check, function(rho) {
        mu <- DeCovarT::generate_mean_signature_matrix(
          n_genes = G,
          n_celltypes = 2L,
          mean_scale = 10,
          target_cosine = rho
        )
        cos_hat <- sum(mu[, 1] * mu[, 2]) /
          sqrt(sum(mu[, 1]^2) * sum(mu[, 2]^2))
        data.frame(G = G, rho_target = rho, cos_realised = cos_hat)
      })
    )
  })
)

knitr::kable(
  cosine_concentration,
  digits = 3,
  col.names = c("Genes (G)", "Target cosine", "Realised cosine")
)
```

With only two genes the cross terms in @eq-mean-inner are large, so the
realised cosine is a warped but strictly increasing map of $\rho$. At
$G=200$, realised cosine is much closer to target levels, which is the
high-dimensional regime assumed by the synthetic generator.

@fig-cosine-geometry makes the construction geometric for the same three
operating points (the cosine levels shared by scenarios A1–C1, A2–C2 and
A3–C3). In the plane of two selected genes (a visual projection), each
panel draws:

1. the shared unit direction $\boldsymbol{u}$ and the private markers
   $\boldsymbol{v}_{1}$, $\boldsymbol{v}_{2}$;
2. the un-normalised blends
   $\tilde{\boldsymbol{\mu}}_{\cdot j}=\sqrt{\rho}\,\boldsymbol{u}+\sqrt{1-\rho}\,\boldsymbol{v}_{j}$
   (dashed);
3. the final columns $\boldsymbol{\mu}_{\cdot j}$ after
   $\ell_2$-normalisation and scaling by $s$ (solid), with the angle
   between them labelled by the realised cosine.

The sequence $\boldsymbol{v}_{j}\rightarrow\tilde{\boldsymbol{\mu}}_{\cdot j}\rightarrow\boldsymbol{\mu}_{\cdot j}$
is the one-shot constructive counterpart of AutoGeneS-style search for a
target collinearity: rather than iterating a multi-objective optimiser,
@eq-mean-cosine places the two cell-type profiles at a controlled angle
in a single pass [@alieeAutoGeneSAutomaticGene2021].

```{r}
#| label: fig-cosine-geometry
#| fig-cap: >-
#|   Constructive cosine dial for $J=2$ cell types in a two-gene projection.
#|   Panels correspond to the three scenario levels
#|   $\rho\in\{0,\,0.3,\,1\}$ (suffixes 1–3 in A1–C3). Shared direction
#|   $\boldsymbol{u}$ (grey), private markers $\boldsymbol{v}_{j}$
#|   (open arrows), blends $\tilde{\boldsymbol{\mu}}_{\cdot j}$ (dashed),
#|   and scaled profiles $\boldsymbol{\mu}_{\cdot j}$ (solid).
#| fig-width: 9
#| fig-height: 3.4
#| echo: false
#| warning: false
#| message: false
#| eval: true
rhos_fig <- c(0, 0.3, 1)
s_fig <- 10
u <- c(1, 1) / sqrt(2)
v1 <- c(1, 0)
v2 <- c(0, 1)

arrow_df <- list()
label_df <- list()
for (rho in rhos_fig) {
  tilde1 <- sqrt(rho) * u + sqrt(1 - rho) * v1
  tilde2 <- sqrt(rho) * u + sqrt(1 - rho) * v2
  mu1 <- s_fig * tilde1 / sqrt(sum(tilde1^2))
  mu2 <- s_fig * tilde2 / sqrt(sum(tilde2^2))
  tilde1_u <- tilde1 / sqrt(sum(tilde1^2))
  tilde2_u <- tilde2 / sqrt(sum(tilde2^2))
  mu1_u <- mu1 / s_fig
  mu2_u <- mu2 / s_fig
  cos_hat <- sum(mu1 * mu2) / (s_fig * s_fig)

  panel <- sprintf(
    "rho = %s  (realised cos = %.2f)",
    format(rho, nsmall = 1),
    cos_hat
  )
  arrow_df[[length(arrow_df) + 1L]] <- data.frame(
    panel = panel,
    rho = rho,
    name = c(
      "u", "v1", "v2",
      "tilde1", "tilde2",
      "mu1", "mu2"
    ),
    kind = c(
      "shared", "private", "private",
      "blend", "blend",
      "profile", "profile"
    ),
    x0 = 0, y0 = 0,
    x1 = c(u[1], v1[1], v2[1], tilde1_u[1], tilde2_u[1], mu1_u[1], mu2_u[1]),
    y1 = c(u[2], v1[2], v2[2], tilde1_u[2], tilde2_u[2], mu1_u[2], mu2_u[2]),
    stringsAsFactors = FALSE
  )
  label_df[[length(label_df) + 1L]] <- data.frame(
    panel = panel,
    label = sprintf("realised cosine = %.2f", cos_hat),
    x = 0.05,
    y = 1.12,
    stringsAsFactors = FALSE
  )
}
arrows <- do.call(rbind, arrow_df)
labels <- do.call(rbind, label_df)
arrows$panel <- factor(arrows$panel, levels = unique(arrows$panel))
labels$panel <- factor(labels$panel, levels = levels(arrows$panel))

print(
  ggplot2::ggplot() +
    ggplot2::geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
    ggplot2::geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.3) +
    ggplot2::geom_segment(
      data = arrows[arrows$kind == "shared", ],
      ggplot2::aes(x = x0, y = y0, xend = x1, yend = y1),
      colour = "grey40",
      linewidth = 0.7,
      arrow = ggplot2::arrow(length = grid::unit(0.18, "cm"))
    ) +
    ggplot2::geom_segment(
      data = arrows[arrows$kind == "private", ],
      ggplot2::aes(x = x0, y = y0, xend = x1, yend = y1),
      colour = "#1B9E77",
      linewidth = 0.7,
      linetype = "22",
      arrow = ggplot2::arrow(length = grid::unit(0.18, "cm"))
    ) +
    ggplot2::geom_segment(
      data = arrows[arrows$kind == "blend", ],
      ggplot2::aes(x = x0, y = y0, xend = x1, yend = y1),
      colour = "#D95F02",
      linewidth = 0.65,
      linetype = "dashed",
      arrow = ggplot2::arrow(length = grid::unit(0.18, "cm"))
    ) +
    ggplot2::geom_segment(
      data = arrows[arrows$kind == "profile", ],
      ggplot2::aes(x = x0, y = y0, xend = x1, yend = y1, colour = name),
      linewidth = 1.05,
      arrow = ggplot2::arrow(length = grid::unit(0.22, "cm"))
    ) +
    ggplot2::geom_text(
      data = labels,
      ggplot2::aes(x = x, y = y, label = label),
      hjust = 0,
      size = 3,
      colour = "grey20"
    ) +
    ggplot2::scale_colour_manual(
      values = c(mu1 = "#7570B3", mu2 = "#E7298A"),
      labels = c(
        expression(bold(mu)[plain("\u00b7") * 1]),
        expression(bold(mu)[plain("\u00b7") * 2])
      ),
      name = NULL
    ) +
    ggplot2::coord_equal(xlim = c(-0.15, 1.2), ylim = c(-0.15, 1.25)) +
    ggplot2::facet_wrap(~panel, nrow = 1) +
    ggplot2::labs(x = "gene 1", y = "gene 2") +
    ggplot2::theme_bw(base_size = 11) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(
        fill = "grey95",
        colour = "grey70"
      )
    )
)
```

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a quantity to maximise [@alieeAutoGeneSAutomaticGene2021]. Cosine is preferred to Pearson correlation so that collinear but unequally scaled profiles remain penalised.

## Network skeleton: `generate_random_network_skeleton()`

A binary adjacency $\boldsymbol{A}\in\{0,1\}^{G\times G}$ is drawn from one of three undirected families via \pkg{igraph}: scale-free preferential attachment (`igraph::sample_pa()`), a stochastic block model (`igraph::sample_sbm()`), or Watts–Strogatz small-world (`igraph::sample_smallworld()`) [@R-igraph; @barabasiEmergenceScalingRandom1999]. Edges encode presence of partial-correlation links, not their strengths. i.i.d. signs with inhibitory fraction `prop_inhibitory` then form the weighted matrix $\boldsymbol{W}$ before the spectral SPD completion.

## Positive-definite precision: `build_normalised_precision()`

Signed edge weights enter through an affine spectral shift of $\boldsymbol{W}$,

$$
\boldsymbol{\Omega}
=
\boldsymbol{W}
+
\bigl(
  \lvert\lambda_{\min}(\boldsymbol{W})\rvert + u
\bigr)
\mathbf{I},
$$

with magnitude $v$ (`precision_scale`) baked into $\boldsymbol{W}$ and diagonal cushion $u$ (`precision_shift`). The shift preserves off-diagonal signs and support while guaranteeing $\boldsymbol{\Omega}\succ 0$.

## Shared covariances: `build_shared_covariance_array()`

Cell types share the same second-order structure,

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Downstream bulk simulation uses the Gaussian convolution $\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

# Perspectives: a single geometric score for $\boldsymbol{\mu}$

The present API exposes two AutoGeneS-style objectives. A natural refinement is to replace them by one scalar that jointly rewards large column norms (and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$

is the volume of the parallelepiped spanned by the columns of $\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when columns lengthen or become more orthogonal—precisely the desired MOO trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$

available in R via `kappa(mean_profiles, exact = TRUE)` (or `1 / kappa(...)`), measures how ill-posed linear recovery of $\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$ is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising $\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes", "volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the constructive $(\rho,s)$ generator for controllable scenarios, and report the chosen scalar alongside the existing pairwise diagnostics.

# References
