---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator `simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture simulation. Mean profiles follow an AutoGeneS-inspired construction [@alieeAutoGeneSAutomaticGene2021] in which a single dial $\rho$ (`target_cosine`) sets pairwise collinearity between cell-type columns of $\boldsymbol{\mu}$, while a fixed scale $s$ (`mean_scale`) sets column norms. Prefer varying $\rho$ across scenarios when precision weights already control second-order dependence.

Each cell type carries its own graph-constrained signed precision $\boldsymbol{\Omega}_j$, mapped to $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}_j^{-1}$. An undirected Gaussian Markov simulation separates four layers (@eq-ggm-pipeline): graph $G_j$, signed weights $W(G_j)$, SPD precision $\Omega_j$, and latent Gaussians (an optional observation layer can add Poisson–log-normal or zero-inflated counts). The **graph** and the **precision** must not be conflated: a binary adjacency is almost never itself positive definite. DeCovarT draws $G_j$ via `generate_random_network_skeleton()`, completes $\Omega_j\succ 0$ with `build_normalised_precision()`, and stacks the inverted slices with `build_covariance_array_from_precision()`. @sec-ggm-networks develops the topology, weight, and SPD design in detail.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that can be passed to `simulate_bulk_mixture()` and then to any deconvolution routine exposed by `deconvolute_ratios()`.

```{mermaid}
%%| label: fig-pipeline-api
%%| fig-cap: "Pipeline from synthetic moment generation to deconvolution."
%%{init: {"theme": "sandstone"}}%%
flowchart LR
  A["Graph G"] --> B["Weights W(G)"]
  B --> C["Precision Ω ≻ 0"]
  C --> D["Σ = Ω^{-1}"]
  E["Means μ_j\n(AutoGeneS-style ρ)"] --> F["simulate_hierarchical_grn_moments()"]
  D --> F
  F --> G["simulate_bulk_mixture()"]
  G --> H["deconvolute_ratios()"]
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  prop_inhibitory = 0.5,
  graph_model = "scale_free"
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
dim(bulk$latent_profiles) # G x J x N latent draws; input mu is shared
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is insufficient for stable deconvolution [@alieeAutoGeneSAutomaticGene2021]. In the scenarios of this package we therefore **hold `mean_scale` fixed** and dial only `target_cosine`, so first-order scale and second-order precision weights stay comparable across runs.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()` {#sec-lr-block-mu}

Notation: $G$ genes, $J$ cell types indexed by $j=1,\ldots,J$, and columns $\boldsymbol{\mu}_{\cdot j}\in\mathbb{R}^{G}$ of the mean signature. Fix the column scale $s$ (`mean_scale`, default $10$ in the nine-scenario grid) and dial only the target cosine $\rho\in[0,1]$ (`target_cosine`).

Let $\boldsymbol{u}=G^{-1/2}\mathbf{1}_{G}$ be the shared unit direction and let $\boldsymbol{v}_{j}$ be the $\ell_2$-normalised indicator of the gene block assigned to type $j$ (disjoint partition of $\{1,\ldots,G\}$). Then $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$ for $j\neq k$. Each column is the normalised blend

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}
=
\sqrt{\rho}\,\boldsymbol{u}
+\sqrt{1-\rho}\,\boldsymbol{v}_{j},
\qquad
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{\tilde{\boldsymbol{\mu}}_{\cdot j}}{
  \|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2
},
\qquad
j=1,\ldots,J.
$$ {#eq-mean-cosine}

Stacking the un-normalised columns yields a **rank-one shared term plus a block-sparse private term**:

$$
\tilde{\boldsymbol{M}}
=
\bigl[
  \tilde{\boldsymbol{\mu}}_{\cdot 1}
  \mid
  \cdots
  \mid
  \tilde{\boldsymbol{\mu}}_{\cdot J}
\bigr]
=
\sqrt{\rho}\,
\boldsymbol{u}\mathbf{1}_{J}^{\mathsf{T}}
+
\sqrt{1-\rho}\,
\mathbf{V},
$$ {#eq-lr-block}

where
$\mathbf{V}=\bigl[\boldsymbol{v}_{1}\mid\cdots\mid\boldsymbol{v}_{J}\bigr]$
has **disjoint supports** across columns (each $\boldsymbol{v}_j$ is non-zero only on block $j$). The shared part is rank one; the private part is block-diagonal in the gene-partition sense. After column normalisation and scaling,

$$
\boldsymbol{\mu}
=
s\,\tilde{\boldsymbol{M}}\,
\mathbf{D}^{-1},
\qquad
\mathbf{D}=\mathrm{diag}\bigl(
  \|\tilde{\boldsymbol{\mu}}_{\cdot 1}\|_2,
  \ldots,
  \|\tilde{\boldsymbol{\mu}}_{\cdot J}\|_2
\bigr),
$$ {#eq-mu-normalised}

so $\boldsymbol{\mu}$ is a low-rank-plus-block construction followed by per-column calibration. This matches synthetic signature designs that combine a common baseline with type-specific marker programmes [@alieeAutoGeneSAutomaticGene2021; @schelkerEstimationImmuneCell2017; @baWhenLessNot2026].

Expanding the un-normalised inner product for $j\neq k$ gives the exact identity

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
=
\rho
+
\sqrt{\rho(1-\rho)}\,
\bigl(
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}
  +
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{k}
\bigr),
$$ {#eq-mean-inner}

because $\boldsymbol{u}^{\mathsf{T}}\boldsymbol{u}=1$ and $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$. Block indicators are **not** orthogonal to the shared direction: if block $j$ contains $L_j$ genes then

$$
\boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}
=
\sqrt{\frac{L_j}{G}},
$$ {#eq-u-vj}

so @eq-mean-inner becomes

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
=
\rho
+
\sqrt{\rho(1-\rho)}\,
c_{jk},
\qquad
c_{jk}
=
\sqrt{\tfrac{L_j}{G}}
+
\sqrt{\tfrac{L_k}{G}}.
$$ {#eq-mean-inner-explicit}

Column-wise renormalisation in @eq-mean-cosine maps this to a realised cosine that is **not** exactly $\rho$ for finite $G$ and $J$:

$$
\cos\bigl(
  \boldsymbol{\mu}_{\cdot j},
  \boldsymbol{\mu}_{\cdot k}
\bigr)
=
\frac{
  \rho
  + \sqrt{\rho(1-\rho)}\,c_{jk}
}{
  1 + \sqrt{\rho(1-\rho)}\,c_{jk}
},
\qquad j\neq k.
$$ {#eq-realised-cosine}

### Asymptotic cosine control {#sec-asymptotic-cosine}

For equal blocks ($L_j\approx G/J$) we have $c_{jk}\approx 2/\sqrt{J}$. Writing the bias relative to the target,

$$
\cos\bigl(
  \boldsymbol{\mu}_{\cdot j},
  \boldsymbol{\mu}_{\cdot k}
\bigr)
-
\rho
=
\frac{
  (1-\rho)\sqrt{\rho(1-\rho)}\,c_{jk}
}{
  1 + \sqrt{\rho(1-\rho)}\,c_{jk}
}
=
O\!\bigl(J^{-1/2}\bigr)
=
o(1)
\quad\text{as }J\to\infty
$$ {#eq-cosine-bias}

(with $G/J$ bounded away from zero). Equivalently,

$$
\lim_{J\to\infty}
\cos\bigl(
  \boldsymbol{\mu}_{\cdot j},
  \boldsymbol{\mu}_{\cdot k}
\bigr)
=
\rho,
$$ {#eq-cosine-limit}

so the cross terms in @eq-mean-inner become negligible relative to $\rho$ and the realised pairwise cosine tracks the dial asymptotically. The scale $s$ sets column norms without changing angles: $\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2\propto s$ at fixed $\rho$. We keep $s=10$ across scenarios and only vary $\rho$. @fig-cosine-convergence illustrates the finite-sample approach to $\rho=0.5$ on a $3\times 3$ grid of $(J,G)$.

@fig-mean-signature-schematic gives a toy schematic for $J=3$ cell types: a uniform shared component, three orthogonal private blocks, and the resulting mean directions.

![Schematic of shared-plus-private mean signature construction for $J=3$ cell types. A rank-one shared direction $\boldsymbol{u}$ (grey) blends with block-sparse private markers $\boldsymbol{v}_j$ (coloured) at weights $\sqrt{\rho}$ and $\sqrt{1-\rho}$; columns are normalised and scaled to $\boldsymbol{\mu}_{\cdot j}$. Illustration generated for this vignette in a Nature Methods-style layout.](figures/fig_mean_signature_shared_private.png){#fig-mean-signature-schematic width=90%}

### Visualising the cosine construction

```{r}
#| label: fig-cosine-convergence
#| fig-cap: >-
#|   Realised mean absolute pairwise cosine vs target $\rho=0.5$ on a
#|   $3\times 3$ grid: $J\in\{3,10,20\}$ nested with $G\in\{20,100,400\}$
#|   (equal gene blocks). Thick black line: target; coloured bands:
#|   harder (red, small $J$) to easier (green, large $J$); arrows:
#|   residual to the target.
#| fig-width: 7.5
#| fig-height: 5.2
#| echo: false
#| eval: true
target_rho <- 0.5
grid <- expand.grid(
  J = c(3L, 10L, 20L),
  G = c(20L, 100L, 400L),
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)
grid <- grid[grid$G >= grid$J, ]
conv_df <- data.frame(
  G = integer(),
  J = integer(),
  mean_abs_cosine = numeric(),
  target = numeric()
)
for (i in seq_len(nrow(grid))) {
  J <- grid$J[[i]]
  mu <- DeCovarT::generate_mean_signature_matrix(
    n_genes = grid$G[[i]],
    n_celltypes = grid$J[[i]],
    mean_scale = 10,
    target_cosine = target_rho
  )
  n_pairs <- J * (J - 1L) / 2L
  acc <- 0
  for (jj in seq_len(J - 1L)) {
    for (kk in (jj + 1L):J) {
      mu_j <- mu[, jj]
      mu_k <- mu[, kk]
      norms <- sqrt(sum(mu_j^2) * sum(mu_k^2))
      cos_jk <- if (norms < .Machine$double.eps) {
        0
      } else {
        sum(mu_j * mu_k) / norms
      }
      acc <- acc + abs(cos_jk)
    }
  }
  conv_df <- rbind(
    conv_df,
    data.frame(
      G = grid$G[[i]],
      J = grid$J[[i]],
      mean_abs_cosine = acc / n_pairs,
      target = target_rho
    )
  )
}
conv_df$J <- factor(conv_df$J, levels = c(3L, 10L, 20L))
conv_df$label <- paste0("J=", conv_df$J, ", G=", conv_df$G)
conv_df$label <- factor(
  conv_df$label,
  levels = rev(unique(conv_df$label[order(as.integer(as.character(conv_df$J)), conv_df$G)]))
)
j_cols <- c("3" = "#F4A582", "10" = "#FEE08B", "20" = "#A6D96A")
band_df <- data.frame(
  J = factor(c(3L, 10L, 20L), levels = levels(conv_df$J)),
  xmin = -Inf,
  xmax = Inf
)
# Map each J band to the y-range of its three labels (discrete positions 1..n).
label_levels <- levels(conv_df$label)
band_rects <- do.call(
  rbind,
  lapply(levels(conv_df$J), function(jj) {
    labs_j <- as.character(conv_df$label[conv_df$J == jj])
    idx <- match(labs_j, label_levels)
    data.frame(
      J = factor(jj, levels = levels(conv_df$J)),
      ymin = min(idx) - 0.45,
      ymax = max(idx) + 0.45,
      xmin = -Inf,
      xmax = Inf
    )
  })
)

ggplot2::ggplot(conv_df, ggplot2::aes(x = mean_abs_cosine, y = label)) +
  ggplot2::geom_rect(
    data = band_rects,
    ggplot2::aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, fill = J),
    inherit.aes = FALSE,
    alpha = 0.35,
    colour = NA
  ) +
  ggplot2::geom_vline(
    xintercept = target_rho,
    colour = "black",
    linewidth = 1.15
  ) +
  ggplot2::geom_segment(
    ggplot2::aes(xend = target_rho, yend = label),
    colour = "grey25",
    linewidth = 0.45,
    arrow = grid::arrow(length = grid::unit(0.14, "cm"), type = "closed")
  ) +
  ggplot2::geom_point(size = 2.6, colour = "#1B4F72") +
  ggplot2::scale_fill_manual(
    values = j_cols,
    name = "Cell types J",
    labels = c("3 (harder)", "10", "20 (easier)")
  ) +
  ggplot2::labs(
    x = "Mean |cosine| (realised)",
    y = NULL
  ) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(legend.position = "bottom")
```

@fig-cosine-geometry visualises the same three operating points ($\rho\in\{0,\,0.3,\,1\}$) for $G=2$ genes and $J=2$ cell types. Cell type 1 is fixed along the positive gene-1 axis; only the **angular** position of cell type 2 is varied so that $\rho=0$ is exactly orthogonal. Independent Gaussian noise is added *only in this vignette chunk* around each centroid ($N=20$ i.i.d. draws per population) to emulate within-type dispersion under an independence assumption, without altering the mean-signature generator itself. Coloured arrows mark each type's mean direction; the shaded wedge is the angle between the two means.

This illustration is inspired by panel B in the AutoGeneS paper [@alieeAutoGeneSAutomaticGene2021], but focuses here on the cosine-angle control induced by @eq-mean-cosine (without modelling centroid-distance optimisation). With only two genes the cross terms in @eq-mean-inner are large, so the realised cosine from the generator is a warped but strictly increasing map of $\rho$; the panels below enforce the target angle geometrically after generation.

```{r}
#| label: fig-cosine-geometry
#| fig-cap: >-
#|   Cosine-control toy for $J=2$ cell types and $G=2$ genes.
#|   Cell type 1 is fixed on the positive gene-1 axis; only the angular
#|   position of cell type 2 varies (so $\rho=0$ is orthogonal).
#|   Coloured arrows: mean directions; shaded wedge: angle between means.
#|   Labels report the realised cosine. Points are $N=20$ i.i.d. Gaussian
#|   draws around each centroid (stars). Styling inspired by panel B of
#|   `AutoGeneS` (@alieeAutoGeneSAutomaticGene2021).
#| fig-width: 9
#| fig-height: 4.6
#| echo: false
#| warning: false
#| message: false
#| eval: true
rhos_fig <- c(0, 0.3, 1)
set.seed(20260804)
n_genes_fig <- 2L
n_celltypes_fig <- 2L
n_samples <- 20L
s_fig <- 10
point_noise_sd <- 0.85
arc_n <- 80L
pop_cols <- c(cell_1 = "#2B8C99", cell_2 = "#F4B400")

scatter_df <- list()
centroid_df <- list()
arrow_df <- list()
wedge_df <- list()
arc_df <- list()
label_df <- list()

for (k in seq_along(rhos_fig)) {
  rho <- rhos_fig[[k]]
  # Cell type 1 on the +x axis; cell type 2 at angle acos(rho).
  theta <- acos(max(-1, min(1, rho)))
  mu <- cbind(
    cell_1 = c(s_fig, 0),
    cell_2 = s_fig * c(cos(theta), sin(theta))
  )
  rownames(mu) <- c("gene_1", "gene_2")

  cos_hat <- sum(mu[, 1] * mu[, 2]) /
    sqrt(sum(mu[, 1]^2) * sum(mu[, 2]^2))
  cos_hat <- max(-1, min(1, cos_hat))
  panel <- sprintf(
    "rho = %s",
    format(rho, nsmall = 1)
  )

  ang1 <- 0
  ang2 <- theta
  ang_seq <- seq(ang1, ang2, length.out = arc_n)
  r_arc <- 0.42 * s_fig

  for (j in seq_len(n_celltypes_fig)) {
    scatter_df[[length(scatter_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = stats::rnorm(n_samples, mean = mu[1L, j], sd = point_noise_sd),
      y = stats::rnorm(n_samples, mean = mu[2L, j], sd = point_noise_sd)
    )
    centroid_df[[length(centroid_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = mu[1L, j],
      y = mu[2L, j]
    )
    arrow_df[[length(arrow_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = 0,
      y = 0,
      xend = mu[1L, j],
      yend = mu[2L, j]
    )
  }

  wedge_df[[length(wedge_df) + 1L]] <- data.frame(
    panel = panel,
    x = c(0, r_arc * cos(ang_seq), 0),
    y = c(0, r_arc * sin(ang_seq), 0)
  )
  arc_df[[length(arc_df) + 1L]] <- data.frame(
    panel = panel,
    x = r_arc * cos(ang_seq),
    y = r_arc * sin(ang_seq)
  )

  label_df[[length(label_df) + 1L]] <- data.frame(
    panel = panel,
    label = sprintf("cos = %.2f", cos_hat)
  )
}

points_df <- do.call(rbind, scatter_df)
centroids <- do.call(rbind, centroid_df)
arrows <- do.call(rbind, arrow_df)
wedges <- do.call(rbind, wedge_df)
arcs <- do.call(rbind, arc_df)
labels <- do.call(rbind, label_df)
panel_levels <- sprintf("rho = %s", format(rhos_fig, nsmall = 1))
for (nm in c(
  "points_df", "centroids", "arrows", "wedges", "arcs", "labels"
)) {
  df <- get(nm)
  df$panel <- factor(df$panel, levels = panel_levels)
  assign(nm, df)
}

ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, colour = "grey88", linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey88", linewidth = 0.3) +
  ggplot2::geom_polygon(
    data = wedges,
    ggplot2::aes(x = x, y = y, group = panel),
    fill = "grey35",
    alpha = 0.12,
    colour = NA
  ) +
  ggplot2::geom_path(
    data = arcs,
    ggplot2::aes(x = x, y = y, group = panel),
    colour = "grey25",
    linewidth = 0.55,
    lineend = "round"
  ) +
  ggplot2::geom_segment(
    data = arrows,
    ggplot2::aes(
      x = x,
      y = y,
      xend = xend,
      yend = yend,
      colour = celltype
    ),
    linewidth = 0.95,
    arrow = grid::arrow(
      length = grid::unit(0.22, "cm"),
      type = "closed"
    )
  ) +
  ggplot2::geom_point(
    data = points_df,
    ggplot2::aes(x = x, y = y, fill = celltype),
    shape = 21,
    colour = "grey20",
    stroke = 0.35,
    alpha = 0.85,
    size = 2.15
  ) +
  ggplot2::geom_point(
    data = centroids,
    ggplot2::aes(x = x, y = y),
    shape = 8,
    size = 3.3,
    colour = "black",
    inherit.aes = FALSE
  ) +
  ggplot2::geom_label(
    data = labels,
    ggplot2::aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.0,
    colour = "grey15",
    fill = "#fffbe9",
    label.size = 0.15
  ) +
  ggplot2::scale_colour_manual(
    values = pop_cols,
    labels = c("Cell type 1", "Cell type 2"),
    name = "Population"
  ) +
  ggplot2::scale_fill_manual(
    values = pop_cols,
    labels = c("Cell type 1", "Cell type 2"),
    name = "Population"
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.08, 0.28))
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.08, 0.28))
  ) +
  ggplot2::coord_equal() +
  ggplot2::facet_wrap(~panel, nrow = 1) +
  ggplot2::labs(
    x = "Expression of gene 1",
    y = "Expression of gene 2"
  ) +
  ggplot2::theme_bw(base_size = 11.5) +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(
      fill = "#f8f8f8",
      colour = "grey75"
    ),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

### Related signature constructions {#sec-related-signatures}

Blending a common baseline with type-specific markers is standard in deconvolution benchmarking. **`AutoGeneS`** [@alieeAutoGeneSAutomaticGene2021] selects genes by jointly minimising inter-type correlation and maximising centroid distance; its simulations use a similar shared-plus-private logic to control signature geometry. **`Schelker et al.`** [@schelkerEstimationImmuneCell2017] build tumour-derived reference gene expression profiles from single-cell RNA-seq clusters and marker genes, then simulate bulk mixtures by combining those columns—empirically a shared baseline plus distinct marker programmes per immune cell type.

The **`DICEPro`** framework [@baWhenLessNot2026] targets incomplete reference matrices in supervised deconvolution. Its simulation engine generates synthetic signature matrices from multivariate Gaussian (or Poisson) models with a prescribed correlation structure; a **`bloc=TRUE`** option enforces **block sparsity** by zeroing expression outside type-specific gene blocks—closely analogous to the private indicators $\boldsymbol{v}_j$ in @eq-lr-block. DICEPro then optimises reference signatures when cell types are missing from the matrix, complementing the controlled $\rho$-dial we use here for method comparison under known ground truth.

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a quantity to maximise [@alieeAutoGeneSAutomaticGene2021]. Cosine is preferred to Pearson correlation so that collinear but unequally scaled profiles remain penalised.

## Perspectives: a single geometric score for $\boldsymbol{\mu}$ {#sec-geometric-score}

The present API exposes two AutoGeneS-style objectives. A natural refinement is to replace them by one scalar that jointly rewards large column norms (and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$ {#eq-gram-volume}

is the volume of the parallelepiped spanned by the columns of $\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when columns lengthen or become more orthogonal—precisely the desired MOO trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$ {#eq-reciprocal-kappa}

available in R via `kappa(mean_profiles, exact = TRUE)` (or `1 / kappa(...)`), measures how ill-posed linear recovery of $\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$ is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising $\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes", "volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the constructive $(\rho,s)$ generator for controllable scenarios, and report the chosen scalar alongside the existing pairwise diagnostics.


# Undirected Gaussian Markov network generation {#sec-ggm-networks}

$$
G \longrightarrow W(G) \longrightarrow \Omega \succ 0
\longrightarrow X_i \sim \mathcal{N}_p(\mu_i,\Omega^{-1})
$$ {#eq-ggm-pipeline}

Here $G$ is an **undirected** graph, $W(G)$ a symmetric weighted matrix with the same off-diagonal support, $\Omega$ a strictly positive-definite precision, and $X_i$ an expression profile. This section covers **undirected** Gaussian Markov / precision-graph simulation only. `generate_random_network_skeleton()` draws $G$; `build_normalised_precision()` completes it to $\Omega\succ 0$ by an affine spectral shift; `build_covariance_array_from_precision()` maps each slice $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}_j^{-1}$ (cell types need not share one network).

```{mermaid}
%%| label: fig-ggm-pipeline
%%| fig-cap: "Undirected simulation pipeline from topology to observations (optional observation layer dashed)."
%%{init: {"theme": "sandstone"}}%%
flowchart LR
  G["Undirected graph G"] --> W["Signed weights W(G)"]
  W --> Omega["SPD precision Ω"]
  Omega --> Latent["Latent Gaussian X"]
  Mu["Mean design μ"] --> Latent
  Latent --> Obs["Observation layer"]
  Latent -.-> Obs
```

## Graph structure generators

$$
\rho_{jk\mid -\{j,k\}}
=
-\frac{\Omega_{jk}}{\sqrt{\Omega_{jj}\Omega_{kk}}}
$$ {#eq-partial-cor}

@tbl-topologies compares the main **undirected** topology families used in GGM benchmarks, with R entry points. @fig-topologies sketches the six families most often used in random-like network structure generation.

| Structure | Definition and controls | R entry points | Advantages | Limitations |
|---------------|---------------|---------------|---------------|---------------|
| **Erdős–Rényi** | Each edge present independently with probability $q$; $q=d_{\mathrm{avg}}/(p-1)$ targets average degree | `huge::huge.generator(graph="random")` [@R-huge]; `BDgraph::graph.sim(graph="random")` [@R-BDgraph]; DeCovarT `graph_model = "erdos_renyi"` | Exact sparsity control in expectation; few nuisance parameters | Narrow degree distribution; little modular organisation |
| **Band / AR** | Edge $j$–$k$ if $1\le\lvert j-k\rvert\le b$ | `huge.generator(graph="band")` [@R-huge]; `BDgraph::bdgraph.sim(graph="AR(1)")` / `"AR(2)"` [@R-BDgraph] | Local dependence and controlled max degree | Artificial gene ordering; no hubs or modules |
| **Hub / star** | $g$ groups; one centre linked to remaining members | `huge.generator(graph="hub")` [@R-huge]; `graph.sim(graph="hub")` / `"star"` [@R-BDgraph]; DeCovarT `graph_model = "hub"` | Stress-tests high-degree regulators | Pure stars omit cross-talk |
| **Scale-free** | Preferential attachment (e.g. Barabási–Albert with $m$ edges per new node) | `huge.generator(graph="scale-free")` [@R-huge]; DeCovarT `graph_model = "scale_free"` | Heterogeneous degrees | Weak modularity under plain BA growth |
| **Cluster / SBM** | Within-block edge probability exceeds between-block | `huge.generator(graph="cluster")` [@R-huge]; DeCovarT `graph_model = "stochastic_block_model"` | Pathway / module structure | Equal blocks can be unrealistically regular |
| **Small-world / lattice / circle** | High clustering with short paths, or fixed neighbourhoods | DeCovarT `graph_model = "small_world"` (Watts–Strogatz); `BDgraph::graph.sim(graph="smallworld")` [@R-BDgraph] | Local modules and spatial organisation | Narrow degrees; needs a defensible ordering |

: Topology families for undirected GGM simulation (R-focused). {#tbl-topologies}

## Biological relevance of the six families

Not every topology is equally useful as a **gene-regulatory** prior. The three simulation studies in @tbl-studies, together with classical network biology, suggest the following reading.

| Family | Biological reading | Why it matters for DeCovarT |
|----------------|-----------------------|----------------------------------|
| **Star / hub** | Master-regulator TF with many targets; one-to-many signalling | Stress-tests recovery of a few high-degree hubs that dominate partial correlations [@wuEstimatingHeterogeneousGene2022; @federicoStructureLearningGene2023] |
| **Scale-free (BA)** | Preferential attachment produces a heavy-tailed degree distribution and several hubs [@barabasiEmergenceScalingRandom1999] | Standard high-dimensional GGM stress test used by SILGGM via `huge` [@zhangSILGGMExtensivePackage2018]; useful for testing degree heterogeneity, but not a default empirical model of gene regulation |
| **Cluster / SBM** | Co-regulated modules / pathways with dense within-block links | Matches pathway organisation; BLGGM’s global block partition is SBM-like before finer within-block motifs [@wuEstimatingHeterogeneousGene2022] |
| **Small-world** | High local clustering with short paths (ring + shortcuts) | Local co-expression neighbourhoods plus long-range regulatory shortcuts; useful when modularity is soft rather than hard blocks [@federicoStructureLearningGene2023] |
| **Band / AR** | Ordered local dependence only | Useful numerical control, but gene indices rarely encode a natural order—limited biological fidelity [@zhangSILGGMExtensivePackage2018] |
| **Erdős–Rényi** | Homogeneous random wiring | Null / baseline sparsity; little pathway or hub structure [@zhangSILGGMExtensivePackage2018] |

: Biological interpretability of undirected topology families. {#tbl-topology-biology}

### The scale-free hypothesis: a useful stress test, not a biological default

The phrase *scale-free network* usually means that the degree distribution follows a power law, $\Pr(K=k)\propto k^{-\alpha}$, at least above a lower cut-off. This statistical statement is distinct from the Barabási–Albert (BA) growth mechanism. Preferential attachment can generate a power law, but observing a heavy-tailed degree distribution does not establish preferential attachment; other mechanisms can produce similar tails, and variants of preferential attachment need not produce a pure power law [@lima-mendezPowerfulLawPower2009; @broidoScalefreeNetworksAre2019].

Whether biological networks are scale-free has therefore remained contested. Early claims often relied on approximately straight log–log degree plots. Such plots are not goodness-of-fit tests: binning, the plotting scale, a small number of hubs, and incomplete or biased sampling can all make non-power-law distributions appear linear. The underlying graph representation matters as well. For example, pool metabolites can become artificial hubs in metabolic graphs, while protein-interaction networks depend strongly on the assay and sampling scheme [@lima-mendezPowerfulLawPower2009]. A power law should instead be fitted to the discrete degree data, tested for plausibility, and compared on the same support with alternatives such as log-normal, exponential, stretched-exponential, and power-law-with-cut-off distributions.

The evidence is not uniformly negative. Using adjacency-spectrum distributions rather than degree alone, Takahashi and colleagues classified eight protein–protein interaction networks as scale-free [@takahashiDiscriminatingDifferentClasses2012]. Importantly, this was a relative model-selection result: BA-like networks fitted better than the two candidate alternatives, Erdős–Rényi and Watts–Strogatz networks. It did not show that a power law was an adequate absolute model, that preferential attachment generated the observed networks, or that unobserved interactions would preserve the classification. The authors explicitly noted both the restricted candidate set and possible sampling artefacts.

A broader test of 928 real-world network data sets reached a more cautious conclusion [@broidoScalefreeNetworksAre2019]. Across all domains, only 4% met the strongest scale-free criterion, whereas a log-normal distribution fitted as well as or better than a power law for 88% of degree distributions. Among the biological networks, 63% showed neither direct nor indirect evidence of scale-free structure, although 6% met the strongest criterion, principally metabolic networks. These proportions depend on the network corpus and the operational definition of scale-freeness, but they reject a universal scale-free law rather than excluding scale-free structure from every biological system.

### Evidence from the literature

@tbl-studies summarises how recent simulation studies combine topology, precision construction, and mean / observation models.

| Source | Graph structures | Precision construction | Mean / observation | Purpose |
|---------------|---------------|---------------|---------------|---------------|
| PLNnetwork [@chiquetVariationalInferenceSparse2018] | Erdős–Rényi, preferential attachment, affiliation | $\Omega = vG + \operatorname{diag}(\lvert\lambda_{\min}(vG)\rvert + u)$ | Latent log-abundance $XB$; compositional / multinomial counts | Count-network estimators under compositionality |
| SILGGM [@zhangSILGGMExtensivePackage2018] | Band, hub, Erdős–Rényi, scale-free via `huge` | Package covariance / precision simulation | Centred Gaussian; focus on precision entries | Large-$p$ inferential and computational stress tests |
| BLGGM [@wuEstimatingHeterogeneousGene2022] | Block mixtures of dense, circle, star, signed modules | Module matrices with structured signed off-diagonals | Cell-type-specific $\mu_k$, $\Omega_k$; expression-dependent dropout | Joint clustering, heterogeneous networks, zero inflation |

: Literature survey of undirected graph → precision → observation designs. {#tbl-studies}

Several conclusions follow.

1.  **Band** is not intrinsically a random-graph model: its support is usually deterministic once the bandwidth is fixed (edge weights and signs may still be random). Hub constructions are often partly deterministic as well. Erdős–Rényi, preferential attachment, and stochastic block models are genuinely stochastic.
2.  A positive off-diagonal entry in $\Omega$ encodes a **negative** partial correlation (@eq-partial-cor). Uniformly positive precision weights therefore induce uniformly negative edgewise partial correlations; random or signed biological weights are needed when activation-like and inhibition-like associations matter.
3.  These models target **undirected** conditional dependence. Federico and colleagues prefer a Markov-network representation when observational data do not justify edge directionality [@federicoStructureLearningGene2023].
4.  **BLGGM as a hybrid topology design.** Wu and Luo’s simulation can be read as two nested layers [@wuEstimatingHeterogeneousGene2022]. At the **global** scale, genes are partitioned into blocks (a stochastic-block–like partition). Within each block, a **finer** local topology is specified—dense clique-like modules, circles, stars / hubs, or signed dense substructures—so that different blocks can stand for distinct regulatory regimes (e.g. dense co-expression programmes, cyclic signalling motifs, master-regulator hubs). Signs can also differ across modules or cell types. This is richer than a flat ER draw, yet still fully undirected.

For DeCovarT, a BA graph should consequently be interpreted as a deliberately demanding **hub-heterogeneity benchmark**. It tests whether covariance estimation and deconvolution recover a few high-degree vertices without claiming that a gene-regulatory network grew by preferential attachment. Results should be compared with density-matched SBM, small-world, star, and Erdős–Rényi scenarios. Conclusions that depend only on the BA scenario should be labelled topology-specific; hub recovery, robustness to vertex removal, and biological realism do not follow from a fitted heavy tail alone.

**BLGGM as a hybrid.** Wu and Luo combine an SBM-like **global** gene partition with **local** dense, circle, star, or signed modules inside blocks [@wuEstimatingHeterogeneousGene2022]. Distinct local motifs can stand for distinct processes (e.g. master-regulator stars vs dense co-expression programmes). DeCovarT’s exported generators currently expose the three families most useful as standalone benchmarks: `scale_free`, `stochastic_block_model`, and `small_world`.


![Six undirected topology families used in GGM simulation: Erdős–Rényi, Barabási–Albert (preferential attachment / scale-free), stochastic block (cluster), band / AR, star / hub, and small-world (Watts–Strogatz).\
\
](figures/fig_network_topologies_six.png){#fig-topologies}

## Weight design and random signs

Topology generators return a binary undirected support $E$. **Signs and magnitudes** are a separate design layer $W(G)$ in @eq-ggm-pipeline: they are assigned *after* (or jointly with) the skeleton, then completed to an SPD precision. Remember that

$$
\operatorname{sign}(\rho_{jk\mid -\{j,k\}})
=
-\operatorname{sign}(\Omega_{jk})
$$ {#eq-sign-flip}

see @eq-partial-cor, so a positive precision weight is an inhibitory partial correlation. @tbl-weights contrasts three standard undirected strategies.

| Approach | What is randomised | How $\Omega$ is obtained | When to use |
|------------------|------------------|------------------|------------------|
| **i.i.d. edge signs** | For each $\{j,k\}\in E$, draw $s_{jk}\in\{-1,+1\}$ (e.g. fair coin or $\mathbb{P}(+)=\pi$) and magnitude $m_{jk}>0$; set $W_{jk}=W_{kj}=s_{jk}m_{jk}$ | Support-preserving SPD map of $W$ (spectral shift, diagonal dominance, …) | Simple signed ER / hub / SBM benchmarks; explicit control of the inhibitory fraction |
| **Partial-correlation scaling** | Fill a symmetric matrix $R$ with $R_{jk}=0$ off $E$ and $R_{jk}\in(-1,1)$ (signed) on $E$; ensure $\rho(R)<1$ | $\Omega = D^{1/2}(I-R)D^{1/2}$ (@eq-partial-scale) | Direct control of partial-correlation signs and strengths |
| **G-Wishart** (`BDgraph::rgwish()`) | Continuous random $\Omega$ on the fixed graph zeros | $\Omega\sim W_G(b,D)$ already SPD [@R-BDgraph; @BDgraph2019] | Heterogeneous random weights without hand-tuned magnitudes; Bayesian-style replicates |

: Three undirected strategies for signed / random edge weights after a fixed skeleton. {#tbl-weights}

### i.i.d. signs on a fixed skeleton

Draw $G$ (ER, band, hub, …). Independently for each undirected edge,

$$
W_{jk}=W_{kj}=s_{jk}\,m_{jk},
\qquad
s_{jk}\in\{-1,+1\},
\quad m_{jk}>0,
$$ {#eq-iid-signs}

then apply a support-preserving SPD completion (next section). Uniform positive loadings $W=vG$ are the special case $s_{jk}\equiv +1$ used by PLN-style and many `huge` simulations [@chiquetVariationalInferenceSparse2018]—all partial correlations then share the same sign. DeCovarT uses i.i.d. signs with inhibitory fraction `prop_inhibitory` and magnitude $v$ (`precision_scale`) baked into $\boldsymbol{W}$ before the spectral SPD completion.

### Partial-correlation scaling and why $I-R$

In a Gaussian Markov model the **matrix of partial correlations** (with ones on the diagonal) relates to the precision by a diagonal congruence. Write the off-diagonal partial correlations as a symmetric matrix $R$ with

$$
R_{jj}=0,
\qquad
R_{jk}
=
\begin{cases}
\rho_{jk\mid -\{j,k\}} & \{j,k\}\in E,\\
0 & \{j,k\}\notin E,
\end{cases}
$$ {#eq-R-support}

and choose magnitudes so that the spectral radius satisfies $\rho(R)<1$. Then

$$
\Omega
=
D^{1/2}(I-R)D^{1/2}
$$ {#eq-partial-scale}

for a positive diagonal $D$ (often $D=I$ after standardising margins).

### G-Wishart draws on a fixed graph

Given adjacency $G$, `BDgraph::rgwish()` samples $\Omega\sim W_G(b,D)$ already positive definite and with $\Omega_{jk}=0$ whenever $(j,k)\notin E$ [@R-BDgraph; @BDgraph2019]. Signs and magnitudes arise from the continuous distribution rather than from an explicit coin-flip layer. This is the most “automatic” signed-weight generator once the skeleton is fixed; the trade-off is less direct control of the inhibitory fraction than @eq-iid-signs or @eq-partial-scale.

## Positive-definiteness strategies

@tbl-pd contrasts constructions that turn a weighted support $W$ (or a partial-correlation design) into $\Omega\succ 0$. DeCovarT’s `build_normalised_precision()` implements the uniform spectral shift below, with diagonal cushion $u$ (`precision_shift`):

$$
\boldsymbol{\Omega}
=
\boldsymbol{W}
+
\bigl(
  \lvert\lambda_{\min}(\boldsymbol{W})\rvert + u
\bigr)
\mathbf{I}.
$$

| Construction | Formula | SPD? | Preserves exact support? | Behaviour |
|---------------|---------------|--------------:|--------------:|---------------|
| Uniform spectral shift | $\Omega=W+\max(0,\varepsilon-\lambda_{\min}(W))I$ | Yes | Yes | Same shift on every eigenvalue; off-diagonals (hence signs) unchanged |
| `huge` / PLN-style loading | Diagonal load from $\lvert\lambda_{\min}\rvert$, baseline, and $u$ | Yes | Yes | $v$ sets edge magnitude; loading sets conditioning [@chiquetVariationalInferenceSparse2018] |
| Partial-correlation scaling | $\Omega=D^{1/2}(I-R)D^{1/2}$ with $\rho(R)<1$ | Yes | Yes | Signs and strengths set on the partial-correlation scale (@eq-partial-scale) |
| Graph Laplacian / M-matrix | $\Omega=L_G+\varepsilon I$ | Yes | Yes | Off-diagonals non-positive ⇒ all edgewise partial correlations positive |
| G-Wishart | $\Omega\sim W_G(b,D)$ | Yes | Yes | Heterogeneous random weights; `BDgraph::rgwish()` [@R-BDgraph] |
| Eigenvalue clipping | Clip $\Lambda$ then reconstruct $Q\Lambda Q^{\top}$ | Yes | **No** | Fills structural zeros |
| Nearest PD projection | e.g. `Matrix::nearPD()` [@R-Matrix] | Yes | **No** (in general) | Repair tool, not a support-preserving truth |

: Positive-definiteness constructions for graph-constrained precisions. {#tbl-pd}

### Why the uniform spectral shift is attractive

If $W=Q\Lambda Q^{\top}$, then (@eq-spectral-shift)

$$
W+cI = Q(\Lambda+cI)Q^{\top}.
$$ {#eq-spectral-shift}

Every eigenvalue increases by $c$, while every off-diagonal of $W$ is unchanged, so the edge set

$$
E=\{(j,k):j<k,\ \Omega_{jk}\neq 0\}
$$ {#eq-edge-set}

is identical before and after loading—and **signs of** $W$ are preserved.

The floor $\varepsilon$ also controls the condition number

$$
\kappa(\Omega)=\frac{\lambda_{\max}(\Omega)}{\lambda_{\min}(\Omega)}.
$$ {#eq-omega-kappa}

Large loading yields a well-conditioned but weakly dependent network; loading only slightly above $-\lambda_{\min}$ strengthens dependence but can make estimation fragile. Benchmarks should report partial-correlation distributions and $\kappa(\Omega)$, not only the adjacency.

## Cell-type covariances: `build_covariance_array_from_precision()`

By default each cell type has its own precision (and thus covariance),

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}_j^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Shared networks remain available by supplying the same adjacency (or the same graph model seed path) for every type. Downstream bulk simulation uses the Gaussian convolution $\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

## Conclusions

> DeCovarT draws `scale_free`, `stochastic_block_model`, and `small_world` skeletons with \pkg{igraph}, then i.i.d. signed weights (`prop_inhibitory`) and a spectral shift to guarantee $\Omega\succ 0$.

### Further high-dimensional designs

Beyond the families in @tbl-topologies:

- **Degree-prescribed / configuration** graphs separate degree heterogeneity from preferential-attachment *mechanisms*.
- **Spatial / geometric** graphs suit spatial transcriptomics or chromatin neighbourhoods.
- **Differential networks:**
  - Start from a shared base $G_0$
  - Control the differential shift $\Omega_k=\Omega_0+\Delta_k$
  - Ensure the SPD step
  - Alternatively, vary the *support, weight, and sign* separately [@federicoStructureLearningGene2023; @wuEstimatingHeterogeneousGene2022].
- **Other distributions,** using the GGM layer as a parameter:
  - **Latent-variable GGMs** — sparse $\Omega_S$ plus low-rank confounding $Bf_i$.
  - **Nonparanormal / Gaussian-copula** margins on a latent Gaussian graph.
  - **Poisson–log-normal / compositional** observation layers on latent log-abundances [@chiquetVariationalInferenceSparse2018].
  - **Zero-inflated mixture GGMs** with cell-type-specific $(\mu_k,\Omega_k)$ and expression-dependent dropout [@wuEstimatingHeterogeneousGene2022].

### Recommended benchmark specification

Use one pipeline for every topology (@eq-benchmark-pipe, @fig-ggm-pipeline):

$$
\text{topology}
\rightarrow
\text{signed weights}
\rightarrow
\text{SPD precision}
\rightarrow
\text{mean design}
\rightarrow
\text{latent Gaussian}
\rightarrow
\text{observation model}
$$ {#eq-benchmark-pipe}

| Axis | Suggested levels |
|-----------------------|------------------------------------------------|
| Dimension, namely number of variables (here, genes) | $p \in \{100,500,1000\}$ |
| Topology | ER, band, hub, scale-free, SBM, small-world |
| Expected degree | $\approx 2,4,8$ (keep graphs sparse) |
| Edge weights | Constant; i.i.d. signed; partial-correlation scaling; G-Wishart |
| Conditioning | Report $\lambda_{\min}$, $\lambda_{\max}$, $\kappa(\Omega)$ |
| Means | Zero; sparse group shifts, driven by cell-type; joint $\mu$–$\Omega$ heterogeneity, for example playing on the *cosine similarity*, or the *condition number*, of the mean profile |

: Compact factorial axes for a reproducible undirected GGM simulation study. {#tbl-benchmark-axes}

**Main insights.** Prefer the uniform spectral shift (@eq-spectral-shift) or partial-correlation scaling (@eq-partial-scale) when support and signs must be exact; add the mean layer **independently** of graph generation.




# Synthetic simulation design: a hybrid multi-topology reference scenario {#sec-scenario-grid}

The earlier factorial design swept one topology and one cosine level at a time, keeping a single **shared** covariance across all $J$ cell types (`simulate_hierarchical_grn_moments()`; @sec-ggm-networks). `scripts/generate_random_markov_network.R` instead builds **one** scenario that stresses several axes of the feature-selection pipeline simultaneously: two cell types that are unrecoverable from $\boldsymbol{\mu}$ alone and must be told apart from **topology**, a third cell type set apart by a compact marker block, and a null block that should survive every selection stage as *discarded*. Notation matches the manuscript: $G=50$ genes, $J=3$ cell types, $N$ bulk samples for $\boldsymbol{Y}\in\mathcal{M}_{G\times N}$.

## Gene and cell-type design

The $G=50$ genes are partitioned into three blocks, and each cell type gets its **own** $G\times G$ precision $\boldsymbol{\Omega}_j$ (not a shared one):

$$
\underbrace{\mathcal{G}_{12}}_{30\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{3}}_{10\text{ genes}}
\ \cup\
\underbrace{\mathcal{G}_{\mathrm{eq}}}_{10\text{ genes}},
\qquad
\boldsymbol{\Omega}_j
=
\mathrm{blockdiag}\bigl(
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{12})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{3})},\,
  \boldsymbol{\Omega}_j^{(\mathcal{G}_{\mathrm{eq}})}
\bigr),
\qquad j=1,2,3.
$$

$\boldsymbol{\mu}$ is built by **two** iterative calls to `generate_mean_signature_matrix()` (@eq-mean-cosine): first on $\mathcal{G}_{12}$ for the pair (cell type 1, cell type 2) with a **high** target cosine $\rho_{12}=0.95$ (near-collinear, per @sec-lr-block-mu); then on a pool of $2\times10$ genes for the pair (merged $\{1,2\}$, cell type 3) with a **low** target cosine $\rho_{3}=0.05$ (near-orthogonal), keeping only the 10 genes whose private block belongs to cell type 3. $\mathcal{G}_{\mathrm{eq}}$ gets a flat baseline directly (identical mean in all three types by construction). Only $\mathcal{G}_{12}$ and $\mathcal{G}_3$ carry a mean signal; cell types 1 and 2 are, by design, mean-indistinguishable on $\mathcal{G}_{12}$.

Following the BLGGM hybrid design (@sec-ggm-networks), each block's **local** topology is set independently per cell type, with no edges between blocks:

| Gene block | # genes | Cell type 1 | Cell type 2 | Cell type 3 |
|---|---|---|---|---|
| `shared_12_vs_3` | 30 | stochastic block model (hub-like modules) | star (single key-driver gene) | Erdős–Rényi (background) |
| `marker_3` | 10 | Erdős–Rényi (background) | Erdős–Rényi (background) | scale-free |
| `equal_all` | 10 | Erdős–Rényi | Erdős–Rényi | Erdős–Rényi |

: Local topology assigned to each (gene block, cell type) pair; no edges connect different blocks (block-diagonal precision support). {#tbl-hybrid-topology-design}

Cell types 1 and 2 therefore differ **only** on `shared_12_vs_3`, and only through $\boldsymbol{\Omega}$: hub/stochastic-block modules (cascading-pathway-like) for cell type 1 versus a single star (one key-driver gene) for cell type 2. `equal_all` is wired identically (Erdős–Rényi) in every cell type, matching its complete lack of mean signal.

```{r}
#| label: setup-hybrid-scenario
#| warning: false
#| message: false
#| eval: true
#| echo: false
#| output: false

if (requireNamespace("DeCovarT", quietly = TRUE)) {
  library(DeCovarT)
} else {
  stop("Package 'DeCovarT' is not installed.")
}

set.seed(20260807)

n_g12 <- 30L
n_g3 <- 10L
n_geq <- 10L
n_genes <- n_g12 + n_g3 + n_geq
celltype_names <- c("celltype_1", "celltype_2", "celltype_3")
idx_g12 <- seq_len(n_g12)
idx_g3 <- n_g12 + seq_len(n_g3)
idx_geq <- n_g12 + n_g3 + seq_len(n_geq)

mean_scale <- 10
mu_12 <- generate_mean_signature_matrix(
  n_genes = n_g12,
  n_celltypes = 2L,
  mean_scale = mean_scale,
  target_cosine = 0.95,
  celltype_names = c("celltype_1", "celltype_2")
)
mu_3_pool <- generate_mean_signature_matrix(
  n_genes = 2L * n_g3,
  n_celltypes = 2L,
  mean_scale = mean_scale,
  target_cosine = 0.05,
  celltype_names = c("celltype_12_merged", "celltype_3")
)
mu_3_block <- mu_3_pool[(n_g3 + 1L):(2L * n_g3), , drop = FALSE]

mu <- matrix(NA_real_, n_genes, 3L, dimnames = list(NULL, celltype_names))
mu[idx_g12, "celltype_1"] <- mu_12[, "celltype_1"]
mu[idx_g12, "celltype_2"] <- mu_12[, "celltype_2"]
mu[idx_g12, "celltype_3"] <- 0.5
mu[idx_g3, "celltype_1"] <- mu_3_block[, "celltype_12_merged"]
mu[idx_g3, "celltype_2"] <- mu_3_block[, "celltype_12_merged"]
mu[idx_g3, "celltype_3"] <- mu_3_block[, "celltype_3"]
mu[idx_geq, ] <- 2

precision_shift <- 0.1
precision_scale <- 0.3
prop_inhibitory <- 0.5

block_defs <- list(
  celltype_1 = list(
    list(idx_g12, "stochastic_block_model", list(block_prob = c(0.4, 0.3, 0.3))),
    list(idx_g3, "erdos_renyi", list()),
    list(idx_geq, "erdos_renyi", list())
  ),
  celltype_2 = list(
    list(idx_g12, "hub", list(n_hubs = 1L)),
    list(idx_g3, "erdos_renyi", list()),
    list(idx_geq, "erdos_renyi", list())
  ),
  celltype_3 = list(
    list(idx_g12, "erdos_renyi", list()),
    list(idx_g3, "scale_free", list(power = 1, edges_per_node = 1L)),
    list(idx_geq, "erdos_renyi", list())
  )
)
topology_label <- c(
  celltype_1 = "SBM (hub-like modules)",
  celltype_2 = "star (single key driver)",
  celltype_3 = "scale-free (marker genes)"
)

spectral_rows <- list()
for (ct in celltype_names) {
  adjacency <- matrix(0L, n_genes, n_genes)
  for (blk in block_defs[[ct]]) {
    idx <- blk[[1]]
    adjacency[idx, idx] <- DeCovarT:::generate_random_network_skeleton(
      n_genes = length(idx),
      graph_model = blk[[2]],
      graph_params = blk[[3]]
    )
  }
  W <- DeCovarT:::assign_iid_signed_weights(
    adjacency,
    prop_inhibitory = prop_inhibitory,
    weight_magnitude = precision_scale
  )
  Omega <- DeCovarT:::build_normalised_precision(W, precision_shift)
  evals <- eigen(Omega, symmetric = TRUE, only.values = TRUE)$values
  upper_w <- W[upper.tri(W) & adjacency != 0L]
  spectral_rows[[ct]] <- data.frame(
    `cell type` = ct,
    topology = topology_label[[ct]],
    lambda_min = min(evals),
    lambda_max = max(evals),
    kappa_Omega = max(evals) / min(evals),
    prop_inhib = if (length(upper_w)) mean(upper_w > 0) else NA_real_,
    check.names = FALSE
  )
}
spectral_df <- do.call(rbind, spectral_rows)
rownames(spectral_df) <- NULL

pairs <- utils::combn(celltype_names, 2L)
pairwise_rows <- list()
for (k in seq_len(ncol(pairs))) {
  ct_j <- pairs[1L, k]
  ct_l <- pairs[2L, k]
  mu_j <- mu[, ct_j]
  mu_l <- mu[, ct_l]
  pairwise_rows[[k]] <- data.frame(
    pair = paste(ct_j, "vs", ct_l),
    cosine = sum(mu_j * mu_l) / sqrt(sum(mu_j^2) * sum(mu_l^2)),
    euclidean = sqrt(sum((mu_j - mu_l)^2))
  )
}
pairwise_df <- do.call(rbind, pairwise_rows)
```

```{r}
#| label: tbl-mean-geometry
#| tbl-cap: "Realised pairwise cosine and Euclidean distance of the 50 x 3 mean signature, over the full G = 50 genes."
#| warning: false
#| message: false
#| echo: false
tinytable::tt(pairwise_df, digits = 3) |>
  tinytable::style_tt(j = 2:3, align = "c")
```

```{r}
#| label: tbl-topology-diagnostics
#| tbl-cap: "Per-cell-type precision spectrum for the 50 x 50 hybrid-topology Omega_j."
#| warning: false
#| message: false
#| echo: false
# Math-friendly column names for HTML (markdown + KaTeX via format_tt).
spectral_tt <- spectral_df
names(spectral_tt) <- c(
  "cell type",
  "topology",
  "$\\lambda_{\\min}$",
  "$\\lambda_{\\max}$",
  "$\\kappa(\\Omega)$",
  "prop inhib"
)
tinytable::tt(spectral_tt, digits = 3) |>
  tinytable::style_tt(j = 3:6, align = "c") |>
  tinytable::format_tt(j = 3:5, markdown = TRUE)
```

Column guide: $\lambda_{\min}$, $\lambda_{\max}$, and $\kappa(\boldsymbol{\Omega})=\lambda_{\max}/\lambda_{\min}$ summarise the precision spectrum of each cell type's *own* $\boldsymbol{\Omega}_j$; `prop_inhib` is the realised fraction of inhibitory precision edges. @tbl-mean-geometry shows cell types 1 and 2 near-collinear (cosine $\approx 0.97$), tracking the local target $\rho_{12}=0.95$ used on `shared_12_vs_3`. Cell type 3, however, sits at only a *moderate* cosine ($\approx0.56$) from the other two—well above the local target $\rho_{3}=0.05$ used on the `marker_3` pool. The gap is exactly the finite-sample effect of @sec-asymptotic-cosine: the flat `celltype_12_merged` background on `shared_12_vs_3` and the identical `equal_all` baseline both contribute a strictly positive term to *every* pairwise inner product over the full $G=50$ vector, diluting the block-local orthogonality of `marker_3` alone. @tbl-topology-diagnostics shows that, despite the shared mean on `shared_12_vs_3`, cell types 1 and 2 carry distinct, well-conditioned precision spectra—exactly the topology-only separation the scenario is designed to require.

@fig-network-topologies renders the three realised networks with `igraph`, coloured by gene block and by precision-edge sign.

![Cell-type-specific block topologies for the 50-gene hybrid scenario: a stochastic-block/hub-like module structure for cell type 1, restricted to the 30 `shared_12_vs_3` genes; a single star (one key-driver gene) for cell type 2 on the same 30 genes; and a scale-free network for cell type 3 restricted to its own 10 `marker_3` genes. The `equal_all` block (grey) is wired as Erdős–Rényi in every cell type. Edge colour encodes the sign of the precision entry (red = inhibitory, teal = activatory); node colour encodes the gene block.](figures/fig_network_topologies.png){#fig-network-topologies width="100%"}

The full pipeline built on this scenario—mean signature, per-cell-type topologies, $N=200$ bulk simulation, and the pre-selection / NSGA-II feature-selection stages it stresses—is reproduced end-to-end in `scripts/generate_random_markov_network.R`.

## Related literature

Pseudobulk simulation tools such as **`muscat`** [@crowellMuscatDetectsSubpopulationspecific2020] aggregate single-cell counts across samples and conditions for differential-state testing; **`scDD`** [@korthauerStatisticalApproachIdentifying2016] partitions genes, relative to a reference condition, into five differential-distribution (DD) patterns: equivalent expression (**EE**), differential expression (**DE**, a mean shift with one mode preserved per condition), differential proportion (**DP**, a shift in the mixing weights of a shared bimodal pattern), differential modality (**DM**, a change in the number of modes without necessarily shifting the overall mean), and differential expression **and** modality combined (**DB**).

The hybrid scenario above instantiates a loose, network-level analogue of three of these patterns, tailored to deconvolution rather than to single-cell distribution testing directly (@fig-dd-taxonomy):

- the 10 `equal_all` genes are **EE**: identical mean and identically Erdős–Rényi-wired second-order structure in every cell type;
- the 10 `marker_3` genes are **DE** against the merged $\{1,2\}$ background: a mean shift confined to cell type 3, layered on a scale-free local topology;
- the 30 `shared_12_vs_3` genes are **DE** against cell type 3 but, between cell types 1 and 2 specifically, carry **no mean shift at all**—only the precision-matrix topology (@tbl-hybrid-topology-design) differs. This is not scDD's original single-gene DM test (unimodal versus bimodal *marginal* densities within one gene), but it plays a structurally similar role here: two populations a mean-only model cannot separate, which a covariance-aware model such as DeCovarT is designed to exploit.

![Schematic differential-distribution taxonomy for the three gene blocks: equivalent expression (EE, `equal_all`), differential expression marking cell type 3 (DE, `marker_3`), and a differential-modality-like contrast between cell types 1 and 2 that is invisible at the mean level and only resolved by network topology (DM-like, `shared_12_vs_3`).](figures/fig_dd_taxonomy_ee_de_dm.png){#fig-dd-taxonomy width="100%"}

DeCovarT's generator is narrower—it fixes $\boldsymbol{\mu}$ through @eq-mean-cosine and @eq-lr-block rather than estimating signatures from real scRNA-seq—but the same low-rank-plus-block intuition underpins many reference-matrix constructions cited above.


# References
