---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../inst/REFERENCES.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator
`simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture
simulation. Mean profiles are constructed so that pairwise collinearity and
centroid separation can be dialled independently, following the AutoGeneS
multi-objective rationale of @aliee_theis21: minimise correlation (here,
cosine similarity) between cell-type columns of
$\boldsymbol{\mu}$ while maximising their pairwise Euclidean distances.
Covariances share a graph-constrained precision
$\boldsymbol{\Omega}$, mapped to
$\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that
can be passed to `simulate_bulk_mixture()` and then to any deconvolution
routine exposed by `deconvolute_ratios()`.

```{r pipeline-mermaid, echo=FALSE, results='asis'}
mmd_path <- "figures/synthetic_grn_pipeline.mmd"
if (!file.exists(mmd_path)) {
  mmd_path <- file.path("vignettes", mmd_path)
}
mmd <- paste(readLines(mmd_path), collapse = "\n")
cat("```mermaid\n", mmd, "\n```\n", sep = "")
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  graph_model = "power_law",
  graph_params = list(power = 1, edges_per_node = 2)
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is
insufficient for stable deconvolution [@aliee_theis21]: low Euclidean
separation makes ratio estimates noise-sensitive, and collinear centroids
inflate variance of $\hat{\boldsymbol{p}}$. The two knobs `target_cosine`
and `mean_scale` map onto that trade-off without running a genetic algorithm
at simulation time.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()`

Each column $\boldsymbol{\mu}_{\cdot j}$ is a normalised blend of a shared
direction $\boldsymbol{u}$ (uniform over genes) and a private marker block
$\boldsymbol{v}_{j}$ (high on a gene partition for type $j$, low elsewhere):

$$
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{
  \sqrt{\rho}\,\boldsymbol{u}
  +\sqrt{1-\rho}\,\boldsymbol{v}_{j}
}{
  \bigl\|
    \sqrt{\rho}\,\boldsymbol{u}
    +\sqrt{1-\rho}\,\boldsymbol{v}_{j}
  \bigr\|_2
},
\qquad
j=1,\ldots,J.
$$

Here $\rho\in[0,1]$ is `target_cosine` and $s>0$ is `mean_scale`. Small
$\rho$ yields complementary centroids; $\rho\to 1$ recovers a shared
direction (cosine one, vanishing Euclidean gap). Small $s$ places centroids
near the origin and shrinks pairwise distances while preserving angles.

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a
quantity to maximise [@aliee_theis21]. Cosine is preferred to Pearson
correlation so that collinear but unequally scaled profiles remain penalised.

## Network skeleton: `generate_random_network_skeleton()`

A binary adjacency $\boldsymbol{A}\in\{0,1\}^{G\times G}$ is drawn from either
a preferential-attachment (power-law / hub) model or a stochastic block model
(community structure). Both return an undirected graph with zero diagonal;
they encode presence of partial-correlation edges, not their strengths.

## Positive-definite precision: `build_normalised_precision()`

Edge weights enter through an affine spectral shift of $\boldsymbol{A}$,
matching the construction used in the manuscript:

$$
\tilde{\boldsymbol{\Omega}}
=
\boldsymbol{A}\,v,
\qquad
\boldsymbol{\Omega}
=
\tilde{\boldsymbol{\Omega}}
+
\bigl(
  \lvert\lambda_{\min}(\tilde{\boldsymbol{\Omega}})\rvert + u
\bigr)
\mathbf{I},
$$

with off-diagonal scale $v$ (`precision_scale`) and diagonal cushion $u$
(`precision_shift`). Larger $v$ strengthens interactions and spreads the
spectrum; larger $u$ improves conditioning. The map
$Y=aX+bI$ leaves eigenvectors unchanged and transforms eigenvalues as
$\lambda_i(Y)=a\lambda_i(X)+b$; choosing
$b=\lvert\lambda_{\min}(\tilde{\boldsymbol{\Omega}})\rvert+u$ with $a=1$
therefore guarantees $\boldsymbol{\Omega}\succ 0$ while preserving the
sparsity pattern of $\boldsymbol{A}$ off the diagonal.

## Shared covariances: `build_shared_covariance_array()`

Cell types share the same second-order structure,

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Downstream bulk
simulation uses the Gaussian convolution
$\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

# Perspectives: a single geometric score for $\boldsymbol{\mu}$

The present API exposes two AutoGeneS-style objectives. A natural refinement
is to replace them by one scalar that jointly rewards large column norms
(and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$

is the volume of the parallelepiped spanned by the columns of
$\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when
columns lengthen or become more orthogonal—precisely the desired MOO
trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$

available in R via `kappa(mean_profiles, exact = TRUE)` (or
`1 / kappa(...)`), measures how ill-posed linear recovery of
$\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$
is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising
$\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a
deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes",
"volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the
constructive $(\rho,s)$ generator for controllable scenarios, and report the
chosen scalar alongside the existing pairwise diagnostics.

# References
