---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator `simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture simulation. Mean profiles follow an AutoGeneS-inspired construction [@alieeAutoGeneSAutomaticGene2021] in which a single dial $\rho$ (`target_cosine`) sets pairwise collinearity between cell-type columns of $\boldsymbol{\mu}$, while a fixed scale $s$ (`mean_scale`) sets column norms. Prefer varying $\rho$ across scenarios when precision weights already control second-order dependence. Covariances share a graph-constrained signed precision $\boldsymbol{\Omega}$, mapped to $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that can be passed to `simulate_bulk_mixture()` and then to any deconvolution routine exposed by `deconvolute_ratios()`.

```{mermaid}
%%| label: fig-pipeline-api
%%| fig-cap: "Pipeline from synthetic moment generation to deconvolution."
%%{init: {"theme": "sandstone"}}%%
flowchart LR
  A["Graph G"] --> B["Weights W(G)"]
  B --> C["Precision Ω ≻ 0"]
  C --> D["Σ = Ω^{-1}"]
  E["Means μ_j\n(AutoGeneS-style ρ)"] --> F["simulate_hierarchical_grn_moments()"]
  D --> F
  F --> G["simulate_bulk_mixture()"]
  G --> H["deconvolute_ratios()"]
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  prop_inhibitory = 0.5,
  graph_model = "scale_free"
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is
insufficient for stable deconvolution [@alieeAutoGeneSAutomaticGene2021]. In
the scenarios of this package we therefore **hold `mean_scale` fixed** and
dial only `target_cosine`, so first-order scale and second-order precision
weights stay comparable across runs.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()`

Notation: $G$ genes, $J$ cell types indexed by $j=1,\ldots,J$, and
columns $\boldsymbol{\mu}_{\cdot j}\in\mathbb{R}^{G}$ of the mean
signature. Fix the column scale $s$ (`mean_scale`, default $10$ in the
nine-scenario grid) and dial only the target cosine $\rho\in[0,1]$
(`target_cosine`).

Let $\boldsymbol{u}=G^{-1/2}\mathbf{1}_{G}$ be the shared unit direction
and let $\boldsymbol{v}_{j}$ be the $\ell_2$-normalised indicator of the
gene block assigned to type $j$ (disjoint partition of
$\{1,\ldots,G\}$). Then $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$
for $j\neq k$. Each column is the normalised blend

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}
=
\sqrt{\rho}\,\boldsymbol{u}
+\sqrt{1-\rho}\,\boldsymbol{v}_{j},
\qquad
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{\tilde{\boldsymbol{\mu}}_{\cdot j}}{
  \|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2
},
\qquad
j=1,\ldots,J.
$$ {#eq-mean-cosine}

Expanding the un-normalised inner product for $j\neq k$ gives the exact
identity

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
=
\rho
+
\sqrt{\rho(1-\rho)}\,
\bigl(
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}
  +
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{k}
\bigr),
$$ {#eq-mean-inner}

because $\boldsymbol{u}^{\mathsf{T}}\boldsymbol{u}=1$ and
$\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$. The cross terms
$\boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}$ equal the (positive) mass of
the shared direction on block $j$; they vanish in the idealised case
where each $\boldsymbol{v}_{j}$ is taken orthogonal to $\boldsymbol{u}$,
in which @eq-mean-inner collapses to $\rho$ and, since
$\|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2=1$, the cosine of
$\boldsymbol{\mu}$ equals $\rho$ exactly. With block indicators the
cross terms are of order $J^{-1/2}$, so after the column
renormalisation in @eq-mean-cosine the realised pairwise cosine tracks
$\rho$ closely once each block contains many genes.

The scale $s$ sets column norms without changing angles:
$\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2\propto s$ at
fixed $\rho$. We keep $s=10$ across scenarios and only vary $\rho$.

### Visualising the cosine construction

@fig-cosine-geometry visualises the same three operating points
($\rho\in\{0,\,0.3,\,1\}$) in a two-gene projection sampled from a
$G=200$ construction. We add small Gaussian perturbations to emulate
biological dispersion around each cell-type centroid while preserving the
targeted cosine geometry.

This illustration is inspired by panel B in the AutoGeneS paper
[@alieeAutoGeneSAutomaticGene2021], but focuses here on the cosine-angle
control induced by @eq-mean-cosine (without modelling centroid-distance
optimisation).

```{r}
#| label: fig-cosine-geometry
#| fig-cap: >-
#|   Cosine-control toy for $J=2$ cell types with $G=200$ genes.
#|   Each panel shows noisy point clouds around centroids (stars), with
#|   realised angle and cosine. Styling is inspired by panel B of AutoGeneS
#|   (@alieeAutoGeneSAutomaticGene2021).
#| fig-width: 9
#| fig-height: 4.6
#| echo: false
#| warning: false
#| message: false
#| eval: true
rhos_fig <- c(0, 0.3, 1)
set.seed(20260804)
g_total <- 200L
s_fig <- 10
n_points <- 35L
direction_noise_sd <- 0.02
point_noise_sd <- 0.33
gene_pairs <- list(c(11L, 12L), c(71L, 72L), c(151L, 152L))

scatter_df <- list()
centroid_df <- list()
label_df <- list()

for (k in seq_along(rhos_fig)) {
  rho <- rhos_fig[[k]]
  mu <- DeCovarT::generate_mean_signature_matrix(
    n_genes = g_total,
    n_celltypes = 2L,
    mean_scale = s_fig,
    target_cosine = rho,
    direction_noise_sd = direction_noise_sd,
    random_seed = 100 + k
  )
  cos_hat <- sum(mu[, 1] * mu[, 2]) /
    sqrt(sum(mu[, 1]^2) * sum(mu[, 2]^2))
  angle_deg <- acos(max(-1, min(1, cos_hat))) * 180 / pi
  pair <- gene_pairs[[k]]
  g_x <- pair[[1]]
  g_y <- pair[[2]]
  panel <- sprintf(
    "rho = %s",
    format(rho, nsmall = 1)
  )

  for (j in 1:2) {
    scatter_df[[length(scatter_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = paste0("cell_", j),
      x = stats::rnorm(n_points, mean = mu[g_x, j], sd = point_noise_sd),
      y = stats::rnorm(n_points, mean = mu[g_y, j], sd = point_noise_sd)
    )
    centroid_df[[length(centroid_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = paste0("cell_", j),
      x = mu[g_x, j],
      y = mu[g_y, j]
    )
  }
  label_df[[length(label_df) + 1L]] <- data.frame(
    panel = panel,
    x = min(mu[g_x, ]) - 0.7,
    y = max(mu[g_y, ]) + 0.8,
    label = sprintf("realised cos = %.2f\nangle = %.1f°", cos_hat, angle_deg)
  )
}

points_df <- do.call(rbind, scatter_df)
centroids <- do.call(rbind, centroid_df)
labels <- do.call(rbind, label_df)
panel_levels <- sprintf("rho = %s", format(rhos_fig, nsmall = 1))
points_df$panel <- factor(points_df$panel, levels = panel_levels)
centroids$panel <- factor(centroids$panel, levels = panel_levels)
labels$panel <- factor(labels$panel, levels = panel_levels)

print(
  ggplot2::ggplot(points_df, ggplot2::aes(x = x, y = y, colour = celltype)) +
    ggplot2::geom_hline(yintercept = 0, colour = "grey88", linewidth = 0.3) +
    ggplot2::geom_vline(xintercept = 0, colour = "grey88", linewidth = 0.3) +
    ggplot2::geom_point(alpha = 0.75, size = 2.1) +
    ggplot2::geom_point(
      data = centroids,
      ggplot2::aes(x = x, y = y),
      shape = 8,
      size = 3.3,
      colour = "black",
      inherit.aes = FALSE
    ) +
    ggplot2::geom_label(
      data = labels,
      ggplot2::aes(x = x, y = y, label = label),
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 1,
      size = 3.2,
      colour = "grey15",
      fill = "#fffbe9",
      label.size = 0.15
    ) +
    ggplot2::scale_colour_manual(
      values = c(cell_1 = "#2B8C99", cell_2 = "#F4B400"),
      labels = c("Cell type 1", "Cell type 2"),
      name = "Population"
    ) +
    ggplot2::coord_equal() +
    ggplot2::facet_wrap(~panel, nrow = 1) +
    ggplot2::labs(
      x = "Expression of selected gene x",
      y = "Expression of selected gene y"
    ) +
    ggplot2::theme_bw(base_size = 11.5) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      strip.background = ggplot2::element_rect(
        fill = "#f8f8f8",
        colour = "grey75"
      ),
      strip.text = ggplot2::element_text(face = "bold")
    )
)
```

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a quantity to maximise [@alieeAutoGeneSAutomaticGene2021]. Cosine is preferred to Pearson correlation so that collinear but unequally scaled profiles remain penalised.

## Network skeleton: `generate_random_network_skeleton()`

A binary adjacency $\boldsymbol{A}\in\{0,1\}^{G\times G}$ is drawn from one of three undirected families via \pkg{igraph}: scale-free preferential attachment (`igraph::sample_pa()`), a stochastic block model (`igraph::sample_sbm()`), or Watts–Strogatz small-world (`igraph::sample_smallworld()`) [@R-igraph; @barabasiEmergenceScalingRandom1999]. Edges encode presence of partial-correlation links, not their strengths. i.i.d. signs with inhibitory fraction `prop_inhibitory` then form the weighted matrix $\boldsymbol{W}$ before the spectral SPD completion.

## Positive-definite precision: `build_normalised_precision()`

Signed edge weights enter through an affine spectral shift of $\boldsymbol{W}$,

$$
\boldsymbol{\Omega}
=
\boldsymbol{W}
+
\bigl(
  \lvert\lambda_{\min}(\boldsymbol{W})\rvert + u
\bigr)
\mathbf{I},
$$

with magnitude $v$ (`precision_scale`) baked into $\boldsymbol{W}$ and diagonal cushion $u$ (`precision_shift`). The shift preserves off-diagonal signs and support while guaranteeing $\boldsymbol{\Omega}\succ 0$.

## Shared covariances: `build_shared_covariance_array()`

Cell types share the same second-order structure,

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Downstream bulk simulation uses the Gaussian convolution $\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

# Perspectives: a single geometric score for $\boldsymbol{\mu}$

The present API exposes two AutoGeneS-style objectives. A natural refinement is to replace them by one scalar that jointly rewards large column norms (and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$

is the volume of the parallelepiped spanned by the columns of $\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when columns lengthen or become more orthogonal—precisely the desired MOO trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$

available in R via `kappa(mean_profiles, exact = TRUE)` (or `1 / kappa(...)`), measures how ill-posed linear recovery of $\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$ is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising $\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes", "volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the constructive $(\rho,s)$ generator for controllable scenarios, and report the chosen scalar alongside the existing pairwise diagnostics.

# References
