---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator `simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture simulation. Mean profiles follow an AutoGeneS-inspired construction [@alieeAutoGeneSAutomaticGene2021] in which a single dial $\rho$ (`target_cosine`) sets pairwise collinearity between cell-type columns of $\boldsymbol{\mu}$, while a fixed scale $s$ (`mean_scale`) sets column norms. Prefer varying $\rho$ across scenarios when precision weights already control second-order dependence. Covariances share a graph-constrained signed precision $\boldsymbol{\Omega}$, mapped to $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that can be passed to `simulate_bulk_mixture()` and then to any deconvolution routine exposed by `deconvolute_ratios()`.

```{r pipeline-mermaid, echo=FALSE, results='asis'}
mmd_path <- "figures/synthetic_grn_pipeline.mmd"
if (!file.exists(mmd_path)) {
  mmd_path <- file.path("vignettes", mmd_path)
}
mmd <- paste(readLines(mmd_path), collapse = "\n")
cat("```mermaid\n", mmd, "\n```\n", sep = "")
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  prop_inhibitory = 0.5,
  graph_model = "scale_free"
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is insufficient for stable deconvolution [@alieeAutoGeneSAutomaticGene2021] when centroids also collapse toward the origin. In the scenarios of this package we therefore **hold `mean_scale` fixed** and dial only `target_cosine`, so that Euclidean norms (and second-order precision weights) stay comparable across runs.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()`

Each column is a normalised blend of a shared unit direction
$\boldsymbol{u}$ and a private marker direction $\boldsymbol{v}_{j}$
(supported on a gene partition for type $j$):

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}
=
\sqrt{\rho}\,\boldsymbol{u}
+\sqrt{1-\rho}\,\boldsymbol{v}_{j},
\qquad
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{\tilde{\boldsymbol{\mu}}_{\cdot j}}{
  \|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2
},
\qquad
j=1,\ldots,J.
$$ {#eq-mean-cosine}

The private vectors are nearly orthogonal after normalisation, so

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
\approx
\rho
\qquad (j\neq k),
$$

and the pairwise cosines of $\boldsymbol{\mu}$ therefore track
$\rho\in[0,1]$ (`target_cosine`). The scale $s$ (`mean_scale`) sets column
norms without changing angles:
$\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2\propto s$ at
fixed $\rho$. When comparing collinearity alone (as in the nine Markov
scenarios), keep $s$ fixed and vary only $\rho$
[@alieeAutoGeneSAutomaticGene2021].

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a quantity to maximise [@alieeAutoGeneSAutomaticGene2021]. Cosine is preferred to Pearson correlation so that collinear but unequally scaled profiles remain penalised.

## Network skeleton: `generate_random_network_skeleton()`

A binary adjacency $\boldsymbol{A}\in\{0,1\}^{G\times G}$ is drawn from one of three undirected families: scale-free preferential attachment and cluster / SBM via `huge::huge.generator()` [@R-huge; @zhangSILGGMExtensivePackage2018], or an internal Watts–Strogatz small-world construction. Edges encode presence of partial-correlation links, not their strengths. i.i.d. signs with inhibitory fraction `prop_inhibitory` then form the weighted matrix $\boldsymbol{W}$ before the spectral SPD completion.

## Positive-definite precision: `build_normalised_precision()`

Signed edge weights enter through an affine spectral shift of $\boldsymbol{W}$,

$$
\boldsymbol{\Omega}
=
\boldsymbol{W}
+
\bigl(
  \lvert\lambda_{\min}(\boldsymbol{W})\rvert + u
\bigr)
\mathbf{I},
$$

with magnitude $v$ (`precision_scale`) baked into $\boldsymbol{W}$ and diagonal cushion $u$ (`precision_shift`). The shift preserves off-diagonal signs and support while guaranteeing $\boldsymbol{\Omega}\succ 0$.

## Shared covariances: `build_shared_covariance_array()`

Cell types share the same second-order structure,

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Downstream bulk simulation uses the Gaussian convolution $\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

# Perspectives: a single geometric score for $\boldsymbol{\mu}$

The present API exposes two AutoGeneS-style objectives. A natural refinement is to replace them by one scalar that jointly rewards large column norms (and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$

is the volume of the parallelepiped spanned by the columns of $\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when columns lengthen or become more orthogonal—precisely the desired MOO trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$

available in R via `kappa(mean_profiles, exact = TRUE)` (or `1 / kappa(...)`), measures how ill-posed linear recovery of $\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$ is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising $\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes", "volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the constructive $(\rho,s)$ generator for controllable scenarios, and report the chosen scalar alongside the existing pairwise diagnostics.

# References
