---
title: "Simulating semi-synthetic pseudo-bulk mixtures for benchmarking"
author: "Bastien Chassagnol"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Simulating semi-synthetic pseudo-bulk mixtures for benchmarking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography:
  - ../inst/packages.bib
  - ../inst/REFERENCES.bib
link-citations: true
format:
  html:
    embed-resources: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Overview

This vignette documents the synthetic first- and second-order moment generator `simulate_hierarchical_grn_moments()` and how its outputs feed bulk mixture simulation. Mean profiles follow an AutoGeneS-inspired construction [@alieeAutoGeneSAutomaticGene2021] in which a single dial $\rho$ (`target_cosine`) sets pairwise collinearity between cell-type columns of $\boldsymbol{\mu}$, while a fixed scale $s$ (`mean_scale`) sets column norms. Prefer varying $\rho$ across scenarios when precision weights already control second-order dependence.

Covariances share a graph-constrained signed precision $\boldsymbol{\Omega}$, mapped to $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$. An undirected Gaussian Markov simulation separates four layers (@eq-ggm-pipeline): graph $G$, signed weights $W(G)$, SPD precision $\Omega$, and latent Gaussians (an optional observation layer can add Poisson–log-normal or zero-inflated counts). The **graph** and the **precision** must not be conflated: a binary adjacency is almost never itself positive definite. DeCovarT draws $G$ via `generate_random_network_skeleton()`, completes $\Omega\succ 0$ with `build_normalised_precision()`, and stacks shared covariances with `build_shared_covariance_array()`. @sec-ggm-networks develops the topology, weight, and SPD design in detail.

# Pipeline API

The exported entry point orchestrates internal helpers and returns moments that can be passed to `simulate_bulk_mixture()` and then to any deconvolution routine exposed by `deconvolute_ratios()`.

```{mermaid}
%%| label: fig-pipeline-api
%%| fig-cap: "Pipeline from synthetic moment generation to deconvolution."
%%{init: {"theme": "sandstone"}}%%
flowchart LR
  A["Graph G"] --> B["Weights W(G)"]
  B --> C["Precision Ω ≻ 0"]
  C --> D["Σ = Ω^{-1}"]
  E["Means μ_j\n(AutoGeneS-style ρ)"] --> F["simulate_hierarchical_grn_moments()"]
  D --> F
  F --> G["simulate_bulk_mixture()"]
  G --> H["deconvolute_ratios()"]
```

```{r moments-example}
library(DeCovarT)

set.seed(42)
moments <- simulate_hierarchical_grn_moments(
  n_genes = 40L,
  n_celltypes = 3L,
  mean_scale = 10,
  target_cosine = 0.1,
  precision_shift = 0.1,
  precision_scale = 0.3,
  prop_inhibitory = 0.5,
  graph_model = "scale_free"
)

moments$objectives

bulk <- simulate_bulk_mixture(
  signature_matrix = moments$mean_profiles,
  Sigma = moments$covariance_matrices,
  p = c(0.5, 0.3, 0.2),
  n = 20
)
dim(bulk$Y)
dim(bulk$latent_profiles) # G x J x N latent draws; input mu is shared
```

Selecting genes (or, here, designing $\boldsymbol{\mu}$) by cosine alone is insufficient for stable deconvolution [@alieeAutoGeneSAutomaticGene2021]. In the scenarios of this package we therefore **hold `mean_scale` fixed** and dial only `target_cosine`, so first-order scale and second-order precision weights stay comparable across runs.

# Mathematical roles of the generators

## Mean signature: `generate_mean_signature_matrix()`

Notation: $G$ genes, $J$ cell types indexed by $j=1,\ldots,J$, and columns $\boldsymbol{\mu}_{\cdot j}\in\mathbb{R}^{G}$ of the mean signature. Fix the column scale $s$ (`mean_scale`, default $10$ in the nine-scenario grid) and dial only the target cosine $\rho\in[0,1]$ (`target_cosine`).

Let $\boldsymbol{u}=G^{-1/2}\mathbf{1}_{G}$ be the shared unit direction and let $\boldsymbol{v}_{j}$ be the $\ell_2$-normalised indicator of the gene block assigned to type $j$ (disjoint partition of $\{1,\ldots,G\}$). Then $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$ for $j\neq k$. Each column is the normalised blend

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}
=
\sqrt{\rho}\,\boldsymbol{u}
+\sqrt{1-\rho}\,\boldsymbol{v}_{j},
\qquad
\boldsymbol{\mu}_{\cdot j}
=
s\,
\frac{\tilde{\boldsymbol{\mu}}_{\cdot j}}{
  \|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2
},
\qquad
j=1,\ldots,J.
$$ {#eq-mean-cosine}

Expanding the un-normalised inner product for $j\neq k$ gives the exact identity

$$
\tilde{\boldsymbol{\mu}}_{\cdot j}^{\mathsf{T}}
\tilde{\boldsymbol{\mu}}_{\cdot k}
=
\rho
+
\sqrt{\rho(1-\rho)}\,
\bigl(
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}
  +
  \boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{k}
\bigr),
$$ {#eq-mean-inner}

because $\boldsymbol{u}^{\mathsf{T}}\boldsymbol{u}=1$ and $\boldsymbol{v}_{j}^{\mathsf{T}}\boldsymbol{v}_{k}=0$. The cross terms $\boldsymbol{u}^{\mathsf{T}}\boldsymbol{v}_{j}$ equal the (positive) mass of the shared direction on block $j$; they vanish in the idealised case where each $\boldsymbol{v}_{j}$ is taken orthogonal to $\boldsymbol{u}$, in which @eq-mean-inner collapses to $\rho$ and, since $\|\tilde{\boldsymbol{\mu}}_{\cdot j}\|_2=1$, the cosine of $\boldsymbol{\mu}$ equals $\rho$ exactly. With block indicators the cross terms are of order $J^{-1/2}$, so after the column renormalisation in @eq-mean-cosine the realised pairwise cosine tracks $\rho$ closely once each block contains many genes.

The scale $s$ sets column norms without changing angles: $\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2\propto s$ at fixed $\rho$. We keep $s=10$ across scenarios and only vary $\rho$.

### Visualising the cosine construction

@fig-cosine-geometry visualises the same three operating points ($\rho\in\{0,\,0.3,\,1\}$) for $G=2$ genes and $J=2$ cell types. The centroids come from `generate_mean_signature_matrix()`, with **cell type 1 held fixed** at its $\rho=1$ mean across panels and only cell type 2's mean moved to realise each target cosine. Independent Gaussian noise is added *only in this vignette chunk* around each centroid ($N=20$ i.i.d. draws per population) to emulate within-type dispersion under an independence assumption, without altering the mean-signature generator itself. Coloured arrows mark each type's mean direction; the shaded wedge is the angle between the two means.

This illustration is inspired by panel B in the AutoGeneS paper [@alieeAutoGeneSAutomaticGene2021], but focuses here on the cosine-angle control induced by @eq-mean-cosine (without modelling centroid-distance optimisation). With only two genes the cross terms in @eq-mean-inner are large, so the realised cosine is a warped but strictly increasing map of $\rho$.

```{r}
#| label: fig-cosine-geometry
#| fig-cap: >-
#|   Cosine-control toy for $J=2$ cell types and $G=2$ genes.
#|   Cell type 1's mean is fixed (from $\rho=1$); only cell type 2 moves.
#|   Coloured arrows: mean directions; shaded wedge: angle between means.
#|   Labels report realised cosine (unitless) and angle (radians).
#|   Points are $N=20$ i.i.d. Gaussian draws around each centroid (stars).
#|   Styling is inspired by panel B of `AutoGeneS`
#|   (@alieeAutoGeneSAutomaticGene2021).
#| fig-width: 9
#| fig-height: 4.6
#| echo: false
#| warning: false
#| message: false
#| eval: true
rhos_fig <- c(0, 0.3, 1)
set.seed(20260804)
n_genes_fig <- 2L
n_celltypes_fig <- 2L
n_samples <- 20L
s_fig <- 10
point_noise_sd <- 0.85
arc_n <- 80L
pop_cols <- c(cell_1 = "#2B8C99", cell_2 = "#F4B400")

mu_ref <- DeCovarT::generate_mean_signature_matrix(
  n_genes = n_genes_fig,
  n_celltypes = n_celltypes_fig,
  mean_scale = s_fig,
  target_cosine = 1,
  gene_names = c("gene_1", "gene_2"),
  celltype_names = c("cell_1", "cell_2")
)
mu_cell1_fixed <- mu_ref[, "cell_1"]

scatter_df <- list()
centroid_df <- list()
arrow_df <- list()
wedge_df <- list()
arc_df <- list()
label_df <- list()

for (k in seq_along(rhos_fig)) {
  rho <- rhos_fig[[k]]
  mu <- DeCovarT::generate_mean_signature_matrix(
    n_genes = n_genes_fig,
    n_celltypes = n_celltypes_fig,
    mean_scale = s_fig,
    target_cosine = rho,
    gene_names = c("gene_1", "gene_2"),
    celltype_names = c("cell_1", "cell_2")
  )
  # Keep cell_1 fixed across panels; only cell_2 carries the cosine change
  mu[, "cell_1"] <- mu_cell1_fixed

  cos_hat <- sum(mu[, 1] * mu[, 2]) /
    sqrt(sum(mu[, 1]^2) * sum(mu[, 2]^2))
  cos_hat <- max(-1, min(1, cos_hat))
  angle_rad <- acos(cos_hat)
  panel <- sprintf(
    "rho = %s",
    format(rho, nsmall = 1)
  )

  ang1 <- atan2(mu[2L, 1L], mu[1L, 1L])
  ang2 <- atan2(mu[2L, 2L], mu[1L, 2L])
  # Shortest signed turn from ang1 to ang2 in (-pi, pi]
  d_ang <- (ang2 - ang1 + pi) %% (2 * pi) - pi
  ang_seq <- ang1 + seq(0, d_ang, length.out = arc_n)
  r_arc <- 0.42 * min(sqrt(colSums(mu^2)))

  for (j in seq_len(n_celltypes_fig)) {
    scatter_df[[length(scatter_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = stats::rnorm(n_samples, mean = mu[1L, j], sd = point_noise_sd),
      y = stats::rnorm(n_samples, mean = mu[2L, j], sd = point_noise_sd)
    )
    centroid_df[[length(centroid_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = mu[1L, j],
      y = mu[2L, j]
    )
    arrow_df[[length(arrow_df) + 1L]] <- data.frame(
      panel = panel,
      celltype = colnames(mu)[[j]],
      x = 0,
      y = 0,
      xend = mu[1L, j],
      yend = mu[2L, j]
    )
  }

  wedge_df[[length(wedge_df) + 1L]] <- data.frame(
    panel = panel,
    x = c(0, r_arc * cos(ang_seq), 0),
    y = c(0, r_arc * sin(ang_seq), 0)
  )
  arc_df[[length(arc_df) + 1L]] <- data.frame(
    panel = panel,
    x = r_arc * cos(ang_seq),
    y = r_arc * sin(ang_seq)
  )

  label_df[[length(label_df) + 1L]] <- data.frame(
    panel = panel,
    label = sprintf("cos = %.2f\nangle = %.2f rad", cos_hat, angle_rad)
  )
}

points_df <- do.call(rbind, scatter_df)
centroids <- do.call(rbind, centroid_df)
arrows <- do.call(rbind, arrow_df)
wedges <- do.call(rbind, wedge_df)
arcs <- do.call(rbind, arc_df)
labels <- do.call(rbind, label_df)
panel_levels <- sprintf("rho = %s", format(rhos_fig, nsmall = 1))
for (nm in c(
  "points_df", "centroids", "arrows", "wedges", "arcs", "labels"
)) {
  df <- get(nm)
  df$panel <- factor(df$panel, levels = panel_levels)
  assign(nm, df)
}

ggplot2::ggplot() +
  ggplot2::geom_hline(yintercept = 0, colour = "grey88", linewidth = 0.3) +
  ggplot2::geom_vline(xintercept = 0, colour = "grey88", linewidth = 0.3) +
  ggplot2::geom_polygon(
    data = wedges,
    ggplot2::aes(x = x, y = y, group = panel),
    fill = "grey35",
    alpha = 0.12,
    colour = NA
  ) +
  ggplot2::geom_path(
    data = arcs,
    ggplot2::aes(x = x, y = y, group = panel),
    colour = "grey25",
    linewidth = 0.55,
    lineend = "round"
  ) +
  ggplot2::geom_segment(
    data = arrows,
    ggplot2::aes(
      x = x,
      y = y,
      xend = xend,
      yend = yend,
      colour = celltype
    ),
    linewidth = 0.95,
    arrow = grid::arrow(
      length = grid::unit(0.22, "cm"),
      type = "closed"
    )
  ) +
  ggplot2::geom_point(
    data = points_df,
    ggplot2::aes(x = x, y = y, fill = celltype),
    shape = 21,
    colour = "grey20",
    stroke = 0.35,
    alpha = 0.85,
    size = 2.15
  ) +
  ggplot2::geom_point(
    data = centroids,
    ggplot2::aes(x = x, y = y),
    shape = 8,
    size = 3.3,
    colour = "black",
    inherit.aes = FALSE
  ) +
  ggplot2::geom_label(
    data = labels,
    ggplot2::aes(x = Inf, y = Inf, label = label),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.2,
    size = 3.0,
    colour = "grey15",
    fill = "#fffbe9",
    label.size = 0.15
  ) +
  ggplot2::scale_colour_manual(
    values = pop_cols,
    labels = c("Cell type 1", "Cell type 2"),
    name = "Population"
  ) +
  ggplot2::scale_fill_manual(
    values = pop_cols,
    labels = c("Cell type 1", "Cell type 2"),
    name = "Population"
  ) +
  ggplot2::scale_x_continuous(
    expand = ggplot2::expansion(mult = c(0.08, 0.28))
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = c(0.08, 0.28))
  ) +
  ggplot2::coord_equal() +
  ggplot2::facet_wrap(~panel, nrow = 1) +
  ggplot2::labs(
    x = "Expression of gene 1",
    y = "Expression of gene 2"
  ) +
  ggplot2::theme_bw(base_size = 11.5) +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.minor = ggplot2::element_blank(),
    strip.background = ggplot2::element_rect(
      fill = "#f8f8f8",
      colour = "grey75"
    ),
    strip.text = ggplot2::element_text(face = "bold")
  )
```

## Objectives: `compute_mean_profile_objectives()`

For columns of $\boldsymbol{\mu}$,

$$
\overline{\lvert\cos\rvert}
=
\frac{2}{J(J-1)}
\sum_{1\le j<k\le J}
\Biggl|
  \frac{
    \boldsymbol{\mu}_{\cdot j}^{\mathsf{T}}
    \boldsymbol{\mu}_{\cdot k}
  }{
    \|\boldsymbol{\mu}_{\cdot j}\|_2\,
    \|\boldsymbol{\mu}_{\cdot k}\|_2
  }
\Biggr|,
$$

$$
D_{\mathrm{Euc}}
=
\sum_{1\le j<k\le J}
\|\boldsymbol{\mu}_{\cdot j}-\boldsymbol{\mu}_{\cdot k}\|_2.
$$

AutoGeneS treats the first as a quantity to minimise and the second as a quantity to maximise [@alieeAutoGeneSAutomaticGene2021]. Cosine is preferred to Pearson correlation so that collinear but unequally scaled profiles remain penalised.

Second-order structure—network skeleton, signed weights, SPD precision, and shared covariances—is developed in @sec-ggm-networks rather than duplicated here.

# Undirected Gaussian Markov network generation {#sec-ggm-networks}

$$
G \longrightarrow W(G) \longrightarrow \Omega \succ 0
\longrightarrow X_i \sim \mathcal{N}_p(\mu_i,\Omega^{-1})
$$ {#eq-ggm-pipeline}

Here $G$ is an **undirected** graph, $W(G)$ a symmetric weighted matrix with the same off-diagonal support, $\Omega$ a strictly positive-definite precision, and $X_i$ an expression profile. This section covers **undirected** Gaussian Markov / precision-graph simulation only. `generate_random_network_skeleton()` draws $G$; `build_normalised_precision()` completes it to $\Omega\succ 0$ by an affine spectral shift; `build_shared_covariance_array()` maps $\boldsymbol{\Sigma}_j=\boldsymbol{\Omega}^{-1}$ for each cell type.

```{mermaid}
%%| label: fig-ggm-pipeline
%%| fig-cap: "Undirected simulation pipeline from topology to observations (optional observation layer dashed)."
%%{init: {"theme": "sandstone"}}%%
flowchart LR
  G["Undirected graph G"] --> W["Signed weights W(G)"]
  W --> Omega["SPD precision Ω"]
  Omega --> Latent["Latent Gaussian X"]
  Mu["Mean design μ"] --> Latent
  Latent --> Obs["Observation layer"]
  Latent -.-> Obs
```

## Graph structure generators

### Evidence from the literature

@tbl-studies summarises how recent simulation studies combine topology, precision construction, and mean / observation models.

| Source | Graph structures | Precision construction | Mean / observation | Purpose |
|---------------|---------------|---------------|---------------|---------------|
| PLNnetwork [@chiquetVariationalInferenceSparse2018] | Erdős–Rényi, preferential attachment, affiliation | $\Omega = vG + \operatorname{diag}(\lvert\lambda_{\min}(vG)\rvert + u)$ | Latent log-abundance $XB$; compositional / multinomial counts | Count-network estimators under compositionality |
| SILGGM [@zhangSILGGMExtensivePackage2018] | Band, hub, Erdős–Rényi, scale-free via `huge` | Package covariance / precision simulation | Centred Gaussian; focus on precision entries | Large-$p$ inferential and computational stress tests |
| BLGGM [@wuEstimatingHeterogeneousGene2022] | Block mixtures of dense, circle, star, signed modules | Module matrices with structured signed off-diagonals | Cell-type-specific $\mu_k$, $\Omega_k$; expression-dependent dropout | Joint clustering, heterogeneous networks, zero inflation |

: Literature survey of undirected graph → precision → observation designs.

Several conclusions follow.

1.  **Band** is not intrinsically a random-graph model: its support is usually deterministic once the bandwidth is fixed (edge weights and signs may still be random). Hub constructions are often partly deterministic as well. Erdős–Rényi, preferential attachment, and stochastic block models are genuinely stochastic.
2.  A positive off-diagonal entry in $\Omega$ encodes a **negative** partial correlation (@eq-partial-cor). Uniformly positive precision weights therefore induce uniformly negative edgewise partial correlations; random or signed biological weights are needed when activation-like and inhibition-like associations matter.
3.  These models target **undirected** conditional dependence. Federico and colleagues prefer a Markov-network representation when observational data do not justify edge directionality [@federicoStructureLearningGene2023].
4.  **BLGGM as a hybrid topology design.** Wu and Luo’s simulation can be read as two nested layers [@wuEstimatingHeterogeneousGene2022]. At the **global** scale, genes are partitioned into blocks (a stochastic-block–like partition). Within each block, a **finer** local topology is specified—dense clique-like modules, circles, stars / hubs, or signed dense substructures—so that different blocks can stand for distinct regulatory regimes (e.g. dense co-expression programmes, cyclic signalling motifs, master-regulator hubs). Signs can also differ across modules or cell types. This is richer than a flat ER draw, yet still fully undirected.

$$
\rho_{jk\mid -\{j,k\}}
=
-\frac{\Omega_{jk}}{\sqrt{\Omega_{jj}\Omega_{kk}}}
$$ {#eq-partial-cor}

@tbl-topologies compares the main **undirected** topology families used in GGM benchmarks, with R entry points. @fig-topologies sketches the six families most often used in random-like network structure generation.

| Structure | Definition and controls | R entry points | Advantages | Limitations |
|---------------|---------------|---------------|---------------|---------------|
| **Erdős–Rényi** | Each edge present independently with probability $q$; $q=d_{\mathrm{avg}}/(p-1)$ targets average degree | `huge::huge.generator(graph="random")` [@R-huge]; `BDgraph::graph.sim(graph="random")` [@R-BDgraph]; DeCovarT `graph_model = "erdos_renyi"` | Exact sparsity control in expectation; few nuisance parameters | Narrow degree distribution; little modular organisation |
| **Band / AR** | Edge $j$–$k$ if $1\le\lvert j-k\rvert\le b$ | `huge.generator(graph="band")` [@R-huge]; `BDgraph::bdgraph.sim(graph="AR(1)")` / `"AR(2)"` [@R-BDgraph] | Local dependence and controlled max degree | Artificial gene ordering; no hubs or modules |
| **Hub / star** | $g$ groups; one centre linked to remaining members | `huge.generator(graph="hub")` [@R-huge]; `graph.sim(graph="hub")` / `"star"` [@R-BDgraph]; DeCovarT `graph_model = "hub"` | Stress-tests high-degree regulators | Pure stars omit cross-talk |
| **Scale-free** | Preferential attachment (e.g. Barabási–Albert with $m$ edges per new node) | `huge.generator(graph="scale-free")` [@R-huge]; DeCovarT `graph_model = "scale_free"` | Heterogeneous degrees | Weak modularity under plain BA growth |
| **Cluster / SBM** | Within-block edge probability exceeds between-block | `huge.generator(graph="cluster")` [@R-huge]; DeCovarT `graph_model = "stochastic_block_model"` | Pathway / module structure | Equal blocks can be unrealistically regular |
| **Small-world / lattice / circle** | High clustering with short paths, or fixed neighbourhoods | DeCovarT `graph_model = "small_world"` (Watts–Strogatz); `BDgraph::graph.sim(graph="smallworld")` [@R-BDgraph] | Local modules and spatial organisation | Narrow degrees; needs a defensible ordering |

: Topology families for undirected GGM simulation (R-focused).

## Biological relevance of the six families

Not every topology is equally useful as a **gene-regulatory** prior. The three simulation studies in @tbl-studies, together with classical network biology, suggest the following reading.

| Family | Biological reading | Why it matters for DeCovarT |
|----------------|-----------------------|----------------------------------|
| **Star / hub** | Master-regulator TF with many targets; one-to-many signalling | Stress-tests recovery of a few high-degree hubs that dominate partial correlations [@wuEstimatingHeterogeneousGene2022; @federicoStructureLearningGene2023] |
| **Scale-free (BA)** | Preferential attachment produces a heavy-tailed degree distribution and several hubs [@barabasiEmergenceScalingRandom1999] | Standard high-dimensional GGM stress test used by SILGGM via `huge` [@zhangSILGGMExtensivePackage2018]; useful for testing degree heterogeneity, but not a default empirical model of gene regulation |
| **Cluster / SBM** | Co-regulated modules / pathways with dense within-block links | Matches pathway organisation; BLGGM’s global block partition is SBM-like before finer within-block motifs [@wuEstimatingHeterogeneousGene2022] |
| **Small-world** | High local clustering with short paths (ring + shortcuts) | Local co-expression neighbourhoods plus long-range regulatory shortcuts; useful when modularity is soft rather than hard blocks [@federicoStructureLearningGene2023] |
| **Band / AR** | Ordered local dependence only | Useful numerical control, but gene indices rarely encode a natural order—limited biological fidelity [@zhangSILGGMExtensivePackage2018] |
| **Erdős–Rényi** | Homogeneous random wiring | Null / baseline sparsity; little pathway or hub structure [@zhangSILGGMExtensivePackage2018] |

: Biological interpretability of undirected topology families.

### The scale-free hypothesis: a useful stress test, not a biological default

The phrase *scale-free network* usually means that the degree distribution follows a power law, $\Pr(K=k)\propto k^{-\alpha}$, at least above a lower cut-off. This statistical statement is distinct from the Barabási–Albert (BA) growth mechanism. Preferential attachment can generate a power law, but observing a heavy-tailed degree distribution does not establish preferential attachment; other mechanisms can produce similar tails, and variants of preferential attachment need not produce a pure power law [@lima-mendezPowerfulLawPower2009; @broidoScalefreeNetworksAre2019].

Whether biological networks are scale-free has therefore remained contested. Early claims often relied on approximately straight log–log degree plots. Such plots are not goodness-of-fit tests: binning, the plotting scale, a small number of hubs, and incomplete or biased sampling can all make non-power-law distributions appear linear. The underlying graph representation matters as well. For example, pool metabolites can become artificial hubs in metabolic graphs, while protein-interaction networks depend strongly on the assay and sampling scheme [@lima-mendezPowerfulLawPower2009]. A power law should instead be fitted to the discrete degree data, tested for plausibility, and compared on the same support with alternatives such as log-normal, exponential, stretched-exponential, and power-law-with-cut-off distributions.

The evidence is not uniformly negative. Using adjacency-spectrum distributions rather than degree alone, Takahashi and colleagues classified eight protein–protein interaction networks as scale-free [@takahashiDiscriminatingDifferentClasses2012]. Importantly, this was a relative model-selection result: BA-like networks fitted better than the two candidate alternatives, Erdős–Rényi and Watts–Strogatz networks. It did not show that a power law was an adequate absolute model, that preferential attachment generated the observed networks, or that unobserved interactions would preserve the classification. The authors explicitly noted both the restricted candidate set and possible sampling artefacts.

A broader test of 928 real-world network data sets reached a more cautious conclusion [@broidoScalefreeNetworksAre2019]. Across all domains, only 4% met the strongest scale-free criterion, whereas a log-normal distribution fitted as well as or better than a power law for 88% of degree distributions. Among the biological networks, 63% showed neither direct nor indirect evidence of scale-free structure, although 6% met the strongest criterion, principally metabolic networks. These proportions depend on the network corpus and the operational definition of scale-freeness, but they reject a universal scale-free law rather than excluding scale-free structure from every biological system.

For DeCovarT, a BA graph should consequently be interpreted as a deliberately demanding **hub-heterogeneity benchmark**. It tests whether covariance estimation and deconvolution recover a few high-degree vertices without claiming that a gene-regulatory network grew by preferential attachment. Results should be compared with density-matched SBM, small-world, star, and Erdős–Rényi scenarios. Conclusions that depend only on the BA scenario should be labelled topology-specific; hub recovery, robustness to vertex removal, and biological realism do not follow from a fitted heavy tail alone.

**BLGGM as a hybrid.** Wu and Luo combine an SBM-like **global** gene partition with **local** dense, circle, star, or signed modules inside blocks [@wuEstimatingHeterogeneousGene2022]. Distinct local motifs can stand for distinct processes (e.g. master-regulator stars vs dense co-expression programmes). DeCovarT’s exported generators currently expose the three families most useful as standalone benchmarks: `scale_free`, `stochastic_block_model`, and `small_world`.

## Package roles (R)

`generate_random_network_skeleton()` draws a binary adjacency $\boldsymbol{A}\in\{0,1\}^{G\times G}$ from one of three undirected families via \pkg{igraph}: scale-free preferential attachment (`igraph::sample_pa()`), a stochastic block model (`igraph::sample_sbm()`), or Watts–Strogatz small-world (`igraph::sample_smallworld()`) [@R-igraph; @barabasiEmergenceScalingRandom1999]. Edges encode presence of partial-correlation links, not their strengths.

- **`igraph`** [@R-igraph] — DeCovarT draws the three undirected skeletons used in simulation (`scale_free` via `sample_pa()`, `stochastic_block_model` via `sample_sbm()`, `small_world` via `sample_smallworld()`).
- **`huge`** [@R-huge] — compact end-to-end GGM benchmark via `huge.generator()`: topology, SPD precision, covariance, and $\mathcal{N}_p$ samples. Supports random, hub, cluster, band, and scale-free graphs [@zhangSILGGMExtensivePackage2018]; useful as a literature comparator, not the DeCovarT skeleton backend.
- **`BDgraph`** [@R-BDgraph] — `graph.sim()` / `bdgraph.sim()` cover overlapping families plus small-world and scale-free; `rgwish()` samples G-Wishart precisions on a fixed graph.
- **`bnlearn`** [@R-bnlearn] — directed / DAG and some undirected structure learning; helpful when the scientific story is causal or when comparing directed versus undirected ground truths.
- Prefer keeping the **same** topology → weight → SPD → mean pipeline when swapping generators, so differences reflect the graph family rather than an incidental change of SPD or sampling code.

![Six undirected topology families used in GGM simulation: Erdős–Rényi, Barabási–Albert (preferential attachment / scale-free), stochastic block (cluster), band / AR, star / hub, and small-world (Watts–Strogatz).\
\
](figures/fig_network_topologies_six.png){#fig-topologies}

## Weight design and random signs

Topology generators return a binary undirected support $E$. **Signs and magnitudes** are a separate design layer $W(G)$ in @eq-ggm-pipeline: they are assigned *after* (or jointly with) the skeleton, then completed to an SPD precision. Remember that

$$
\operatorname{sign}(\rho_{jk\mid -\{j,k\}})
=
-\operatorname{sign}(\Omega_{jk})
$$ {#eq-sign-flip}

see @eq-partial-cor, so a positive precision weight is an inhibitory partial correlation. @tbl-weights contrasts three standard undirected strategies.

| Approach | What is randomised | How $\Omega$ is obtained | When to use |
|------------------|------------------|------------------|------------------|
| **i.i.d. edge signs** | For each $\{j,k\}\in E$, draw $s_{jk}\in\{-1,+1\}$ (e.g. fair coin or $\mathbb{P}(+)=\pi$) and magnitude $m_{jk}>0$; set $W_{jk}=W_{kj}=s_{jk}m_{jk}$ | Support-preserving SPD map of $W$ (spectral shift, diagonal dominance, …) | Simple signed ER / hub / SBM benchmarks; explicit control of the inhibitory fraction |
| **Partial-correlation scaling** | Fill a symmetric matrix $R$ with $R_{jk}=0$ off $E$ and $R_{jk}\in(-1,1)$ (signed) on $E$; ensure $\rho(R)<1$ | $\Omega = D^{1/2}(I-R)D^{1/2}$ (@eq-partial-scale) | Direct control of partial-correlation signs and strengths |
| **G-Wishart** (`BDgraph::rgwish()`) | Continuous random $\Omega$ on the fixed graph zeros | $\Omega\sim W_G(b,D)$ already SPD [@R-BDgraph; @BDgraph2019] | Heterogeneous random weights without hand-tuned magnitudes; Bayesian-style replicates |

: Three undirected strategies for signed / random edge weights after a fixed skeleton.

### i.i.d. signs on a fixed skeleton

Draw $G$ (ER, band, hub, …). Independently for each undirected edge,

$$
W_{jk}=W_{kj}=s_{jk}\,m_{jk},
\qquad
s_{jk}\in\{-1,+1\},
\quad m_{jk}>0,
$$ {#eq-iid-signs}

then apply a support-preserving SPD completion (next section). Uniform positive loadings $W=vG$ are the special case $s_{jk}\equiv +1$ used by PLN-style and many `huge` simulations [@chiquetVariationalInferenceSparse2018]—all partial correlations then share the same sign. DeCovarT uses i.i.d. signs with inhibitory fraction `prop_inhibitory` and magnitude $v$ (`precision_scale`) baked into $\boldsymbol{W}$ before the spectral SPD completion.

### Partial-correlation scaling and why $I-R$

In a Gaussian Markov model the **matrix of partial correlations** (with ones on the diagonal) relates to the precision by a diagonal congruence. Write the off-diagonal partial correlations as a symmetric matrix $R$ with

$$
R_{jj}=0,
\qquad
R_{jk}
=
\begin{cases}
\rho_{jk\mid -\{j,k\}} & \{j,k\}\in E,\\
0 & \{j,k\}\notin E,
\end{cases}
$$ {#eq-R-support}

and choose magnitudes so that the spectral radius satisfies $\rho(R)<1$. Then

$$
\Omega
=
D^{1/2}(I-R)D^{1/2}
$$ {#eq-partial-scale}

for a positive diagonal $D$ (often $D=I$ after standardising margins).

**Why subtract from the identity?** The precision encodes *conditional* dependence. For a correlation-type matrix, $I-R$ places **ones on the diagonal** (self-precision after scaling) and **negatives of the partial correlations** off the diagonal, matching @eq-sign-flip: if $R_{jk}>0$ (positively partially correlated variables), then $(I-R)_{jk}<0$, hence $\Omega_{jk}<0$.

### G-Wishart draws on a fixed graph

Given adjacency $G$, `BDgraph::rgwish()` samples $\Omega\sim W_G(b,D)$ already positive definite and with $\Omega_{jk}=0$ whenever $(j,k)\notin E$ [@R-BDgraph; @BDgraph2019]. Signs and magnitudes arise from the continuous distribution rather than from an explicit coin-flip layer. This is the most “automatic” signed-weight generator once the skeleton is fixed; the trade-off is less direct control of the inhibitory fraction than @eq-iid-signs or @eq-partial-scale.

## Positive-definiteness strategies

@tbl-pd contrasts constructions that turn a weighted support $W$ (or a partial-correlation design) into $\Omega\succ 0$. DeCovarT’s `build_normalised_precision()` implements the uniform spectral shift below, with diagonal cushion $u$ (`precision_shift`):

$$
\boldsymbol{\Omega}
=
\boldsymbol{W}
+
\bigl(
  \lvert\lambda_{\min}(\boldsymbol{W})\rvert + u
\bigr)
\mathbf{I}.
$$

| Construction | Formula | SPD? | Preserves exact support? | Behaviour |
|---------------|---------------|--------------:|--------------:|---------------|
| Uniform spectral shift | $\Omega=W+\max(0,\varepsilon-\lambda_{\min}(W))I$ | Yes | Yes | Same shift on every eigenvalue; off-diagonals (hence signs) unchanged |
| `huge` / PLN-style loading | Diagonal load from $\lvert\lambda_{\min}\rvert$, baseline, and $u$ | Yes | Yes | $v$ sets edge magnitude; loading sets conditioning [@chiquetVariationalInferenceSparse2018] |
| Partial-correlation scaling | $\Omega=D^{1/2}(I-R)D^{1/2}$ with $\rho(R)<1$ | Yes | Yes | Signs and strengths set on the partial-correlation scale (@eq-partial-scale) |
| Graph Laplacian / M-matrix | $\Omega=L_G+\varepsilon I$ | Yes | Yes | Off-diagonals non-positive ⇒ all edgewise partial correlations positive |
| G-Wishart | $\Omega\sim W_G(b,D)$ | Yes | Yes | Heterogeneous random weights; `BDgraph::rgwish()` [@R-BDgraph] |
| Eigenvalue clipping | Clip $\Lambda$ then reconstruct $Q\Lambda Q^{\top}$ | Yes | **No** | Fills structural zeros |
| Nearest PD projection | e.g. `Matrix::nearPD()` [@R-Matrix] | Yes | **No** (in general) | Repair tool, not a support-preserving truth |

: Positive-definiteness constructions for graph-constrained precisions.

### Why the uniform spectral shift is attractive

If $W=Q\Lambda Q^{\top}$, then (@eq-spectral-shift)

$$
W+cI = Q(\Lambda+cI)Q^{\top}.
$$ {#eq-spectral-shift}

Every eigenvalue increases by $c$, while every off-diagonal of $W$ is unchanged, so the edge set

$$
E=\{(j,k):j<k,\ \Omega_{jk}\neq 0\}
$$ {#eq-edge-set}

is identical before and after loading—and **signs of** $W$ are preserved.

The floor $\varepsilon$ also controls the condition number

$$
\kappa(\Omega)=\frac{\lambda_{\max}(\Omega)}{\lambda_{\min}(\Omega)}.
$$ {#eq-omega-kappa}

Large loading yields a well-conditioned but weakly dependent network; loading only slightly above $-\lambda_{\min}$ strengthens dependence but can make estimation fragile. Benchmarks should report partial-correlation distributions and $\kappa(\Omega)$, not only the adjacency.

## Shared covariances: `build_shared_covariance_array()`

Cell types share the same second-order structure,

$$
\boldsymbol{\Sigma}_j
=
\boldsymbol{\Omega}^{-1},
\qquad
j=1,\ldots,J,
$$

stacked as an array in $\mathcal{M}_{G\times G\times J}$. Downstream bulk simulation uses the Gaussian convolution $\boldsymbol{y}\mid(\boldsymbol{\zeta},\boldsymbol{p})\sim
\mathcal{N}_{G}(\boldsymbol{\mu}\boldsymbol{p},
\sum_j p_j^{2}\boldsymbol{\Sigma}_j)$.

## Conclusions

> DeCovarT draws `scale_free`, `stochastic_block_model`, and `small_world` skeletons with \pkg{igraph}, then i.i.d. signed weights (`prop_inhibitory`) and a spectral shift to guarantee $\Omega\succ 0$.

### Further high-dimensional designs

Beyond the families in @tbl-topologies:

- **Degree-prescribed / configuration** graphs separate degree heterogeneity from preferential-attachment *mechanisms*.
- **Spatial / geometric** graphs suit spatial transcriptomics or chromatin neighbourhoods.
- **Differential networks:**
  - Start from a shared base $G_0$
  - Control the differential shift $\Omega_k=\Omega_0+\Delta_k$
  - Ensure the SPD step
  - Alternatively, vary the *support, weight, and sign* separately [@federicoStructureLearningGene2023; @wuEstimatingHeterogeneousGene2022].
- **Other distributions,** using the GGM layer as a parameter:
  - **Latent-variable GGMs** — sparse $\Omega_S$ plus low-rank confounding $Bf_i$.
  - **Nonparanormal / Gaussian-copula** margins on a latent Gaussian graph.
  - **Poisson–log-normal / compositional** observation layers on latent log-abundances [@chiquetVariationalInferenceSparse2018].
  - **Zero-inflated mixture GGMs** with cell-type-specific $(\mu_k,\Omega_k)$ and expression-dependent dropout [@wuEstimatingHeterogeneousGene2022].

### Recommended benchmark specification

Use one pipeline for every topology (@eq-benchmark-pipe, @fig-ggm-pipeline):

$$
\text{topology}
\rightarrow
\text{signed weights}
\rightarrow
\text{SPD precision}
\rightarrow
\text{mean design}
\rightarrow
\text{latent Gaussian}
\rightarrow
\text{observation model}
$$ {#eq-benchmark-pipe}

| Axis | Suggested levels |
|-----------------------|------------------------------------------------|
| Dimension, namely number of variables (here, genes) | $p \in \{100,500,1000\}$ |
| Topology | ER, band, hub, scale-free, SBM, small-world |
| Expected degree | $\approx 2,4,8$ (keep graphs sparse) |
| Edge weights | Constant; i.i.d. signed; partial-correlation scaling; G-Wishart |
| Conditioning | Report $\lambda_{\min}$, $\lambda_{\max}$, $\kappa(\Omega)$ |
| Means | Zero; sparse group shifts, driven by cell-type; joint $\mu$–$\Omega$ heterogeneity, for example playing on the *cosine similarity*, or the *condition number*, of the mean profile |

: Compact factorial axes for a reproducible undirected GGM simulation study.

**Main insights.** Prefer the uniform spectral shift (@eq-spectral-shift) or partial-correlation scaling (@eq-partial-scale) when support and signs must be exact; add the mean layer **independently** of graph generation.

## Nine factorial simulation scenarios

The script `scripts/generate_random_network_skeleton.R` builds a $3\times 3$ grid of undirected Markov-network moments. Notation matches the manuscript: $G$ genes and $N$ bulk samples for $\boldsymbol{Y}\in\mathcal{M}_{G\times N}$.

$$
\begin{aligned}
\text{topology}
&\in
\{\text{scale-free (A)},\
\text{SBM (B)},\
\text{small-world (C)}\},
\\
\rho
&\in
\{0.05,\,0.35,\,0.70\}
\quad\text{(cosine levels 1–3)},
\\
\text{prop. inhibitory edges}
&=
\tfrac12,
\qquad
\text{target }\kappa(\boldsymbol{\Omega})
\approx
50.
\end{aligned}
$$ {#eq-scenario-grid}

Scenario IDs are $\mathtt{A1}$–$\mathtt{C3}$ (topology letter × cosine level). @tbl-scenarios reports realised diagnostics for each run (seeded).

```{r}
#| label: tbl-scenarios
#| tbl-cap: "Nine undirected Markov-network scenarios (3 topologies × 3 cosine levels)."
#| warning: false
#| message: false
#| eval: true
#| echo: false

if (requireNamespace("DeCovarT", quietly = TRUE)) {
  library(DeCovarT)
} else {
  stop("Package 'DeCovarT' is not installed.")
}

G <- 60L
N <- 80L
n_celltypes <- 3L
mean_scale <- 10
precision_scale <- 0.3
prop_inhibitory <- 0.5
target_kappa <- 50
cosine_levels <- c(0.05, 0.35, 0.70)

topology_specs <- list(
  A = list(
    topology = "scale_free",
    graph_model = "scale_free",
    graph_params = list()
  ),
  B = list(
    topology = "stochastic_block_model",
    graph_model = "stochastic_block_model",
    graph_params = list(block_prob = c(0.5, 0.25, 0.25))
  ),
  C = list(
    topology = "small_world",
    graph_model = "small_world",
    graph_params = list(nei = 2L, p = 0.05)
  )
)

calibrate_u <- function(adjacency) {
  shifts <- c(0.05, 0.1, 0.2, 0.5, 1, 2, 5)
  best_u <- shifts[[1]]
  best_gap <- Inf
  for (u in shifts) {
    W <- DeCovarT:::assign_iid_signed_weights(
      adjacency,
      prop_inhibitory = prop_inhibitory,
      weight_magnitude = precision_scale
    )
    Omega <- DeCovarT:::build_normalised_precision(W, u)
    gap <- abs(log(kappa(Omega, exact = TRUE)) - log(target_kappa))
    if (gap < best_gap) {
      best_gap <- gap
      best_u <- u
    }
  }
  best_u
}

rows <- list()
seed0 <- 100L
for (t_idx in seq_along(topology_specs)) {
  t_code <- names(topology_specs)[[t_idx]]
  spec <- topology_specs[[t_code]]
  for (c_idx in seq_along(cosine_levels)) {
    scenario_id <- paste0(t_code, c_idx)
    set.seed(seed0 + 10L * t_idx + c_idx)
    A <- DeCovarT:::generate_random_network_skeleton(
      n_genes = G,
      graph_model = spec$graph_model,
      graph_params = spec$graph_params
    )
    u <- calibrate_u(A)
    set.seed(seed0 + 10L * t_idx + c_idx)
    moments <- simulate_hierarchical_grn_moments(
      n_genes = G,
      n_celltypes = n_celltypes,
      mean_scale = mean_scale,
      target_cosine = cosine_levels[[c_idx]],
      precision_shift = u,
      precision_scale = precision_scale,
      prop_inhibitory = prop_inhibitory,
      graph_model = spec$graph_model,
      graph_params = spec$graph_params
    )
    Omega <- moments$graph_structure$normalised_precision
    W <- moments$graph_structure$weighted_adjacency
    mu <- moments$mean_profiles
    evals <- eigen(Omega, symmetric = TRUE, only.values = TRUE)$values
    upper_w <- W[upper.tri(W) & A == 1]
    rows[[scenario_id]] <- data.frame(
      ID = scenario_id,
      topology = spec$topology,
      rho = cosine_levels[[c_idx]],
      lambda_min = min(evals),
      lambda_max = max(evals),
      kappa_Omega = max(evals) / min(evals),
      prop_inhib = if (length(upper_w)) mean(upper_w > 0) else NA_real_,
      mean_abs_cos = moments$objectives$mean_abs_cosine,
      recip_kappa_mu = 1 / kappa(mu, exact = TRUE),
      gram_det = sqrt(max(0, det(crossprod(mu)))),
      stringsAsFactors = FALSE
    )
  }
}

scenario_df <- do.call(rbind, rows)
rownames(scenario_df) <- NULL

if (requireNamespace("tinytable", quietly = TRUE)) {
  scenario_df |>
    tinytable::tt(digits = 3) |>
    tinytable::group_tt(
      i = c(
        rep("A - scale-free (Barabasi-Albert)", 3L),
        rep("B - stochastic block model", 3L),
        rep("C - small-world (Watts-Strogatz)", 3L)
      )
    ) |>
    tinytable::format_tt(
      j = c(
        "rho",
        "lambda_min",
        "lambda_max",
        "kappa_Omega",
        "prop_inhib",
        "mean_abs_cos",
        "recip_kappa_mu",
        "gram_det"
      ),
      digits = 3
    )
} else {
  knitr::kable(scenario_df, digits = 3)
}
```

Column guide: $\lambda_{\min}$, $\lambda_{\max}$, and $\kappa(\boldsymbol{\Omega})=\lambda_{\max}/\lambda_{\min}$ summarise the precision spectrum; `prop_inhib` is the realised fraction of inhibitory precision edges; `mean_abs_cos` is the AutoGeneS cosine objective on $\boldsymbol{\mu}$; `recip_kappa_mu` is $\kappa_2(\boldsymbol{\mu})^{-1}$; `gram_det` is $\sqrt{\det(\boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu})}$.

# Perspectives: a single geometric score for $\boldsymbol{\mu}$ {#sec-geometric-score}

The present API exposes two AutoGeneS-style objectives. A natural refinement is to replace them by one scalar that jointly rewards large column norms (and hence Euclidean separation) and penalises alignment.

Two candidates are immediate. The Gram determinant

$$
V(\boldsymbol{\mu})
=
\sqrt{
  \det\bigl(
    \boldsymbol{\mu}^{\mathsf{T}}\boldsymbol{\mu}
  \bigr)
}
$$ {#eq-gram-volume}

is the volume of the parallelepiped spanned by the columns of $\boldsymbol{\mu}$. It vanishes under exact collinearity and grows when columns lengthen or become more orthogonal—precisely the desired MOO trade-off in one number.

Equivalently, the reciprocal condition number

$$
\kappa_2(\boldsymbol{\mu})^{-1}
=
\frac{\sigma_{\min}(\boldsymbol{\mu})}{\sigma_{\max}(\boldsymbol{\mu})},
$$ {#eq-reciprocal-kappa}

available in R via `kappa(mean_profiles, exact = TRUE)` (or `1 / kappa(...)`), measures how ill-posed linear recovery of $\boldsymbol{p}$ from $\boldsymbol{y}\approx\boldsymbol{\mu}\boldsymbol{p}$ is. Maximising $\kappa_2(\boldsymbol{\mu})^{-1}$ (or minimising $\kappa_2(\boldsymbol{\mu})$) collapses cosine and distance into a deconvolution-centric criterion without a Pareto front.

A practical next step is to expose an optional `mean_objective = c("autogenes", "volume", "condition")` in `simulate_hierarchical_grn_moments()`, keep the constructive $(\rho,s)$ generator for controllable scenarios, and report the chosen scalar alongside the existing pairwise diagnostics.

# References
